### 2014-02-04 tpw_rate.tex
Wipe the template

% gjilguid2e.tex
% V2.0 released 1998 December 18
% V2.1 released 2003 October 7 -- Gregor Hutton, updated the web address for the style files.

\documentclass{gji}
\usepackage{timet}

\title[TPW Rates]
  {Rates of true polar wander in convecting planets}
\author[I. Rose and B. Buffett]
  {Ian Rose$^1$ and Bruce Buffett$^1$ \\
  $^1$ Department of Earth \& Planetary Science, University of California, Berkeley, CA 94720, USA.  E-mail: ian.rose@berkeley.edu
  }
\date{}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\volume{}
\pubyear{}

\let\leqslant=\leq

\begin{document}

\label{firstpage}

\maketitle

\begin{summary}
Abstract goes where?
\end{summary}

\begin{keywords}
True polar wander; Earth's rotational stability; mantle convection; scaling analyses;
\end{keywords}

\section{Introduction}

\begin{acknowledgments}
\end{acknowledgments}

%\begin{thebibliography}{}
%\end{thebibliography}

%\bsp % ``This paper has been produced using the Blackwell
     %   Publishing GJI \LaTeXe\ class file.''

\label{lastpage}

\end{document}

### 2014-02-05 tpw_rate.tex
Add keywords
Update .gitignore

@@
 % gjilguid2e.tex
-% V2.0 released 1998 December 18
-% V2.1 released 2003 October 7 -- Gregor Hutton, updated the web address for the style files.
 
+%\documentclass[onecolumn]{gji}
 \documentclass{gji}
 \usepackage{timet}
+\usepackage{amsmath}
 
@@
 \author[I. Rose and B. Buffett]
-  {Ian Rose$^1$ and Bruce Buffett$^1$ \\
+  {Ian Rose$^1$ and Bruce A. Buffett$^1$ \\
   $^1$ Department of Earth \& Planetary Science, University of California, Berkeley, CA 94720, USA.  E-mail: ian.rose@berkeley.edu
@@
 \begin{keywords}
-True polar wander; Earth's rotational stability; mantle convection; scaling analyses;
+Earth rotation variations; Mantle processes; Dynamics: convection currents and mantle plumes; Planetary interiors; Numerical solutions; Paleomagnetism applied to tectonics.
 \end{keywords}
@@
 
+\section{Governing equations}
+
+As mantle convection and rotational dynamics of planetary bodies are usually considered separately, some extra care must be taken to establish self-consistent governing equations.  
+Here we consider a planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.  The equations for mass, momentum, and energy then read
+
+\begin{equation}
+\nabla \cdot {\bf v} = 0
+\label{conserve_mass}
+\end{equation}
+
+\begin{equation}
+\begin{aligned}
+ \rho \left[ \frac{D \bf v}{D t} + \frac{d \bf \Omega}{dt} \times {\bf r} +  2 {\bf \Omega} \times {\bf v} + {\bf \Omega \times \Omega \times r} \right] \\ = - \nabla P + \nabla \cdot \left( \eta \nabla_s {\bf v} \right) + \rho {\bf g}
+\label{navier_stokes}
+\end{aligned}
+\end{equation}
+
+\begin{equation}
+\frac{\partial T}{\partial t} + {\bf v} \cdot \nabla T = \kappa \nabla^2 T
+\label{energy}
+\end{equation}
+
+along with the equation of state
+
+\begin{equation}
+\rho = \rho_0 \left( 1 - \alpha T \right)
+\label{eos}
+\end{equation}
+
+Dimensional analysis of this system (c.f. \cite{barenblatt1996scaling}) requires seven nondimensional numbers to characterize it.
+
+\section{Scaling}
+
+\section{Rates of true polar wander}
+
+\section{Numerical results}
+
+\section{Discussion}
+
 \begin{acknowledgments}
@@
 
-%\begin{thebibliography}{}
-%\end{thebibliography}
-
-%\bsp % ``This paper has been produced using the Blackwell
-     %   Publishing GJI \LaTeXe\ class file.''
+\bibliographystyle{gji}
+\bibliography{tpw_rate.bib}
 

### 2014-02-09 tpw_rate.tex
Start adding some tables

@@
 \begin{aligned}
- \rho \left[ \frac{D \bf v}{D t} + \frac{d \bf \Omega}{dt} \times {\bf r} +  2 {\bf \Omega} \times {\bf v} + {\bf \Omega \times \Omega \times r} \right] \\ = - \nabla P + \nabla \cdot \left( \eta \nabla_s {\bf v} \right) + \rho {\bf g}
+ \rho_0 \left[ \frac{D \bf v}{D t} + \frac{d \bf \Omega}{dt} \times {\bf r} +  2 {\bf \Omega} \times {\bf v} + {\bf \Omega \times \Omega \times r} \right] \\ = - \nabla P + \nabla \cdot \left( \eta \nabla_s {\bf v} \right) + \rho {\bf g}
 \label{navier_stokes}
@@
 
-Dimensional analysis of this system (c.f. \cite{barenblatt1996scaling}) requires seven nondimensional numbers to characterize it.
+Dimensional analysis of this system (c.f. \cite{barenblatt1996scaling}) requires six nondimensional numbers to characterize it.
+
+\begin{table}
+\caption{Parameters for rotating mantle convection}
+\label{parameters}
+\begin{tabular}{@{}lcc}
+Symbol & Definition\\
+\hline
+$R_i$ & inner radius \\
+$R_o$ & outer radius \\
+$\Omega_0$ & reference rotation rate \\
+$\eta$ & viscosity \\
+$\kappa$ & thermal diffusivity \\
+$\alpha$ & thermal expansivity \\
+$g_0$ & reference gravity \\
+$\rho_0$ & reference density \\
+$\Delta T$ & temperature difference \\ 
+\end{tabular}
+\end{table}
+
+\begin{table}
+\caption{Nondimensional numbers with approximate Earthlike values}
+\label{nondim}
+\begin{tabular}{@{}lcccc}
+Symbol &  Number & Definition & Approximate value \\
+\hline
+Ra & Rayleigh &  $\rho_0 g_0 \alpha \Delta T R_o^3/\eta \kappa$ & $10^7$\\
+Pr & Prandtl & $\eta/\rho_0 \kappa$ & $10^{23}$ \\
+Ek & Ekman & $\eta/\rho_0 \Omega_0 R_{o}^2$ & $10^8$ \\
+Fr & Froude & $\Omega_0^2 R_o/g_0$ & $10^{-3}$ \\
+$\Gamma$ & Density deficit & $\alpha \Delta T$ & $10^{-2}$ \\
+A & Aspect ratio & $R_i/R_o$ & $0.55$ \\
+\end{tabular}
+\end{table}
 

### 2014-02-20 tpw_rate.tex
Add extra option for more citation options
Add some appendix text about MacCullagh

@@
 
-%\documentclass[onecolumn]{gji}
-\documentclass{gji}
+%\documentclass[extra,onecolumn]{gji}
+\documentclass[extra]{gji}
 \usepackage{timet}
@@
 \let\leqslant=\leq
+
+%in case I want to add detail to any derivation
+\newif\ifdetail
+\detailfalse
 
@@
 \begin{aligned}
- \rho_0 \left[ \frac{D \bf v}{D t} + \frac{d \bf \Omega}{dt} \times {\bf r} +  2 {\bf \Omega} \times {\bf v} + {\bf \Omega \times \Omega \times r} \right] \\ = - \nabla P + \nabla \cdot \left( \eta \nabla_s {\bf v} \right) + \rho {\bf g}
+ \rho_0 \left[ \frac{D \bf v}{D t} + \frac{d \bf \Omega}{dt} \times {\bf r} +  2 {\bf \Omega} \times {\bf v} + {\bf \Omega \times \Omega \times r} \right] \\ = - \nabla P + \nabla \cdot \left( \eta \varepsilon ({\bf v}) \right) + \rho {\bf g}
 \label{navier_stokes}
@@
 
-along with the equation of state
+ where the parameters are defined in Table~\ref{parameters} and $\varepsilon({\bf v}) = 1/2 \; ( \partial v_i / \partial x_j + \partial v_j / \partial x_i )$ corresponds to the symmetrized velocity gradient.
+In addition we use the simple equation of state
 
@@
 
-Dimensional analysis of this system (c.f. \cite{barenblatt1996scaling}) requires six nondimensional numbers to characterize it.
+Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires six nondimensional numbers to characterize it.
 
@@
 
+\appendix
+
+\section{Degree-two moments}
+
+We may explicitly draw a connection between the moment of inertia of a rotating object and it's degree-two density structure.  The moment of inertia tensor may be written in index notation
+\begin{equation}
+I_{ij} = \int_V \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) d^3 {\bf r}
+\label{inertia}
+\end{equation}
+
+where $\bf r$ is the Eulerian coordinate, $\rho$ is the density, and $V$ is the volume of the material.
+The degree-two density structure may be expressed as a quadrupole-moment tensor (cf.
+\citet{jackson1998classical}):
+
+\begin{equation}
+Q_{ij} = \int_V \rho \left( 3 r_i r_j - r_q r_q \delta_{ij} \right) d^3 {\bf r}
+\end{equation}
+
+The commutator of these tensors, $\bf QI - IQ$, is zero, indicating that they may be simultaneously diagonalized.
+
+\ifdetail
+We may demonstrate that the commutator is zero by using the definition of the commutator and expanding:
+\begin{equation} 
+\begin{aligned}
+{\bf IQ-QI} = &I_{ik} Q_{kj} - Q_{ik} I_{kj} \\
+ = &\int_V \rho \left( r_q r_q \delta_{ik} - r_i r_k \right) d^3 {\bf r} \int_V \rho \left( 3 r_k r_j - r_q r_q \delta_{kj} \right) d^3 {\bf r} \\
+ &- \int_V \rho \left( r_q r_q \delta_{kj} - r_k r_j \right) d^3 {\bf r} \int_V \rho \left( 3 r_i r_k - r_q r_q \delta_{ik} \right) d^3 {\bf r} \\
+ = &3 \delta_{ik} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_k r_j d^3 {\bf r} \\
+ &-\delta_{kj} \delta_{ik} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
+ &-3 \int_V \rho r_i r_k d^3 {\bf r} \int_V \rho r_k r_j d^3 {\bf r} \\
+ &+\delta_{kj} \int_V \rho r_i r_k d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
+ - &3 \delta_{kj} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_k d^3 {\bf r} \\
+ &+\delta_{ik} \delta_{kj} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
+ &+3 \int_V \rho r_k r_j d^3 {\bf r} \int_V \rho r_i r_k d^3 {\bf r} \\
+ &-\delta_{ik} \int_V \rho r_k r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
+ = \;&3 \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_j d^3 {\bf r} \\
+ &+ \int_V \rho r_i r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
+ - &3 \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_j d^3 {\bf r} \\
+ &- \int_V \rho r_i r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
+ & = 0
+\end{aligned}
+\end{equation}
+\else
+\fi
+
+We may therefore go into the principle coordinate system of both tensors to find
+\begin{equation}
+{\bf I} = \begin{bmatrix}
+\int_V \rho (y^2+z^z)\\
+\int_V \rho (x^2+z^2) \\
+\int_V \rho (x^2+y^2) 
+\end{bmatrix}^T {\bf 1}
+\end{equation}
+
+\begin{equation}
+{\bf Q} = \begin{bmatrix}
+\int_V \rho (2 x^2 - y^2 - z^z)\\
+\int_V \rho (2 y^2 - x^2 -z^2)\\
+\int_V \rho (2 z^2 - x^2 - y^2) 
+\end{bmatrix}^T {\bf 1}
+\end{equation}
+
+where $\bf 1$ is the identity matrix.  If we let the diagonal elements of $\bf I$ be $\lambda_1, \lambda_2, \lambda_3$, we may rewrite $\bf Q$ as
+
+\begin{equation}
+{\bf Q} = \begin{bmatrix}
+(\lambda_2-\lambda_1) + (\lambda_3-\lambda_1)\\
+(\lambda_1-\lambda_2) + (\lambda_3-\lambda_2)\\
+(\lambda_1-\lambda_3) + (\lambda_2-\lambda_3)\\
+\end{bmatrix}^T {\bf 1}
+\label{maccullagh}
+\end{equation}
+
+which shows the connection between the degree-two component of the density field and the principle moments of inertia.  This may be seen as a form of MacCullagh's formula \citep{stacey1977physics}
+
 \label{lastpage}

### 2014-02-25 tpw_rate.tex
Start adding text about rotational dynamics

@@
 
+\section{Rates of true polar wander}
+
+We may write the nonlinear Liouville equations 
+\begin{equation}
+{\bf \Omega} \times {\bf \dot{\Omega} } = \frac{1}{\varepsilon I_0 \tau} {\bf \Omega} \times \left( {\bf E \cdot \Omega} \right)
+\end{equation}
+
+introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
+\begin{equation}
+ \dot{\mitbf{\omega}}  = \frac{1}{\varepsilon I_0 \tau} \left[ {\bf E \cdot \mitbf{\omega}} - \left( {\bf \mitbf{\omega} \cdot E \cdot \mitbf{\omega} } \right) \mitbf{\omega} \right]
+\end{equation}
+
+Note that the quantity in brackets is identical in form to the shear stress on a plane in classical elastostatics.
+If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
+\begin{equation}
+\begin{aligned}
+\dot{\theta} &= - \frac{1}{2 \varepsilon I_0 \tau} \sin{2 \theta} \left[ (\lambda_3-\lambda_1) \cos^2{\phi} + (\lambda_3-\lambda_2) \sin^2{\phi} \right] \\
+\dot{\phi} &= \;\; \frac{1}{2 \varepsilon I_0 \tau} \sin{2 \phi} \; (\lambda_2 - \lambda_1)
+\end{aligned}
+\label{milankovitch}
+\end{equation}
+
+These equations are a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.  
+Written this way, it is clear that the important quantities are $\theta$, $\phi$, and the differences of the eigenvalues of the convective moment $\bf E$.
+Thus it remains to derive estimates for these.
+
 \section{Scaling}
 
-\section{Rates of true polar wander}
+The nonhydrostatic moment of inertia is defined
+\begin{equation}
+J_{ij} = \int_{V_S} \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) 
+\end{equation}
+
+where $V_S$ is the reference spherical volume of the mantle.
+If we substitute in Equation \ref{eos} we find
+
+\begin{equation}
+J_{ij} = I_0 \delta_{ij} + \int_{V_S} \rho_0 \alpha T \left( r_q r_q \delta_{ij} - r_i r_j \right) = I_0 \delta_{ij} + E_{ij} 
+\end{equation}
+
+Nondimensionalizing the integral for $E_ij$ we find
+
+\begin{equation}
+J_{ij} = I_0 \delta_{ij} + \int_{V_S^\prime} \alpha T^\prime \left( r_q^\prime r_q^\prime \delta_{ij} - r_i^\prime r_j^\prime \right) 
+\end{equation}
 
@@
 
-which shows the connection between the degree-two component of the density field and the principle moments of inertia.  This may be seen as a form of MacCullagh's formula \citep{stacey1977physics}
+which shows the connection between the degree-two component of the density field and the principle moments of inertia.  This may be seen as a form of MacCullagh's formula \citep{stacey1977physics}.
 

### 2014-02-25 tpw_rate.tex
Work on the governing equations

@@
 As mantle convection and rotational dynamics of planetary bodies are usually considered separately, some extra care must be taken to establish self-consistent governing equations.  
-Here we consider a planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.  The equations for mass, momentum, and energy then read
+Here we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.  The equations for mass, momentum, and energy then read
 
@@
 \begin{aligned}
- \rho_0 \left[ \frac{D \bf v}{D t} + \frac{d \bf \Omega}{dt} \times {\bf r} +  2 {\bf \Omega} \times {\bf v} + {\bf \Omega \times \Omega \times r} \right] \\ = - \nabla P + \nabla \cdot \left( \eta \varepsilon ({\bf v}) \right) + \rho {\bf g}
+ \rho{\bf \Omega \times \Omega \times r}= - \nabla P + \nabla \cdot \left( \eta \varepsilon ({\bf v}) \right) + \rho {\bf g}
 \label{navier_stokes}
@@
 
-Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires six nondimensional numbers to characterize it.
+Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure. 
+Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires five nondimensional numbers to characterize it.
+Convenient choices for these numbers are shown in Table \ref{nondim} along with approximate Earthlike values.
+In addition to the more usual Rayleigh and Prandtl numbers we include a Froude number, characterizing the ratio of centrifugal to gravitational forces, and $\Gamma$, a nondimensional characterization of density variations.
 
@@
 $R_i$ & inner radius \\
-$R_o$ & outer radius \\
+$R$ & outer radius \\
 $\Omega_0$ & reference rotation rate \\
@@
 \hline
-Ra & Rayleigh &  $\rho_0 g_0 \alpha \Delta T R_o^3/\eta \kappa$ & $10^7$\\
+Ra & Rayleigh &  $\rho_0 g_0 \alpha \Delta T R^3/\eta \kappa$ & $10^7$\\
 Pr & Prandtl & $\eta/\rho_0 \kappa$ & $10^{23}$ \\
-Ek & Ekman & $\eta/\rho_0 \Omega_0 R_{o}^2$ & $10^8$ \\
-Fr & Froude & $\Omega_0^2 R_o/g_0$ & $10^{-3}$ \\
+Fr & Froude & $\Omega_0^2 R/g_0$ & $10^{-3}$ \\
 $\Gamma$ & Density deficit & $\alpha \Delta T$ & $10^{-2}$ \\
-A & Aspect ratio & $R_i/R_o$ & $0.55$ \\
+A & Aspect ratio & $R_i/R$ & $0.55$ \\
 \end{tabular}
@@
 
-Nondimensionalizing the integral for $E_ij$ we find
+Nondimensionalizing the integral for $E_{ij}$ we find
 

### 2014-05-08 tpw_rate.tex
Add some detail stuff about nondimensionalization

@@
 Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires five nondimensional numbers to characterize it.
+
+\ifdetail
+\subsection{Buckingham Pi theorem and nondimensionalization}
+The Buckingham Pi theorem (essentially an application of the rank-nullspace theorem from linear algebra) allows us to determine the number of nondimensional numbers for this problem.  
+We count the number of fundamental units for the problem, in this case mass (kg), length (m), time (s), and temperature (K), as well as the number of parameters in use for this problem.  
+These parameters are listed in Table \ref{parameters}, and subtracting the ...
+
+\begin{table}
+\caption{Nondimensionalizations for variables, primed variables denote dimensional variables}
+\label{nondim_convert}
+\begin{tabular}{@{}lll}
+$x^\prime$ &=& $R \;\; x$ \\
+$t^\prime$ &=& $R^2/\kappa \;\; t$ \\
+$T^\prime$ &=& $\Delta T \;\; T$ \\
+$\rho^\prime$ &=& $\rho_0 \;\; \rho$\\
+$v^\prime$ &=& $\kappa/R \;\; v$ \\
+$P^\prime$ &=& $\eta \kappa/R^2 \;\; P$ \\
+$\Omega^\prime$ &=& $\Omega_0 \;\; \Omega$ \\
+$g^\prime$ &=& $g_0 \;\; g$
+\end{tabular}
+\end{table}
+
+If we plug these nondimensionalizations into the governing equations we find that the incompressibility constraint remains unchanged.  The temperature equation and equation of state become
+
+\begin{equation}
+\frac{\partial T}{\partial t} + {\bf v} \cdot \nabla T = \nabla^2 T
+\end{equation}
+
+\begin{equation}
+\rho = 1 - \alpha \Delta T \;\;  T
+\end{equation}
+
+We may furthermore define a hydrostatic reference state where the density is the reference density everywhere and the velocity is zero, i.e.
+
+\begin{equation}
+ \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}
+\label{navier_stokes}
+\end{equation}
+
+Subtracting this from the momentum equation and defining a dynamic pressure $P^* = P - P_0$, we find
+
+\begin{equation}
+\begin{aligned}
+ - & \Omega_0^2  \alpha  \Delta T T R \; {\bf \Omega \times \Omega \times r} = - \eta \kappa / R^2 \; \nabla P^* \\ 
+&+ \eta \kappa / R^2 \; \nabla \cdot \left( \eta \varepsilon ({\bf v}) \right) - \alpha \Delta T g_0 \; {\bf g}
+\end{aligned}
+\end{equation}
+
+\begin{equation}
+\begin{aligned}
+ - & \mathrm{Ra \; Fr}\; T \; {\bf \Omega \times \Omega \times r} = - \nabla P^* \\ 
+&+ \; \nabla \cdot \left( \eta \varepsilon ({\bf v}) \right) - \mathrm{Ra} \; T \; {\bf g}
+\end{aligned}
+\end{equation}
+
+\fi
+
 Convenient choices for these numbers are shown in Table \ref{nondim} along with approximate Earthlike values.

### 2014-05-13 tpw_rate.tex
Add some messy text

@@
 In addition to the more usual Rayleigh and Prandtl numbers we include a Froude number, characterizing the ratio of centrifugal to gravitational forces, and $\Gamma$, a nondimensional characterization of density variations.
+Nondimensionalizing with these parameters we obtain
+\begin{equation}
+\begin{aligned}
+ - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\bf \Omega \times \Omega \times r} = 0
+\end{aligned}
+\end{equation}
+
+where we have introduced a dynamic pressure $P^* = P - P_0$.
 
@@
 
+Without loss of generality we may choose the coordinate system such that $\phi=0$:
+\begin{equation}
+\dot{\theta} = - \frac{1}{2 \varepsilon \tau}\sin{2 \theta} \frac{(\lambda_3-\lambda_1)}{I_0}
+\end{equation}
+
 These equations are a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.  
-Written this way, it is clear that the important quantities are $\theta$, $\phi$, and the differences of the eigenvalues of the convective moment $\bf E$.
+Written this way, it is clear that the important quantities are $\theta$ and the differences between the eigenvalues of the convective moment $\bf E$.
+We may rewrite this in terms of the nondimensional numbers defined above, where $\varepsilon \sim \mathrm{Fr}$ and $\tau \sim \eta / \rho g R \sim \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}$.
+
+\begin{equation}
+\dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma \mathrm{Fr}} \sin{2 \theta} \frac{(\lambda_3-\lambda_1)}{I_0}
+\end{equation}
+
 Thus it remains to derive estimates for these.

### 2014-05-13 tpw_rate.tex
Add some abstract text

@@
 \begin{summary}
-Abstract goes where?
+Mass redistribution in the convecting mantle of a planet can cause perturbations in its moment of inertia tensor.
+Conservation of angular momentum dictates that these perturbations can change the direction of the rotation vector of the planet, a process known as true polar wander (TPW).
+Although the existence of TPW on Earth is well-verified, its rate and magnitude over geologic time scales remain controversial.
+Here we present scaling analyses and novel numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors.
+
+Our numerical model couples the angular and linear momentum equations by implementing a free surface and movable mesh in the CIG-sponsored mantle convection code ASPECT.
+We allow the mantle to deform self consistently to centrifugal forces and update the rotation axis by solving the angular momentum equation at every time step.
+Our simulation and scaling indicate that the important parameters are the Rayleigh number, the rotation rate, and the size of relative density fluctuations (i.e.
+thermal expansivity times the temperature variations).
+We identify two timescales, one for the longevity of convective fluctuations, and one for the adjustment time of the hydrostatic bulge.
+These timescales control the rate and magnitude of TPW, as well as the likelihood of hypothesized "inertial interchange" TPW events.
+
+Finally, we discuss implications for large-scale TPW in Earth's past, which has been suggested to be important for life and climate history.
 \end{summary}

### 2014-05-13 tpw_rate.tex
more work on abstract

--- 2014-05-13 tpw_rate.tex
@@
 \begin{summary}
-Abstract goes where?
+Mass redistribution in the convecting mantle of a planet can cause perturbations in its moment of inertia tensor. 
+Conservation of angular momentum dictates that these perturbations can change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). 
+Although the existence of TPW on Earth is well-verified, its rate and magnitude over geologic time scales remain controversial. 
+Here we present scaling analyses and numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors. 
+For simple rotating convection, the most important parameters are the Rayleigh number, the rotation rate, and the size of relative density fluctuations (i.e.
+thermal expansivity times the temperature variations).
+We identify timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander. 
+These timescales, as well as the relative sizes of convective anomalies, control the rate and magnitude of TPW.
+This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and when they are likely to occur.
+Finally, we discuss implications for large-scale TPW in Earth's past, which has been suggested to be important for life and climate history.
 \end{summary}

### 2014-07-01 tpw_rate.tex
Change to `u`
Add optional detail for Milankovitch theorem

@@
 \detailfalse
+%\detailtrue
 
@@
 \begin{equation}
-\nabla \cdot {\bf v} = 0
+\nabla \cdot {\bf u} = 0
 \label{conserve_mass}
@@
 \begin{aligned}
- \rho{\bf \Omega \times \Omega \times r}= - \nabla P + \nabla \cdot \left( \eta \varepsilon ({\bf v}) \right) + \rho {\bf g}
+ \rho{\bf \Omega \times \Omega \times r}= - \nabla P + \nabla \cdot \left( \eta \varepsilon ({\bf u}) \right) + \rho {\bf g}
 \label{navier_stokes}
@@
 \begin{equation}
-\frac{\partial T}{\partial t} + {\bf v} \cdot \nabla T = \kappa \nabla^2 T
+\frac{\partial T}{\partial t} + {\bf u} \cdot \nabla T = \kappa \nabla^2 T
 \label{energy}
@@
  \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}
-\label{navier_stokes}
 \end{equation}
@@
 If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
+
+\ifdetail
+
+And here is the tedious algebra.
+Let the prefactors $\frac{1}{\varepsilon I_0 \tau}$ be denoted by $A$, and the pole of the coordinate system ($\theta = 0$) associated with the eigenvector for $\lambda_3$.
+Therefore we find that $\mitbf{\omega} = \left[ \sin{\theta} \cos{\phi} \;\; \sin{\theta} \sin{\phi} \;\; \cos{\theta} \right]^T$.
+Plugging this in, we find a system of three equations, which we want to solve for $\dot{\theta}$ and $\dot{\phi}$:
+
+\begin{equation}
+\begin{aligned}
+ & \cos{\theta}\cos{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \dot{\phi} = \\
+  &A \sin{\theta}\cos{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+ &\cos{\theta}\sin{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \dot{\phi} = \\
+  &A \sin{\theta}\sin{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+ - &\sin{\theta} \dot{\theta} = \\
+  &A \cos{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+\end{aligned}
+\end{equation}
+
+Now multiply the top equation by $\cos{\phi}$ and the second equation by $\sin{\phi}$ and add them together to get
+
+\begin{equation}
+\begin{aligned}
+ & \cos{\theta}\cos^2{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \cos{\phi} \dot{\phi} + \cos{\theta}\sin^2{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \sin{\phi} \dot{\phi} = \\
+  &A \sin{\theta}\cos^2{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] + \\
+  &A \sin{\theta}\sin^2{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+ & \cos{\theta}\dot{\theta} =  
+   A \sin{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
+  &A \sin{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+\end{aligned}
+\end{equation}
+
+Now multiply the above by $\cos{\theta}$ and the third equation by $-\sin{\theta}$ and add them to get
+
+\begin{equation}
+\begin{aligned}
+ & \cos^2{\theta}\dot{\theta} + \sin^2{\theta} \dot{\theta} =  
+   A \sin{\theta} \cos{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
+  &A \sin{\theta} \cos{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] - \\
+  &A \cos{\theta} \sin{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+\end{aligned}
+\end{equation}
+
+After a lot of cancellation, we find
+
+\begin{equation}
+\begin{aligned}
+ \dot{\theta} &= A \sin{\theta} \cos{\theta} \left[ - \lambda_3 + \lambda_1 \cos^2{\phi} + \lambda_2 \sin^2{\phi} \right] \\
+              &= - A \sin{\theta} \cos{\theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
+              &= -\frac{A}{2} \sin{2 \theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
+\end{aligned}
+\end{equation}
+
+Likewise we may multiply the first equation by $-\sin{\phi}$ and the second equation by $\cos{\phi}$ and add them to get
+
+\begin{equation}
+\begin{aligned}
+ & \sin{\theta} \dot{\phi} = A \sin{\theta}\cos{\phi}\sin{\phi} \left( \lambda_2 - \lambda_1 \right ) \\
+& \dot{\phi} = \frac{A}{2} \sin{2 \phi} (\lambda_2-\lambda_1)
+\end{aligned}
+\end{equation}
+
+\else
+\fi
+
 \begin{equation}

### 2014-07-02 tpw_rate.tex
Text in rotational dynamics

@@
 
-\section{Rates of true polar wander}
+\section{Rotational dynamics}
 
@@
 \dot{\theta} = - \frac{1}{2 \varepsilon \tau}\sin{2 \theta} \frac{(\lambda_3-\lambda_1)}{I_0}
+\label{simple_milankovitch}
 \end{equation}
@@
 Written this way, it is clear that the important quantities are $\theta$ and the differences between the eigenvalues of the convective moment $\bf E$.
-We may rewrite this in terms of the nondimensional numbers defined above, where $\varepsilon \sim \mathrm{Fr}$ and $\tau \sim \eta / \rho g R \sim \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}$.
+It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} = (\lambda_i - \lambda_j)/I_0$.  
+With this we may rewrite this in terms of the nondimensional numbers defined above, where 
+$\varepsilon \sim \mathrm{Fr}$ and $\tau \sim \eta / \rho g R = \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}$.
 
 \begin{equation}
-\dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma \mathrm{Fr}} \sin{2 \theta} \frac{(\lambda_3-\lambda_1)}{I_0}
+\frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma \mathrm{Fr}} \Lambda_{31} \sin{2 \theta}
 \end{equation}
 
-Thus it remains to derive estimates for these.
+At this point we do not have estimates for the characteristic magnitudes of $\Lambda_{ij}$ or $\theta$, both of which are crucial for estimating characteristic rates of TPW.
+They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between where the 
+convective moment
 

### 2014-07-02 tpw_rate.tex
Adjust formulae
Use equiv
A bunch of nonsense

@@
 \usepackage{amsmath}
+\usepackage{graphicx}
 
@@
 Written this way, it is clear that the important quantities are $\theta$ and the differences between the eigenvalues of the convective moment $\bf E$.
-It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} = (\lambda_i - \lambda_j)/I_0$.  
+It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
 With this we may rewrite this in terms of the nondimensional numbers defined above, where 
@@
 \section{Scaling}
+\subsection{An estimate of $\Lambda_{ij}$}
+
+\begin{figure}
+\label{misfit}
+\includegraphics[]{figures/eigengap.pdf}
+\caption{Oooga}
+\end{figure}
+
+\begin{figure}
+\label{misfit}
+\includegraphics[width=10cm]{figures/perturb.pdf}
+\caption{Oooga}
+\end{figure}
 
@@
 
+For thermal convection, the convective part of the moment of inertia $\bf E$ is directly related to the degree-two structure of the temperature field (see Appendix \ref{appendix:moments}).
+Therefore, an estimate of one allows an estimate of the other.
+
+The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
+the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is the broadening of the spatial 
+and temporal spectra \citep{mclaughlin1982transition}.
+
+We can expand the temperature structure of the convecting planet with a set of orthogonal basis functions $X_{lmn}$, 
+where $l$, $m$, and $n$ are wavenumbers in the $x$, $y$, and $z$ directions, respectively.  
+\begin{equation} 
+T( x, y ,z )  = {\displaystyle \sum_{l m n} } T_{lmn} X_{lmn} (x , y, z)
+\end{equation}
+
+For chaotic thermal convection  there will be some power in every mode $X_{lmn}$ up to some wavenumber.  
+At that critical wavenumber thermal diffusivity will
+
+\subsection{An estimate of the mismatch $\theta$}
+
+\citet{davis1970rotation}
+
+\begin{figure}
+\label{misfit}
+\includegraphics[width=10cm]{figures/misfit.pdf}
+\caption{Oooga}
+\end{figure}
+
 \section{Numerical results}
@@
 \section{Degree-two moments}
+\label{appendix:moments}
 
@@
 \begin{equation}
-{\bf I} = \begin{bmatrix}
+{\bf I} = {\bf 1} \begin{bmatrix}
 \int_V \rho (y^2+z^z)\\
@@
 \int_V \rho (x^2+y^2) 
-\end{bmatrix}^T {\bf 1}
+\end{bmatrix}
 \end{equation}
@@
 \begin{equation}
-{\bf Q} = \begin{bmatrix}
+{\bf Q} = {\bf 1} \begin{bmatrix}
 \int_V \rho (2 x^2 - y^2 - z^z)\\
@@
 \int_V \rho (2 z^2 - x^2 - y^2) 
-\end{bmatrix}^T {\bf 1}
+\end{bmatrix}
 \end{equation}
@@
 \begin{equation}
-{\bf Q} = \begin{bmatrix}
+{\bf Q} = {\bf 1} \begin{bmatrix}
 (\lambda_2-\lambda_1) + (\lambda_3-\lambda_1)\\
@@
 (\lambda_1-\lambda_3) + (\lambda_2-\lambda_3)\\
-\end{bmatrix}^T {\bf 1}
+\end{bmatrix}
 \label{maccullagh}

### 2014-07-02 figures/eigengap.py

@@ -1,0 +1,123 @@
+import matplotlib
+matplotlib.use('agg')
+
+import numpy as np
+import scipy as sp
+import scipy.signal as signal
+import scipy.fftpack as fftpack
+import scipy.interpolate as interp
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+import matplotlib.ticker
+import glob
+
+#c = ['#0A5F02', '#1c567a', '#814292', '#d7383b', '#fdae61', '#c0f8b8']
+c = ['#e41a1c', '#377eb8', '#4daf4a', '#984ea3', '#ff7f00', '#ffff33'] 
+#c = ['#1b9e77', '#d95f02', '#7570b3', '#e7298a', '#66a61e']
+
+rho = 3300
+alpha_0 = 4.e-5
+delta_T_0 = 200.0
+k = 4.1
+cp = 1250.
+kappa = k/rho/cp
+eta_0 = 1.e21
+D = 6371.e3
+g = 9.81
+
+omega_E = 7.29e-5
+omega = 4.e-5
+
+frac = 50
+
+def get_spin_spectrum(text_file):
+
+  #load the data
+  data = np.loadtxt(text_file,skiprows=50, usecols=(1,15,16,17,18))
+  #get rid of duplicate times (why might these exist?)
+  stats = []
+  for i in range(len(data)-1):
+    if data[i,0] != data[i+1, 0]:
+      stats.append(data[i,:])
+  stats = np.array(stats) 
+        
+  eig1tck = interp.splrep(stats[:,0],stats[:,3], k=3 )
+  eig2tck = interp.splrep(stats[:,0],stats[:,4], k=3)
+  axistck = interp.splrep(stats[:,0],stats[:,1], k=3)
+  mismatchtck = interp.splrep(stats[:,0],stats[:,2], k=3)
+
+  times = np.linspace(stats[0,0], stats[-1,0], len(stats[:,0]))
+  lambda1 = interp.splev(times, eig1tck, der=0)
+  lambda2 = interp.splev(times, eig2tck, der=0)
+  axis = interp.splev(times, axistck, der=0)
+  mismatch = interp.splev(times, mismatchtck, der=0)
+
+  # nondimensionalize
+  diff = (lambda1-lambda2)/( np.mean(lambda1 + lambda2)/2.0 )
+  times = times * kappa / D / D
+ 
+  frequency = times[-1]/len(times)
[…]

### 2014-07-02 figures/misfit.py

@@ -1,0 +1,64 @@
+
+import matplotlib
+matplotlib.use('agg')
+
+import matplotlib.pyplot as plt
+import matplotlib.ticker
+import numpy as np
+
+c = ['#0A5F02', '#1c567a', '#814292', '#d7383b', '#fdae61', '#c0f8b8']
+data = np.loadtxt('statistics_500', skiprows=50, usecols=(1,15,16,17,18))
+
+rho = 3300
+alpha_0 = 4.e-5
+delta_T_0 = 200.0
+k = 4.1
+cp = 1250.
+kappa = k/rho/cp
+eta_0 = 1.e21
+D = 6371.e3
+g = 9.81
+
+
+time = data[:,0]*kappa/D/D
+mismatch = data[:,2]
+eig1 = data[:,3]
+eig2 = data[:,4]
+eigmean = np.mean(eig1+eig2)/2.0
+eig1 = eig1/eigmean - 1.
+eig2 = eig2/eigmean - 1.
+
+spin = data[:,1]
+
+pos = np.where(np.abs(np.diff(spin)) >= 10.0)[0]
+time[pos] = np.nan
+spin[pos] = np.nan
+
+fig = plt.figure()
+fig.subplots_adjust(wspace=0.0, hspace=0.0)
+
+ax = fig.add_subplot(211)
+
+ax.plot(time, eig1, color=c[3], label=r'$\lambda_1$')
+ax.plot(time, eig2, color=c[1], label=r'$\lambda_2$')
+plt.xlim(0.0, 0.5e-10)
+ax.legend(loc='upper right')
+ax.xaxis.set_major_formatter(matplotlib.ticker.NullFormatter())
+ax.yaxis.set_major_locator(matplotlib.ticker.MaxNLocator(nbins=5, prune='both'))
+plt.ylabel(r'Relative moment')
+
+ax = fig.add_subplot(212)
+
+ax.plot(time, spin, color=c[0], label=r'Spin axis')
+ax.plot(time, mismatch, color=c[2], label=r'$\theta$')
+plt.xlim(0.0, 0.5e-10)
+ax.legend(loc='upper right')
+ax.yaxis.set_major_locator(matplotlib.ticker.MaxNLocator(nbins=6, prune='both'))
+plt.xlabel(r"Time")
+plt.ylabel(r"Angle ($^\circ$)")
+
+
+
+
+plt.show()
+plt.savefig("misfit.pdf")

### 2014-07-02 figures/perturb.sh

@@ -1,0 +1,78 @@
+
+#draw ellipse
+gmt psxy -R-2/2/-2/2 -Jx3 -Se -Wthicker,black -K > perturb.ps <<EOF
+0 0 20 3.6 2.8
+EOF
+
+gmt psxy -R -J -Sv0.3+e -Wthicker,red -K -O >> perturb.ps <<EOF
+0 0 20 1.8 1.8
+EOF
+gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
+0 0 110 1.4 1.4
+EOF
+
+gmt pstext -R -J -K  -O >>perturb.ps <<EOF
+.0 0.36 @~l@-3
+EOF
+
+gmt pstext -R -J -K -O >>perturb.ps <<EOF
+.32 0.26 @~l@-1
+EOF
+
+gmt pstext -R -J -K -O >>perturb.ps <<EOF
+.46 0.10 @~q
+EOF
+
+gmt psxy -R -J -Sv0.3+e -Wthicker,black -K -O >> perturb.ps <<EOF
+0 0 4 2.4 1 
+EOF
+
+gmt pstext -R -J -K -O >>perturb.ps <<EOF
+0.9 .05 @~W
+EOF
+
+gmt psxy -R -J -Sc.1i -Gdarkgray -K -O >>perturb.ps <<EOF
+-0.2 0.3
+EOF
+
+#draw ellipse
+gmt psxy -R-2/2/-2/2 -Jx3 -X2i -Se -Wthicker,black -K -O >> perturb.ps <<EOF
+0 0 70 3.2 3.2
+EOF
+
+gmt psxy -R -J -Sv0.3+e -Wthicker,red -K -O >> perturb.ps <<EOF
+0 0 80 1.6 1.6
+EOF
+gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
+0 0 170 1.6 1.6
+EOF
+
+gmt pstext -R -J -K  -O >>perturb.ps <<EOF
+-0.36 0.2 @~l@-3
+EOF
+
+gmt pstext -R -J -K -O >>perturb.ps <<EOF
+.0 0.46 @~l@-1
+EOF
+
+gmt pstext -R -J -K -O >>perturb.ps <<EOF
+.1 0.10 @~q
+EOF
+
[…]

### 2014-07-02 perturb.sh

@@ -1,78 +1,0 @@
-
-#draw ellipse
-gmt psxy -R-2/2/-2/2 -Jx3 -Se -Wthicker,black -K > perturb.ps <<EOF
-0 0 20 3.6 2.8
-EOF
-
-gmt psxy -R -J -Sv0.3+e -Wthicker,red -K -O >> perturb.ps <<EOF
-0 0 20 1.8 1.8
-EOF
-gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
-0 0 110 1.4 1.4
-EOF
-
-gmt pstext -R -J -K  -O >>perturb.ps <<EOF
-.0 0.36 @~l@-3
-EOF
-
-gmt pstext -R -J -K -O >>perturb.ps <<EOF
-.32 0.26 @~l@-1
-EOF
-
-gmt pstext -R -J -K -O >>perturb.ps <<EOF
-.46 0.10 @~q
-EOF
-
-gmt psxy -R -J -Sv0.3+e -Wthicker,black -K -O >> perturb.ps <<EOF
-0 0 4 2.4 1 
-EOF
-
-gmt pstext -R -J -K -O >>perturb.ps <<EOF
-0.9 .05 @~W
-EOF
-
-gmt psxy -R -J -Sc.1i -Gdarkgray -K -O >>perturb.ps <<EOF
--0.2 0.3
-EOF
-
-#draw ellipse
-gmt psxy -R-2/2/-2/2 -Jx3 -X2i -Se -Wthicker,black -K -O >> perturb.ps <<EOF
-0 0 70 3.2 3.2
-EOF
-
-gmt psxy -R -J -Sv0.3+e -Wthicker,red -K -O >> perturb.ps <<EOF
-0 0 80 1.6 1.6
-EOF
-gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
-0 0 170 1.6 1.6
-EOF
-
-gmt pstext -R -J -K  -O >>perturb.ps <<EOF
--0.36 0.2 @~l@-3
-EOF
-
-gmt pstext -R -J -K -O >>perturb.ps <<EOF
-.0 0.46 @~l@-1
-EOF
-
-gmt pstext -R -J -K -O >>perturb.ps <<EOF
-.1 0.10 @~q
-EOF
-
[…]

### 2014-09-30 tpw_rate.tex
TEXT

@@
 \begin{table}
+\centering
 \caption{Nondimensionalizations for variables, primed variables denote dimensional variables}
@@
 \begin{table}
+\centering
 \caption{Parameters for rotating mantle convection}
@@
 \begin{table}
+\centering
 \caption{Nondimensional numbers with approximate Earthlike values}
@@
 \frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma \mathrm{Fr}} \Lambda_{31} \sin{2 \theta}
+\label{scaled_rotation}
 \end{equation}
@@
 They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between where the 
-convective moment
+rotation axis it and where the convective moment wants it to be.
 
 \section{Scaling}
+
 \subsection{An estimate of $\Lambda_{ij}$}
@@
 \begin{figure}
+\centering
 \label{misfit}
-\includegraphics[]{figures/eigengap.pdf}
+\includegraphics[width=20cm]{figures/eigengap.pdf}
 \caption{Oooga}
@@
 \begin{figure}
+\centering
 \label{misfit}
@@
 
-We can expand the temperature structure of the convecting planet with a set of orthogonal basis functions $X_{lmn}$, 
-where $l$, $m$, and $n$ are wavenumbers in the $x$, $y$, and $z$ directions, respectively.  
+We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $X_{lmn}$, 
+where $l$, $m$, and $n$ are wavenumbers.  
 \begin{equation} 
@@
 
-For chaotic thermal convection  there will be some power in every mode $X_{lmn}$ up to some wavenumber.  
-At that critical wavenumber thermal diffusivity will
+The quantity $\Lambda_{ij}$ is the normalized degree-two temperature field.
+The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $X_{lmn}$ is constrained by
+
+\begin{equation}
+\lVert T(x,y,z) \rVert_\infty = \lVert {\displaystyle \sum_{l m n} } T_{lmn} X_{lmn} (x , y, z) \rVert_\infty = 1
+\end{equation}
+
+This is a strong constraint, but it gives very little information about the distribution of power across the $T_{lmn}$.
+We can, however, make hypotheses about the power spectrum in two different regimes: that of steady/quasisteady flow (relatively low $\mathrm{Ra}$) and that of chaotic flow (relatively high $\mathrm{Ra}$).
 
@@
 
-\citet{davis1970rotation}
+The mismatch between the current axis and the principle axis of the convective moment is perhaps the most important parameter in Equation \ref{scaled_rotation}.  
+Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
+
+An estimate for $\theta$ must concern the stability of the convective moment of inertia. 
+Convection is continuously redistributing mass throughout the mantle, which will perturb the convective moment of inertia.  
+If the convective moment is stable to small perturbations, then $\theta$ should be small, but if it is not stable to perturbations, then $\theta$ could be up to $90^\circ$. 
+Given a random perturbation to that moment, we would like to give a bound on how big the perturbation to $\theta$ is. 
+This may be done by application of a theorem due to \citet{davis1970rotation}.
+
+Let $\delta$ be the perturbation to the convective moment of inertia, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that moment.  
+Then the rotation of the principal axes of the tensor is bounded by
+
+\begin{equation}
+\sin(2 \theta) \le \frac{ 2 \lVert \delta \rVert}{ \displaystyle \min_{i \neq j} \lVert \lambda_i - \lambda_j \rVert }
+\end{equation}
+
+That is to say, if there is a large difference between the eigenvalues, this stabilizes axes to perturbations.  
+If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principle axes.
+
+Furthermore, the quantity $(\lambda_i-\lambda_j)$ which appears in the denominator is precisely the same as the quantity which we estimated in the previous section to be $\sim \mathrm{Ra}^{-1}$.
+Therefore, as the Rayleigh number increases, the characteristic gap between the eigenvalues of the convective moment becomes smaller.  
+Additionally, the timescale of fluctuations in these values goes down. 
+Overall, this makes the principle axes of high Rayleigh number systems much less stable.
+This result agrees with the qualitative results of
 
 \begin{figure}
+\centering
 \label{misfit}
-\includegraphics[width=10cm]{figures/misfit.pdf}
+\includegraphics[width=20cm]{figures/misfit.pdf}
 \caption{Oooga}
 \end{figure}
-
-\section{Numerical results}
 
@@
 
-which shows the connection between the degree-two component of the density field and the principle moments of inertia.  This may be seen as a form of MacCullagh's formula \citep{stacey1977physics}.
+which shows the connection between the degree-two component of the density field and the principle moments of inertia.  This may be seen as a form of MacCullagh's formula \citep{munk1960rotation}.
 

### 2014-09-30 tpw_rate.tex
TEXT

@@
 
-%\documentclass[extra,onecolumn]{gji}
-\documentclass[extra]{gji}
+\documentclass[extra,onecolumn]{gji}
+%\documentclass[extra]{gji}
 \usepackage{timet}
@@
 \label{misfit}
-\includegraphics[width=20cm]{figures/eigengap.pdf}
+\includegraphics[width=0.9\textwidth]{figures/eigengap.pdf}
 \caption{Oooga}
@@
 \label{misfit}
-\includegraphics[width=10cm]{figures/perturb.pdf}
-\caption{Oooga}
+\includegraphics[width=0.9\textwidth]{figures/perturb.pdf}
+\caption{Graphical demonstration of the $\sin{2 \theta}$ theorem of \citet{davis1970rotation}.
+Two spheroidal bodies with eigenvalues $\lambda_2 > \lambda_1$ start out with the rotation axis $\Omega$ aligned with the $\lambda_2$ axis.
+However, on the left the eigengap $\lVert \lambda_2 - \lambda_1 \rVert$ is large, while on the right it is small.
+A perturbation is instantaneously added to both bodies, which effects a small rotation of the principle axes on the left, but a large one on the right.}
 \end{figure}
@@
 
-Furthermore, the quantity $(\lambda_i-\lambda_j)$ which appears in the denominator is precisely the same as the quantity which we estimated in the previous section to be $\sim \mathrm{Ra}^{-1}$.
+Furthermore, the quantity $(\lambda_i-\lambda_j)$ which appears in the denominator is precisely the same as the quantity which we estimated in the previous section to scale with $\sim \mathrm{Ra}^{-1}$.
 Therefore, as the Rayleigh number increases, the characteristic gap between the eigenvalues of the convective moment becomes smaller.  
@@
 Overall, this makes the principle axes of high Rayleigh number systems much less stable.
-This result agrees with the qualitative results of
+This result is consistent with the result of \citet{richards1997explanation}.
 
@@
 \label{misfit}
-\includegraphics[width=20cm]{figures/misfit.pdf}
-\caption{Oooga}
+\includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
+\caption{Top: }
 \end{figure}
@@
 \section{Discussion}
+
+We have developed a framework for discussing the rates of true polar wander in terms of a small number of nondimensional numbers.
 

### 2014-10-01 tpw_rate.tex
STUFF

@@
 \usepackage{timet}
+\usepackage{url}
 \usepackage{amsmath}
@@
 \section{Introduction}
+\label{sec:intro}
 
 \section{Governing equations}
+\label{sec:govern}
 
@@
 Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure. 
-Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires five nondimensional numbers to characterize it.
+Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires \_\_\_\_? nondimensional numbers to characterize it.
 
@@
 $\kappa$ & thermal diffusivity \\
-$\alpha$ & thermal expansivity \\
 $g_0$ & reference gravity \\
-$\rho_0$ & reference density \\
-$\Delta T$ & temperature difference \\ 
+$\delta \rho = \rho_0 \alpha \Delta T$ & buoyancy parameter \\ 
 \end{tabular}
@@
 \section{Rotational dynamics}
+\label{sec:rotation}
 
@@
 {\bf \Omega} \times {\bf \dot{\Omega} } = \frac{1}{\varepsilon I_0 \tau} {\bf \Omega} \times \left( {\bf E \cdot \Omega} \right)
+\label{eq:liouville}
 \end{equation}
@@
 
-At this point we do not have estimates for the characteristic magnitudes of $\Lambda_{ij}$ or $\theta$, both of which are crucial for estimating characteristic rates of TPW.
+At this point we do not have estimates for the characteristic magnitudes of $\Lambda_{ij}$ or $\theta$, both of which are crucial for predicting characteristic rates of TPW.
 They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between where the 
@@
 \section{Scaling}
+\label{sec:scaling}
+
+The quantites $\Lambda_{i,j}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
+We consider relatively slowly rotating bodies here ($\mathrm{Fr} << 1$), so that the rotation does not affect the style of convection.  
+In this regime, therefore, we neglect the dependence on $\mathrm{Fr}$.
+
+We support our scaling analyses with numerical simulations of 2D mantle convection using the finite element code \texttt{Aspect} \citep{kronbichler2012high}, based on the \texttt{deal.II} library \citep{dealII81}.
+\texttt{Aspect} provides modern numerical methods, flexible geometry, and powerful support for integrating user-defined modules and postprocessors.
+We performed a number of simulations at different Rayleigh numbers, integrating Equation \ref{eq:liouville} at every time step.
 
@@
 \caption{Oooga}
-\end{figure}
-
-\begin{figure}
-\centering
-\label{misfit}
-\includegraphics[width=0.9\textwidth]{figures/perturb.pdf}
-\caption{Graphical demonstration of the $\sin{2 \theta}$ theorem of \citet{davis1970rotation}.
-Two spheroidal bodies with eigenvalues $\lambda_2 > \lambda_1$ start out with the rotation axis $\Omega$ aligned with the $\lambda_2$ axis.
-However, on the left the eigengap $\lVert \lambda_2 - \lambda_1 \rVert$ is large, while on the right it is small.
-A perturbation is instantaneously added to both bodies, which effects a small rotation of the principle axes on the left, but a large one on the right.}
 \end{figure}
@@
 \begin{equation} 
-T( x, y ,z )  = {\displaystyle \sum_{l m n} } T_{lmn} X_{lmn} (x , y, z)
+T( x, y ,z, t )  = {\displaystyle \sum_{l m n}^\infty } T_{lmn}(t) X_{lmn} (x , y, z)
+\label{T_series}
 \end{equation}
@@
 \begin{equation}
-\lVert T(x,y,z) \rVert_\infty = \lVert {\displaystyle \sum_{l m n} } T_{lmn} X_{lmn} (x , y, z) \rVert_\infty = 1
+\lVert T(x,y,z, t) \rVert_\infty = \lVert {\displaystyle \sum_{l m n}^\infty } T_{lmn} (t) X_{lmn} (x , y, z) \rVert_\infty = 1
 \end{equation}
@@
 This is a strong constraint, but it gives very little information about the distribution of power across the $T_{lmn}$.
-We can, however, make hypotheses about the power spectrum in two different regimes: that of steady/quasisteady flow (relatively low $\mathrm{Ra}$) and that of chaotic flow (relatively high $\mathrm{Ra}$).
+We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow (relatively low $\mathrm{Ra}$) and that of chaotic flow (relatively high $\mathrm{Ra}$).
+
+At high Rayleigh number we expect there to be a lengthscale which characterizes the smallest length of features in the convecting system.  
+At lengthscales smaller than this diffusion will tend to wipe out heterogeneities, and there will be no power in those modes.  
+Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in \ref{T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:
+
+\begin{equation}
+n_{\mathrm{max}}, l_{\mathrm{max}}, m_{\mathrm{max}} \sim \frac{R}{d}
+\end{equation}
+
+The total number of modes that are accessible to the system are therefore
+
+\begin{equation}
+N_{\mathrm{modes}} = n_{\mathrm{max}} \times l_{\mathrm{max}} \times n_{\mathrm{max}} \sim \left( \frac{R}{d} \right)^{3}
+\end{equation}
+
+The value of each $T_{lmn}(t)$ will in general be some complex function of time, but for a given style of convection we expect there to be some average value.
+For chaotic flow the power should be spread out amongst the modes accessible to it.
+We may make the hypothesis that each of the modes are roughly as likely as any of the others, which implies
+
+\begin{equation}
+T_{\mathrm{degree 2}}(t) \sim \frac{1}{N_{\mathrm{modes}}} \sim \left( \frac{d}{R}\right)^3
+\end{equation}
+
+Any of a number of scaling laws can provide an estimate for the characteristic length scale of a convecting system which may depending on rheology, geometry, or density structure.
+The simplest, based on boundary layer theory \citep{turcotte1967finite}, finds $d/R ~ \mathrm{Ra}^{-1/3}$, thus furnishing us with an estimate of the power in the degree-two part of the field as a function of Rayleigh number:
+
+\begin{equation}
+T_{\mathrm{degree 2}}(t) \sim \mathrm{Ra}^{-1}
+\end{equation}
+
+This result has a simple interpretation.
+As the Rayleigh number of the system increases, the smallest lengthscale of convective features gets smaller.
+The total power in the temperature field is spread across a larger spectrum, leaving less total power for the degree-two part, which is what drives TPW.
+
+With an estimate for the power in the degree-two part of the temperature field, we may finally estimate $\Lambda_{ij}$:
+
+\begin{equation}
+\Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }
+\end{equation}
 
@@
 \begin{equation}
-\sin(2 \theta) \le \frac{ 2 \lVert \delta \rVert}{ \displaystyle \min_{i \neq j} \lVert \lambda_i - \lambda_j \rVert }
+\lVert \sin(2 \theta) \rVert \le \frac{ 2 \lVert \delta \rVert}{ \displaystyle \min_{i \neq j} \lVert \lambda_i - \lambda_j \rVert }
+\label{eq:kahan}
 \end{equation}
@@
 If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principle axes.
+This is illustrated in Figure \ref{fig:perturb}
 
-Furthermore, the quantity $(\lambda_i-\lambda_j)$ which appears in the denominator is precisely the same as the quantity which we estimated in the previous section to scale with $\sim \mathrm{Ra}^{-1}$.
+We may arrive at a simple estimate of the growth rate of the angle $\theta$ by differentiating Equation \ref{eq:kahan}, tentatively holding $\lVert \lambda_i - \lambda_j \rVert$ fixed:
+
+\begin{equation}
+\lVert \dot{\theta} \rVert \le \frac{ \lVert \dot{\delta} \rVert}{ \displaystyle \min_{i \neq j} \lVert \lambda_i - \lambda_j \rVert }
+\label{eq:grow_perturbation}
+\end{equation}
+
+\begin{figure}
+\centering
+\label{fig:perturb}
+\includegraphics[width=0.9\textwidth]{figures/perturb.pdf}
+\caption{Graphical demonstration of the $\sin{2 \theta}$ theorem of \citet{davis1970rotation}.
+Two spheroidal bodies with eigenvalues $\lambda_2 > \lambda_1$ start out with the rotation axis $\Omega$ aligned with the $\lambda_2$ axis.
+However, on the left the eigengap $\lVert \lambda_2 - \lambda_1 \rVert$ is large, while on the right it is small.
+A perturbation is instantaneously added to both bodies, which effects a small rotation of the principle axes on the left, but a large one on the right.}
+\end{figure}
+
+Furthermore, the quantity $(\lambda_i-\lambda_j)$ which appears in the denominator of these equations is precisely the same as the quantity which we estimated in the previous section to scale with $\sim \mathrm{Ra}^{-1}$.
 Therefore, as the Rayleigh number increases, the characteristic gap between the eigenvalues of the convective moment becomes smaller.  
@@
 Overall, this makes the principle axes of high Rayleigh number systems much less stable.
-This result is consistent with the result of \citet{richards1997explanation}.
+This is consistent with the result of \citet{richards1997explanation}.
 
@@
 
+Figure \ref{misfit} demonstrates the importance of this ``eigengap'' for the rates of true polar wander.
+
 \section{Discussion}
 
+The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
+As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
+There is a constant competition between growth of the mismatch angle $\theta$ through Equation \ref{eq:grow_perturbation} and its relaxation through Equation \ref{simple_milankovitch}
+We may plug in the estimate for $\Lambda_{ij}$ into equation \ref{simple_milankovitch} to find the strikingly simple expression
+\begin{equation}
+\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
+\label{eq:simplest_milankovitch}
+\end{equation}
+Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the scaling. 
+This is somewhat of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
+However, we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.
+
+More important is the observation that the most important parameters are the Froude number, which acts as the brakes on the system, and $\theta(\mathrm{Ra})$.
+Whereas the rest of the expression has at most a weak dependence on $\mathrm{Ra}$, this is expected to be strongly dependent on it.
+
+Indeed, we can identify two endmember behaviors of Equation \ref{eq:simplest_milankovitch}.
+When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.
+This is the regime considered in \citet{steinberger1997changes}, \citet{roberts2007cause}, and \citet{zhong2007supercontinent}, and can be considered the ``slow TPW'' regime.
+
+When convection is more chaotic, however, there may be large excursions in $\theta$, which can drastically increase the wander rate.  
+If $\theta=90^\circ$, this corresponds to IITPW \citep{kirschvink1997evidence}.  
+This, however, is a special case in the large $\theta$, ``fast TPW'' regime.
+
+\section{Conclusion}
 We have developed a framework for discussing the rates of true polar wander in terms of a small number of nondimensional numbers.
@@
 
+\section {TODO}
+
+\begin{enumerate}
+\item Figure out the best way of nondimensionalizing
+\item What to do about $\alpha \Delta T$?
+\item Connect nondimensionalization with Equation \ref{eq:liouville} via correspondence principle
+\item Make notation everywhere consistent
+\item More discussion of Davis+Kahan?
+\item Make better figures
+\item Make the argument about degree-two scaling much sharper
+\item Can I write equation \ref{eq:liouville} in terms of \ref{maccullagh}?
+\item Discuss elastic lithosphere, Creveling
+\item Discuss high-viscosity lower mantle
+\item Earth history, exoplanets?
+\end{enumerate}
+
 \label{lastpage}

### 2015-01-26 tpw_rate.tex
Work on intro

@@
 
-\documentclass[extra,onecolumn]{gji}
-%\documentclass[extra]{gji}
+%\documentclass[extra,onecolumn]{gji}
+\documentclass[extra,mreferee]{gji}
 \usepackage{timet}
@@
 \label{sec:intro}
+
+A rotating, quasistatic body like a planetary mantle will tend to spin about the axis of its maximum moment of inertia.
+Convection in a planetary mantle continuously redistributes mass, which can change the moment of ineratia tensor, necessitating a change in the spin axis of the planet to conserve angular momentum, a process known as true polar wander (TPW).
+
+TPW was first considered in detail by \citet{darwin1887influence}, and the theory has been subsequently developed by many \citep[e.g.][]{munk1960rotation, goldreich1969some, ricard1993polar}. 
+Despite this, the ability of internal mass anomalies to drive large-scale TPW remains controversial. 
+Paleomagnetic data have been interpreted to requre up to $\sim 1^\circ$/Myr rates of TPW \citep{mitchell2011sutton}, but the ability of the mantle to respond at such rates has been questioned \citep{tsai2007theoretical}.
+
+The primary uncertainties in assigning a maximum TPW rate to a convecting planet are the size of convective anomalies, which drive the rotational adjustment, and the viscosity structure of the mantle, which retards it.
+These two uncertainties are not unrelated: they are both expected to be functions of the geometric and material properties of the mantle.
+As such, they do not vary independently. 
+This suggests that an approach rooted in dimensional analysis and fluid dynamics can clarify the rates and magnitudes of TPW.
+
+Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment if inertia variations \citep[e.g.]{tsai2007theoretical, creveling2012mechanisms}.
+\citet{richards1999polar} coupled full thermal convection models to a polar wander model, but did not address in detail the scaling relationships between the two.
+
+Herein we perform a scaling analysis for a minimal system of a rotating, convecting mantle.
 
@@
 The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
-the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is the broadening of the spatial 
+the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is a broadening of the spatial 
 and temporal spectra \citep{mclaughlin1982transition}.
@@
 As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
-There is a constant competition between growth of the mismatch angle $\theta$ through Equation \ref{eq:grow_perturbation} and its relaxation through Equation \ref{simple_milankovitch}
+There is a constant competition between growth of the mismatch angle $\theta$ through Equation \ref{eq:grow_perturbation} and its relaxation through Equation \ref{simple_milankovitch}.
 We may plug in the estimate for $\Lambda_{ij}$ into equation \ref{simple_milankovitch} to find the strikingly simple expression

### 2015-01-26 tpw_rate.tex
Work on scaling

@@
 
-Herein we perform a scaling analysis for a minimal system of a rotating, convecting mantle.
+Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle.
 
@@
 The quantites $\Lambda_{i,j}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
-We consider relatively slowly rotating bodies here ($\mathrm{Fr} << 1$), so that the rotation does not affect the style of convection.  
+We consider relatively slowly rotating bodies here ($\mathrm{Fr} \ll 1$), so that the rotation does not affect the style of convection.  
 In this regime, therefore, we neglect the dependence on $\mathrm{Fr}$.
@@
 \texttt{Aspect} provides modern numerical methods, flexible geometry, and powerful support for integrating user-defined modules and postprocessors.
-We performed a number of simulations at different Rayleigh numbers, integrating Equation \ref{eq:liouville} at every time step.
+We performed a number of simulations at different Rayleigh numbers, integrating Equation \ref{eq:liouville} in time.
 
@@
 \includegraphics[width=0.9\textwidth]{figures/eigengap.pdf}
-\caption{Oooga}
+\caption{ Left: Time series of the normalized difference between moments $\Lambda_{12}$ for convection in a 2D annulus at several different Rayleigh numbers.
+As the Rayliehg number increases, the average value of the relative moment decreases due to less low-degree coherence in the temperature structure.
+Right:  average value of $\Lambda_{12}$ for the different Rayleigh numbers.
+Also shown is a line with slope $\mathrm{Ra}^{-2/3}$, which is predicted from the scaling analysis (the exponent is $-2/3$ instead of $-1$ due to the reduced dimensionality of the simulations).}
 \end{figure}
@@
 
-We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $X_{lmn}$, 
-where $l$, $m$, and $n$ are wavenumbers.  
+We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
+where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of radial polynomials.  
 \begin{equation} 
-T( x, y ,z, t )  = {\displaystyle \sum_{l m n}^\infty } T_{lmn}(t) X_{lmn} (x , y, z)
+T( r , \theta, \phi, t )  = {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi)
 \label{T_series}
@@
 \begin{equation}
-\lVert T(x,y,z, t) \rVert_\infty = \lVert {\displaystyle \sum_{l m n}^\infty } T_{lmn} (t) X_{lmn} (x , y, z) \rVert_\infty = 1
+\lVert T(x,y,z, t) \rVert_\infty = \lVert {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \rVert_\infty = 1
 \end{equation}
@@
 
+At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes (possibly, but not necessarily degree-two).
+
 At high Rayleigh number we expect there to be a lengthscale which characterizes the smallest length of features in the convecting system.  
 At lengthscales smaller than this diffusion will tend to wipe out heterogeneities, and there will be no power in those modes.  
-Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in \ref{T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:
+Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in Equation \ref{T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:
 
@@
 Any of a number of scaling laws can provide an estimate for the characteristic length scale of a convecting system which may depending on rheology, geometry, or density structure.
-The simplest, based on boundary layer theory \citep{turcotte1967finite}, finds $d/R ~ \mathrm{Ra}^{-1/3}$, thus furnishing us with an estimate of the power in the degree-two part of the field as a function of Rayleigh number:
+The simplest, based on boundary layer theory \citep{turcotte1967finite}, finds $d/R \sim \mathrm{Ra}^{-1/3}$, thus furnishing us with an estimate of the power in the degree-two part of the field as a function of Rayleigh number:
 
@@
 
-Let $\delta$ be the perturbation to the convective moment of inertia, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that moment.  
+Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.  
 Then the rotation of the principal axes of the tensor is bounded by
@@
 Overall, this makes the principle axes of high Rayleigh number systems much less stable.
-This is consistent with the result of \citet{richards1997explanation}.
+This is consistent with the result of \citet{richards1999polar}.
+
+In the limit that the eigengap becomes zero, the rotation of the pinciple axes can be arbitrary. 
+This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}.
 
@@
 \includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
-\caption{Top: }
+\caption{Top: Time series of principle moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.
+Bottom: Time series of spin axis and mismatch angle $\theta$.
+When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger.}
 \end{figure}
-
-Figure \ref{misfit} demonstrates the importance of this ``eigengap'' for the rates of true polar wander.
 

### 2015-01-26 tpw_rate.tex
Change epsilon

@@
  - & \Omega_0^2  \alpha  \Delta T T R \; {\bf \Omega \times \Omega \times r} = - \eta \kappa / R^2 \; \nabla P^* \\ 
-&+ \eta \kappa / R^2 \; \nabla \cdot \left( \eta \varepsilon ({\bf v}) \right) - \alpha \Delta T g_0 \; {\bf g}
+&+ \eta \kappa / R^2 \; \nabla \cdot \left( \eta \epsilon ({\bf v}) \right) - \alpha \Delta T g_0 \; {\bf g}
 \end{aligned}
@@
  - & \mathrm{Ra \; Fr}\; T \; {\bf \Omega \times \Omega \times r} = - \nabla P^* \\ 
-&+ \; \nabla \cdot \left( \eta \varepsilon ({\bf v}) \right) - \mathrm{Ra} \; T \; {\bf g}
+&+ \; \nabla \cdot \left( \eta \epsilon ({\bf v}) \right) - \mathrm{Ra} \; T \; {\bf g}
 \end{aligned}
@@
 \begin{equation}
-{\bf \Omega} \times {\bf \dot{\Omega} } = \frac{1}{\varepsilon I_0 \tau} {\bf \Omega} \times \left( {\bf E \cdot \Omega} \right)
+{\bf \Omega} \times {\bf \dot{\Omega} } = \frac{1}{\epsilon I_0 \tau} {\bf \Omega} \times \left( {\bf E \cdot \Omega} \right)
 \label{eq:liouville}
@@
 \begin{equation}
- \dot{\mitbf{\omega}}  = \frac{1}{\varepsilon I_0 \tau} \left[ {\bf E \cdot \mitbf{\omega}} - \left( {\bf \mitbf{\omega} \cdot E \cdot \mitbf{\omega} } \right) \mitbf{\omega} \right]
+ \dot{\mitbf{\omega}}  = \frac{1}{\epsilon I_0 \tau} \left[ {\bf E \cdot \mitbf{\omega}} - \left( {\bf \mitbf{\omega} \cdot E \cdot \mitbf{\omega} } \right) \mitbf{\omega} \right]
 \end{equation}
@@
 And here is the tedious algebra.
-Let the prefactors $\frac{1}{\varepsilon I_0 \tau}$ be denoted by $A$, and the pole of the coordinate system ($\theta = 0$) associated with the eigenvector for $\lambda_3$.
+Let the prefactors $\frac{1}{\epsilon I_0 \tau}$ be denoted by $A$, and the pole of the coordinate system ($\theta = 0$) associated with the eigenvector for $\lambda_3$.
 Therefore we find that $\mitbf{\omega} = \left[ \sin{\theta} \cos{\phi} \;\; \sin{\theta} \sin{\phi} \;\; \cos{\theta} \right]^T$.
@@
 \begin{aligned}
-\dot{\theta} &= - \frac{1}{2 \varepsilon I_0 \tau} \sin{2 \theta} \left[ (\lambda_3-\lambda_1) \cos^2{\phi} + (\lambda_3-\lambda_2) \sin^2{\phi} \right] \\
-\dot{\phi} &= \;\; \frac{1}{2 \varepsilon I_0 \tau} \sin{2 \phi} \; (\lambda_2 - \lambda_1)
+\dot{\theta} &= - \frac{1}{2 \epsilon I_0 \tau} \sin{2 \theta} \left[ (\lambda_3-\lambda_1) \cos^2{\phi} + (\lambda_3-\lambda_2) \sin^2{\phi} \right] \\
+\dot{\phi} &= \;\; \frac{1}{2 \epsilon I_0 \tau} \sin{2 \phi} \; (\lambda_2 - \lambda_1)
 \end{aligned}
@@
 \begin{equation}
-\dot{\theta} = - \frac{1}{2 \varepsilon \tau}\sin{2 \theta} \frac{(\lambda_3-\lambda_1)}{I_0}
+\dot{\theta} = - \frac{1}{2 \epsilon \tau}\sin{2 \theta} \frac{(\lambda_3-\lambda_1)}{I_0}
 \label{simple_milankovitch}
@@
 With this we may rewrite this in terms of the nondimensional numbers defined above, where 
-$\varepsilon \sim \mathrm{Fr}$ and $\tau \sim \eta / \rho g R = \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}$.
+$\epsilon \sim \mathrm{Fr}$ and $\tau \sim \eta / \rho g R = \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}$.
 
@@
 At this point we do not have estimates for the characteristic magnitudes of $\Lambda_{ij}$ or $\theta$, both of which are crucial for predicting characteristic rates of TPW.
-They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between where the 
-rotation axis it and where the convective moment wants it to be.
+They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principle axis of the convective moment.
 
@@
 
-The quantites $\Lambda_{i,j}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
+The quantites $\Lambda_{ij}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
 We consider relatively slowly rotating bodies here ($\mathrm{Fr} \ll 1$), so that the rotation does not affect the style of convection.  
@@
 \begin{enumerate}
-\item Figure out the best way of nondimensionalizing
-\item What to do about $\alpha \Delta T$?
+\item Figure out the best way of nondimensionalizing ($\alpha \Delta T$)
 \item Connect nondimensionalization with Equation \ref{eq:liouville} via correspondence principle
@@
 \item Make better figures
+\item Integrate numerical results
 \item Make the argument about degree-two scaling much sharper
@@
 \item Discuss high-viscosity lower mantle
-\item Earth history, exoplanets?
+\item Earth-like parameters
 \end{enumerate}

### 2015-02-11 tpw_rate.tex
Disorganized work

@@
 Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment if inertia variations \citep[e.g.]{tsai2007theoretical, creveling2012mechanisms}.
-\citet{richards1999polar} coupled full thermal convection models to a polar wander model, but did not address in detail the scaling relationships between the two.
+\citet{richards1999polar} coupled thermal convection models to a polar wander model, but did not address in detail the scaling relationships between the two.
 
@@
 \subsection{Buckingham Pi theorem and nondimensionalization}
-The Buckingham Pi theorem (essentially an application of the rank-nullspace theorem from linear algebra) allows us to determine the number of nondimensional numbers for this problem.  
+The Buckingham Pi theorem (essentially an application of the rank-nullity theorem) allows us to determine the number of nondimensional numbers for this problem.  
 We count the number of fundamental units for the problem, in this case mass (kg), length (m), time (s), and temperature (K), as well as the number of parameters in use for this problem.  
@@
 
+This can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
+
+\begin{equation}
+-\int_V {\bf r} \times \nabla P^* + \int_V {\bf r} \times \nabla^2 {\bf v} - \mathrm{Ra} \int_V T {\bf r \times g} + \mathrm{Ra \; Fr} \int_V T {\bf r \times \Omega \times \Omega \times r} = 0
+\end{equation}
+
+The first three terms represent pressure, viscous torques, and gravitational torques on the mantle.  
+Convection in the outer core, atmospheres, and oceans is not strong enough to provide significant pressure and viscous torques over geologic timescales, and a self-gravitating body cannot self-torque (citation...).
+Therefore, the we can neglect those terms, and we are left with
+
+\begin{equation}
+\begin{aligned}
+\int_V T {\bf r \times \Omega \times \Omega \times r} = 0 \\
+\end{aligned}
+\end{equation}
+
 \section{Rotational dynamics}
 \label{sec:rotation}
+
+\begin{equation}
+{\bf \Omega \times C \cdot \Omega} = 0
+\end{equation}
+
+Following \citet{ricard1993polar}, the total moment of inertia of the mantle may be separated into three parts, a spherically symmetric contribution, a rotationally deformed contribution, and a contribution due to the internal mass anomalies:
+
+\begin{equation}
+C_{ij} = I_0 \delta_{ij} + J_{ij} + E_{ij}
+\end{equation}
+
+where ${\bf E} = \left[ \delta(t) + k^L(t) \right] * {\bf C}$ is the convective part convolved with the surface response ot internal density anomalies, and $
 

### 2015-02-13 figures/eigengap.py
Update statistics files, work on figures

@@ -33,7 +33,7 @@ frac = 50
 def get_spin_spectrum(text_file):
 
   #load the data
-  data = np.loadtxt(text_file,skiprows=50, usecols=(1,15,16,17,18))
+  data = np.loadtxt(text_file,skiprows=50, usecols=(1,17,18,19,20))
   #get rid of duplicate times (why might these exist?)
   stats = []
   for i in range(len(data)-1):
@@ -89,9 +89,9 @@ for f, i in zip( output_files, range(len(output_files)) ):
 
 
   times, diff, acorrdiff, peak = get_spin_spectrum(f)
-  diff = diff/alpha/delta_T
-  acorrdiff = acorrdiff/alpha/delta_T/alpha/delta_T
   tmax = times[-1]
+  tmin = times[0]
+  times = times-tmin
   
   Rayleighs.append(Ra)
   amplitudes.append(diff.mean())
@@ -99,8 +99,8 @@ for f, i in zip( output_files, range(len(output_files)) ):
 
   ax = plt.subplot(1,2,1)
   ax.plot(times, diff,  label=r'$\mathrm{Ra} = %.1f \times 10^{%i}$' % ( Ra/np.power(10., np.floor(np.log10(Ra))) , np.floor(np.log10(Ra))))
-  ax.set_xlim(0,5.)
-  ax.set_ylim(0, 1.0e-2)
+  ax.set_xlim(0,7)
+  ax.set_ylim(0, 1.0e-4)
   ax.set_xlabel("Time (Gyr)")
   ax.set_ylabel(r'Relative moment $(\lambda_2-\lambda_1)/I_0$')
 
@@ -117,16 +117,16 @@ ax.set_yscale('log')
 ax.xaxis.set_major_formatter(matplotlib.ticker.LogFormatterMathtext())
 ax.yaxis.set_major_formatter(matplotlib.ticker.LogFormatterMathtext())
 
-Ra=np.logspace(7.3, 8.3)
-Moment = 350.*np.power(Ra, -2./3.)
+Ra=np.logspace(6.6, 8.3)
+Moment = 2.5*np.power(Ra, -2./3.)
 
 ax.plot(Ra, Moment, '--', c='0.5', linewidth=4)
 
 clist = plt.rcParams['axes.color_cycle']
 ax.scatter(Rayleighs, np.array(amplitudes), c = clist, s=100)
 
[…]

### 2015-02-13 figures/misfit.py

@@ -1,13 +1,14 @@
-
 import matplotlib
-
 import matplotlib.pyplot as plt
 import matplotlib.ticker
 import numpy as np
+
+import scipy.signal as signal
+import scipy.interpolate as interp
+
 plt.style.use(['ian'])
 
-#c = ['#0A5F02', '#1c567a', '#814292', '#d7383b', '#fdae61', '#c0f8b8']
-data = np.loadtxt('statistics_500', skiprows=50, usecols=(1,15,16,17,18))
+data = np.loadtxt('statistics_500', skiprows=50, usecols=(1,17,18,19,20))
 
 rho = 3300
 alpha_0 = 4.e-5
@@ -19,15 +20,20 @@ eta_0 = 1.e21
 D = 6371.e3
 g = 9.81
 
+time = data[:,0]/1.e9
+t_start = 9.
+time = time - t_start
+tmin = 0.
+tmax = 2.
 
-time = data[:,0]*kappa/D/D
-tmax = 0.5e-10
 mismatch = data[:,2]
+#Interpolate using splines for a bit nicer-looking curves
+theta = interp.UnivariateSpline( time, mismatch, s=8.e4)
+
 eig1 = data[:,3]
 eig2 = data[:,4]
-eigmean = np.mean(eig1[np.where( time<tmax )] + eig2[np.where(time<tmax)])/2.
-eig1 = (eig1-eigmean)/eigmean
-eig2 = (eig2-eigmean)/eigmean
+eigmean = np.mean(eig1 + eig2)/2.
+eigdiff = (eig1-eig2)/eigmean
 
 spin = data[:,1]
 
@@ -36,14 +42,11 @@ time[pos] = np.nan
 spin[pos] = np.nan
 
 fig = plt.figure()
-#fig.subplots_adjust(wspace=0.0, hspace=0.0)
-
 ax = fig.add_subplot(211)
 
-ax.plot(time, eig1, label=r'$\lambda_1$')
-ax.plot(time, eig2, label=r'$\lambda_2$')
-plt.xlim(0.0, tmax)
-plt.ylim(-4.e-5, 4.e-5)
+ax.plot(time, eigdiff, label=r'$(\lambda_1-\lambda_2)/I_0$')
+plt.xlim(tmin, tmax)
+plt.ylim(0., 2.e-5)
 ax.legend(loc='upper right')
 ax.xaxis.set_major_formatter(matplotlib.ticker.NullFormatter())
[…]

### 2015-02-26 tpw_rate.tex
Work on rotational dynamics

@@
 
-Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment if inertia variations \citep[e.g.]{tsai2007theoretical, creveling2012mechanisms}.
+Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment if inertia variations \citep[e.g.][]{tsai2007theoretical, creveling2012mechanisms}.
 \citet{richards1999polar} coupled thermal convection models to a polar wander model, but did not address in detail the scaling relationships between the two.
@@
 Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle.
+
+\section{Rotational dynamics}
+
+\subsection{The Liouville equation}
+Conservation of angular momentum for a torque-free system in a rotating reference frame requires
+
+\begin{equation}
+\frac{d {\bf H}}{dt} + {\bf \Omega} \times {\bf H} = 0
+\label{euler}
+\end{equation}
+
+known as Euler's equations, where ${\bf H} = {\bf I} \cdot {\bf \Omega}$ is the angular momentum vector, $\bf I$ is the moment of inertia tensor, and $\bf \Omega$ is the angular velocity.
+On a dynamic planet $\bf I$ may be a function of time, so to conserve angular momentum $\bf \Omega$ must be as well.
+In this case Equation \ref{euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
+For a slowly convecting fluid such as a planetary mantle the inertial term $\partial {\bf H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
+\begin{equation}
+{\bf \Omega } (t)\times ( {\bf I}(t) \cdot {\bf \Omega }(t) ) = 0
+\label{quasistatic_liouville}
+\end{equation}
+
+Note that solving Equation \ref{quasistatic_liouville} is equivalent to solving an eigenvalue problem for ${\bf I}$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
+In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
+The moment of inertia tensor above includes all contributions to the mass structure of the planet, including
+\begin{itemize}
+\item the spherically symmetric mass distribution
+\item the rotational deformation
+\item internal density anomalies
+\item surface deflections due to internal density anomalies
+\end{itemize}
+
+The moment of inertia tensor is commonly separated into three parts \citep{ricard1993polar}:
+
+\begin{equation}
+I_{ij}(t) = I_0 \delta_{ij} + J_{ij}(t) + C_{ij}(t)
+\label{separation}
+\end{equation}
+
+wher $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $C_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
+If we plug this decomposition into Equation \ref{quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
+
+\begin{equation}
+{\bf \Omega} \times ({\bf J} \cdot {\bf \Omega}) = -{\bf \Omega } \times ( {\bf C} \cdot {\bf \Omega} )
+\end{equation}
+
+This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches of the convective part of the moment of inertia ($\bf C$) and the rotational deformation part of the moment of inertia ($\bf J$).
+If $\bf J$ and $\bf C$ are coaxial, then the mismatch must be zero, and solving the polar wander problem reduces to solving an eigenvalue problem for $\bf C$.
+Our goal is to characterize this balance.
+
+\subsection{Rotational deformation}
+
+The part of the moment of inertia due to the elastic rotational deformation is traditionally related to the degree-two part of the gravity field via MacCullagh's formula \citep{munk1960rotation}:
+
+\begin{equation}
+J_{ij} = \frac{k a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right)
+\label{elastic_deformation}
+\end{equation}
+
+where $k$ is an elastic Love number, $a$ is the semimajor axis of the planet, and $G$ is the gravitational constant.
+This result may be extended to a viscoelastic rheology via the viscoelastic correspondence principle \citep[e.g.][]{peltier1974impulse}:
+
+\begin{equation}
+J_{ij} = \frac{k(t) a^5}{3 G} * \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right)
+\label{viscoelastic_deformation}
+\end{equation}
+
+where $k$ is now a time-dependent viscoelastic Love number which is convolved with the time-dependent rotation vector.
+
+The infinite time limit of Equation \ref{viscoelastic_deformation} allows us to calculate the fluid limit of the Love number $k_f$:
+\begin{equation}
+k_f = \frac{3 G (C-A)}{\Omega^2 a^5}
+\end{equation}
+
+where $C$ and $A$ are the polar and equatorial moments of inertia, respectively.
+
+Following \citet{ricard1993polar}
+..
+..(fill in gaps)
+
+\begin{equation}
+J_{ij} = \frac{k_f a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right) -
+ \frac{k_f a^5 T_1}{3G} \left(\dot{\Omega}_i \Omega_j + \Omega_i \dot{\Omega}_j - \frac{2}{3} \Omega_q \dot{\Omega}_q \delta_{ij} \right)
+\end{equation}
+
+The two terms of this equation have simple interpretations.  
+The first term corresponds to the fluid limit of rotational deformation (in the absence of any long-term elastic strength).  
+The second term represents the lag in the moment of inertia due to the viscous adjustment of the rotational bulge, where $T_1$ is the characteristic time constant for this adjustment.
+Since the first term represents the fluid limit of rotational deformation, it automatically satisfies Equation \ref{quasistatic_liouville}, and hence does not contribute to the polar wander problem.
 
@@
 
-where ${\bf E} = \left[ \delta(t) + k^L(t) \right] * {\bf C}$ is the convective part convolved with the surface response ot internal density anomalies, and $
+where ${\bf E} = \left[ \delta(t) + k^L(t) \right] * {\bf C}$ is the convective part convolved with the surface response ot internal density anomalies, and
 

### 2015-03-09 tpw_rate.tex
Work

@@
 These two uncertainties are not unrelated: they are both expected to be functions of the geometric and material properties of the mantle.
-As such, they do not vary independently. 
-This suggests that an approach rooted in dimensional analysis and fluid dynamics can clarify the rates and magnitudes of TPW.
+As such, they do not vary independently, and first-order questions about the propensity for planets to experience TPW remain: how are rates of TPW expected to vary with the vigor of convection? 
+Are other planetary bodies more or less likely than Earth to experience TPW?  And are these rates expected to vary in Earth history?
 
+These questions suggest that an approach rooted in dimensional analysis and fluid dynamics can clarify the rates and magnitudes of TPW.
 Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment if inertia variations \citep[e.g.][]{tsai2007theoretical, creveling2012mechanisms}.
@@
 \begin{equation}
-I_{ij}(t) = I_0 \delta_{ij} + J_{ij}(t) + C_{ij}(t)
+I_{ij}(t) = I_0 \delta_{ij} + J_{ij}(t) + E_{ij}(t)
 \label{separation}
@@
 
-wher $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $C_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
+wher $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
 If we plug this decomposition into Equation \ref{quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
@@
 \begin{equation}
-{\bf \Omega} \times ({\bf J} \cdot {\bf \Omega}) = -{\bf \Omega } \times ( {\bf C} \cdot {\bf \Omega} )
+{\bf \Omega} \times ({\bf J} \cdot {\bf \Omega}) = -{\bf \Omega } \times ( {\bf E} \cdot {\bf \Omega} )
+\label{disequilibrium}
 \end{equation}
 
-This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches of the convective part of the moment of inertia ($\bf C$) and the rotational deformation part of the moment of inertia ($\bf J$).
-If $\bf J$ and $\bf C$ are coaxial, then the mismatch must be zero, and solving the polar wander problem reduces to solving an eigenvalue problem for $\bf C$.
-Our goal is to characterize this balance.
+This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches of the convective part of the moment of inertia ($\bf E$) and the rotational deformation part of the moment of inertia ($\bf J$).
+If $\bf J$ and $\bf E$ are coaxial, then the mismatch must be zero, and solving the polar wander problem reduces to solving an eigenvalue problem for $\bf E$.
+Our goal is to characterize this balance from a perspective of scaling and fluid dynamics.
 
@@
 
-The infinite time limit of Equation \ref{viscoelastic_deformation} allows us to calculate the fluid limit of the Love number $k_f$:
+The infinite-time limit of Equation \ref{viscoelastic_deformation} allows us to calculate the fluid limit of the Love number $k_f$:
 \begin{equation}
 k_f = \frac{3 G (C-A)}{\Omega^2 a^5}
+\label{fluid_love}
 \end{equation}
@@
 where $C$ and $A$ are the polar and equatorial moments of inertia, respectively.
+The infinite-time limit, however, does not allow for any disequilibrium between $J_{ij}$ and $C_{ij}$.
 
-Following \citet{ricard1993polar}
-..
-..(fill in gaps)
+\citet{ricard1993polar} obtain an approximation to Equation \ref{viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.
+This introduces a new parameter, termed $T_1$, which can be seen as a weighted tidal relaxation time for the system.  
+This simple approximation in the Laplace domain allows for an analytical transformation back into the time domain, and neglecting second order terms in $\dot{\Omega}$ we find:
 
@@
  \frac{k_f a^5 T_1}{3G} \left(\dot{\Omega}_i \Omega_j + \Omega_i \dot{\Omega}_j - \frac{2}{3} \Omega_q \dot{\Omega}_q \delta_{ij} \right)
+\label{rotational_deformation}
 \end{equation}
@@
 
-\section{Governing equations}
-\label{sec:govern}
+\subsection{The convective moment of inertia}
+
+The term on the right-hand side of equation \ref{disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
+This, as well, may be parameterized using a viscoelastic Love number approach:
+\begin{equation} 
+{\bf E} = \left[ \delta(t) + k^L(t) \right] * {\bf C}
+\end{equation}
+
+where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, and $C_{ij}$ is the moment of inertia without that response.
+Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:
+\begin{equation}
+E_{ij} = (1+k^L_f) C_{ij}
+\end{equation}
+
+..
+..
+(Text on free surface?)
+
+\subsection{Rate of true polar wander}
+
+We are in a position to address the rates of true polar wander for a given convective moment $\bf E$.
+A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d/dt (\Omega^2) = 2 {\Omega_i} \dot{ \Omega}_i = 0)$
+Substituting Equation \ref{rotational_deformation} into Equation \ref{disequilibrium} we find
+
+\begin{equation}
+\frac{k_f a^5 T_1}{3G}{\bf \Omega} \times {\bf \dot{\Omega} } = {\bf \Omega} \times \left( {\bf E \cdot \Omega} \right)
+\label{eq:liouville}
+\end{equation}
+
+Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation \ref{fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
+\begin{equation}
+ \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ {\bf E \cdot \mitbf{\omega}} - \left( {\bf \mitbf{\omega} \cdot E \cdot \mitbf{\omega} } \right) \mitbf{\omega} \right]
+\end{equation}
+
+Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
+If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
+
+\ifdetail
+
+And here is the tedious algebra.
+Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$, and the pole of the coordinate system ($\theta = 0$) associated with the eigenvector for $\lambda_3$.
+Therefore we find that $\mitbf{\omega} = \left[ \sin{\theta} \cos{\phi} \;\; \sin{\theta} \sin{\phi} \;\; \cos{\theta} \right]^T$.
+Plugging this in, we find a system of three equations, which we want to solve for $\dot{\theta}$ and $\dot{\phi}$:
+
+\begin{equation}
+\begin{aligned}
+ & \cos{\theta}\cos{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \dot{\phi} = \\
+  &A \sin{\theta}\cos{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+ &\cos{\theta}\sin{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \dot{\phi} = \\
+  &A \sin{\theta}\sin{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+ - &\sin{\theta} \dot{\theta} = \\
+  &A \cos{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+\end{aligned}
+\end{equation}
+
+Now multiply the top equation by $\cos{\phi}$ and the second equation by $\sin{\phi}$ and add them together to get
+
+\begin{equation}
+\begin{aligned}
+ & \cos{\theta}\cos^2{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \cos{\phi} \dot{\phi} + \cos{\theta}\sin^2{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \sin{\phi} \dot{\phi} = \\
+  &A \sin{\theta}\cos^2{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] + \\
+  &A \sin{\theta}\sin^2{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+ & \cos{\theta}\dot{\theta} =  
+   A \sin{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
+  &A \sin{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+\end{aligned}
+\end{equation}
+
+Now multiply the above by $\cos{\theta}$ and the third equation by $-\sin{\theta}$ and add them to get
+
+\begin{equation}
+\begin{aligned}
+ & \cos^2{\theta}\dot{\theta} + \sin^2{\theta} \dot{\theta} =  
+   A \sin{\theta} \cos{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
+  &A \sin{\theta} \cos{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] - \\
+  &A \cos{\theta} \sin{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+\end{aligned}
+\end{equation}
+
+After a lot of cancellation, we find
+
+\begin{equation}
+\begin{aligned}
+ \dot{\theta} &= A \sin{\theta} \cos{\theta} \left[ - \lambda_3 + \lambda_1 \cos^2{\phi} + \lambda_2 \sin^2{\phi} \right] \\
+              &= - A \sin{\theta} \cos{\theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
+              &= -\frac{A}{2} \sin{2 \theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
+\end{aligned}
+\end{equation}
+
+Likewise we may multiply the first equation by $-\sin{\phi}$ and the second equation by $\cos{\phi}$ and add them to get
+
+\begin{equation}
+\begin{aligned}
+ & \sin{\theta} \dot{\phi} = A \sin{\theta}\cos{\phi}\sin{\phi} \left( \lambda_2 - \lambda_1 \right ) \\
+& \dot{\phi} = \frac{A}{2} \sin{2 \phi} (\lambda_2-\lambda_1)
+\end{aligned}
+\end{equation}
+
+\else
+\fi
+
+\begin{equation}
+\begin{aligned}
+\dot{\theta} &= - \frac{1}{(C-A)T_1} \sin{2 \theta} \left[ (\lambda_3-\lambda_1) \cos^2{\phi} + (\lambda_3-\lambda_2) \sin^2{\phi} \right] \\
+\dot{\phi} &= \;\; \frac{1}{(C-A)T_1} \sin{2 \phi} \; (\lambda_2 - \lambda_1)
+\end{aligned}
+\label{milankovitch}
+\end{equation}
+
+Without loss of generality we may choose the coordinate system such that $\phi=0$:
+\begin{equation}
+\dot{\theta} = - \frac{1}{(C-A)T_1} (\lambda_3-\lambda_1) \sin{2 \theta}
+\label{simple_milankovitch}
+\end{equation}
+
+These equations are a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.  
+Written this way, it is clear that the important quantities are $\theta$ and the differences between the eigenvalues of the convective moment $\bf E$, both of which depend on the structure and dynamics of mantle convection.
+They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principle axis of the convective moment.
+
+In order to estimate $\theta$ and $\lambda_i$ we must examine the linear momentume equation.
+
+\section{Linear momentum equation}
+\label{sec:linear}
 
@@
 \begin{aligned}
- \rho{\bf \Omega \times \Omega \times r}= - \nabla P + \nabla \cdot \left( \eta \varepsilon ({\bf u}) \right) + \rho {\bf g}
+ \rho{\bf \Omega \times \Omega \times r}= - \nabla P + \nabla^2 {\bf u} + \rho {\bf g}
 \label{navier_stokes}
@@
 
- where the parameters are defined in Table~\ref{parameters} and $\varepsilon({\bf v}) = 1/2 \; ( \partial v_i / \partial x_j + \partial v_j / \partial x_i )$ corresponds to the symmetrized velocity gradient.
+where the parameters are defined in Table~\ref{parameters}.
 In addition we use the simple equation of state

### 2015-03-18 tpw_rate.tex
Disorganized work

@@
 
-In order to estimate $\theta$ and $\lambda_i$ we must examine the linear momentume equation.
+In order to estimate $\theta$ and $\lambda_i$ we must examine the linear momentum equation.
 
@@
 As mantle convection and rotational dynamics of planetary bodies are usually considered separately, some extra care must be taken to establish self-consistent governing equations.  
-Here we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.  The equations for mass, momentum, and energy then read
+For simplicty we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.
+The equations for mass, momentum, and energy then read
 
@@
 Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
-Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires \_\_\_\_? nondimensional numbers to characterize it.
+Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires three nondimensional numbers to characterize it.
+Convenient choices for these numbers are listed in Table \ref{nondim}, along with approximate Earth-like values for them.
+
+In particular, the most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
+This second nondimensional number does not have a uniformly agreed-upon name, but has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, which we adopt here.
+Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
+It would be straightforward to include them, but does not affect the overall treatment of the effects of rotation.
 
@@
 
-Convenient choices for these numbers are shown in Table \ref{nondim} along with approximate Earthlike values.
-In addition to the more usual Rayleigh and Prandtl numbers we include a Froude number, characterizing the ratio of centrifugal to gravitational forces, and $\Gamma$, a nondimensional characterization of density variations.
 Nondimensionalizing with these parameters we obtain
@@
 \begin{aligned}
- - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\bf \Omega \times \Omega \times r} = 0
+ - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times r} = 0
 \end{aligned}
@@
 Ra & Rayleigh &  $\rho_0 g_0 \alpha \Delta T R^3/\eta \kappa$ & $10^7$\\
-Pr & Prandtl & $\eta/\rho_0 \kappa$ & $10^{23}$ \\
 Fr & Froude & $\Omega_0^2 R/g_0$ & $10^{-3}$ \\
-$\Gamma$ & Density deficit & $\alpha \Delta T$ & $10^{-2}$ \\
-A & Aspect ratio & $R_i/R$ & $0.55$ \\
+A & Aspect ratio & $R_i/R$ & $0.54$ \\
 \end{tabular}
@@
 \begin{equation}
--\int_V {\bf r} \times \nabla P^* + \int_V {\bf r} \times \nabla^2 {\bf v} - \mathrm{Ra} \int_V T {\bf r \times g} + \mathrm{Ra \; Fr} \int_V T {\bf r \times \Omega \times \Omega \times r} = 0
+-\int_V {\bf r} \times \nabla P^* + \int_V {\bf r} \times \nabla^2 {\bf v} - \mathrm{Ra} \int_V T {\bf r \times g} + \mathrm{Ra \; Fr} \int_V T {\bf r \times \mitbf{\omega} \times \mitbf{\omega} \times r} = 0
 \end{equation}
@@
 \begin{aligned}
-\int_V T {\bf r \times \Omega \times \Omega \times r} = 0 \\
+\int_V T {\bf r \times \mitbf{\omega} \times \mitbf{\omega} \times r} = 0 \\
 \end{aligned}
@@
 
-\section{Rotational dynamics}
-\label{sec:rotation}
+\section{Scaling}
+\label{sec:scaling}
 
+It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
+This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
+We also define a new nondimensional number (which will come in handy in intermediate steps) corresponding to the fractional density changes due to temperature variations: $\Gamma = \alpha \Delta T$.
+With this we may rewrite Equation \ref{simple_milankovitch} in terms of the nondimensional numbers defined above.
+The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep[e.g.][]{??}.
 \begin{equation}
-{\bf \Omega \times C \cdot \Omega} = 0
+(C-A) \sim I_0 \mathrm{Fr}
 \end{equation}
 
-Following \citet{ricard1993polar}, the total moment of inertia of the mantle may be separated into three parts, a spherically symmetric contribution, a rotationally deformed contribution, and a contribution due to the internal mass anomalies:
-
+The time constant $T_1$ is related to the viscous tidal relaxation time of the planet.
+It is frequently represented as a weighted average of the different relaxation modes of the planet \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
 \begin{equation}
-C_{ij} = I_0 \delta_{ij} + J_{ij} + E_{ij}
+T_1 \sim \eta / \rho g R = \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}
 \end{equation}
-
-where ${\bf E} = \left[ \delta(t) + k^L(t) \right] * {\bf C}$ is the convective part convolved with the surface response ot internal density anomalies, and
-
-We may write the nonlinear Liouville equations 
-\begin{equation}
-{\bf \Omega} \times {\bf \dot{\Omega} } = \frac{1}{\epsilon I_0 \tau} {\bf \Omega} \times \left( {\bf E \cdot \Omega} \right)
-\label{eq:liouville}
-\end{equation}
-
-introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
-\begin{equation}
- \dot{\mitbf{\omega}}  = \frac{1}{\epsilon I_0 \tau} \left[ {\bf E \cdot \mitbf{\omega}} - \left( {\bf \mitbf{\omega} \cdot E \cdot \mitbf{\omega} } \right) \mitbf{\omega} \right]
-\end{equation}
-
-Note that the quantity in brackets is identical in form to the shear stress on a plane in classical elastostatics.
-If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
-
-\ifdetail
-
-And here is the tedious algebra.
-Let the prefactors $\frac{1}{\epsilon I_0 \tau}$ be denoted by $A$, and the pole of the coordinate system ($\theta = 0$) associated with the eigenvector for $\lambda_3$.
-Therefore we find that $\mitbf{\omega} = \left[ \sin{\theta} \cos{\phi} \;\; \sin{\theta} \sin{\phi} \;\; \cos{\theta} \right]^T$.
-Plugging this in, we find a system of three equations, which we want to solve for $\dot{\theta}$ and $\dot{\phi}$:
-
-\begin{equation}
-\begin{aligned}
- & \cos{\theta}\cos{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \dot{\phi} = \\
-  &A \sin{\theta}\cos{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
- &\cos{\theta}\sin{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \dot{\phi} = \\
-  &A \sin{\theta}\sin{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
- - &\sin{\theta} \dot{\theta} = \\
-  &A \cos{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
-\end{aligned}
-\end{equation}
-
-Now multiply the top equation by $\cos{\phi}$ and the second equation by $\sin{\phi}$ and add them together to get
-
-\begin{equation}
-\begin{aligned}
- & \cos{\theta}\cos^2{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \cos{\phi} \dot{\phi} + \cos{\theta}\sin^2{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \sin{\phi} \dot{\phi} = \\
-  &A \sin{\theta}\cos^2{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] + \\
-  &A \sin{\theta}\sin^2{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
- & \cos{\theta}\dot{\theta} =  
-   A \sin{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
-  &A \sin{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
-\end{aligned}
-\end{equation}
-
-Now multiply the above by $\cos{\theta}$ and the third equation by $-\sin{\theta}$ and add them to get
-
-\begin{equation}
-\begin{aligned}
- & \cos^2{\theta}\dot{\theta} + \sin^2{\theta} \dot{\theta} =  
-   A \sin{\theta} \cos{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
-  &A \sin{\theta} \cos{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] - \\
-  &A \cos{\theta} \sin{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
-\end{aligned}
-\end{equation}
-
-After a lot of cancellation, we find
-
-\begin{equation}
-\begin{aligned}
- \dot{\theta} &= A \sin{\theta} \cos{\theta} \left[ - \lambda_3 + \lambda_1 \cos^2{\phi} + \lambda_2 \sin^2{\phi} \right] \\
-              &= - A \sin{\theta} \cos{\theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
-              &= -\frac{A}{2} \sin{2 \theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
-\end{aligned}
-\end{equation}
-
-Likewise we may multiply the first equation by $-\sin{\phi}$ and the second equation by $\cos{\phi}$ and add them to get
-
-\begin{equation}
-\begin{aligned}
- & \sin{\theta} \dot{\phi} = A \sin{\theta}\cos{\phi}\sin{\phi} \left( \lambda_2 - \lambda_1 \right ) \\
-& \dot{\phi} = \frac{A}{2} \sin{2 \phi} (\lambda_2-\lambda_1)
-\end{aligned}
-\end{equation}
-
-\else
-\fi
-
-\begin{equation}
-\begin{aligned}
-\dot{\theta} &= - \frac{1}{2 \epsilon I_0 \tau} \sin{2 \theta} \left[ (\lambda_3-\lambda_1) \cos^2{\phi} + (\lambda_3-\lambda_2) \sin^2{\phi} \right] \\
-\dot{\phi} &= \;\; \frac{1}{2 \epsilon I_0 \tau} \sin{2 \phi} \; (\lambda_2 - \lambda_1)
-\end{aligned}
-\label{milankovitch}
-\end{equation}
-
-Without loss of generality we may choose the coordinate system such that $\phi=0$:
-\begin{equation}
-\dot{\theta} = - \frac{1}{2 \epsilon \tau}\sin{2 \theta} \frac{(\lambda_3-\lambda_1)}{I_0}
-\label{simple_milankovitch}
-\end{equation}
-
-These equations are a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.  
-Written this way, it is clear that the important quantities are $\theta$ and the differences between the eigenvalues of the convective moment $\bf E$.
-It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
-With this we may rewrite this in terms of the nondimensional numbers defined above, where 
-$\epsilon \sim \mathrm{Fr}$ and $\tau \sim \eta / \rho g R = \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}$.
 
@@
 
-\section{Scaling}
-\label{sec:scaling}
-
 The quantites $\Lambda_{ij}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
-We consider relatively slowly rotating bodies here ($\mathrm{Fr} \ll 1$), so that the rotation does not affect the style of convection.  
-In this regime, therefore, we neglect the dependence on $\mathrm{Fr}$.
+We consider relatively slowly rotating bodies here ($\mathrm{Fr} \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
+In this regime, therefore, we neglect the dependence on $\mathrm{Fr}$, and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra})$ and $\theta(\mathrm{Ra})$.
 
@@
 
-The nonhydrostatic moment of inertia is defined
-\begin{equation}
-J_{ij} = \int_{V_S} \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) 
-\end{equation}
+TODO: Needs work with notational consistency/connecting to previous scaling...
 
-where $V_S$ is the reference spherical volume of the mantle.
-If we substitute in Equation \ref{eos} we find
+Fluctuations in the nonhydrostatic moment of inertia are due to temperature fluctuations in the interior and their spatial structure.
 
 \begin{equation}
-J_{ij} = I_0 \delta_{ij} + \int_{V_S} \rho_0 \alpha T \left( r_q r_q \delta_{ij} - r_i r_j \right) = I_0 \delta_{ij} + E_{ij} 
+C_{ij} = \int_{V_S} \rho_0 \alpha T \left( r_q r_q \delta_{ij} - r_i r_j \right) 
 \end{equation}
+where $V_S$ is the reference spherical volume of the mantle.
+
+These internal density loads may be convolved with surface responses to get the convective moment of inertia $\bf E$.
 
@@
 \begin{enumerate}
-\item Figure out the best way of nondimensionalizing ($\alpha \Delta T$)
-\item Connect nondimensionalization with Equation \ref{eq:liouville} via correspondence principle
 \item Make notation everywhere consistent
@@
 \item Make the argument about degree-two scaling much sharper
-\item Can I write equation \ref{eq:liouville} in terms of \ref{maccullagh}?
 \item Discuss elastic lithosphere, Creveling

### 2015-04-02 tpw_rate.tex
Disorganized work

@@
 \begin{aligned}
- \rho{\bf \Omega \times \Omega \times r}= - \nabla P + \nabla^2 {\bf u} + \rho {\bf g}
+- \nabla P + \eta \nabla^2 {\bf g} =  \rho {\bf g} -  \rho{\bf \Omega \times \Omega \times r}
 \label{navier_stokes}
@@
 Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
-It would be straightforward to include them, but does not affect the overall treatment of the effects of rotation.
+It would be straightforward to include them, but does not affect the overall treatment of this scaling.
+
+Customarily a hydrostatic reference state is defined as the solution to \ref{navier_stokes} where there is no flow:
+
+\begin{equation}
+\begin{aligned}
+- \nabla P_0 =  \rho_0 {\bf g} -  \rho_0 {\bf \Omega \times \Omega \times r}
+\label{hydrostatic}
+\end{aligned}
+\end{equation}
+
+where $P_0$ is the reference pressure.
+We can define a ``dynamic pressure'' $P^* = P - P_0$.
+
+Nondimensionalizing with the parameters in Table \ref{nondim} we find
+\begin{equation}
+\begin{aligned}
+ - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times {\bf r}} = 0
+\end{aligned}
+\end{equation}
+
+The figure for the planet which is consistent with the hydrostatic state we call $V_H$.
+The actual figure of the planet's mantle, $V$, may have some disequilibrium with this hydrostatic shape: $V = V_H + \Delta V $.
+The introduction of $\Delta V$ requires another nondimensional number to characterize it, and we find that the quantity $\Gamma = \alpha \Delta T$ is convenient.
+
+Equation \ref{navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
+
+\begin{equation}
+\begin{aligned}
+-\int_V {\bf r} \times \nabla P + \int_V \eta {\bf r} \times \nabla^2 {\bf u} - \int_V \rho {\bf r \times g} + \int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} &= 0 \\
+\end{aligned}
+\end{equation}
+
+The first three terms represent pressure, viscous torques, and gravitational torques on the mantle.  
+Convection in the outer core, atmospheres, and oceans is not strong enough to provide significant pressure and viscous torques over geologic timescales, and a self-gravitating body cannot self-torque (citation...).
+Therefore, the we can neglect those terms, and we are left with
+
+\begin{equation}
+\begin{aligned}
+\int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} &= 0 \\
+\end{aligned}
+\end{equation}
+
+which is simply a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.
+Splitting the volume into $V_H$ and $\Delta V$ and substituting \ref{eos} we find
+
+\begin{equation}
+\begin{aligned}
+&\int_{V_H} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} + 
+\int_{V_H} \rho_0 \alpha T {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} +  \\
+&\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} + 
+\int_{\Delta V} \rho_0 \alpha T {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} = 0  \\
+\end{aligned}
+\end{equation}
+
+The first term of this equation is zero due to the hydrostatic equation.  
+The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\alpha \Delta T$.
 
@@
 
-Nondimensionalizing with these parameters we obtain
-\begin{equation}
-\begin{aligned}
- - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times r} = 0
-\end{aligned}
-\end{equation}
-
-where we have introduced a dynamic pressure $P^* = P - P_0$.
-
 \begin{table}
@@
 
-This can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
-
-\begin{equation}
--\int_V {\bf r} \times \nabla P^* + \int_V {\bf r} \times \nabla^2 {\bf v} - \mathrm{Ra} \int_V T {\bf r \times g} + \mathrm{Ra \; Fr} \int_V T {\bf r \times \mitbf{\omega} \times \mitbf{\omega} \times r} = 0
-\end{equation}
-
-The first three terms represent pressure, viscous torques, and gravitational torques on the mantle.  
-Convection in the outer core, atmospheres, and oceans is not strong enough to provide significant pressure and viscous torques over geologic timescales, and a self-gravitating body cannot self-torque (citation...).
-Therefore, the we can neglect those terms, and we are left with
-
-\begin{equation}
-\begin{aligned}
-\int_V T {\bf r \times \mitbf{\omega} \times \mitbf{\omega} \times r} = 0 \\
-\end{aligned}
-\end{equation}
-
 \section{Scaling}
@@
 
-The time constant $T_1$ is related to the viscous tidal relaxation time of the planet.
-It is frequently represented as a weighted average of the different relaxation modes of the planet \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
+The time constant $T_1$ is a viscous tidal relaxation time of the planetary mantle.
+It is frequently represented as a weighted average of the different relaxation modes \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
 \begin{equation}

### 2015-04-02 tpw_rate.tex
Work

@@
 \usepackage{graphicx}
+\usepackage{color, soul}
 
@@
 
+\hl{This text is still causing me difficulty. Nondimensionalizing the hydrostatic equation does indeed result in an $\alpha T$ dependence.  }
+
 Equation \ref{navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
@@
 which is simply a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.
+
+\hl{I suppose this could stop here. Another couple of lines of text will get to the quasistatic Liouville equation.  However, it does not really make a clear connection to the nondimensionalization }
+
 Splitting the volume into $V_H$ and $\Delta V$ and substituting \ref{eos} we find
@@
 The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\alpha \Delta T$.
+Removing these, we find
+\begin{equation}
+\begin{aligned}
+\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} = 
+-\int_{V_H} \rho_0 \alpha T {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }}
+\end{aligned}
+\end{equation}
+
+This equation may be identified with \ref{disequilibrium}, where the mismatch of the convective moment of intertia with the spin axis is balanced by disequilibrium in the rotational deformation.
+
+\hl{(I am not sure about this strategy for connecting the linear momentum equation with the angular momentum equation.
+In particular, I could go much further, and divide $V$ into more pieces: $V = V_S + \Delta V_H + \Delta V_L + \Delta V_D$, where we have, repsectively, the spherical reference volume, the reference hydrostatic volume, the lag between the current figure and the hydrostatic figure, and the deflections of the surface.
+This would produce a lot of terms...
+The goal here is to show that this set of equations is consistent with the Liouville approach above, and that this set of nondimensional numbers is appropriate.)}
 

### 2015-06-08 tpw_rate.tex
Disorganized work

--- 2015-04-02 tpw_rate.tex
@@
 On a dynamic planet $\bf I$ may be a function of time, so to conserve angular momentum $\bf \Omega$ must be as well.
-In this case Equation \ref{euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
+In this case Equation \eqref{euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
 For a slowly convecting fluid such as a planetary mantle the inertial term $\partial {\bf H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
@@
 
-Note that solving Equation \ref{quasistatic_liouville} is equivalent to solving an eigenvalue problem for ${\bf I}$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
+Note that solving Equation \eqref{quasistatic_liouville} is equivalent to solving an eigenvalue problem for ${\bf I}$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
 In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
@@
 wher $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
-If we plug this decomposition into Equation \ref{quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
+If we plug this decomposition into Equation \eqref{quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
 
@@
 
-The infinite-time limit of Equation \ref{viscoelastic_deformation} allows us to calculate the fluid limit of the Love number $k_f$:
+The infinite-time limit of Equation \eqref{viscoelastic_deformation} allows us to calculate the fluid limit of the Love number $k_f$:
 \begin{equation}
@@
 
-\citet{ricard1993polar} obtain an approximation to Equation \ref{viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.
+\citet{ricard1993polar} obtain an approximation to Equation \eqref{viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.
 This introduces a new parameter, termed $T_1$, which can be seen as a weighted tidal relaxation time for the system.  
@@
 The second term represents the lag in the moment of inertia due to the viscous adjustment of the rotational bulge, where $T_1$ is the characteristic time constant for this adjustment.
-Since the first term represents the fluid limit of rotational deformation, it automatically satisfies Equation \ref{quasistatic_liouville}, and hence does not contribute to the polar wander problem.
+Since the first term represents the fluid limit of rotational deformation, it automatically satisfies Equation \eqref{quasistatic_liouville}, and hence does not contribute to the polar wander problem.
 
@@
 
-The term on the right-hand side of equation \ref{disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
+The term on the right-hand side of equation \eqref{disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
 This, as well, may be parameterized using a viscoelastic Love number approach:
@@
 A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d/dt (\Omega^2) = 2 {\Omega_i} \dot{ \Omega}_i = 0)$
-Substituting Equation \ref{rotational_deformation} into Equation \ref{disequilibrium} we find
+Substituting Equation \eqref{rotational_deformation} into Equation \eqref{disequilibrium} we find
 
@@
 
-Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation \ref{fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
+Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation \eqref{fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
 \begin{equation}
@@
 
-As mantle convection and rotational dynamics of planetary bodies are usually considered separately, some extra care must be taken to establish self-consistent governing equations.  
+Mantle convection and rotational dynamics of planetary bodies are usually considered separately, yet the processes are based on a common set of governing equations. 
+As such, some extra care must be taken to ensure that the equations we consider are self-consistent. 
 For simplicty we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.
@@
 \begin{aligned}
-- \nabla P + \eta \nabla^2 {\bf g} =  \rho {\bf g} -  \rho{\bf \Omega \times \Omega \times r}
+- \nabla P + \eta \nabla^2 {\bf u} =  \rho {\bf g} -  \rho{\bf \Omega \times \Omega \times r}
 \label{navier_stokes}
@@
 \end{equation}
-
-where the parameters are defined in Table~\ref{parameters}.
+here the parameters are defined in Table~\eqref{parameters}.
 In addition we use the simple equation of state
-
 \begin{equation}
-\rho = \rho_0 \left( 1 - \alpha T \right)
+\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right)
 \label{eos}
 \end{equation}
-
+where $\bf u$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
+The other parameters are defined in Table \eqref{parameters}.
 Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
 Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires three nondimensional numbers to characterize it.
-Convenient choices for these numbers are listed in Table \ref{nondim}, along with approximate Earth-like values for them.
+Convenient choices for these numbers are listed in Table \eqref{nondim}, along with approximate Earth-like values for them.
 
@@
 Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
-It would be straightforward to include them, but does not affect the overall treatment of this scaling.
+It would be straightforward to include them, and they do not affect the overall treatment of this scaling.
 
-Customarily a hydrostatic reference state is defined as the solution to \ref{navier_stokes} where there is no flow:
+Customarily a hydrostatic reference state is defined as the solution to \eqref{navier_stokes} where there is no flow:
 
@@
 
-Nondimensionalizing with the parameters in Table \ref{nondim} we find
+Nondimensionalizing with the parameters in Table \eqref{nondim} we find
 \begin{equation}
@@
 
-Equation \ref{navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
+Equation \eqref{navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
 
@@
 The first three terms represent pressure, viscous torques, and gravitational torques on the mantle.  
-Convection in the outer core, atmospheres, and oceans is not strong enough to provide significant pressure and viscous torques over geologic timescales, and a self-gravitating body cannot self-torque (citation...).
+Convection in the outer core, atmospheres, and oceans is not strong enough to provide significant pressure and viscous torques over geologic timescales, and a self-gravitating body cannot self-torque \citep{braginsky1995equations}.
 Therefore, the we can neglect those terms, and we are left with
@@
 which is simply a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.
-Splitting the volume into $V_H$ and $\Delta V$ and substituting \ref{eos} we find
+
+Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eos} we find
 
@@
 &\int_{V_H} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} + 
-\int_{V_H} \rho_0 \alpha T {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} +  \\
+\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} +  \\
 &\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} + 
-\int_{\Delta V} \rho_0 \alpha T {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} = 0  \\
+\int_{\Delta V} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} = 0  \\
 \end{aligned}
@@
 The first term of this equation is zero due to the hydrostatic equation.  
-The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\alpha \Delta T$.
+The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma$.
+Removing these, we find
+\begin{equation}
+\begin{aligned}
+\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} = 
+-\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }}
+\end{aligned}
+\end{equation}
+
+This equation may be identified with \eqref{disequilibrium}, where the mismatch of the convective moment of intertia with the spin axis is balanced by disequilibrium in the rotational deformation.
 
@@
 We count the number of fundamental units for the problem, in this case mass (kg), length (m), time (s), and temperature (K), as well as the number of parameters in use for this problem.  
-These parameters are listed in Table \ref{parameters}, and subtracting the ...
+These parameters are listed in Table \eqref{parameters}, and subtracting the ...
 
@@
 $g_0$ & reference gravity \\
+$T_0$ & reference temperature \\
 $\delta \rho = \rho_0 \alpha \Delta T$ & buoyancy parameter \\ 
@@
 Fr & Froude & $\Omega_0^2 R/g_0$ & $10^{-3}$ \\
-A & Aspect ratio & $R_i/R$ & $0.54$ \\
+A & aspect ratio & $R_i/R$ & $0.54$ \\
+$\Gamma$ & density deficit &$ \alpha \Delta T$ & $10^{-2}$ \\
 \end{tabular}
@@
 We also define a new nondimensional number (which will come in handy in intermediate steps) corresponding to the fractional density changes due to temperature variations: $\Gamma = \alpha \Delta T$.
-With this we may rewrite Equation \ref{simple_milankovitch} in terms of the nondimensional numbers defined above.
+With this we may rewrite Equation \eqref{simple_milankovitch} in terms of the nondimensional numbers defined above.
 The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep[e.g.][]{??}.
@@
 \texttt{Aspect} provides modern numerical methods, flexible geometry, and powerful support for integrating user-defined modules and postprocessors.
-We performed a number of simulations at different Rayleigh numbers, integrating Equation \ref{eq:liouville} in time.
+We performed a number of simulations at different Rayleigh numbers, integrating Equation \eqref{eq:liouville} in time.
 
@@
 \caption{ Left: Time series of the normalized difference between moments $\Lambda_{12}$ for convection in a 2D annulus at several different Rayleigh numbers.
-As the Rayliehg number increases, the average value of the relative moment decreases due to less low-degree coherence in the temperature structure.
+As the Rayleigh number increases, the average value of the relative moment decreases due to less low-degree coherence in the temperature structure.
 Right:  average value of $\Lambda_{12}$ for the different Rayleigh numbers.
@@
 At lengthscales smaller than this diffusion will tend to wipe out heterogeneities, and there will be no power in those modes.  
-Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in Equation \ref{T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:
+Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in Equation \eqref{T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:
 
@@
 
-The mismatch between the current axis and the principle axis of the convective moment is perhaps the most important parameter in Equation \ref{scaled_rotation}.  
+The mismatch between the current axis and the principle axis of the convective moment is perhaps the most important parameter in Equation \eqref{scaled_rotation}.  
 Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
@@
 If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principle axes.
-This is illustrated in Figure \ref{fig:perturb}
+This is illustrated in Figure \eqref{fig:perturb}
 
-We may arrive at a simple estimate of the growth rate of the angle $\theta$ by differentiating Equation \ref{eq:kahan}, tentatively holding $\lVert \lambda_i - \lambda_j \rVert$ fixed:
+We may arrive at a simple estimate of the growth rate of the angle $\theta$ by differentiating Equation \eqref{eq:kahan}, tentatively holding $\lVert \lambda_i - \lambda_j \rVert$ fixed:
 
@@
 As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
-There is a constant competition between growth of the mismatch angle $\theta$ through Equation \ref{eq:grow_perturbation} and its relaxation through Equation \ref{simple_milankovitch}.
-We may plug in the estimate for $\Lambda_{ij}$ into equation \ref{simple_milankovitch} to find the strikingly simple expression
+There is a constant competition between growth of the mismatch angle $\theta$ through Equation \eqref{eq:grow_perturbation} and its relaxation through Equation \eqref{simple_milankovitch}.
+We may plug in the estimate for $\Lambda_{ij}$ into equation \eqref{simple_milankovitch} to find the strikingly simple expression
 \begin{equation}
@@
 
-Indeed, we can identify two endmember behaviors of Equation \ref{eq:simplest_milankovitch}.
+Indeed, we can identify two endmember behaviors of Equation \eqref{eq:simplest_milankovitch}.
 When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.

### 2015-06-09 tpw_rate.tex
Werk

@@
 Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle.
+We support our analysis with numerical simulations of 2D mantle convection using the finite element code \texttt{Aspect} \citep{kronbichler2012high}, based on the \texttt{deal.II} library \citep{dealII81}.
+\texttt{Aspect} provides modern numerical methods, flexible geometry, and powerful support for integrating user-defined modules and postprocessors.
 
@@
 \subsection{The convective moment of inertia}
+\label{sec:convective_moment}
 
@@
 
-..
-..
-(Text on free surface?)
+An alternative to the loading Love number approach to surface deflections is to calculate the surface deflections directly using mantle convection simulations with a true free surface boundary condition.
 
@@
 
-In order to estimate $\theta$ and $\lambda_i$ we must examine the linear momentum equation.
+In order to estimate $\theta$ and $\lambda_i$ we must examine the dynamics of convection in the planetary interior.
 
-\section{Linear momentum equation}
-\label{sec:linear}
+\section{Internal dynamics}
+\label{sec:internal}
 
@@
 
-Customarily a hydrostatic reference state is defined as the solution to \eqref{navier_stokes} where there is no flow:
+In this case the dynamics can be characterized in terms of deviations from a reference hydrostatic state, which includes the dynamic pressure $P^* = P - P_0$ and density perturbations $\delta \rho = \rho- \rho_0 = - \rho_0 \alpha (T-T_0)$.
+In addition, we expect deviations in the figure of the planet from its hydrostatic shape, which we denote by $V = V_H + \Delta V$, where $V_H$ is the hydrostatic figure and $\Delta V$ is the deviation.
+The introduction of $\Delta V$ requires another nondimensional number to characterize it, and we find that the quantity $\Gamma = \alpha \Delta T$ is convenient.
+Finally, we define $\mitbf{\Omega} = \Omega_0 \mitbf{\omega}$, where $\mitbf{\omega}$ is a unit vector in the direction of $\mitbf{\Omega}$.
+
+By definition the hydrostatic reference state is a solution to \eqref{navier_stokes} where there is no flow:
 
@@
 
-where $P_0$ is the reference pressure.
-We can define a ``dynamic pressure'' $P^* = P - P_0$.
-
-Nondimensionalizing with the parameters in Table \eqref{nondim} we find
+Nondimensionalizing with the parameters in Table \eqref{nondim} and removing the reference state we find the nondimensional momentum equation:
 \begin{equation}
@@
 \end{equation}
-
-The figure for the planet which is consistent with the hydrostatic state we call $V_H$.
-The actual figure of the planet's mantle, $V$, may have some disequilibrium with this hydrostatic shape: $V = V_H + \Delta V $.
-The introduction of $\Delta V$ requires another nondimensional number to characterize it, and we find that the quantity $\Gamma = \alpha \Delta T$ is convenient.
 
@@
 
-This equation may be identified with \eqref{disequilibrium}, where the mismatch of the convective moment of intertia with the spin axis is balanced by disequilibrium in the rotational deformation.
+This equation may be identified with \eqref{disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by a the mismatch of the convective moment of intertia with the spin axis (right side).
+Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table \eqref{nondim}.
 
@@
 
-If we plug these nondimensionalizations into the governing equations we find that the incompressibility constraint remains unchanged.  The temperature equation and equation of state become
+If we plug these nondimensionalizations into the governing equations we find that the incompressibility constraint remains unchanged.  
+The temperature equation and equation of state become
 
@@
 $g_0$ & reference gravity \\
+$I_0$ & reference moment of inertia \\
 $T_0$ & reference temperature \\
@@
 This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
-We also define a new nondimensional number (which will come in handy in intermediate steps) corresponding to the fractional density changes due to temperature variations: $\Gamma = \alpha \Delta T$.
 With this we may rewrite Equation \eqref{simple_milankovitch} in terms of the nondimensional numbers defined above.
-The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep[e.g.][]{??}.
+The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep{munk1960rotation}.
 \begin{equation}
@@
 \begin{equation}
-T_1 \sim \eta / \rho g R = \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}
+T_1 \sim \frac{ \eta }{ \rho_0 g_0 R} = \frac{R^2}{\kappa} \frac{\Gamma}{\mathrm{Ra} }
 \end{equation}
+Plugging these scalings into equation \eqref{simple_milankovitch} we find
 
@@
 
-We support our scaling analyses with numerical simulations of 2D mantle convection using the finite element code \texttt{Aspect} \citep{kronbichler2012high}, based on the \texttt{deal.II} library \citep{dealII81}.
-\texttt{Aspect} provides modern numerical methods, flexible geometry, and powerful support for integrating user-defined modules and postprocessors.
-We performed a number of simulations at different Rayleigh numbers, integrating Equation \eqref{eq:liouville} in time.
-
 \subsection{An estimate of $\Lambda_{ij}$}
@@
 
-TODO: Needs work with notational consistency/connecting to previous scaling...
-
-Fluctuations in the nonhydrostatic moment of inertia are due to temperature fluctuations in the interior and their spatial structure.
+Fluctuations in the nonhydrostatic moment of inertia are due to temperature fluctuations in the interior and their spatial structure:
 
 \begin{equation}
-C_{ij} = \int_{V_S} \rho_0 \alpha T \left( r_q r_q \delta_{ij} - r_i r_j \right) 
+C_{ij} = \int_{V_S} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) 
+\label{eq:temperature_fluctuations}
 \end{equation}
 where $V_S$ is the reference spherical volume of the mantle.
+As discussed in Section \ref{sec:convective_moment}, these internal density loads may be convolved with their surface responses to get the convective moment of inertia $E_{ij}$.
+The surface response is a function of the viscosity structure of the planet and the wavelength of the interal load, but the multiplicative factor $(1+k^L_f)$ is generally an order-one parameter, so we neglect it for this scaling.
 
-These internal density loads may be convolved with surface responses to get the convective moment of inertia $\bf E$.
+For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the quadrupole structure of the temperature field (see Appendix \ref{appendix:moments}).
+Therefore, an estimate of one allows an estimate of the other.
+We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
+where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials.  
+\begin{equation} 
+T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi)
+\label{T_series}
+\end{equation}
+Where the $T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$.
 
-Nondimensionalizing the integral for $E_{ij}$ we find
-
+The transform in Equation \eqref{eq:temperature_fluctuations} is picking out a degree-two spherical harmonic in laterally, and a degree-two polynomial radially.
+We therefore want to estimate the power in a single mode of the temperature expansion, allowing us identify
 \begin{equation}
-J_{ij} = I_0 \delta_{ij} + \int_{V_S^\prime} \alpha T^\prime \left( r_q^\prime r_q^\prime \delta_{ij} - r_i^\prime r_j^\prime \right) 
+\Lambda_{12} \sim \Gamma T_{202}
 \end{equation}
 
-For thermal convection, the convective part of the moment of inertia $\bf E$ is directly related to the degree-two structure of the temperature field (see Appendix \ref{appendix:moments}).
-Therefore, an estimate of one allows an estimate of the other.
-
-The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
-the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is a broadening of the spatial 
-and temporal spectra \citep{mclaughlin1982transition}.
-
-We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
-where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of radial polynomials.  
-\begin{equation} 
-T( r , \theta, \phi, t )  = {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi)
-\label{T_series}
-\end{equation}
-
-The quantity $\Lambda_{ij}$ is the normalized degree-two temperature field.
 The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $X_{lmn}$ is constrained by
@@
 We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow (relatively low $\mathrm{Ra}$) and that of chaotic flow (relatively high $\mathrm{Ra}$).
+The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
+the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is a broadening of the spatial 
+and temporal spectra \citep{mclaughlin1982transition}.
 
-At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes (possibly, but not necessarily degree-two).
+At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes which are largely influenced by the aspect ratio.
+This spectrum may or may not have a lot of power in degree-two, does not depend strongly on time.
 
@@
 Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in Equation \eqref{T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:
-
 \begin{equation}
@@
 \end{equation}
-
 The total number of modes that are accessible to the system are therefore
@@
 \end{equation}
+
+Other power spectra for the temperature field are possible. Isoviscous models tend to be ``bluer,'' and models with viscosity stratification tend to be ``redder'' \citep{richards1999polar}.
+Nevertheless, at high Rayleigh number the expectation is that the power is distributed across many length scales.
 
@@
 However, on the left the eigengap $\lVert \lambda_2 - \lambda_1 \rVert$ is large, while on the right it is small.
-A perturbation is instantaneously added to both bodies, which effects a small rotation of the principle axes on the left, but a large one on the right.}
+A negative mass perturbation is instantaneously added to both bodies, which effects a small rotation of the principle axes on the left, but a large one on the right.}
 \end{figure}
@@
 This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}.
+We suggest, then, that a more vigorously convecting planet is much more likely to experience rapid TPW events.
 
@@
 \end{equation}
-Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the scaling. 
-This is somewhat of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
+Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
+At high Rayleigh numbers response timescales for relaxation of the mismatch angle go down.
+At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
+That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
+Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
 However, we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.
@@
 \begin{enumerate}
-\item Make notation everywhere consistent
-\item More discussion of Davis+Kahan?
 \item Make better figures

### 2015-08-04 figures/eigengap.py
Catch errors in figure scripts

@@ -9,7 +9,12 @@ import matplotlib.cm as cm
 import matplotlib.ticker
 import glob
 
-plt.style.use('ian')
+try:
+  plt.style.use('ian')
+except AttributeError:
+  print "old matplotlib version?"
+except ValueError:
+  print "Cannot find the requested stylesheet"
 
 #c = ['#0A5F02', '#1c567a', '#814292', '#d7383b', '#fdae61', '#c0f8b8']
 #c = ['#e41a1c', '#377eb8', '#4daf4a', '#984ea3', '#ff7f00', '#ffff33'] 

### 2015-08-04 figures/misfit.py

@@ -6,7 +6,12 @@ import numpy as np
 import scipy.signal as signal
 import scipy.interpolate as interp
 
-plt.style.use(['ian'])
+try:
+  plt.style.use('ian')
+except AttributeError:
+  print "old matplotlib version?"
+except ValueError:
+  print "Cannot find the requested stylesheet"
 
 data = np.loadtxt('statistics_500', skiprows=50, usecols=(1,17,18,19,20))
 

### 2015-08-04 tpw_rate.tex
Fix up labels

@@
 \frac{d {\bf H}}{dt} + {\bf \Omega} \times {\bf H} = 0
-\label{euler}
+\label{eq:euler}
 \end{equation}
@@
 On a dynamic planet $\bf I$ may be a function of time, so to conserve angular momentum $\bf \Omega$ must be as well.
-In this case Equation \eqref{euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
+In this case Equation \eqref{eq:euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
 For a slowly convecting fluid such as a planetary mantle the inertial term $\partial {\bf H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
@@
 {\bf \Omega } (t)\times ( {\bf I}(t) \cdot {\bf \Omega }(t) ) = 0
-\label{quasistatic_liouville}
+\label{eq:quasistatic_liouville}
 \end{equation}
 
-Note that solving Equation \eqref{quasistatic_liouville} is equivalent to solving an eigenvalue problem for ${\bf I}$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
+Note that solving Equation \eqref{eq:quasistatic_liouville} is equivalent to solving an eigenvalue problem for ${\bf I}$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
 In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
@@
 I_{ij}(t) = I_0 \delta_{ij} + J_{ij}(t) + E_{ij}(t)
-\label{separation}
+\label{eq:separation}
 \end{equation}
@@
 wher $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
-If we plug this decomposition into Equation \eqref{quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
+If we plug this decomposition into Equation \eqref{eq:quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
 
@@
 {\bf \Omega} \times ({\bf J} \cdot {\bf \Omega}) = -{\bf \Omega } \times ( {\bf E} \cdot {\bf \Omega} )
-\label{disequilibrium}
+\label{eq:disequilibrium}
 \end{equation}
@@
 J_{ij} = \frac{k a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right)
-\label{elastic_deformation}
+\label{eq:elastic_deformation}
 \end{equation}
@@
 J_{ij} = \frac{k(t) a^5}{3 G} * \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right)
-\label{viscoelastic_deformation}
+\label{eq:viscoelastic_deformation}
 \end{equation}
@@
 
-The infinite-time limit of Equation \eqref{viscoelastic_deformation} allows us to calculate the fluid limit of the Love number $k_f$:
+The infinite-time limit of Equation \eqref{eq:viscoelastic_deformation} allows us to calculate the fluid limit of the Love number $k_f$:
 \begin{equation}
 k_f = \frac{3 G (C-A)}{\Omega^2 a^5}
-\label{fluid_love}
+\label{eq:fluid_love}
 \end{equation}
@@
 
-\citet{ricard1993polar} obtain an approximation to Equation \eqref{viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.
+\citet{ricard1993polar} obtain an approximation to Equation \eqref{eq:viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.
 This introduces a new parameter, termed $T_1$, which can be seen as a weighted tidal relaxation time for the system.  
@@
  \frac{k_f a^5 T_1}{3G} \left(\dot{\Omega}_i \Omega_j + \Omega_i \dot{\Omega}_j - \frac{2}{3} \Omega_q \dot{\Omega}_q \delta_{ij} \right)
-\label{rotational_deformation}
+\label{eq:rotational_deformation}
 \end{equation}
@@
 The second term represents the lag in the moment of inertia due to the viscous adjustment of the rotational bulge, where $T_1$ is the characteristic time constant for this adjustment.
-Since the first term represents the fluid limit of rotational deformation, it automatically satisfies Equation \eqref{quasistatic_liouville}, and hence does not contribute to the polar wander problem.
+Since the first term represents the fluid limit of rotational deformation, it automatically satisfies Equation \eqref{eq:quasistatic_liouville}, and hence does not contribute to the polar wander problem.
 
@@
 
-The term on the right-hand side of equation \eqref{disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
+The term on the right-hand side of equation \eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
 This, as well, may be parameterized using a viscoelastic Love number approach:
@@
 A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d/dt (\Omega^2) = 2 {\Omega_i} \dot{ \Omega}_i = 0)$
-Substituting Equation \eqref{rotational_deformation} into Equation \eqref{disequilibrium} we find
+Substituting Equation \eqref{eq:rotational_deformation} into Equation \eqref{eq:disequilibrium} we find
 
@@
 
-Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation \eqref{fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
+Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation \eqref{eq:fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
 \begin{equation}
@@
 \end{aligned}
-\label{milankovitch}
+\label{eq:milankovitch}
 \end{equation}
@@
 \dot{\theta} = - \frac{1}{(C-A)T_1} (\lambda_3-\lambda_1) \sin{2 \theta}
-\label{simple_milankovitch}
+\label{eq:simple_milankovitch}
 \end{equation}
@@
 \nabla \cdot {\bf u} = 0
-\label{conserve_mass}
+\label{eq:conserve_mass}
 \end{equation}
@@
 - \nabla P + \eta \nabla^2 {\bf u} =  \rho {\bf g} -  \rho{\bf \Omega \times \Omega \times r}
-\label{navier_stokes}
+\label{eq:navier_stokes}
 \end{aligned}
@@
 \frac{\partial T}{\partial t} + {\bf u} \cdot \nabla T = \kappa \nabla^2 T
-\label{energy}
+\label{eq:energy}
 \end{equation}
-here the parameters are defined in Table~\eqref{parameters}.
+here the parameters are defined in Table~\ref{tab:parameters}.
 In addition we use the simple equation of state
@@
 \rho = \rho_0 \left( 1 - \alpha (T-T_0) \right)
-\label{eos}
+\label{eq:eos}
 \end{equation}
 where $\bf u$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
-The other parameters are defined in Table \eqref{parameters}.
+The other parameters are defined in Table \ref{tab:parameters}.
 Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
 Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires three nondimensional numbers to characterize it.
-Convenient choices for these numbers are listed in Table \eqref{nondim}, along with approximate Earth-like values for them.
+Convenient choices for these numbers are listed in Table \ref{tab:nondim}, along with approximate Earth-like values for them.
 
@@
 
-By definition the hydrostatic reference state is a solution to \eqref{navier_stokes} where there is no flow:
+By definition the hydrostatic reference state is a solution to \eqref{eq:navier_stokes} where there is no flow:
 
@@
 - \nabla P_0 =  \rho_0 {\bf g} -  \rho_0 {\bf \Omega \times \Omega \times r}
-\label{hydrostatic}
+\label{eq:hydrostatic}
 \end{aligned}
@@
 
-Nondimensionalizing with the parameters in Table \eqref{nondim} and removing the reference state we find the nondimensional momentum equation:
+Nondimensionalizing with the parameters in Table \ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
 \begin{equation}
@@
 
-Equation \eqref{navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
+Equation \eqref{eq:navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
 
@@
 
-Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eos} we find
+Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eq:eos} we find
 
@@
 
-This equation may be identified with \eqref{disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by a the mismatch of the convective moment of intertia with the spin axis (right side).
-Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table \eqref{nondim}.
+This equation may be identified with \eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by a the mismatch of the convective moment of intertia with the spin axis (right side).
+Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table \ref{tab:nondim}.
 
@@
 We count the number of fundamental units for the problem, in this case mass (kg), length (m), time (s), and temperature (K), as well as the number of parameters in use for this problem.  
-These parameters are listed in Table \eqref{parameters}, and subtracting the ...
+These parameters are listed in Table \ref{tab:parameters}, and subtracting the ...
 
@@
 \caption{Nondimensionalizations for variables, primed variables denote dimensional variables}
-\label{nondim_convert}
+\label{tab:nondim_convert}
 \begin{tabular}{@{}lll}
@@
 \caption{Parameters for rotating mantle convection}
-\label{parameters}
+\label{tab:parameters}
 \begin{tabular}{@{}lcc}
@@
 \caption{Nondimensional numbers with approximate Earthlike values}
-\label{nondim}
+\label{tab:nondim}
 \begin{tabular}{@{}lcccc}
@@
 This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
-With this we may rewrite Equation \eqref{simple_milankovitch} in terms of the nondimensional numbers defined above.
+With this we may rewrite Equation \eqref{eq:simple_milankovitch} in terms of the nondimensional numbers defined above.
 The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep{munk1960rotation}.
@@
 \end{equation}
-Plugging these scalings into equation \eqref{simple_milankovitch} we find
+Plugging these scalings into equation \eqref{eq:simple_milankovitch} we find
 
@@
 \frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma \mathrm{Fr}} \Lambda_{31} \sin{2 \theta}
-\label{scaled_rotation}
+\label{eq:scaled_rotation}
 \end{equation}
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
-\label{misfit}
+\label{fig:eigengap}
 \includegraphics[width=0.9\textwidth]{figures/eigengap.pdf}
@@
 Also shown is a line with slope $\mathrm{Ra}^{-2/3}$, which is predicted from the scaling analysis (the exponent is $-2/3$ instead of $-1$ due to the reduced dimensionality of the simulations).}
-\end{figure}
+\end{figure*}
 
@@
 T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi)
-\label{T_series}
+\label{eq:T_series}
 \end{equation}
@@
 At lengthscales smaller than this diffusion will tend to wipe out heterogeneities, and there will be no power in those modes.  
-Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in Equation \eqref{T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:
+Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in Equation \eqref{eq:T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:
 \begin{equation}
@@
 
-The mismatch between the current axis and the principle axis of the convective moment is perhaps the most important parameter in Equation \eqref{scaled_rotation}.  
+The mismatch between the current axis and the principle axis of the convective moment is perhaps the most important parameter in Equation \eqref{eq:scaled_rotation}.  
 Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
@@
 If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principle axes.
-This is illustrated in Figure \eqref{fig:perturb}
+This is illustrated in Figure \ref{fig:perturb}
 
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
@@
 A negative mass perturbation is instantaneously added to both bodies, which effects a small rotation of the principle axes on the left, but a large one on the right.}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
-\label{misfit}
+\label{fig:misfit}
 \includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
@@
 When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger.}
-\end{figure}
+\end{figure*}
 
@@
 As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
-There is a constant competition between growth of the mismatch angle $\theta$ through Equation \eqref{eq:grow_perturbation} and its relaxation through Equation \eqref{simple_milankovitch}.
-We may plug in the estimate for $\Lambda_{ij}$ into equation \eqref{simple_milankovitch} to find the strikingly simple expression
+There is a constant competition between growth of the mismatch angle $\theta$ through Equation \eqref{eq:grow_perturbation} and its relaxation through Equation \eqref{eq:simple_milankovitch}.
+We may plug in the estimate for $\Lambda_{ij}$ into equation \eqref{eq:simple_milankovitch} to find the strikingly simple expression
 \begin{equation}
@@
 I_{ij} = \int_V \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) d^3 {\bf r}
-\label{inertia}
+\label{eq:inertia}
 \end{equation}
@@
 \end{bmatrix}
-\label{maccullagh}
+\label{eq:maccullagh}
 \end{equation}

### 2015-08-04 tpw_rate.tex
Address comments

@@
 Despite this, the ability of internal mass anomalies to drive large-scale TPW remains controversial. 
-Paleomagnetic data have been interpreted to requre up to $\sim 1^\circ$/Myr rates of TPW \citep{mitchell2011sutton}, but the ability of the mantle to respond at such rates has been questioned \citep{tsai2007theoretical}.
+Paleomagnetic data have been interpreted to require up to $\sim 1^\circ$/Myr rates of TPW \citep{mitchell2011sutton}, but the ability of the mantle to respond at such rates has been questioned \citep{tsai2007theoretical}.
 
@@
 known as Euler's equations, where ${\bf H} = {\bf I} \cdot {\bf \Omega}$ is the angular momentum vector, $\bf I$ is the moment of inertia tensor, and $\bf \Omega$ is the angular velocity.
-On a dynamic planet $\bf I$ may be a function of time, so to conserve angular momentum $\bf \Omega$ must be as well.
+On a dynamic planet $\bf I$ may be a function of time, so to conserve angular momentum $\bf \Omega$ must be also vary with time.
 In this case Equation \eqref{eq:euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
@@
 In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
-The moment of inertia tensor above includes all contributions to the mass structure of the planet, including
-\begin{itemize}
-\item the spherically symmetric mass distribution
-\item the rotational deformation
-\item internal density anomalies
-\item surface deflections due to internal density anomalies
-\end{itemize}
+The moment of inertia tensor in \eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
+of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
+Here we are interested in the moment of inertia due to mantle convection, and thus neglect the contributions of surface
+density anomalies (though these can be quite important, \hl{ citations for Glaciers, Tharsis, etc} ).
 
-The moment of inertia tensor is commonly separated into three parts \citep{ricard1993polar}:
+For mantle convection problems the moment of inertia tensor is commonly separated into three parts \citep{ricard1993polar}:
 
@@
 
-wher $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
+where $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
 If we plug this decomposition into Equation \eqref{eq:quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
@@
 where $C$ and $A$ are the polar and equatorial moments of inertia, respectively.
-The infinite-time limit, however, does not allow for any disequilibrium between $J_{ij}$ and $C_{ij}$.
+The infinite-time limit, however, does not allow for any disequilibrium between $J_{ij}$ and $E_{ij}$.
 
@@
 We are in a position to address the rates of true polar wander for a given convective moment $\bf E$.
-A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d/dt (\Omega^2) = 2 {\Omega_i} \dot{ \Omega}_i = 0)$
+A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
 Substituting Equation \eqref{eq:rotational_deformation} into Equation \eqref{eq:disequilibrium} we find
@@
 They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principle axis of the convective moment.
-
-In order to estimate $\theta$ and $\lambda_i$ we must examine the dynamics of convection in the planetary interior.
+Both $\theta$ and $\lambda_i$ depend upon the dynamics of the convecting planetary mantle.
+In order to derive estimates for them, we must consider those dynamics.
 
@@
 Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
-Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires three nondimensional numbers to characterize it.
+Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires four nondimensional numbers to characterize it.
 Convenient choices for these numbers are listed in Table \ref{tab:nondim}, along with approximate Earth-like values for them.
 
-In particular, the most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
+In particular, the two most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
 This second nondimensional number does not have a uniformly agreed-upon name, but has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, which we adopt here.
@@
 
-which is simply a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.
+which is a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.
 
@@
 
-This equation may be identified with \eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by a the mismatch of the convective moment of intertia with the spin axis (right side).
+This equation may be identified with \eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
 Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table \ref{tab:nondim}.
@@
 \begin{equation}
-C_{ij} = \int_{V_S} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) 
+C_{ij} = \int_{V_S} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) dV
 \label{eq:temperature_fluctuations}
@@
 
-The transform in Equation \eqref{eq:temperature_fluctuations} is picking out a degree-two spherical harmonic in laterally, and a degree-two polynomial radially.
+The integral in Equation \eqref{eq:temperature_fluctuations} is picking out a degree-two spherical harmonic in the lateral dimensions, and a degree-two polynomial in radius.
 We therefore want to estimate the power in a single mode of the temperature expansion, allowing us identify
@@
 
-The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $X_{lmn}$ is constrained by
+The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by
 
@@
 
-At high Rayleigh number we expect there to be a lengthscale which characterizes the smallest length of features in the convecting system.  
-At lengthscales smaller than this diffusion will tend to wipe out heterogeneities, and there will be no power in those modes.  
-Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in Equation \eqref{eq:T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:
+At high Rayleigh number we expect the shortest lengthscales to be limited by the effects of thermal diffusion,
+which tends to wipe out thermal heterogeneity at small scales.  Consequently, there will be no power in modes
+with shorter lenghtscales than that allowed by diffusion. 
+Therefore we expect, to a good approximation, that the infinite sum in Equation \eqref{eq:T_series} can be truncated at some maximum wavenumber, set by the smallest lengthscale $d$:
 \begin{equation}
@@
 \begin{equation}
-N_{\mathrm{modes}} = n_{\mathrm{max}} \times l_{\mathrm{max}} \times n_{\mathrm{max}} \sim \left( \frac{R}{d} \right)^{3}
+N_{\mathrm{modes}} = n_{\mathrm{max}} \times l_{\mathrm{max}} \times m_{\mathrm{max}} \sim \left( \frac{R}{d} \right)^{3}
 \end{equation}
@@
 T_{\mathrm{degree 2}}(t) \sim \mathrm{Ra}^{-1}
+\label{eq:degree_two_of_ra}
 \end{equation}
-
+For two-dimensional simulations there is a reduced dimensionality when calculating the number of modes,
+so $N_\mathrm{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\mathrm{degree2}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
 This result has a simple interpretation.

### 2015-08-04 figures/misfit.py
Stop showing

@@ -70,8 +70,4 @@ ax.yaxis.set_major_locator(matplotlib.ticker.MaxNLocator(nbins=3, prune=None))
 plt.xlabel(r"Time")
 plt.ylabel(r"Angle ($^\circ$)")
 
-
-
-
 plt.savefig("misfit.pdf", bbox_inches='tight')
-plt.show()

### 2015-08-04 figures/eigengap.py
Remove for now

@@ -77,6 +77,7 @@ def get_spin_spectrum(text_file):
 output_files = [f for f in glob.glob("statistics_*")]
 sort_key = lambda s: float(s.split('_')[1])
 output_files.sort(key=sort_key)
+output_files.remove('statistics_50')
 print output_files
 
 amplitudes = []

### 2015-08-06 tpw_rate.tex
Add some text on elastic lithosphere

@@
  \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}
-\end{equation}
 
@@
 \section{Discussion}
+\label{sec:discussion}
 
@@
 Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
-However, we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.
+However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d/R \sim \mathrm{Ra}^{-\beta}$.
+With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
+\begin{equation}
+\dot{\theta} \propto \mathrm{Ra}^{1-3\beta}
+\end{equation}
+In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.
 
@@
 
+Thus far we have restricted our discussion to planets with lithospheres lacking long-term elastic strength.
+For this case the long-time limit of the planetary figure is coaxial with the convective 
+moment of inertia. This assumption is not necessarily true in all cases.
+Earth's lithosphere is pervasively fractured and hydrated, and may not have strength on geologic timescales.
+However, a planet with a stagnant lid (such as Mars) may have some strength, preventing the figure of the planet 
+from reaching the fluid limit of Equation~\eqref{eq:fluid_love}.
+
+The theory of TPW response for the case of elastic lithospheres has been developed in, among other places, 
+\citet{matsuyama2006rotational}, \citet{creveling2012mechanisms}, and \citet{chan2014time}.
+The formalism developed in Section~\ref{sec:scaling} can still be applied to this case, 
+though the response to internal variations in the moment of inertia becomes more limited.
+
 \section{Conclusion}
-We have developed a framework for discussing the rates of true polar wander in terms of a small number of nondimensional numbers.
+We have developed a framework for approximating the rates of true polar wander in terms of a small number of nondimensional numbers.
 
@@
 \begin{enumerate}
-\item Make better figures
-\item Integrate numerical results
 \item Make the argument about degree-two scaling much sharper
-\item Discuss elastic lithosphere, Creveling
 \item Discuss high-viscosity lower mantle

### 2015-08-06 tpw_rate.tex
Moar text

@@
 Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
-However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d/R \sim \mathrm{Ra}^{-\beta}$.
+However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \sim \mathrm{Ra}^{-\beta}$.
 With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
@@
 
-More important is the observation that the most important parameters are the Froude number, which acts as the brakes on the system, and $\theta(\mathrm{Ra})$.
+We then suggest that the most important parameters are the Froude number, which acts as the brakes on the system, and $\theta(\mathrm{Ra})$.
 Whereas the rest of the expression has at most a weak dependence on $\mathrm{Ra}$, this is expected to be strongly dependent on it.
@@
 This is the regime considered in \citet{steinberger1997changes}, \citet{roberts2007cause}, and \citet{zhong2007supercontinent}, and can be considered the ``slow TPW'' regime.
-
 When convection is more chaotic, however, there may be large excursions in $\theta$, which can drastically increase the wander rate.  
@@
 This, however, is a special case in the large $\theta$, ``fast TPW'' regime.
+
+As an example, we may consider the early Earth, when the mantle was presumably hotter and less viscous, leading to a 
+higher Rayleigh number. We would then predict that convection was more vigorous, leading to 
+a less stable $\theta(t)$, and thus more TPW and more frequent ``fast TPW'' events.
 
@@
 The formalism developed in Section~\ref{sec:scaling} can still be applied to this case, 
-though the response to internal variations in the moment of inertia becomes more limited.
+though the response to internal variations in the moment of inertia becomes more limited 
+(and potentially richer, as in the oscillitory motions suggested by \citet{creveling2012mechanisms}).
 
 \section{Conclusion}
-We have developed a framework for approximating the rates of true polar wander in terms of a small number of nondimensional numbers.
+We have developed a framework for discussing the rates of true polar wander for a convecting planet 
+from a perspective of scaling and fluid dynamics.
+We identified a small number of dimensionless parameters which describe the system, and showed how they affect the overall dynamics of the system.
+
+The most important parameters are the Rayleigh number and the Froude number.
+The Froude number acts as a damper to TPW.
+The Rayleigh number is more complicated, since it is a control on both the forcing 
+of TPW and the response, which act in opposite directions.
+This perspective allows us to consider not only the polar wandering of Phanerozoic Earth,
+but also allows us to hypothesize about polar wandering during the Archean and Proterozoic, or 
+on other planetary bodies.
 

### 2015-08-27 tpw_rate.tex
Remove confusing sentence
Mostly formatting

@@
 
-Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle.
-We support our analysis with numerical simulations of 2D mantle convection using the finite element code \texttt{Aspect} \citep{kronbichler2012high}, based on the \texttt{deal.II} library \citep{dealII81}.
-\texttt{Aspect} provides modern numerical methods, flexible geometry, and powerful support for integrating user-defined modules and postprocessors.
+Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle,  which we support with numerical simulations.
 
@@
 \end{equation}
-
 known as Euler's equations, where ${\bf H} = {\bf I} \cdot {\bf \Omega}$ is the angular momentum vector, $\bf I$ is the moment of inertia tensor, and $\bf \Omega$ is the angular velocity.
@@
 \end{equation}
-
 Note that solving Equation \eqref{eq:quasistatic_liouville} is equivalent to solving an eigenvalue problem for ${\bf I}$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
@@
 For mantle convection problems the moment of inertia tensor is commonly separated into three parts \citep{ricard1993polar}:
-
 \begin{equation}
@@
 \end{equation}
-
 where $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
 If we plug this decomposition into Equation \eqref{eq:quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
-
 \begin{equation}
@@
 \end{equation}
-
 This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches of the convective part of the moment of inertia ($\bf E$) and the rotational deformation part of the moment of inertia ($\bf J$).
-If $\bf J$ and $\bf E$ are coaxial, then the mismatch must be zero, and solving the polar wander problem reduces to solving an eigenvalue problem for $\bf E$.
 Our goal is to characterize this balance from a perspective of scaling and fluid dynamics.
@@
 The part of the moment of inertia due to the elastic rotational deformation is traditionally related to the degree-two part of the gravity field via MacCullagh's formula \citep{munk1960rotation}:
-
 \begin{equation}
@@
 \end{equation}
-
 where $k$ is an elastic Love number, $a$ is the semimajor axis of the planet, and $G$ is the gravitational constant.
 This result may be extended to a viscoelastic rheology via the viscoelastic correspondence principle \citep[e.g.][]{peltier1974impulse}:
-
 \begin{equation}
@@
 \end{equation}
-
 where $k$ is now a time-dependent viscoelastic Love number which is convolved with the time-dependent rotation vector.
@@
 \end{equation}
-
 where $C$ and $A$ are the polar and equatorial moments of inertia, respectively.
@@
 \citet{ricard1993polar} obtain an approximation to Equation \eqref{eq:viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.
-This introduces a new parameter, termed $T_1$, which can be seen as a weighted tidal relaxation time for the system.  
+This introduces a new parameter, termed $T_1$, which can be seen as a weighted relaxation time for the system.  
 This simple approximation in the Laplace domain allows for an analytical transformation back into the time domain, and neglecting second order terms in $\dot{\Omega}$ we find:
-
 \begin{equation}
@@
 \end{equation}
-
 The two terms of this equation have simple interpretations.  
@@
 \end{equation}
-
 An alternative to the loading Love number approach to surface deflections is to calculate the surface deflections directly using mantle convection simulations with a true free surface boundary condition.
@@
 Substituting Equation \eqref{eq:rotational_deformation} into Equation \eqref{eq:disequilibrium} we find
-
 \begin{equation}
@@
 \end{equation}
-
 Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation \eqref{eq:fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
@@
 \end{equation}
-
 Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
 If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
-
 \ifdetail
@@
 Plugging this in, we find a system of three equations, which we want to solve for $\dot{\theta}$ and $\dot{\phi}$:
-
 \begin{equation}
@@
 \end{equation}
-
 Now multiply the top equation by $\cos{\phi}$ and the second equation by $\sin{\phi}$ and add them together to get
-
 \begin{equation}
@@
 \end{equation}
-
 Now multiply the above by $\cos{\theta}$ and the third equation by $-\sin{\theta}$ and add them to get
-
 \begin{equation}
@@
 \end{equation}
-
 After a lot of cancellation, we find
-
 \begin{equation}
@@
 \end{equation}
-
 Likewise we may multiply the first equation by $-\sin{\phi}$ and the second equation by $\cos{\phi}$ and add them to get
-
 \begin{equation}
@@
 \fi
-
 \begin{equation}
@@
 \end{equation}
-
 These equations are a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.  
@@
 \begin{aligned}
--\int_V {\bf r} \times \nabla P + \int_V \eta {\bf r} \times \nabla^2 {\bf u} - \int_V \rho {\bf r \times g} + \int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} &= 0 \\
+-\int_V {\bf r} \times \nabla P \,dV + \int_V \eta {\bf r} \times \nabla^2 {\bf u} \,dV - \int_V \rho {\bf r \times g} \,dV 
+   + \int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV &= 0 \\
 \end{aligned}
@@
 Convection in the outer core, atmospheres, and oceans is not strong enough to provide significant pressure and viscous torques over geologic timescales, and a self-gravitating body cannot self-torque \citep{braginsky1995equations}.
-Therefore, the we can neglect those terms, and we are left with
+Therefore we can neglect those terms, and we are left with
 
@@
 \begin{aligned}
-\int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} &= 0 \\
+\int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV &= 0 \\
 \end{aligned}
@@
 \begin{aligned}
-&\int_{V_H} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} + 
-\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} +  \\
-&\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} + 
-\int_{\Delta V} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} = 0  \\
+&\int_{V_H} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
+\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV +  \\
+&\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
+\int_{\Delta V} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 0  \\
 \end{aligned}
@@
 \begin{aligned}
-\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} = 
--\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }}
+\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 
+-\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV
 \end{aligned}
@@
 
-The time constant $T_1$ is a viscous tidal relaxation time of the planetary mantle.
+The time constant $T_1$ is a viscous relaxation time of the planetary mantle.
 It is frequently represented as a weighted average of the different relaxation modes \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
@@
 \begin{equation}
-C_{ij} = \int_{V_S} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) dV
+C_{ij} = \int_{V_S} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) \,dV
 \label{eq:temperature_fluctuations}
@@
 We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
-where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials.  
+where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials:
 \begin{equation} 
@@
 \end{equation}
-Where the $T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$.
+where the $T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$.
 

### 2015-08-27 tpw_rate.tex
Fix figure numbering

@@
 Plugging these scalings into equation \eqref{eq:simple_milankovitch} we find
-
 \begin{equation}
@@
 \centering
-\label{fig:eigengap}
 \includegraphics[width=0.9\textwidth]{figures/eigengap.pdf}
@@
 Also shown is a line with slope $\mathrm{Ra}^{-2/3}$, which is predicted from the scaling analysis (the exponent is $-2/3$ instead of $-1$ due to the reduced dimensionality of the simulations).}
+\label{fig:eigengap}
 \end{figure*}
@@
 \end{equation}
-
 The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by
@@
 \begin{equation}
-\lVert T(x,y,z, t) \rVert_\infty = \lVert {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \rVert_\infty = 1
+\max \left( T(x,y,z, t) \right) = \max \left( \displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \right) = 1
 \end{equation}
-
 This is a strong constraint, but it gives very little information about the distribution of power across the $T_{lmn}$.  
@@
 \end{equation}
-
 The value of each $T_{lmn}(t)$ will in general be some complex function of time, but for a given style of convection we expect there to be some average value.
@@
 \end{equation}
-For two-dimensional simulations there is a reduced dimensionality when calculating the number of modes,
+For two-dimensional simulations shown below there is a reduced dimensionality when calculating the number of modes,
 so $N_\mathrm{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\mathrm{degree2}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
@@
 With an estimate for the power in the degree-two part of the temperature field, we may finally estimate $\Lambda_{ij}$:
-
 \begin{equation}
@@
 \centering
-\label{fig:perturb}
 \includegraphics[width=0.9\textwidth]{figures/perturb.pdf}
@@
 A negative mass perturbation is instantaneously added to both bodies, which effects a small rotation of the principle axes on the left, but a large one on the right.}
+\label{fig:perturb}
 \end{figure*}
@@
 \centering
-\label{fig:misfit}
 \includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
@@
 When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger.}
+\label{fig:misfit}
 \end{figure*}
@@
 moment of inertia. This assumption is not necessarily true in all cases.
-Earth's lithosphere is pervasively fractured and hydrated, and may not have strength on geologic timescales.
-However, a planet with a stagnant lid (such as Mars) may have some strength, preventing the figure of the planet 
+Earth's lithosphere is pervasively fractured and hydrated, and may not have much strength when subjected to rotational changes on geologic timescales.
+However, a planet with a stagnant lid (such as Mars) may have considerable strength, preventing the figure of the planet 
 from reaching the fluid limit of Equation~\eqref{eq:fluid_love}.

### 2015-08-27 tpw_rate.tex
comments

@@
 \end{equation}
-Note that solving Equation \eqref{eq:quasistatic_liouville} is equivalent to solving an eigenvalue problem for ${\bf I}$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
+Equation \eqref{eq:quasistatic_liouville} indicates that ${\bf \Omega}$ and ${\bf I}$ are parallel, so a solution for ${\bf \Omega}(t)$ is equivalent to solving an eigenvalue problem for ${\bf I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
 In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
@@
 of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
-Here we are interested in the moment of inertia due to mantle convection, and thus neglect the contributions of surface
-density anomalies (though these can be quite important, \hl{ citations for Glaciers, Tharsis, etc} ).
+Here we are interested in contributions from mantle convection, so we neglect other processes.
 
@@
 \end{equation}
-
 Nondimensionalizing with the parameters in Table \ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
@@
 \end{equation}
-
 Equation \eqref{eq:navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
-
 \begin{equation}
@@
 \end{equation}
-
 which is a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.
@@
 Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eq:eos} we find
-
 \begin{equation}
@@
 \end{equation}
-
 The first term of this equation is zero due to the hydrostatic equation.  
@@
 \end{equation}
-
 This equation may be identified with \eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
@@
 \end{equation}
+
 For two-dimensional simulations shown below there is a reduced dimensionality when calculating the number of modes,
@@
 \begin{equation}
-\dot{\theta} \propto \mathrm{Ra}^{1-3\beta}
+\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{\mathrm{Fr}} \sin{2 \theta}
 \end{equation}
@@
 
-We then suggest that the most important parameters are the Froude number, which acts as the brakes on the system, and $\theta(\mathrm{Ra})$.
-Whereas the rest of the expression has at most a weak dependence on $\mathrm{Ra}$, this is expected to be strongly dependent on it.
+We then suggest that the most important parameters are the Froude number, which acts as the brakes on the system, and the mismatch angle $\theta(\mathrm{Ra})$.
+Whereas the rest of the expression has at most a weak dependence on $\mathrm{Ra}$, $\theta$ is expected to be strongly dependent on it.
 

### 2015-08-27 figures/eigengap.py
Update figure

@@ -7,6 +7,7 @@ import scipy.interpolate as interp
 import matplotlib.pyplot as plt
 import matplotlib.cm as cm
 import matplotlib.ticker
+from itertools import cycle
 import glob
 
 try:
@@ -77,7 +78,6 @@ def get_spin_spectrum(text_file):
 output_files = [f for f in glob.glob("statistics_*")]
 sort_key = lambda s: float(s.split('_')[1])
 output_files.sort(key=sort_key)
-output_files.remove('statistics_50')
 print output_files
 
 amplitudes = []
@@ -86,7 +86,10 @@ Rayleighs = []
 
 #plt.subplots_adjust(wspace=0.0, hspace=0.0)
 
-for f, i in zip( output_files, range(len(output_files)) ):
+colors = cycle(plt.rcParams['axes.color_cycle'])
+ax = plt.subplot(1,2,1)
+
+for i,f in enumerate(output_files):
 
   delta_T = float(f.split('_')[1])
   alpha = alpha_0
@@ -103,18 +106,21 @@ for f, i in zip( output_files, range(len(output_files)) ):
   amplitudes.append(diff.mean())
   timescales.append(peak)
 
-  ax = plt.subplot(1,2,1)
-  ax.plot(times, diff,  label=r'$\mathrm{Ra} = %.1f \times 10^{%i}$' % ( Ra/np.power(10., np.floor(np.log10(Ra))) , np.floor(np.log10(Ra))))
-  ax.set_xlim(0,7)
-  ax.set_ylim(0, 1.0e-4)
-  ax.set_xlabel("Time (Gyr)")
-  ax.set_ylabel(r'Relative moment $(\lambda_2-\lambda_1)/I_0$')
+  #plot a subset of the timeseries (plotting all of them got too busy)
+  if f in ['statistics_50', 'statistics_200', 'statistics_500']:
+    ax.plot(times, diff, next(colors),\
+            label=r'$\mathrm{Ra} = %.1f \times 10^{%i}$' % ( Ra/np.power(10., np.floor(np.log10(Ra))) , np.floor(np.log10(Ra))))
+  else:
+    next(colors)
 
 
 
-plt.legend(loc="upper right")
+ax.set_xlim(0,4)
+ax.set_ylim(0, 8.0e-5)
+ax.set_xlabel("Time (Gyr)")
+ax.set_ylabel(r'$(\lambda_2-\lambda_1)/I_0$')
[…]

### 2015-08-28 tpw_rate.tex
Text

@@
  \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ {\bf E \cdot \mitbf{\omega}} - \left( {\bf \mitbf{\omega} \cdot E \cdot \mitbf{\omega} } \right) \mitbf{\omega} \right]
+\label{eq:liouville_ode}
 \end{equation}
@@
  \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}
+\end{equation}
 
@@
 
-For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the quadrupole structure of the temperature field (see Appendix \ref{appendix:moments}).
+For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-twoo part of the temperature field (see Appendix \ref{appendix:moments}).
 Therefore, an estimate of one allows an estimate of the other.
@@
 
-The integral in Equation \eqref{eq:temperature_fluctuations} is picking out a degree-two spherical harmonic in the lateral dimensions, and a degree-two polynomial in radius.
+The integral in Equation \eqref{eq:temperature_fluctuations} is picking out degree-two spherical harmonics in the lateral dimensions, and low order polynomials in radius.
 We therefore want to estimate the power in a single mode of the temperature expansion, allowing us identify
 \begin{equation}
-\Lambda_{12} \sim \Gamma T_{202}
+\Lambda_{12} \sim \Gamma T_{\mathrm{degree \textendash two}}
 \end{equation}
@@
 \begin{equation}
-T_{\mathrm{degree 2}}(t) \sim \frac{1}{N_{\mathrm{modes}}} \sim \left( \frac{d}{R}\right)^3
+T_{\mathrm{degree \textendash two}}(t) \sim \frac{1}{N_{\mathrm{modes}}} \sim \left( \frac{d}{R}\right)^3
 \end{equation}
@@
 \begin{equation}
-T_{\mathrm{degree 2}}(t) \sim \mathrm{Ra}^{-1}
+T_{\mathrm{degree \textendash two}}(t) \sim \mathrm{Ra}^{-1}
 \label{eq:degree_two_of_ra}
@@
 
-For two-dimensional simulations shown below there is a reduced dimensionality when calculating the number of modes,
-so $N_\mathrm{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\mathrm{degree2}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
+We performed a series of numerical simulations of mantle convection at different Rayleigh numbers to test this scaling.
+We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII81}. 
+In order to test a wide range of Rayleigh numbers, we ran the simulations in a 2D annulus, tracking the eigenvalues of the 
+moment of inertia tensor and integrating Equation~\eqref{eq:liouville_ode} in time.
+For the 2D simulations there is a reduced dimensionality when calculating the number of modes,
+so $N_\mathrm{modes} \sim \left(R/d \right)^2$.
+This leads us to a scaling of $T_{\mathrm{degree\textendash two}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
 This result has a simple interpretation.
@@
 If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principle axes.
-This is illustrated in Figure \ref{fig:perturb}
+This is illustrated schematically in Figure \ref{fig:perturb}
 
@@
 In the limit that the eigengap becomes zero, the rotation of the pinciple axes can be arbitrary. 
-This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}.
-We suggest, then, that a more vigorously convecting planet is much more likely to experience rapid TPW events.
+This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}, where 
+the mismatch angle is $90^\circ$. However, the eigengap does not need to be zero for there to be large displacements polar wander, 
+and if it does go to zero, the wander does not need to be $90^\circ$.
+
+Figure~\ref{fig:misfit} shows a representative timeseries for annular convection, where 
+we track the eigenvalues of the moment of inertia and integrate Equation~\eqref{eq:liouville_ode} in time.
+Since it is a 2D model, the spin axis may be represented by a single angle.
+When the eigengap gets small, the misfit angle becomes much larger, and the rate of polar wander becomes much faster.
+At $\sim$1 Gyr the eigengap goes to zero, and the misfit angle goes to approximately $90^\circ$, 
+an IITPW event. But there are several other events where the eigengap becomes small, and there are 
+still large TPW events associated with them.
+
+We suggest, then, that $90^\circ$ IITPW events are simply a special case of a broad class of events which can occur
+when the principal moments are close to each other.  Furthermore, a more vigorously convecting planet is much more likely to experience rapid TPW events
+as it has a lower characteristic gap between principal moments and the gap is much more likely to go to zero.
 
@@
 Bottom: Time series of spin axis and mismatch angle $\theta$.
-When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger.}
+When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger.
+At $\sim$1 Gyr the gap goes to zero and there is a nearly $90^\circ$ TPW event.
+However, there are several other large TPW events which happen when the eigengap is small.}
 \label{fig:misfit}
@@
 \begin{equation}
-I_{ij} = \int_V \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) d^3 {\bf r}
+I_{ij} = \int_V \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) dV
 \label{eq:inertia}
 \end{equation}
-
 where $\bf r$ is the Eulerian coordinate, $\rho$ is the density, and $V$ is the volume of the material.  
-The degree-two density structure may be expressed as a quadrupole-moment tensor (cf.
-\citet{jackson1998classical}):
-
+It is useful to enter the principal axes of the moment of inertia:
 \begin{equation}
-Q_{ij} = \int_V \rho \left( 3 r_i r_j - r_q r_q \delta_{ij} \right) d^3 {\bf r}
+{\bf I} = \bf{1} \begin{bmatrix}
+\lambda_1 \\ \lambda_2 \\ \lambda_3
+\end{bmatrix} = 
+{\bf 1} \begin{bmatrix}
+\int_V \rho (y^2+z^z) dV\\
+\int_V \rho (x^2+z^2) dV\\
+\int_V \rho (x^2+y^2) dV
+\end{bmatrix}
 \end{equation}
-
-The commutator of these tensors, $\bf QI - IQ$, is zero, indicating that they may be simultaneously diagonalized.
-
-\ifdetail
-We may demonstrate that the commutator is zero by using the definition of the commutator and expanding:
+where $\bf 1$ is the identity matrix, and $\lambda_1$, $\lambda_2$, and $\lambda_3$ are the principal moments.
+From equation~\eqref{eq:milankovitch} we see that the important quantities are the differences between the 
+principal moments, $(\lambda_3-\lambda_1)$, $(\lambda_3-\lambda_2)$ and $(\lambda_2-\lambda_1)$.
+These quantities may be rewritten in terms of degree-two real sperical harmonics \citep[e.g.][]{dahlen1999theoretical}.
+The relevant (fully normalized) harmonics are, in Cartesian coordinates:
 \begin{equation}
 \begin{aligned}
-{\bf IQ-QI} = &I_{ik} Q_{kj} - Q_{ik} I_{kj} \\
- = &\int_V \rho \left( r_q r_q \delta_{ik} - r_i r_k \right) d^3 {\bf r} \int_V \rho \left( 3 r_k r_j - r_q r_q \delta_{kj} \right) d^3 {\bf r} \\
- &- \int_V \rho \left( r_q r_q \delta_{kj} - r_k r_j \right) d^3 {\bf r} \int_V \rho \left( 3 r_i r_k - r_q r_q \delta_{ik} \right) d^3 {\bf r} \\
- = &3 \delta_{ik} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_k r_j d^3 {\bf r} \\
- &-\delta_{kj} \delta_{ik} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
- &-3 \int_V \rho r_i r_k d^3 {\bf r} \int_V \rho r_k r_j d^3 {\bf r} \\
- &+\delta_{kj} \int_V \rho r_i r_k d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
- - &3 \delta_{kj} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_k d^3 {\bf r} \\
- &+\delta_{ik} \delta_{kj} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
- &+3 \int_V \rho r_k r_j d^3 {\bf r} \int_V \rho r_i r_k d^3 {\bf r} \\
- &-\delta_{ik} \int_V \rho r_k r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
- = \;&3 \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_j d^3 {\bf r} \\
- &+ \int_V \rho r_i r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
- - &3 \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_j d^3 {\bf r} \\
- &- \int_V \rho r_i r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
- & = 0
+Y_{20} &= \frac{1}{4} \sqrt{\frac{ 5}{\pi}} \frac{ 2 z^2 - x^2 - y^2}{r^2} \\ 
+Y_{22} &= \frac{1}{4} \sqrt{\frac{15}{\pi}} \frac{ x^2 - y^2}{r^2}
 \end{aligned}
 \end{equation}
-\else
+Solving for $( \lambda_i - \lambda_j )$ in terms of these harmonics, we find
+\ifdetail
+It is useful to define unnormalized verions of the spherical harmonics as intermediate products:
+\begin{equation}
+\begin{aligned}
+Y_{20}^* &= 2 z^2 - x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{5} } r^2 Y_{20} \\ 
+Y_{22}^* &= x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{15} } r^2 Y_{22}
+\end{aligned}
+\end{equation}
+These make it easy to rewrite the differences between the $\lambda_n$:
+\begin{equation}
+\begin{aligned}
+(\lambda_2 - \lambda_1) &= \int_V \rho Y_{22}^* \, dV \\
+(\lambda_3 - \lambda_1) &= \frac{1}{2} \int_V  \rho \left( Y_{22}^* - Y_{20}^* \right) dV \\
+(\lambda_3 - \lambda_2) &= -\frac{1}{2} \int_V \rho \left( Y_{22}^* + Y_{20}^*  \right) dV \\
+\end{aligned}
+\end{equation}
+We may split this into integrations in radius and integrations over the surface of a sphere
+\begin{equation}
+\begin{aligned}
+(\lambda_2 - \lambda_1) &= \int_V \rho Y_{22}^* \, dV \\
+(\lambda_3 - \lambda_1) &= \frac{1}{2} \int_V  \rho \left( Y_{22}^* - Y_{20}^* \right) dV \\
+(\lambda_3 - \lambda_2) &= -\frac{1}{2} \int_V \rho \left( Y_{22}^* + Y_{20}^*  \right) dV \\
+\end{aligned}
+\end{equation}
+Substituting back in for the normalized spherical harmonics we find
 \fi
-
-We may therefore go into the principle coordinate system of both tensors to find
 \begin{equation}
-{\bf I} = {\bf 1} \begin{bmatrix}
-\int_V \rho (y^2+z^z)\\
-\int_V \rho (x^2+z^2) \\
-\int_V \rho (x^2+y^2) 
-\end{bmatrix}
+\begin{aligned}
+(\lambda_2 - \lambda_1) &= 4 \sqrt{\frac{\pi}{15} } \int_V   \rho r^2  Y_{22} \,dV \\
+(\lambda_3 - \lambda_1) &= 2 \sqrt{ \frac{\pi}{15} } \int_V  \rho r^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV \\
+(\lambda_3 - \lambda_2) &= -2 \sqrt{ \frac{\pi}{15} } \int_V  \rho r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV \\
+\end{aligned}
 \end{equation}
-
-\begin{equation}
-{\bf Q} = {\bf 1} \begin{bmatrix}
-\int_V \rho (2 x^2 - y^2 - z^z)\\
-\int_V \rho (2 y^2 - x^2 -z^2)\\
-\int_V \rho (2 z^2 - x^2 - y^2) 
-\end{bmatrix}
-\end{equation}
-
-where $\bf 1$ is the identity matrix.  If we let the diagonal elements of $\bf I$ be $\lambda_1, \lambda_2, \lambda_3$, we may rewrite $\bf Q$ as
-
-\begin{equation}
-{\bf Q} = {\bf 1} \begin{bmatrix}
-(\lambda_2-\lambda_1) + (\lambda_3-\lambda_1)\\
-(\lambda_1-\lambda_2) + (\lambda_3-\lambda_2)\\
-(\lambda_1-\lambda_3) + (\lambda_2-\lambda_3)\\
-\end{bmatrix}
-\label{eq:maccullagh}
-\end{equation}
-
-which shows the connection between the degree-two component of the density field and the principle moments of inertia.  This may be seen as a form of MacCullagh's formula \citep{munk1960rotation}.
-
-\section {TODO}
-
-\begin{enumerate}
-\item Make the argument about degree-two scaling much sharper
-\item Discuss high-viscosity lower mantle
-\item Earth-like parameters
-\end{enumerate}
+Up to the normalization constants, these expressions are idential to multipole expansions, 
+picking out the degree-two part of the density field laterally, and low-order polynomials radially.
+When density is a function of temperature, then we may directly identify this with the same expansion of the temperature field.
 

### 2015-08-28 figures/misfit.py

@@ -49,7 +49,7 @@ spin[pos] = np.nan
 fig = plt.figure()
 ax = fig.add_subplot(211)
 
-ax.plot(time, eigdiff, label=r'$(\lambda_1-\lambda_2)/I_0$')
+ax.plot(time, eigdiff, label=r'$(\lambda_2-\lambda_1)/I_0$')
 plt.xlim(tmin, tmax)
 plt.ylim(0., 2.e-5)
 ax.legend(loc='upper right')

### 2015-09-02 tpw_rate.tex
Caption
Disorganized work

@@
 \newif\ifdetail
-\detailfalse
-%\detailtrue
+%\detailfalse
+\detailtrue
 
@@
 \section{Rotational dynamics}
+\label{sec:rotational_dynamics}
 
 \subsection{The Liouville equation}
+\label{sec:liouville}
 Conservation of angular momentum for a torque-free system in a rotating reference frame requires
@@
 \end{equation}
-Equation \eqref{eq:quasistatic_liouville} indicates that ${\bf \Omega}$ and ${\bf I}$ are parallel, so a solution for ${\bf \Omega}(t)$ is equivalent to solving an eigenvalue problem for ${\bf I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
+Equation \eqref{eq:quasistatic_liouville} indicates that ${\bf \Omega}$ and ${\bf H}$ are parallel, so a solution for ${\bf \Omega}(t)$ is equivalent to solving an eigenvalue problem for ${\bf I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
 In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
@@
 \end{equation}
-
 where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, and $C_{ij}$ is the moment of inertia without that response.
@@
 An alternative to the loading Love number approach to surface deflections is to calculate the surface deflections directly using mantle convection simulations with a true free surface boundary condition.
+We have recently implemented the ability to conduct simulations with a free surface in the CIG sponsored mantle convection software \texttt{ASPECT} \citep{rose2015free}.
 
 \subsection{Rate of true polar wander}
+\label{sec:tpw_rate}
 
@@
 Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
-Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires four nondimensional numbers to characterize it.
+Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires four nondimensional numbers to characterize it
+(a fifth one, defined by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations).
 Convenient choices for these numbers are listed in Table \ref{tab:nondim}, along with approximate Earth-like values for them.
@@
 \int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV &= 0 \\
+\label{eq:quasistatic_angular_momentum}
 \end{aligned}
 \end{equation}
-which is a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.
+which may be rewritten via the Jacobi identity to find
+\ifdetail
+\begin{equation}
+\begin{aligned}
+- \int_V \rho (\mitbf{\Omega}\times \mathbf{r}) \times ( \mathbf{r} \times \mitbf{\Omega} ) \,dV + 
+- \int_V \rho \mitbf{\Omega}\times(\mitbf{\Omega}\times \mathbf{r}) \times \mathbf{r} \,dV &= 0 \\
+\end{aligned}
+\end{equation}
+\fi
+\begin{equation}
+\begin{aligned}
+ \mitbf{\Omega} \times \int_V \rho \mathbf{r} \times (\mitbf{\Omega}\times \mathbf{r}) \,dV &= 0 \\
+\end{aligned}
+\end{equation}
+This can be directly identified with Equation~\eqref{eq:quasistatic_liouville},
+and is a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.
 
-Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eq:eos} we find
+Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eq:eos} into \eqref{eq:quasistatic_angular_momentum}, we find
 \begin{equation}
@@
 $\kappa$ & thermal diffusivity \\
+$\alpha$ &  thermal expansivity \\ 
 $g_0$ & reference gravity \\
@@
 $T_0$ & reference temperature \\
-$\delta \rho = \rho_0 \alpha \Delta T$ & buoyancy parameter \\ 
+$\rho_0$ & reference density \\ 
+$\Delta T$ & temperature drop across mantle
 \end{tabular}
@@
 \centering
-\caption{Nondimensional numbers with approximate Earthlike values}
+\caption{Nondimensional numbers with approximate Earth-like values}
 \label{tab:nondim}
@@
 
+Having drawn the connection between the angular momentum equation (Section~\ref{sec:rotational_dynamics}) and the linear momentum 
+equation (Section~\ref{sec:internal}), we would like to apply insights from fluid dynamics and scaling of the 
+linear momentum equation to the problem of TPW. We would therefore like to rewrite Equation~\eqref{eq:simple_milankovitch}
+in terms of the nondimensional numbers identified in Section~\ref{sec:internal}.
+Specifically, we will need estimates for $(C-A)$, $T_1$, $(\lambda_3-\lambda_1)$, and $\theta$.
+
 It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
 This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
-With this we may rewrite Equation \eqref{eq:simple_milankovitch} in terms of the nondimensional numbers defined above.
 The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep{munk1960rotation}.
@@
 
-For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-twoo part of the temperature field (see Appendix \ref{appendix:moments}).
+For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix \ref{appendix:moments}).
 Therefore, an estimate of one allows an estimate of the other.
@@
 where the $T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$.
-
-The integral in Equation \eqref{eq:temperature_fluctuations} is picking out degree-two spherical harmonics in the lateral dimensions, and low order polynomials in radius.
-We therefore want to estimate the power in a single mode of the temperature expansion, allowing us identify
+Inserting the temperature expansion~\eqref{eq:T_series} into~\eqref{eq:temperature_fluctuations} results in 
+a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
+Orthogonality of the basis functions for the expansion results in the integral picking out degree-two 
+spherical harmonics in the lateral dimensions, and only having power in the lowest few radial functions $R_n(r)$.
+Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
+We want to estimate the power in a single mode of the temperature expansion, allowing us identify
 \begin{equation}
-\Lambda_{12} \sim \Gamma T_{\mathrm{degree \textendash two}}
+\Lambda_{31} \sim \Gamma T_{\mathrm{degree \textendash two}}
 \end{equation}
@@
 At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes which are largely influenced by the aspect ratio.
-This spectrum may or may not have a lot of power in degree-two, does not depend strongly on time.
+This spectrum may or may not have a lot of power in degree-two, and does not depend strongly on time.
 
@@
 We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII81}.
+which allows for flexible implementation of 
 In order to test a wide range of Rayleigh numbers, we ran the simulations in a 2D annulus, tracking the eigenvalues of the 
@@
 \end{equation}
-
 Other power spectra for the temperature field are possible. Isoviscous models tend to be ``bluer,'' and models with viscosity stratification tend to be ``redder'' \citep{richards1999polar}.
@@
 \end{equation} 
-
 That is to say, if there is a large difference between the eigenvalues, this stabilizes axes to perturbations.  
@@
 When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger.
-At $\sim$1 Gyr the gap goes to zero and there is a nearly $90^\circ$ TPW event.
+At $\sim$1 Gyr the gap goes to zero and there is a nearly $90^\circ$ TPW event, with $\sim80^\circ$ degrees of polar wander in $\sim$30 Myr.
 However, there are several other large TPW events which happen when the eigengap is small.}
@@
 \ifdetail
+
 It is useful to define unnormalized verions of the spherical harmonics as intermediate products:
@@
 \end{equation}
-We may split this into integrations in radius and integrations over the surface of a sphere
-\begin{equation}
-\begin{aligned}
-(\lambda_2 - \lambda_1) &= \int_V \rho Y_{22}^* \, dV \\
-(\lambda_3 - \lambda_1) &= \frac{1}{2} \int_V  \rho \left( Y_{22}^* - Y_{20}^* \right) dV \\
-(\lambda_3 - \lambda_2) &= -\frac{1}{2} \int_V \rho \left( Y_{22}^* + Y_{20}^*  \right) dV \\
-\end{aligned}
-\end{equation}
 Substituting back in for the normalized spherical harmonics we find
@@
 (\lambda_3 - \lambda_2) &= -2 \sqrt{ \frac{\pi}{15} } \int_V  \rho r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV \\
+\label{eq:degree_two_moments}
 \end{aligned}
@@
 When density is a function of temperature, then we may directly identify this with the same expansion of the temperature field.
+We can insert the equation of state~\eqref{eq:eos} into Equation~\eqref{eq:degree_two_moments} and integrate over a reference spherical volume $V_S$.
+This allows us to drop the terms which integrate to zero due to the orthogonality of spherical harmonics, and we are left with:
+\begin{equation}
+\begin{aligned}
+(\lambda_2 - \lambda_1) &= -4 \sqrt{\frac{\pi}{15} } \alpha \rho_0 \int_{V_S} T r^2  Y_{22} \,dV \\
+(\lambda_3 - \lambda_1) &= -2 \sqrt{ \frac{\pi}{15} } \alpha \rho_0 \int_{V_S} T r^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV \\
+(\lambda_3 - \lambda_2) &= 2 \sqrt{ \frac{\pi}{15} } \alpha \rho_0 \int_{V_S}  T r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV \\
+\label{eq:degree_two_moments}
+\end{aligned}
+\end{equation}
+We would like to normalize the differences in eigenvalues by the reference momentn $I_0$,
+which can also be written in terms of real spherical harmonics:
+\ifdetail
+\begin{equation}
+\begin{aligned}
+\int_{V_S} \rho_0 \frac{2 r^2 - Y_{20}^*}{3} &= \int_{V_S} \rho_0 (x^2 + y^2) dV = I_0 \\
+\end{aligned}
+\end{equation}
+\begin{equation}
+I_0 = \frac{2}{3} \int_{V_S} \rho_0 r^2 \left(1 - 2 \sqrt{\frac{\pi}{5}} Y_{20}\right) \,dV
+\end{equation}
+\fi
+\begin{equation}
+I_0 = \frac{2}{3} \int_{V_S} \rho_0 r^2 \,dV
+\end{equation}
+with this we can write expressions for $\Lambda_{ij}$:
+\begin{equation}
+\begin{aligned}
+\Lambda_{21} &= \alpha \Delta T \frac{-6 \sqrt{\frac{\pi}{15} } \int_{V_S} T^\prime r^2  Y_{22} \,dV}{\int_{V_S} r^2 \, dV} \\
+\Lambda_{31} &= \alpha \Delta T \frac{-3 \sqrt{ \frac{\pi}{15} } \int_{V_S} T^\prime r^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV}{\int_{V_S} r^2 \, dV}\\
+\Lambda_{32} &= \alpha \Delta T \frac{3 \sqrt{ \frac{\pi}{15} }  \int_{V_S}  T^\prime r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV}{\int_{V_S} r^2 \, dV} \\
+\end{aligned}
+\end{equation}
 

### 2015-09-02 tpw_rate.tex
Update reference

@@
 \newif\ifdetail
-%\detailfalse
-\detailtrue
+\detailfalse
+%\detailtrue
 
@@
 We performed a series of numerical simulations of mantle convection at different Rayleigh numbers to test this scaling.
-We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII81}.
+We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII82}.
 which allows for flexible implementation of 

### 2015-09-02 tpw_rate.tex
More text

@@
 Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
-We want to estimate the power in a single mode of the temperature expansion, allowing us identify
+We want to estimate the power in those few modes, which we denote by $T_{\mathrm{degree \textendash two}}$: 
 \begin{equation}
@@
 Other power spectra for the temperature field are possible. Isoviscous models tend to be ``bluer,'' and models with viscosity stratification tend to be ``redder'' \citep{richards1999polar}.
-Nevertheless, at high Rayleigh number the expectation is that the power is distributed across many length scales.
+Present day Earth seems to have a fairly ``red'' spectrum, with large low-degree seismic anomalies due to 
+Cenozoic subduction history and the lower mantle LLSVPs \citep{dziewonski2010mantle}.
+Nevertheless, at high Rayleigh number the expectation is that the power will be distributed across many length scales.
 
@@
 
+For Cenozoic Earth we can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch}.
+Typical values for the time constant $T_1$ are of order $30 \, \mathrm{kyr}$.
+Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies)
+are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia is $10^{-3} I_0$ \citep{chambat2001mean}.
+The largest question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
+\citet{richards1997explanation} have argued that the convective planform of Earth has been stable for the last
+few hundred million years. On the other hand, we know that there have been large reorganizations of that planform 
+during Earth history, so this may not hold in general \citep{evans2003true}.
+If the mismatch angle can become large, then we may estimate the maximum polar wander wander rate
+\begin{equation}
+\max ( \dot{\theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 90^\circ / \mathrm{Myr}
+\end{equation}
+which is similar to the timescales discussed by \citet{cambiotti2011new}.
+
 Thus far we have restricted our discussion to planets with lithospheres lacking long-term elastic strength.
@@
 \end{equation}
-We would like to normalize the differences in eigenvalues by the reference momentn $I_0$,
-which can also be written in terms of real spherical harmonics:
+We would like to normalize the differences in eigenvalues by the reference moment $I_0$:
 \ifdetail
@@
 \end{equation}
-with this we can write expressions for $\Lambda_{ij}$:
+\ifdetail
 \begin{equation}
@@
 \end{equation}
+\begin{equation}
+\begin{aligned}
+\Lambda_{21} &= \Gamma \frac{-6 \sqrt{\frac{\pi}{15} } \int_{V_S^\prime} T^\prime {r^\prime}^2  Y_{22} \,dV^\prime}{\int_{V_S^\prime} {r^\prime}^2 \,dV^\prime} \\
+\Lambda_{31} &= \Gamma \frac{-3 \sqrt{ \frac{\pi}{15} } \int_{V_S^\prime} T^\prime {r^\prime}^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV^\prime}{\int_{V_S} {r^\prime}^2 \, dV^\prime}\\
+\Lambda_{32} &= \Gamma \frac{3 \sqrt{ \frac{\pi}{15} }  \int_{V_S^\prime}  T^\prime {r^\prime}^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV^\prime}{\int_{V_S} {r^\prime}^2 \, dV^\prime} \\
+\end{aligned}
+\end{equation}
+\fi
+Dividing by $I_0$ and nodimensionalizing the integrals results in a factor of $\Gamma = \alpha \Delta T$ and a normalized set of 
+degree-two coefficients for the temperature field, which we abbreviate as $T_{\mathrm{degree \textendash two}}$:
+\begin{equation}
+\Lambda_{ij} \sim \Gamma T_{\mathrm{degree \textendash two}}
+\end{equation}
 

### 2015-09-08 tpw_rate.tex
Address comments, cleanup

@@
 These questions suggest that an approach rooted in dimensional analysis and fluid dynamics can clarify the rates and magnitudes of TPW.
-Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment if inertia variations \citep[e.g.][]{tsai2007theoretical, creveling2012mechanisms}.
+Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment of inertia variations \citep[e.g.][]{tsai2007theoretical, creveling2012mechanisms}.
 \citet{richards1999polar} coupled thermal convection models to a polar wander model, but did not address in detail the scaling relationships between the two.
@@
 \subsection{Rotational deformation}
+\label{sec:rotational_deformation}
 
@@
 The first term of this equation is zero due to the hydrostatic equation.  
-The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma$.
+The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma\equiv \alpha \Delta T$.
 Removing these, we find
@@
 \subsection{Buckingham Pi theorem and nondimensionalization}
+\label{sec:buckingham}
 The Buckingham Pi theorem (essentially an application of the rank-nullity theorem) allows us to determine the number of nondimensional numbers for this problem.
@@
 The temperature equation and equation of state become
-
 \begin{equation}
@@
 \end{equation}
-
 We may furthermore define a hydrostatic reference state where the density is the reference density everywhere and the velocity is zero, i.e.
@@
 \end{equation}
-
 Subtracting this from the momentum equation and defining a dynamic pressure $P^* = P - P_0$, we find
@@
 This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
-The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep{munk1960rotation}.
+The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep{munk1960rotation}:
 \begin{equation}
@@
 \subsection{An estimate of $\Lambda_{ij}$}
+\label{sec:lambda}
 
@@
 At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes which are largely influenced by the aspect ratio.
-This spectrum may or may not have a lot of power in degree-two, and does not depend strongly on time.
+This spectrum may or may not have a lot of power in the degree-two modes, and does not depend strongly on time.
 
@@
 which tends to wipe out thermal heterogeneity at small scales.  Consequently, there will be no power in modes
-with shorter lenghtscales than that allowed by diffusion. 
+with shorter lengthscales than that allowed by diffusion. 
 Therefore we expect, to a good approximation, that the infinite sum in Equation \eqref{eq:T_series} can be truncated at some maximum wavenumber, set by the smallest lengthscale $d$:
@@
 We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII82}.
-which allows for flexible implementation of 
+which allows for flexible implementation of different rheologies, geometries, and postprocessors.
 In order to test a wide range of Rayleigh numbers, we ran the simulations in a 2D annulus, tracking the eigenvalues of the 
@@
 \subsection{An estimate of the mismatch $\theta$}
+\label{sec:theta}
 
@@
 Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
-At high Rayleigh numbers response timescales for relaxation of the mismatch angle go down.
+Response timescales for relaxation of the mismatch angle go down at high Rayleigh numbers.
 At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
@@
 \frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{\mathrm{Fr}} \sin{2 \theta}
+\label{eq:simplest_milankovitch_defer_scaling}
 \end{equation}
@@
 
-Indeed, we can identify two endmember behaviors of Equation \eqref{eq:simplest_milankovitch}.
+Indeed, we can identify two endmember behaviors of Equation \eqref{eq:simplest_milankovitch_defer_scaling}.
 When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.

### 2015-09-08 tpw_rate.tex
comments, cleanup

@@
 On a dynamic planet $\bf I$ may be a function of time, so to conserve angular momentum $\bf \Omega$ must be also vary with time.
-In this case Equation \eqref{eq:euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
+In this case Equation~\eqref{eq:euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
 For a slowly convecting fluid such as a planetary mantle the inertial term $\partial {\bf H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
@@
 \end{equation}
-Equation \eqref{eq:quasistatic_liouville} indicates that ${\bf \Omega}$ and ${\bf H}$ are parallel, so a solution for ${\bf \Omega}(t)$ is equivalent to solving an eigenvalue problem for ${\bf I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
+Equation~\eqref{eq:quasistatic_liouville} indicates that ${\bf \Omega}$ and ${\bf H}$ are parallel, so a solution for ${\bf \Omega}(t)$ is equivalent to solving an eigenvalue problem for ${\bf I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
 In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
-The moment of inertia tensor in \eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
+The moment of inertia tensor in~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
 of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
@@
 where $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
-If we plug this decomposition into Equation \eqref{eq:quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
+If we plug this decomposition into Equation~\eqref{eq:quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
 \begin{equation}
@@
 
-The infinite-time limit of Equation \eqref{eq:viscoelastic_deformation} allows us to calculate the fluid limit of the Love number $k_f$:
+The infinite-time limit of Equation~\eqref{eq:viscoelastic_deformation} allows us to calculate the fluid limit of the Love number $k_f$:
 \begin{equation}
@@
 
-\citet{ricard1993polar} obtain an approximation to Equation \eqref{eq:viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.
+\citet{ricard1993polar} obtain an approximation to Equation~\eqref{eq:viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.
 This introduces a new parameter, termed $T_1$, which can be seen as a weighted relaxation time for the system.  
@@
 The second term represents the lag in the moment of inertia due to the viscous adjustment of the rotational bulge, where $T_1$ is the characteristic time constant for this adjustment.
-Since the first term represents the fluid limit of rotational deformation, it automatically satisfies Equation \eqref{eq:quasistatic_liouville}, and hence does not contribute to the polar wander problem.
+Since the first term represents the fluid limit of rotational deformation, it automatically satisfies Equation~\eqref{eq:quasistatic_liouville}, and hence does not contribute to the polar wander problem.
 
@@
 
-The term on the right-hand side of equation \eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
+The term on the right-hand side of equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
 This, as well, may be parameterized using a viscoelastic Love number approach:
@@
 \end{equation}
-An alternative to the loading Love number approach to surface deflections is to calculate the surface deflections directly using mantle convection simulations with a true free surface boundary condition.
-We have recently implemented the ability to conduct simulations with a free surface in the CIG sponsored mantle convection software \texttt{ASPECT} \citep{rose2015free}.
+An alternative to the Love number formulism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
+A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2015free}
+permits more general treatments of mantle rheology.
 
@@
 A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
-Substituting Equation \eqref{eq:rotational_deformation} into Equation \eqref{eq:disequilibrium} we find
+Substituting Equation~\eqref{eq:rotational_deformation} into Equation~\eqref{eq:disequilibrium} we find
 \begin{equation}
@@
 \end{equation}
-Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation \eqref{eq:fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
+Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
 \begin{equation}
@@
 \end{equation}
-here the parameters are defined in Table~\ref{tab:parameters}.
+where $\bf u$ is the velocity, $P$ is the pressure, and $T$ is the temperature
+(see Table~\ref{tab:parameters} for definitions of the other parameters}.
 In addition we use the simple equation of state
@@
 \end{equation}
-where $\bf u$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
 The other parameters are defined in Table \ref{tab:parameters}.
@@
 In addition, we expect deviations in the figure of the planet from its hydrostatic shape, which we denote by $V = V_H + \Delta V$, where $V_H$ is the hydrostatic figure and $\Delta V$ is the deviation.
-The introduction of $\Delta V$ requires another nondimensional number to characterize it, and we find that the quantity $\Gamma = \alpha \Delta T$ is convenient.
+The introduction of $\Delta V$ requires another nondimensional number to characterize it, and we find that the quantity $\Gamma \equiv \alpha \Delta T$ is convenient.
 Finally, we define $\mitbf{\Omega} = \Omega_0 \mitbf{\omega}$, where $\mitbf{\omega}$ is a unit vector in the direction of $\mitbf{\Omega}$.
 
-By definition the hydrostatic reference state is a solution to \eqref{eq:navier_stokes} where there is no flow:
+By definition the hydrostatic reference state is a solution to~\eqref{eq:navier_stokes} where there is no flow:
 
@@
 \end{equation}
-Equation \eqref{eq:navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
+
+We can explicitly draw the connection between the angular and linear momentum equations by returning to the dimensional
+Equation~\eqref{eq:navier_stokes}, crossing it with $\mathbf{r}$ and integrating over the volume of the mantle:
 \begin{equation}
 \begin{aligned}
--\int_V {\bf r} \times \nabla P \,dV + \int_V \eta {\bf r} \times \nabla^2 {\bf u} \,dV - \int_V \rho {\bf r \times g} \,dV 
-   + \int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV &= 0 \\
+-\int_V \mathbf{r} \times \nabla P \,dV + \int_V \eta \mathbf{r} \times \nabla^2 \mathbf{u} \,dV - \int_V \rho \mathbf{r} \times \mathbf{g} \,dV 
+   + \int_V \rho \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}  \,dV &= 0 \\
 \end{aligned}
@@
 \begin{aligned}
-\int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV &= 0 \\
+\int_V \rho \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV &= 0 \\
 \label{eq:quasistatic_angular_momentum}
@@
 
-Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eq:eos} into \eqref{eq:quasistatic_angular_momentum}, we find
+Splitting the volume into $V_H$ and $\Delta V$ and substituting~\eqref{eq:eos} into~\eqref{eq:quasistatic_angular_momentum}, we find
 \begin{equation}
 \begin{aligned}
-&\int_{V_H} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
-\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV +  \\
-&\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
-\int_{\Delta V} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 0  \\
+&\int_{V_H} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV + 
+\int_{V_H} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV +  \\
+&\int_{\Delta V} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV + 
+\int_{\Delta V} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 0  \\
 \end{aligned}
@@
 \begin{aligned}
-\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 
--\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV
+\int_{\Delta V} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 
+-\int_{V_H} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV
 \end{aligned}
 \end{equation}
-This equation may be identified with \eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
+This equation may be identified with~\eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
 Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table \ref{tab:nondim}.
@@
 \begin{equation}
-\frac{\partial T}{\partial t} + {\bf v} \cdot \nabla T = \nabla^2 T
+\frac{\partial T}{\partial t} + \mathbf{v} \cdot \nabla T = \nabla^2 T
 \end{equation}
@@
 \begin{equation}
- \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}
+ \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}= - \nabla P_0 + \rho_0 \mathbf{g}
 \end{equation}
@@
 \end{equation}
-Plugging these scalings into equation \eqref{eq:simple_milankovitch} we find
+Plugging these scalings into Equation~\eqref{eq:simple_milankovitch} we find
 \begin{equation}
@@
 Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
-We want to estimate the power in those few modes, which we denote by $T_{\mathrm{degree \textendash two}}$: 
+We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$: 
 \begin{equation}
-\Lambda_{31} \sim \Gamma T_{\mathrm{degree \textendash two}}
+\Lambda_{31} \sim \Gamma T_{\text{degree-two}}
 \end{equation}
@@
 with shorter lengthscales than that allowed by diffusion. 
-Therefore we expect, to a good approximation, that the infinite sum in Equation \eqref{eq:T_series} can be truncated at some maximum wavenumber, set by the smallest lengthscale $d$:
+Therefore we expect, to a good approximation, that the infinite sum in Equation~\eqref{eq:T_series} can be truncated at some maximum wavenumber, set by the smallest lengthscale $d$:
 \begin{equation}
-n_{\mathrm{max}}, l_{\mathrm{max}}, m_{\mathrm{max}} \sim \frac{R}{d}
+n_{\text{max}}, l_{\text{max}}, m_{\text{max}} \sim \frac{R}{d}
 \end{equation}
@@
 \begin{equation}
-N_{\mathrm{modes}} = n_{\mathrm{max}} \times l_{\mathrm{max}} \times m_{\mathrm{max}} \sim \left( \frac{R}{d} \right)^{3}
+N_{\text{modes}} = n_{\text{max}} \times l_{\text{max}} \times m_{\text{max}} \sim \left( \frac{R}{d} \right)^{3}
 \end{equation}
@@
 \begin{equation}
-T_{\mathrm{degree \textendash two}}(t) \sim \frac{1}{N_{\mathrm{modes}}} \sim \left( \frac{d}{R}\right)^3
+T_{\text{degree-two}}(t) \sim \frac{1}{N_{\text{modes}}} \sim \left( \frac{d}{R}\right)^3
 \end{equation}
@@
 \begin{equation}
-T_{\mathrm{degree \textendash two}}(t) \sim \mathrm{Ra}^{-1}
+T_{\text{degree-two}}(t) \sim \mathrm{Ra}^{-1}
 \label{eq:degree_two_of_ra}
@@
 For the 2D simulations there is a reduced dimensionality when calculating the number of modes,
-so $N_\mathrm{modes} \sim \left(R/d \right)^2$.
-This leads us to a scaling of $T_{\mathrm{degree\textendash two}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
+so $N_\text{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\text{degree-two}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
 This result has a simple interpretation.
@@
 
-The mismatch between the current axis and the principle axis of the convective moment is perhaps the most important parameter in Equation \eqref{eq:scaled_rotation}.  
+The mismatch between the current axis and the principle axis of the convective moment is perhaps the most important parameter in Equation~\eqref{eq:scaled_rotation}.  
 Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
@@
 
-We may arrive at a simple estimate of the growth rate of the angle $\theta$ by differentiating Equation \eqref{eq:kahan}, tentatively holding $\lVert \lambda_i - \lambda_j \rVert$ fixed:
+We may arrive at a simple estimate of the growth rate of the angle $\theta$ by differentiating Equation~\eqref{eq:kahan}, tentatively holding $\lVert \lambda_i - \lambda_j \rVert$ fixed:
 
@@
 As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
-There is a constant competition between growth of the mismatch angle $\theta$ through Equation \eqref{eq:grow_perturbation} and its relaxation through Equation \eqref{eq:simple_milankovitch}.
-We may plug in the estimate for $\Lambda_{ij}$ into equation \eqref{eq:simple_milankovitch} to find the strikingly simple expression
+There is a constant competition between growth of the mismatch angle $\theta$ through 
+Equation~\eqref{eq:grow_perturbation} and its relaxation through Equation~\eqref{eq:simple_milankovitch}.
+We may plug in the estimate for $\Lambda_{ij}$ into equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
 \begin{equation}
@@
 
-Indeed, we can identify two endmember behaviors of Equation \eqref{eq:simplest_milankovitch_defer_scaling}.
+Indeed, we can identify two endmember behaviors of Equation~\eqref{eq:simplest_milankovitch_defer_scaling}.
 When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.
@@
 during Earth history, so this may not hold in general \citep{evans2003true}.
-If the mismatch angle can become large, then we may estimate the maximum polar wander wander rate
+Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
+Allowing for such a large mismatch angle ($\theta = 45^\circ$) we may estimate the maximum polar wander rate
 \begin{equation}
@@
 picking out the degree-two part of the density field laterally, and low-order polynomials radially.
-When density is a function of temperature, then we may directly identify this with the same expansion of the temperature field.
-We can insert the equation of state~\eqref{eq:eos} into Equation~\eqref{eq:degree_two_moments} and integrate over a reference spherical volume $V_S$.
+When density is a function of temperature, we can insert the equation of state~\eqref{eq:eos} into 
+Equation~\eqref{eq:degree_two_moments} and integrate over a reference spherical volume $V_S$.
 This allows us to drop the terms which integrate to zero due to the orthogonality of spherical harmonics, and we are left with:
@@
 (\lambda_3 - \lambda_2) &= 2 \sqrt{ \frac{\pi}{15} } \alpha \rho_0 \int_{V_S}  T r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV \\
-\label{eq:degree_two_moments}
+\label{eq:degree_two_moments_of_temperature}
 \end{aligned}
 \end{equation}
-We would like to normalize the differences in eigenvalues by the reference moment $I_0$:
+We normalize the differences in eigenvalues by the reference moment $I_0$:
 \ifdetail
@@
 Dividing by $I_0$ and nodimensionalizing the integrals results in a factor of $\Gamma = \alpha \Delta T$ and a normalized set of 
-degree-two coefficients for the temperature field, which we abbreviate as $T_{\mathrm{degree \textendash two}}$:
+degree-two coefficients for the temperature field, which we abbreviate as $T_{\text{degree-two}}$:
 \begin{equation}
-\Lambda_{ij} \sim \Gamma T_{\mathrm{degree \textendash two}}
+\Lambda_{ij} \sim \Gamma T_{\text{degree-two}}
 \end{equation}

### 2015-09-08 tpw_rate.tex
fix

@@
 where $\bf u$ is the velocity, $P$ is the pressure, and $T$ is the temperature
-(see Table~\ref{tab:parameters} for definitions of the other parameters}.
+(see Table~\ref{tab:parameters} for definitions of the other parameters).
 In addition we use the simple equation of state

### 2015-09-08 tpw_rate.tex
consistently use mathbf

@@
 \begin{equation}
-\frac{d {\bf H}}{dt} + {\bf \Omega} \times {\bf H} = 0
+\frac{d \mathbf{H}}{dt} + \mitbf{\Omega} \times \mathbf{H} = 0
 \label{eq:euler}
 \end{equation}
-known as Euler's equations, where ${\bf H} = {\bf I} \cdot {\bf \Omega}$ is the angular momentum vector, $\bf I$ is the moment of inertia tensor, and $\bf \Omega$ is the angular velocity.
-On a dynamic planet $\bf I$ may be a function of time, so to conserve angular momentum $\bf \Omega$ must be also vary with time.
+known as Euler's equations, where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity.
+On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
 In this case Equation~\eqref{eq:euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
-For a slowly convecting fluid such as a planetary mantle the inertial term $\partial {\bf H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
+For a slowly convecting fluid such as a planetary mantle the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
 \begin{equation}
-{\bf \Omega } (t)\times ( {\bf I}(t) \cdot {\bf \Omega }(t) ) = 0
+\mitbf{\Omega } (t)\times ( \mathbf{I}(t) \cdot \mitbf{\Omega }(t) ) = 0
 \label{eq:quasistatic_liouville}
 \end{equation}
-Equation~\eqref{eq:quasistatic_liouville} indicates that ${\bf \Omega}$ and ${\bf H}$ are parallel, so a solution for ${\bf \Omega}(t)$ is equivalent to solving an eigenvalue problem for ${\bf I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
+Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for $\mathbf{I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
 In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
@@
 \begin{equation}
-{\bf \Omega} \times ({\bf J} \cdot {\bf \Omega}) = -{\bf \Omega } \times ( {\bf E} \cdot {\bf \Omega} )
+\mathbf{\Omega} \times (\mathbf{J} \cdot \mitbf{\Omega}) = -\mitbf{\Omega} \times ( \mathbf{E} \cdot \mitbf{\Omega} )
 \label{eq:disequilibrium}
 \end{equation}
-This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches of the convective part of the moment of inertia ($\bf E$) and the rotational deformation part of the moment of inertia ($\bf J$).
+This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches
+of the convective part of the moment of inertia ($\mathbf{E}$) and the rotational deformation part of the moment of inertia ($\mathbf{J}$).
 Our goal is to characterize this balance from a perspective of scaling and fluid dynamics.
@@
 \begin{equation} 
-{\bf E} = \left[ \delta(t) + k^L(t) \right] * {\bf C}
+\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
 \end{equation}
@@
 
-We are in a position to address the rates of true polar wander for a given convective moment $\bf E$.
+We are in a position to address the rates of true polar wander for a given convective moment $\mathbf{E}$.
 A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
@@
 \begin{equation}
-\frac{k_f a^5 T_1}{3G}{\bf \Omega} \times {\bf \dot{\Omega} } = {\bf \Omega} \times \left( {\bf E \cdot \Omega} \right)
+\frac{k_f a^5 T_1}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right)
 \label{eq:liouville}
 \end{equation}
-Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
+Introducing a unit vector $\mitbf{\omega} = \mitbf{\Omega}/\|\mitbf{\Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for $\dot{\mitbf{\omega}}$:
 \begin{equation}
- \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ {\bf E \cdot \mitbf{\omega}} - \left( {\bf \mitbf{\omega} \cdot E \cdot \mitbf{\omega} } \right) \mitbf{\omega} \right]
+ \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right]
 \label{eq:liouville_ode}
@@
 Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
-If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
+If we enter the coordinate system of the convective moment of inertia $\mathbf{E}$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
 \ifdetail
@@
 These equations are a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.  
-Written this way, it is clear that the important quantities are $\theta$ and the differences between the eigenvalues of the convective moment $\bf E$, both of which depend on the structure and dynamics of mantle convection.
+Written this way, it is clear that the important quantities are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.
 They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principle axis of the convective moment.
@@
 \begin{equation}
-\nabla \cdot {\bf u} = 0
+\nabla \cdot \mathbf{u} = 0
 \label{eq:conserve_mass}
@@
 \begin{aligned}
-- \nabla P + \eta \nabla^2 {\bf u} =  \rho {\bf g} -  \rho{\bf \Omega \times \Omega \times r}
+- \nabla P + \eta \nabla^2 \mathbf{u} =  \rho \mathbf{g} -  \rho \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}
 \label{eq:navier_stokes}
@@
 \begin{equation}
-\frac{\partial T}{\partial t} + {\bf u} \cdot \nabla T = \kappa \nabla^2 T
+\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \kappa \nabla^2 T
 \label{eq:energy}
 \end{equation}
-where $\bf u$ is the velocity, $P$ is the pressure, and $T$ is the temperature
+where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature
 (see Table~\ref{tab:parameters} for definitions of the other parameters).
@@
 \begin{aligned}
-- \nabla P_0 =  \rho_0 {\bf g} -  \rho_0 {\bf \Omega \times \Omega \times r}
+- \nabla P_0 =  \rho_0 \mathbf{g} -  \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}
 \label{eq:hydrostatic}
@@
 \begin{aligned}
- - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times {\bf r}} = 0
+ - \nabla P^* + \; \nabla^2 \mathbf{v} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0
 \end{aligned}
@@
 $\rho^\prime$ &=& $\rho_0 \;\; \rho$\\
-$v^\prime$ &=& $\kappa/R \;\; v$ \\
+$u^\prime$ &=& $\kappa/R \;\; u$ \\
 $P^\prime$ &=& $\eta \kappa/R^2 \;\; P$ \\
@@
 \begin{equation}
-\frac{\partial T}{\partial t} + \mathbf{v} \cdot \nabla T = \nabla^2 T
+\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \nabla^2 T
 \end{equation}
@@
 \begin{aligned}
- - & \Omega_0^2  \alpha  \Delta T T R \; {\bf \Omega \times \Omega \times r} = - \eta \kappa / R^2 \; \nabla P^* \\ 
-&+ \eta \kappa / R^2 \; \nabla \cdot \left( \eta \epsilon ({\bf v}) \right) - \alpha \Delta T g_0 \; {\bf g}
+ - & \Omega_0^2  \alpha  \Delta T T R \; \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} = - \eta \kappa / R^2 \; \nabla P^* \\ 
+&+ \eta \kappa / R^2 \; \nabla \cdot \left( \eta \epsilon (\mathbf{u}) \right) - \alpha \Delta T g_0 \; \mathbf{g}
 \end{aligned}
@@
 \begin{aligned}
- - & \mathrm{Ra \; Fr}\; T \; {\bf \Omega \times \Omega \times r} = - \nabla P^* \\ 
-&+ \; \nabla \cdot \left( \eta \epsilon ({\bf v}) \right) - \mathrm{Ra} \; T \; {\bf g}
+ - & \mathrm{Ra \; Fr}\; T \; \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} = - \nabla P^* \\ 
+&+ \; \nabla \cdot \left( \eta \epsilon (\mathbf{u}) \right) - \mathrm{Ra} \; T \; \mathbf{g}
 \end{aligned}
@@
 \end{equation}
-where $\bf r$ is the Eulerian coordinate, $\rho$ is the density, and $V$ is the volume of the material.  
+where $\mathbf{r}$ is the Eulerian coordinate, $\rho$ is the density, and $V$ is the volume of the material.  
 It is useful to enter the principal axes of the moment of inertia:
 \begin{equation}
-{\bf I} = \bf{1} \begin{bmatrix}
+\mathbf{I} = \mathbf{1} \begin{bmatrix}
 \lambda_1 \\ \lambda_2 \\ \lambda_3
 \end{bmatrix} = 
-{\bf 1} \begin{bmatrix}
+\mathbf{1} \begin{bmatrix}
 \int_V \rho (y^2+z^z) dV\\
@@
 \end{equation}
-where $\bf 1$ is the identity matrix, and $\lambda_1$, $\lambda_2$, and $\lambda_3$ are the principal moments.
+where $\mathbf{1}$ is the identity matrix, and $\lambda_1$, $\lambda_2$, and $\lambda_3$ are the principal moments.
 From equation~\eqref{eq:milankovitch} we see that the important quantities are the differences between the 

### 2015-09-08 figures/eigengap.py
Adjust figures

@@ -38,6 +38,7 @@ frac = 50
 
 def get_spin_spectrum(text_file):
 
+  print text_file
   #load the data
   data = np.loadtxt(text_file,skiprows=50, usecols=(1,17,18,19,20))
   #get rid of duplicate times (why might these exist?)
@@ -47,10 +48,11 @@ def get_spin_spectrum(text_file):
       stats.append(data[i,:])
   stats = np.array(stats) 
         
-  eig1tck = interp.splrep(stats[:,0],stats[:,3], k=3 )
-  eig2tck = interp.splrep(stats[:,0],stats[:,4], k=3)
-  axistck = interp.splrep(stats[:,0],stats[:,1], k=3)
-  mismatchtck = interp.splrep(stats[:,0],stats[:,2], k=3)
+  s = 3.e51 #magic number for some slight smoothing
+  eig1tck = interp.splrep(stats[:,0],stats[:,3], k=3, s=s)
+  eig2tck = interp.splrep(stats[:,0],stats[:,4], k=3, s=s)
+  axistck = interp.splrep(stats[:,0],stats[:,1], k=3, s=s)
+  mismatchtck = interp.splrep(stats[:,0],stats[:,2], k=3, s=s)
 
   times = np.linspace(stats[0,0], stats[-1,0], len(stats[:,0]))
   lambda1 = interp.splev(times, eig1tck, der=0)
@@ -58,20 +60,12 @@ def get_spin_spectrum(text_file):
   axis = interp.splev(times, axistck, der=0)
   mismatch = interp.splev(times, mismatchtck, der=0)
 
-  # nondimensionalize
+  # normalize
   diff = (lambda1-lambda2)/( np.mean(lambda1 + lambda2)/2.0 )
+  # convert to Gyr
   times = times/1.e9 #* kappa / D / D
  
-  frequency = times[-1]/len(times)
-  autocorr = lambda x: np.correlate(x,x, mode='same')
-  psd = lambda x : signal.welch(x,  fs=frequency,scaling='spectrum', return_onesided=True, nfft=100, nperseg=10)
-
-  acorrdiff = autocorr(diff)
-  p = psd(diff)
-  peak = p[0][np.argmax(p[1])]
-  print text_file, peak
-
-  return times, diff, acorrdiff, peak
+  return times, diff
 
 
 
[…]

### 2015-09-08 figures/misfit.py

@@ -49,13 +49,13 @@ spin[pos] = np.nan
 fig = plt.figure()
 ax = fig.add_subplot(211)
 
-ax.plot(time, eigdiff, label=r'$(\lambda_2-\lambda_1)/I_0$')
+ax.plot(time, eigdiff, label=r'$\Lambda_{21} = (\lambda_2-\lambda_1)/I_0$')
 plt.xlim(tmin, tmax)
 plt.ylim(0., 2.e-5)
 ax.legend(loc='upper right')
 ax.xaxis.set_major_formatter(matplotlib.ticker.NullFormatter())
 ax.yaxis.set_major_locator(matplotlib.ticker.MaxNLocator(nbins=4, prune=None))
-plt.ylabel('Relative moment')
+plt.ylabel(r'Eigengap $\Lambda_{21}')
 
 ax = fig.add_subplot(212)
 
@@ -67,7 +67,7 @@ plt.xlim(tmin, tmax)
 plt.ylim(0., 180.)
 ax.legend(loc='upper right')
 ax.yaxis.set_major_locator(matplotlib.ticker.MaxNLocator(nbins=3, prune=None))
-plt.xlabel(r"Time")
+plt.xlabel(r"Time (Gyr)")
 plt.ylabel(r"Angle ($^\circ$)")
 
 plt.savefig("misfit.pdf", bbox_inches='tight')

### 2015-09-08 figures/misfit.py
change axis label

@@ -55,7 +55,7 @@ plt.ylim(0., 2.e-5)
 ax.legend(loc='upper right')
 ax.xaxis.set_major_formatter(matplotlib.ticker.NullFormatter())
 ax.yaxis.set_major_locator(matplotlib.ticker.MaxNLocator(nbins=4, prune=None))
-plt.ylabel(r'Eigengap $\Lambda_{21}')
+plt.ylabel(r'Eigengap')
 
 ax = fig.add_subplot(212)
 

### 2015-09-08 figures/perturb.sh
add some annotation

@@ -4,33 +4,41 @@ gmt psxy -R-2/2/-2/2 -Jx3 -Se -Wthicker,black -K > perturb.ps <<EOF
 0 0 20 3.6 2.8
 EOF
 
+# red axis
 gmt psxy -R -J -Sv0.3+e -Wthicker,red -K -O >> perturb.ps <<EOF
 0 0 20 1.8 1.8
 EOF
+#blue axis
 gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
 0 0 110 1.4 1.4
 EOF
 
+#lambda 1
 gmt pstext -R -J -K  -O >>perturb.ps <<EOF
 .0 0.36 @~l@-1@-@~
 EOF
 
+#lambda 2
 gmt pstext -R -J -K -O >>perturb.ps <<EOF
 .32 0.26 @~l@-2@-@~
 EOF
 
+#theta
 gmt pstext -R -J -K -O >>perturb.ps <<EOF
 .46 0.10 @~q@~
 EOF
 
+#Rotation axis
 gmt psxy -R -J -Sv0.3+e -Wthicker,black -K -O >> perturb.ps <<EOF
 0 0 4 2.4 1 
 EOF
 
+#Omega
 gmt pstext -R -J -K -O >>perturb.ps <<EOF
 0.9 .05 @~W@~
 EOF
 
+#perturbation
 gmt psxy -R -J -Sc.1i -Gdarkgray -K -O >>perturb.ps <<EOF
 -0.2 0.3
 EOF
@@ -40,33 +48,41 @@ gmt psxy -R-2/2/-2/2 -Jx3 -X2i -Se -Wthicker,black -K -O >> perturb.ps <<EOF
 0 0 70 3.2 3.2
 EOF
 
+#red axis
 gmt psxy -R -J -Sv0.3+e -Wthicker,red -K -O >> perturb.ps <<EOF
 0 0 80 1.6 1.6
 EOF
+#blue axis
 gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
 0 0 170 1.6 1.6
 EOF
 
+#lambda 1
 gmt pstext -R -J -K  -O >>perturb.ps <<EOF
 -0.36 0.2 @~l@-1@-@~
 EOF
 
+#lambda 2
 gmt pstext -R -J -K -O >>perturb.ps <<EOF
[…]

### 2015-09-08 figures/perturb.sh
Flip axis

@@ -10,12 +10,12 @@ gmt psxy -R -J -Sv0.3+e -Wthicker,red -K -O >> perturb.ps <<EOF
 EOF
 #blue axis
 gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
-0 0 110 1.4 1.4
+0 0 290 1.4 1.4
 EOF
 
 #lambda 1
 gmt pstext -R -J -K  -O >>perturb.ps <<EOF
-.0 0.36 @~l@-1@-@~
+.04 -0.32 @~l@-1@-@~
 EOF
 
 #lambda 2
@@ -54,17 +54,17 @@ gmt psxy -R -J -Sv0.3+e -Wthicker,red -K -O >> perturb.ps <<EOF
 EOF
 #blue axis
 gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
-0 0 170 1.6 1.6
+0 0 350 1.6 1.6
 EOF
 
 #lambda 1
 gmt pstext -R -J -K  -O >>perturb.ps <<EOF
--0.36 0.2 @~l@-1@-@~
+.36 -0.15 @~l@-1@-@~
 EOF
 
 #lambda 2
 gmt pstext -R -J -K -O >>perturb.ps <<EOF
-.0 0.46 @~l@-2@-@~
+-.02 0.46 @~l@-2@-@~
 EOF
 
 #theta

### 2015-09-08 tpw_rate.tex
Remove unused package
principal

@@
 \usepackage{graphicx}
-\usepackage{color, soul}
 
@@
 Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for $\mathbf{I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
-In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
+In practice, this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
 The moment of inertia tensor in~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
@@
 Written this way, it is clear that the important quantities are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.
-They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principle axis of the convective moment.
+They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principal axis of the convective moment.
 Both $\theta$ and $\lambda_i$ depend upon the dynamics of the convecting planetary mantle.
@@
 This can be directly identified with Equation~\eqref{eq:quasistatic_liouville},
-and is a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.
+and is a statement that a quasistatic body will rotate around the principal axis of its total moment of inertia.
 
@@
 At this point we do not have estimates for the characteristic magnitudes of $\Lambda_{ij}$ or $\theta$, both of which are crucial for predicting characteristic rates of TPW.
-They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principle axis of the convective moment.
+They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principal axis of the convective moment.
 
@@
 
-The mismatch between the current axis and the principle axis of the convective moment is perhaps the most important parameter in Equation~\eqref{eq:scaled_rotation}.  
+The mismatch between the current axis and the principal axis of the convective moment is perhaps the most important parameter in Equation~\eqref{eq:scaled_rotation}.  
 Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
@@
 That is to say, if there is a large difference between the eigenvalues, this stabilizes axes to perturbations.  
-If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principle axes.
+If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principal axes.
 This is illustrated schematically in Figure \ref{fig:perturb}
@@
 However, on the left the eigengap $\lVert \lambda_2 - \lambda_1 \rVert$ is large, while on the right it is small.
-A negative mass perturbation is instantaneously added to both bodies, which effects a small rotation of the principle axes on the left, but a large one on the right.}
+A negative mass perturbation is instantaneously added to both bodies, which effects a small rotation of the principal axes on the left, but a large one on the right.}
 \label{fig:perturb}
@@
 Additionally, the timescale of fluctuations in these values goes down. 
-Overall, this makes the principle axes of high Rayleigh number systems much less stable.
+Overall, this makes the principal axes of high Rayleigh number systems much less stable.
 This is consistent with the result of \citet{richards1999polar}.
@@
 \includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
-\caption{Top: Time series of principle moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.
+\caption{Top: Time series of principal moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.
 Bottom: Time series of spin axis and mismatch angle $\theta$.
@@
 The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
-As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
+As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principal axes of the convective moment of inertia. 
 There is a constant competition between growth of the mismatch angle $\theta$ through 

### 2015-09-08 tpw_rate.tex
More edits

@@
 In practice, this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
-The moment of inertia tensor in~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
+The moment of inertia tensor in Equation~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
 of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
@@
 Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
-Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires four nondimensional numbers to characterize it
-(a fifth one, defined by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations).
+Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires three nondimensional numbers to characterize it
+(a fourth one, defined by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations).
 Convenient choices for these numbers are listed in Table \ref{tab:nondim}, along with approximate Earth-like values for them.
@@
 
-Fluctuations in the nonhydrostatic moment of inertia are due to temperature fluctuations in the interior and their spatial structure:
+Fluctuations in the nonhydrostatic moment of inertia are due to temperature fluctuations in the mantle and their spatial structure:
 
@@
 We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
-where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials:
+where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
+$T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$:
 \begin{equation} 
@@
 \end{equation}
-where the $T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$.
 Inserting the temperature expansion~\eqref{eq:T_series} into~\eqref{eq:temperature_fluctuations} results in 
@@
 Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
-However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \sim \mathrm{Ra}^{-\beta}$.
+However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
 With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
@@
 The Froude number acts as a damper to TPW.
-The Rayleigh number is more complicated, since it is a control on both the forcing 
+The dependence on the Rayleigh number is more complicated, since it is a control on both the forcing 
 of TPW and the response, which act in opposite directions.
+Overall, however, we expect that more vigorously convecting planets should be less rotationally stable, and experience more TPW.
 This perspective allows us to consider not only the polar wandering of Phanerozoic Earth,

### 2015-09-09 tpw_rate.tex
Add support note
Address comments, update estimate of rate

@@
 Despite this, the ability of internal mass anomalies to drive large-scale TPW remains controversial. 
-Paleomagnetic data have been interpreted to require up to $\sim 1^\circ$/Myr rates of TPW \citep{mitchell2011sutton}, but the ability of the mantle to respond at such rates has been questioned \citep{tsai2007theoretical}.
+Paleomagnetic data have been interpreted to require up to $3^\circ-12^\circ$/Myr rates of TPW \citep{mitchell2011sutton}, but the ability of the mantle to respond at such rates has been questioned \citep{tsai2007theoretical}.
 
@@
 \end{equation}
-The other parameters are defined in Table \ref{tab:parameters}.
 Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
@@
 \begin{equation}
-\max \left( T(x,y,z, t) \right) = \max \left( \displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \right) = 1
+\max \left( \displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \right) = 1
 \end{equation}
@@
 \end{equation}
-In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.
+In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
+$d(\mathrm{Ra})$ will still result in a weak dependence of the prefactor in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} 
+on the Rayleigh number.
 
 We then suggest that the most important parameters are the Froude number, which acts as the brakes on the system, and the mismatch angle $\theta(\mathrm{Ra})$.
-Whereas the rest of the expression has at most a weak dependence on $\mathrm{Ra}$, $\theta$ is expected to be strongly dependent on it.
+Whereas the rest of the expression has a weak dependence on $\mathrm{Ra}$, $\theta$ is expected to be strongly dependent on it.
 
@@
 For Cenozoic Earth we can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch}.
-Typical values for the time constant $T_1$ are of order $30 \, \mathrm{kyr}$.
+Typical values for the time constant $T_1$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
 Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies)
-are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia is $10^{-3} I_0$ \citep{chambat2001mean}.
+are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
 The largest question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
@@
 few hundred million years. On the other hand, we know that there have been large reorganizations of that planform 
-during Earth history, so this may not hold in general \citep{evans2003true}.
+during Earth history, so this recent geologic stability may not hold in general \citep{evans2003true}.
 Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
@@
 \begin{equation}
-\max ( \dot{\theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 90^\circ / \mathrm{Myr}
+\max ( \dot{\theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 6^\circ / \mathrm{Myr}
 \end{equation}
-which is similar to the timescales discussed by \citet{cambiotti2011new}.
+which is similar to the timescales discussed by \citet{cambiotti2011new}, and within the range 
+suggested by paleomagnetic observations.
+The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
+which results in a corresponding uncertainty for the relaxation time $T_1$ and the maximum polar wander rate.
 
@@
 \begin{acknowledgments}
+This work was supported by National Science Foundation grant EAR-1246670.
 \end{acknowledgments}

### 2015-11-02 tpw_rate.tex
Fix typo

@@
 A rotating, quasistatic body like a planetary mantle will tend to spin about the axis of its maximum moment of inertia.
-Convection in a planetary mantle continuously redistributes mass, which can change the moment of ineratia tensor, necessitating a change in the spin axis of the planet to conserve angular momentum, a process known as true polar wander (TPW).
+Convection in a planetary mantle continuously redistributes mass, which can change the moment of inertia tensor, necessitating a change in the spin axis of the planet to conserve angular momentum, a process known as true polar wander (TPW).
 

### 2016-01-19 figures/reference_frames.py
Add rotation indication arrows

@@ -8,6 +8,9 @@ from matplotlib import rc
 
 rc('text', usetex=True)
 
+d2r = np.pi/180.
+r2d = 180./np.pi
+
 class Arrow3D(FancyArrowPatch):
     def __init__(self, xs, ys, zs, *args, **kwargs):
         FancyArrowPatch.__init__(self, (0,0), (0,0), *args, **kwargs)
@@ -19,20 +22,75 @@ class Arrow3D(FancyArrowPatch):
         self.set_positions((xs[0],ys[0]),(xs[1],ys[1]))
         FancyArrowPatch.draw(self, renderer)
 
-def Rx(phi):
-    return np.array([[1, 0, 0],
-                     [0, np.cos(phi), -np.sin(phi)],
-                     [0, np.sin(phi), np.cos(phi)]])
+def construct_euler_rotation_matrix(alpha, beta, gamma):
+    """
+    Make a 3x3 matrix which represents a rigid body rotation,
+    with alpha being the first rotation about the z axis,
+    beta being the second rotation about the y axis, and
+    gamma being the third rotation about the z axis.
+ 
+    All angles are assumed to be in radians
+    """
+    rot_alpha = np.array( [ [np.cos(alpha), -np.sin(alpha), 0.],
+                            [np.sin(alpha), np.cos(alpha), 0.],
+                            [0., 0., 1.] ] )
+    rot_beta = np.array( [ [np.cos(beta), 0., np.sin(beta)],
+                           [0., 1., 0.],
+                           [-np.sin(beta), 0., np.cos(beta)] ] )
+    rot_gamma = np.array( [ [np.cos(gamma), -np.sin(gamma), 0.],
+                            [np.sin(gamma), np.cos(gamma), 0.],
+                            [0., 0., 1.] ] )
+    rot = np.dot( rot_gamma, np.dot( rot_beta, rot_alpha ) )
+    return rot
 
-def Ry(theta):
-    return np.array([[np.cos(theta), 0, np.sin(theta)],
-                     [0, 1, 0],
-                     [-np.sin(theta), 0, np.cos(theta)]])
+def spherical_to_cartesian( longitude, latitude, norm ):
+    assert(np.all(longitude >= 0.) and np.all(longitude <= 360.))
+    assert(np.all(latitude >= -90.) and np.all(latitude <= 90.))
[…]

### 2016-01-19 figures/reference_frames.py
Adjustments to arrows

@@ -67,7 +67,7 @@ def plot_rotation_arrow( axes, arrow, ccw=True, frac = 0.75, *args, **kwargs):
 
     # Make points around the arrow
     colat = 4.
-    azimuths = np.linspace(0., 300., 301.)
+    azimuths = np.linspace(0., 330., 331.)
     lats = np.ones_like(azimuths)*(90. - colat)
     norms = np.ones_like(azimuths)*norm[0]*frac
     points = spherical_to_cartesian(azimuths, lats, norms)
@@ -75,19 +75,19 @@ def plot_rotation_arrow( axes, arrow, ccw=True, frac = 0.75, *args, **kwargs):
     rotated_points = np.dot( rotation_matrix, points )
 
     # plot the points
-    dline = 20
+    dline = 40
     line = axes.plot(rotated_points[0,:-dline], rotated_points[1,:-dline], rotated_points[2,:-dline], lw=1, color='k')
     # add the arrowhead
     if ccw is True:
         head = Arrow3D( [rotated_points[0,-1-dline], rotated_points[0,-1]],
                         [rotated_points[1,-1-dline], rotated_points[1,-1]],
                         [rotated_points[2,-1-dline], rotated_points[2,-1]], 
-                        arrowstyle='-|>', mutation_scale=20, color='k')
+                        arrowstyle='-|>', mutation_scale=10, color='k')
     else:
         head = Arrow3D( [rotated_points[0,dline], rotated_points[0,0]],
                         [rotated_points[1,dline], rotated_points[1,0]],
                         [rotated_points[2,dline], rotated_points[2,0]], 
-                        arrowstyle='-|>', mutation_scale=20, color='k')
+                        arrowstyle='-|>', mutation_scale=10, color='k')
     axes.add_artist(head)
     return line,head
 

### 2016-01-21 tpw_rate.tex
Fix typo
Enormous derivation of magnitude of polar wander rate

@@
 \begin{equation}
-\frac{k_f a^5 T_1}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right)
+\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right)
 \label{eq:liouville}
@@
 
-And here is the tedious algebra.
-Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$, and the pole of the coordinate system ($\theta = 0$) associated with the eigenvector for $\lambda_3$.
-Therefore we find that $\mitbf{\omega} = \left[ \sin{\theta} \cos{\phi} \;\; \sin{\theta} \sin{\phi} \;\; \cos{\theta} \right]^T$.
-Plugging this in, we find a system of three equations, which we want to solve for $\dot{\theta}$ and $\dot{\phi}$:
+And here is the tedious algebra.  Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$.
+We are interested in the rate of true polar wander, which is the angular change of the pole
+per unit time. We can transform equation~\ref{eq:liouville_ode} into a scalar equation by dotting it
+with $\dot{\mitbf{\omega}}$, since we are interested in the magnitude of the rate, rather than
+the direction at the moment.
+
 \begin{equation}
 \begin{aligned}
- & \cos{\theta}\cos{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \dot{\phi} = \\
-  &A \sin{\theta}\cos{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
- &\cos{\theta}\sin{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \dot{\phi} = \\
-  &A \sin{\theta}\sin{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
- - &\sin{\theta} \dot{\theta} = \\
-  &A \cos{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+\dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}} &= 
+A^2 \left[ \mathbf{E}\cdot \mitbf{\omega} - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)\mitbf{\omega}\right]
+\cdot \left[ \mathbf{E}\cdot \mitbf{\omega} - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)\mitbf{\omega}\right] \\
+\dot{\mitbf{\omega}}^2  &= 
+A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)\cdot \left(\mathbf{E}\cdot \mitbf{\omega}\right)
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right) \left( \mitbf{\omega}\cdot \mathbf{E} \cdot \mitbf{\omega} \right) 
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right) \left( \mitbf{\omega}\cdot \mathbf{E} \cdot \mitbf{\omega} \right) 
+ + \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 \mitbf{\omega}\cdot \mitbf{\omega} 
+\right]\\
+\dot{\mitbf{\omega}}^2  &= 
+A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 
+\right]\\
 \end{aligned}
 \end{equation}
-Now multiply the top equation by $\cos{\phi}$ and the second equation by $\sin{\phi}$ and add them together to get
+Since this is a scalar equation, we can arbitrarily choose a coordinate system in which to evaluate it.
+It is most convenient to choose the principal axes of $\mathbf{E}$, where
+\begin{equation}
+\mathbf{E} = 
+\begin{bmatrix}
+\lambda_1 & 0 & 0 \\
+0 & \lambda_2 & 0 \\
+0 & 0 & \lambda_3 \\
+\end{bmatrix}
+\end{equation}
+\begin{equation}
+\mitbf{\omega} = 
+\begin{bmatrix}
+\sin{\theta} \cos{\phi} \\
+\sin{\theta} \sin{\phi} \\
+\cos{\theta} \\
+\end{bmatrix}
+\end{equation}
+
+As a warmup, consider the case where $\phi = 0$, then  
+\begin{equation}
+\mitbf{\omega} = 
+\begin{bmatrix}
+\sin{\theta}  \\
+0 \\
+\cos{\theta} \\
+\end{bmatrix}
+\end{equation}
+We can evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
+\begin{equation}
+\mathbf{E}\cdot \mitbf{\omega} = 
+\begin{bmatrix}
+\lambda_1 \sin{\theta} \\
+0 \\
+\lambda_3 \cos{\theta}
+\end{bmatrix}
+\end{equation}
+
+\begin{equation}
+\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
+\lambda_1 \sin^2{\theta} + \lambda_3 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
+\lambda_1^2 \sin^4{\theta} + \lambda_3^2 \cos^4{\theta} + 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta}
+\end{equation}
+Therefore, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
 \begin{equation}
 \begin{aligned}
- & \cos{\theta}\cos^2{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \cos{\phi} \dot{\phi} + \cos{\theta}\sin^2{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \sin{\phi} \dot{\phi} = \\
-  &A \sin{\theta}\cos^2{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] + \\
-  &A \sin{\theta}\sin^2{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
- & \cos{\theta}\dot{\theta} =  
-   A \sin{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
-  &A \sin{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
+ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
+ &= \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
+ - \lambda_1^2 \sin^4{\theta} - \lambda_3^2 \cos^4{\theta} - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
+ &= \lambda_1^2 \sin^2{\theta} (1 - \sin^2{\theta}) + \lambda_3^2 \cos^2{\theta} (1 - \cos^2{\theta})
+  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
+ &= \lambda_1^2 \sin^2{\theta} \cos^2{\theta} + \lambda_3^2 \cos^2{\theta} \sin^2{\theta}
+  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
+ &= (\lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \sin^2{\theta} \cos^2{\theta} \\
+ &= \frac{1}{2} (\lambda_3 - \lambda_1)^2 \sin{2 \theta} \\
 \end{aligned}
 \end{equation}
-Now multiply the above by $\cos{\theta}$ and the third equation by $-\sin{\theta}$ and add them to get
+
+Now consider the more general case whhere $\phi \ne 0$.
+Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
+\begin{equation}
+\mathbf{E}\cdot \mitbf{\omega} = 
+\begin{bmatrix}
+\lambda_1 \sin{\theta} \cos{\phi}\\
+\lambda_2 \sin{\theta} \sin{\phi} \\
+\lambda_3 \cos{\theta}
+\end{bmatrix}
+\end{equation}
+
+\begin{equation}
+\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
+\lambda_1 \sin^2{\theta} \cos^2{\phi} + \lambda_2 \sin^2{\theta} \sin^2{\phi} + \lambda_3 \cos^2{\theta}
+\end{equation}
+
 \begin{equation}
 \begin{aligned}
- & \cos^2{\theta}\dot{\theta} + \sin^2{\theta} \dot{\theta} =  
-   A \sin{\theta} \cos{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
-  &A \sin{\theta} \cos{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] - \\
-  &A \cos{\theta} \sin{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
+\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
+&\lambda_1^2 \sin^4{\theta} \cos^4{\phi} + 
+\lambda_2^2 \sin^4{\theta} \sin^4{\phi} +
+\lambda_3^2 \cos^4{\theta} \\
+&+ 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
+&+ 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
+&+ 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
 \end{aligned}
 \end{equation}
-After a lot of cancellation, we find
+Again, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
 \begin{equation}
 \begin{aligned}
- \dot{\theta} &= A \sin{\theta} \cos{\theta} \left[ - \lambda_3 + \lambda_1 \cos^2{\phi} + \lambda_2 \sin^2{\phi} \right] \\
-              &= - A \sin{\theta} \cos{\theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
-              &= -\frac{A}{2} \sin{2 \theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
+\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
+ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
+&=
+\lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta} \\
+&-\lambda_1^2 \sin^4{\theta} \cos^4{\phi} - 
+\lambda_2^2 \sin^4{\theta} \sin^4{\phi} -
+\lambda_3^2 \cos^4{\theta} \\
+&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
+&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
+&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
+&=
+\sin^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} \right] \\
+&- \sin^2{\theta} \left[ \lambda_1^2 \sin^2{\theta} \cos^2{\phi}(1-\sin^2{\phi}) + \lambda_2^2 \sin^2{\theta} \sin^2{\phi}(1-\cos^2{\phi}) \right] \\
+&+ \lambda_3^2 \cos^2{\theta}(1-\cos^2{\theta}) \\
+&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
+&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
+&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
+&=
+\sin^2{\theta} \left[\lambda_1^2 \cos^2{\phi} (1 - \sin^2{\theta}) + \lambda_2^2 \sin^2{\phi} (1-\sin^2{\theta}) \right] \\
+&+ \sin^2{\theta} \left[ \lambda_1^2 \sin^2{\theta} \cos^2{\phi}\sin^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi}\cos^2{\phi} \right] \\
+&+ \lambda_3^2 \cos^2{\theta}\sin^2{\theta} \\
+&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
+&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
+&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
+&=
+\sin^2{\theta}\cos^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} \right] \\
+&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left[ \lambda_1^2 + \lambda_2^2 -2 \lambda_1\lambda_2 \right] \\
+&+ \lambda_3^2 \cos^2{\theta}\sin^2{\theta} \\
+&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
+&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
+&=
+\sin^2{\theta}\cos^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} + \lambda_3^2 - 2 \lambda_1 \lambda_3 \cos^2{\phi} - 2 \lambda_2 \lambda_3 \sin^2{\phi} \right] \\
+&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
+&=
+\sin^2{\theta}\cos^2{\theta} \left[( \lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \cos^2{\phi} + (\lambda_2^2 + \lambda_3^2 - 2 \lambda_2 \lambda_3 ) \sin^2{\phi} \right] \\
+&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
+&=
+\sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] \\
+&+ \sin^2{\theta} \cos^2{\theta} \tan^2{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
+&=
+\sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi}
++ \tan^2{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \right] \\
 \end{aligned}
 \end{equation}
-Likewise we may multiply the first equation by $-\sin{\phi}$ and the second equation by $\cos{\phi}$ and add them to get
+
+At long last, we get
 \begin{equation}
 \begin{aligned}
- & \sin{\theta} \dot{\phi} = A \sin{\theta}\cos{\phi}\sin{\phi} \left( \lambda_2 - \lambda_1 \right ) \\
-& \dot{\phi} = \frac{A}{2} \sin{2 \phi} (\lambda_2-\lambda_1)
+\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
+\frac{1}{4}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi}
++ \frac{1}{4}\tan^2{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \right] \\
 \end{aligned}
@@
 \begin{aligned}
- - \nabla P^* + \; \nabla^2 \mathbf{v} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0
+ - \nabla P^* + \; \nabla^2 \mathbf{u} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0
 \end{aligned}

### 2016-01-21 tpw_rate.tex
Work on corrected scaling

@@
 Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
-If we enter the coordinate system of the convective moment of inertia $\mathbf{E}$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
+The rate of polar wander is the angular rate of change of the spin axis $\left| \dot{\mitbf{\omega}} \right| = \dot{\Theta}$ in a body-fixed geographic reference frame.
+This can be calculated by dotting Equation~\eqref{eq:liouville_ode} with itself:
+\begin{equation}
+\dot{\Theta}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right]
+\label{eq:polar_wander_rate_tensor}
+\end{equation}
+
+\begin{figure*}
+\centering
+\includegraphics[width=0.9\textwidth]{figures/reference_frames.pdf}
+\caption{ }
+\label{fig:eigengap}
+\end{figure*}
+
+The polar wander rate $\dot{\Theta}$ is measured in the body-fixed geographic frame, 
+but the physics of the right-hand-side of Equation~\eqref{eq:polar_wander_rate_tensor} 
+is more naturally expressed in the reference frame of the principal axes of $\mathbf{E}$,
+which is, in general, rotating with respect to the geographic frame.
+However, the right-hand-side is scalar-valued, and as such is invariant to rotations.
+Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
+with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of 
+$\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$.
+Plugging these in, and after some tedious algebra, we find
 \ifdetail
@@
 \sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] \\
-&+ \sin^2{\theta} \cos^2{\theta} \tan^2{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
+&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
 &=
-\sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi}
-+ \tan^2{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \right] \\
+\sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right]
++ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
 \end{aligned}
@@
 \begin{aligned}
-\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
-\frac{1}{4}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi}
-+ \frac{1}{4}\tan^2{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \right] \\
+\frac{\dot{\mitbf{\omega}}^2}{A^2}  = 
+&\frac{1}{4}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
+&\frac{1}{4}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \\
 \end{aligned}
@@
 \fi
-\begin{equation}
-\begin{aligned}
-\dot{\theta} &= - \frac{1}{(C-A)T_1} \sin{2 \theta} \left[ (\lambda_3-\lambda_1) \cos^2{\phi} + (\lambda_3-\lambda_2) \sin^2{\phi} \right] \\
-\dot{\phi} &= \;\; \frac{1}{(C-A)T_1} \sin{2 \phi} \; (\lambda_2 - \lambda_1)
-\end{aligned}
-\label{eq:milankovitch}
-\end{equation}
-
-Without loss of generality we may choose the coordinate system such that $\phi=0$:
-\begin{equation}
-\dot{\theta} = - \frac{1}{(C-A)T_1} (\lambda_3-\lambda_1) \sin{2 \theta}
-\label{eq:simple_milankovitch}
-\end{equation}
-These equations are a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.  
-Written this way, it is clear that the important quantities are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.
+
+\begin{equation}
+\begin{aligned}
+\dot{\Theta}^2 = 
+&\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
+&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \\
+\label{eq:polar_wander_rate}
+\end{aligned}
+\end{equation}
+This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.
+Two special cases of this equation are of note. First, if $\theta=\phi=45^\circ$, then it can be expressed in terms of
+the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator ($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
+\begin{equation}
+\begin{aligned}
+\dot{\Theta}^2
+&= \frac{1}{4(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
+&= \frac{E_{\mathrm{II}}}{4(C-A)^2 T_1^2} \\
+\end{aligned}
+\end{equation}
+Second, if the rotation vector $\mitbf{\omega}$ lies in the plane of the $\lambda_1$ axis ($\phi = 0$), then 
+Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
+\begin{equation}
+\vert \dot{\Theta} \vert = 
+\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) 
+\label{eq:polar_wander_rate_simplified}
+\end{equation}
+
+Written this way, it is clear that the important quantities for estimating $\dot{\Theta}$ are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.
 They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principal axis of the convective moment.
@@
 As such, some extra care must be taken to ensure that the equations we consider are self-consistent. 
+Furthermore, since our goal is to establish a scaling for TPW rate, we must identify a minimal set of nondimensional numbers which describe its physics.
 For simplicty we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.

### 2016-01-21 tpw_rate.tex
Change title
Begin to rework some scaling.

@@
 \usepackage{graphicx}
+\usepackage{color,soul}
 
 \title[TPW Rates]
-  {Rates of true polar wander in convecting planets}
+  {Scaling rates of true polar wander in convecting planets and moons}
 \author[I. Rose and B. Buffett]
@@
 \frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) 
-\label{eq:polar_wander_rate_simplified}
+\label{eq:simple_milankovitch}
 \end{equation}
@@
 
-\subsection{An estimate of the mismatch $\theta$}
+\subsection{Estimating the mismatch $\theta$}
 \label{sec:theta}
@@
 Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
+
+There are two processes at work which control the evolution of $\theta$, its growth through perturbations in the convective moment and 
+its decay via relaxation of the pole towards the maximum moment of inertia.
+We can interrogate these by constructing Equation~\eqref{eq:liouville_ode} in the reference frame
+of the principal axes of the convective moment $\mathbf{E}$.
+This reference frame rotates with respect to the geographic frame, which we can
+describe by introducing a rotation vector $\mathbf{\Psi}$.
+We can then write the rate of change of $\mitbf{\omega}$ in the rotating frame (denoted by primes)
+\begin{equation}
+\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime
+\end{equation}
+Equation~\eqref{eq:liouville_ode} then becomes
+\begin{equation}
+ \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime
+\label{eq:liouville_ode_new_frame}
+\end{equation}
+
+\hl{This equation can be rewritten in terms of $\dot{\theta}$. It is useful to make the simplifying assumption that $\phi=0$.}
+\hl{This then becomes an equation for the time evolution of $\theta$, which has a decay component due to polar wander, and a growth component due to $\Psi$}
+\hl{What this buys us is a single equation that has both decay (for which we have already estimated the timescale) and growth, which comes next.}
+\hl{The following needs to be rewritten to use the notation of $\Psi$ instead of $\theta$.}
 

### 2016-02-22 tpw_rate.tex
Change Fr to m

@@
 In particular, the two most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
-This second nondimensional number does not have a uniformly agreed-upon name, but has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, which we adopt here.
+This second nondimensional number does not have a uniformly agreed-upon name: it has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence},
+and in the geodesy community has commonly been termed $m$ \citep[e.g.][]{nakiboglu1982hydrostatic, chambat2010flattening}, which we adopt here.
 Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
@@
 \begin{aligned}
- - \nabla P^* + \; \nabla^2 \mathbf{u} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0
+ - \nabla P^* + \; \nabla^2 \mathbf{u} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra} \; m \; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0
 \end{aligned}
@@
 \begin{aligned}
- - & \mathrm{Ra \; Fr}\; T \; \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} = - \nabla P^* \\ 
+ - & \mathrm{Ra} \; m \; T \; \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} = - \nabla P^* \\ 
 &+ \; \nabla \cdot \left( \eta \epsilon (\mathbf{u}) \right) - \mathrm{Ra} \; T \; \mathbf{g}
@@
 \begin{tabular}{@{}lcccc}
-Symbol &  Number & Definition & Approximate value \\
+Symbol &  Name & Definition & Approximate value \\
 \hline
 Ra & Rayleigh &  $\rho_0 g_0 \alpha \Delta T R^3/\eta \kappa$ & $10^7$\\
-Fr & Froude & $\Omega_0^2 R/g_0$ & $10^{-3}$ \\
+$m$ & Froude & $\Omega_0^2 R/g_0 = \Omega_0^2 R^3/GM$ & $10^{-3}$ \\
 A & aspect ratio & $R_i/R$ & $0.54$ \\
@@
 This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
-The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep{munk1960rotation}:
-\begin{equation}
-(C-A) \sim I_0 \mathrm{Fr}
+The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces($m$ \citep{munk1960rotation}:
+\begin{equation}
+(C-A) \sim I_0 m
 \end{equation}
@@
 \begin{equation}
-\frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma \mathrm{Fr}} \Lambda_{31} \sin{2 \theta}
+\frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma m} \Lambda_{31} \sin{2 \theta}
 \label{eq:scaled_rotation}
@@
 
-The quantites $\Lambda_{ij}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
-We consider relatively slowly rotating bodies here ($\mathrm{Fr} \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
-In this regime, therefore, we neglect the dependence on $\mathrm{Fr}$, and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra})$ and $\theta(\mathrm{Ra})$.
+The quantites $\Lambda_{ij}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $m$, and $\Gamma$.  
+We consider relatively slowly rotating bodies here ($m \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
+In this regime, therefore, we neglect the dependence on $m$, and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra})$ and $\theta(\mathrm{Ra})$.
 
@@
 
-The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
+The preceding results clarify the complex relationship between the Rayleigh number, $m$, and the rates of TPW for a convecting planet.
 As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principal axes of the convective moment of inertia. 
@@
 \begin{equation}
-\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
+\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{m} \sin{2 \theta}
 \label{eq:simplest_milankovitch}
@@
 \begin{equation}
-\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{\mathrm{Fr}} \sin{2 \theta}
+\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}
 \label{eq:simplest_milankovitch_defer_scaling}
@@
 
-We then suggest that the most important parameters are the Froude number, which acts as the brakes on the system, and the mismatch angle $\theta(\mathrm{Ra})$.
+We then suggest that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta(\mathrm{Ra})$.
 Whereas the rest of the expression has a weak dependence on $\mathrm{Ra}$, $\theta$ is expected to be strongly dependent on it.
@@
 
-The most important parameters are the Rayleigh number and the Froude number.
-The Froude number acts as a damper to TPW.
+The most important parameters are the Rayleigh number and the $m$, which acts as a damper to TPW.
 The dependence on the Rayleigh number is more complicated, since it is a control on both the forcing 

### 2016-02-22 tpw_rate.tex
Address some comments

@@
 
-For mantle convection problems the moment of inertia tensor is commonly separated into three parts \citep{ricard1993polar}:
+For mantle convection problems the moment of inertia tensor is commonly separated into three parts \citep{sabadini1981pleistocene, spada1992excitation}:
 \begin{equation}
@@
 where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature
-(see Table~\ref{tab:parameters} for definitions of the other parameters).
+(he remaining parameters are defined in Table~\ref{tab:parameters}).
+The gravity vector $\mathbf{g}$ is given by $\mathbf{g} = -\nabla V$, where the 
+gravitational potential $V$ comes from solving Poisson's equation
+\begin{equation}
+\nabla^2 V = 4 \pi G \rho
+\end{equation}.
+For the purposes of scaling, we assume that gravity is the approximately constant
+vector $g_0 \mathbf{\hat{r}}$ (as is the case for Earth's mantle).
 In addition we use the simple equation of state
@@
 $R$ & outer radius \\
+$G$ & gravitational constant \\
+$V$ & gravitational potential
+$M$ & mass of the planet \\
 $\Omega_0$ & reference rotation rate \\

### 2016-02-22 tpw_rate.tex
Expand text on infinite time limit of love number

@@
 
-The infinite-time limit of Equation~\eqref{eq:viscoelastic_deformation} allows us to calculate the fluid limit of the Love number $k_f$:
+The infinite-time limit of Equation~\eqref{eq:viscoelastic_deformation} for a 
+constant rotation vector around the $z$-axis implies
+\begin{equation}
+\begin{aligned}
+J_{zz} = C = &\frac{2}{3} \frac{k_f a^5 \Omega^2}{3 G} \\
+J_{xx} = J_{yy} = A = -&\frac{1}{3} \frac{k_f a^5 \Omega^2}{3 G} \\
+\end{aligned}
+\end{equation}
+where $C$ and $A$ are the polar and equatorial moments of inertia, respectively, 
+and $k_f$ is the fluid limit of $k$. 
+We can solve for $k_f$ in terms of $C-A$:
 \begin{equation}
@@
 \end{equation}
-where $C$ and $A$ are the polar and equatorial moments of inertia, respectively.
-The infinite-time limit, however, does not allow for any disequilibrium between $J_{ij}$ and $E_{ij}$.
+This fluid-limit representation of the deformation does not allow for any disequilibrium between $J_{ij}$ and $E_{ij}$,
+and so cannot by itself capture the physics of TPW.
 

### 2016-02-22 tpw_rate.tex
Format fix

@@
 $G$ & gravitational constant \\
-$V$ & gravitational potential
+$V$ & gravitational potential \\
 $M$ & mass of the planet \\

### 2016-02-23 tpw_rate.tex
Address some comments

@@
 \end{equation}
-known as Euler's equations, where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity.
+where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity.
 On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
-In this case Equation~\eqref{eq:euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
+In this case Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
 For a slowly convecting fluid such as a planetary mantle the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
@@
 of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
-Here we are interested in contributions from mantle convection, so we neglect other processes.
+Here we are interested in contributions from mantle convection, so we restrict our attention to these processes.
 
@@
 \end{equation}
-Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
+Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure
+(in the latter case the boundary conditions on $P$ must be modified).
 Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires three nondimensional numbers to characterize it
@@
 At high Rayleigh number we expect the shortest lengthscales to be limited by the effects of thermal diffusion,
-which tends to wipe out thermal heterogeneity at small scales.  Consequently, there will be no power in modes
+which tends to wipe out thermal heterogeneity at small scales.  Consequently, there will be little power in modes
 with shorter lengthscales than that allowed by diffusion. 
@@
 \end{equation}
-The total number of modes that are accessible to the system are therefore
+Of course, convective mixing can produce smaller scales, but the power in these scales
+is greatly reduced by diffusion.
+Thus total number of modes that are accessible to the system are
 
@@
 For the 2D simulations there is a reduced dimensionality when calculating the number of modes,
-so $N_\text{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\text{degree-two}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
+so $N_\text{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\text{degree-two}} \sim \mathrm{Ra}^{-2/3}$, 
+which is shown as a dashed line in Figure~\ref{fig:eigengap}.
 This result has a simple interpretation.
@@
 describe by introducing a rotation vector $\mathbf{\Psi}$.
-We can then write the rate of change of $\mitbf{\omega}$ in the rotating frame (denoted by primes)
+We can then write the rate of change of $\mitbf{\omega}$ in the principal axis frame (denoted by primes)
 \begin{equation}

### 2016-02-23 figures/reference_frames.py
Edits to reference frame

@@ -75,7 +75,7 @@ def plot_rotation_arrow( axes, arrow, ccw=True, frac = 0.75, *args, **kwargs):
     rotated_points = np.dot( rotation_matrix, points )
 
     # plot the points
-    dline = 40
+    dline = 10
     line = axes.plot(rotated_points[0,:-dline], rotated_points[1,:-dline], rotated_points[2,:-dline], lw=1, color='k')
     # add the arrowhead
     if ccw is True:
@@ -159,14 +159,21 @@ ax.plot( guideline[0,:], guideline[1,:], guideline[2,:], 'k--', alpha=alpha_val)
 # plot relative rotation
 psi_arrow = Arrow3D([o[0], psi[0]], [o[1], psi[1]], [o[2], psi[2]], **arrow_prop_dict)
 ax.add_artist(psi_arrow)
-plot_rotation_arrow(ax, psi_arrow)
+plot_rotation_arrow(ax, psi_arrow, ccw=False)
 # add helper lines for psi
 span = np.linspace(0, 1., 100)
 guideline = np.array([np.cos(beta)*np.sin(alpha)*np.ones_like(span), np.sin(beta)*np.sin(alpha)*np.ones_like(span), np.cos(alpha)*span])
 ax.plot( guideline[0,:], guideline[1,:], guideline[2,:], 'k--', alpha=alpha_val)
 guideline = np.array([np.cos(beta)*np.sin(alpha)*(span), np.sin(beta)*np.sin(alpha)*(span), np.zeros_like(span)])
 ax.plot( guideline[0,:], guideline[1,:], guideline[2,:], 'k--', alpha=alpha_val)
-guideline = np.array([omega * s + psi * (1.-s) for s in span] )
+
+tmpvec = np.array([ np.cos(beta), np.sin(beta), 0.])
+guideline = np.array([tmpvec * s + psi * (1.-s) for s in span] )
+for i,p in enumerate(guideline):
+  guideline[i] = 0.45*p/norm(p)
+ax.plot( guideline[:,0], guideline[:,1], guideline[:,2], 'k')
+tmpvec2 = np.array([ 1., 0., 0.])
+guideline = np.array([tmpvec * s + tmpvec2 * (1.-s) for s in span] )
 for i,p in enumerate(guideline):
[…]

### 2016-02-23 tpw_rate.tex
Add acknowledgement
Flesh out scaling and rework reference frame stuff

@@
 \includegraphics[width=0.9\textwidth]{figures/reference_frames.pdf}
-\caption{ }
-\label{fig:eigengap}
+\caption{Relevant vectors and angles for the TPW analysis.
+The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively.
+The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $z$-axis.
+For simplicity, we assume the longitude $\psi$ of the rotation axis is zero.
+True polar wander in the geographic reference frame moves the rotation axis towards the $z$-axis, shown by $\dot{\mitbf{\omega}}$.
+However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$.
+This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$, with colatitude $\alpha$ and longitude $\beta$.
+The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$ frame, shown as $\dot{\mitbf{\omega}}^\prime$.
+Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Section~\ref{sec:theta}}
+\label{fig:reference_frames}
 \end{figure*}
@@
 \begin{equation}
-\nabla^2 V = 4 \pi G \rho
-\end{equation}.
+\nabla^2 V = 4 \pi G \rho.
+\end{equation}
 For the purposes of scaling, we assume that gravity is the approximately constant
@@
 As discussed in Section \ref{sec:convective_moment}, these internal density loads may be convolved with their surface responses to get the convective moment of inertia $E_{ij}$.
-The surface response is a function of the viscosity structure of the planet and the wavelength of the interal load, but the multiplicative factor $(1+k^L_f)$ is generally an order-one parameter, so we neglect it for this scaling.
+The surface response is a function of the viscosity structure of the planet and the wavelength and depth of the interal load.
+The factor $(1+k^L_f)$ is generally an order-one parameter 
+(strictly speaking, dynamic compensation usually makes it less than one, \citep[e.g.][]{richards1984geoid}). 
+For the purposes of scaling it is reasonable to neglect this multiplicative factor.
 
@@
 \end{equation}
-
-\hl{This equation can be rewritten in terms of $\dot{\theta}$. It is useful to make the simplifying assumption that $\phi=0$.}
-\hl{This then becomes an equation for the time evolution of $\theta$, which has a decay component due to polar wander, and a growth component due to $\Psi$}
-\hl{What this buys us is a single equation that has both decay (for which we have already estimated the timescale) and growth, which comes next.}
-\hl{The following needs to be rewritten to use the notation of $\Psi$ instead of $\theta$.}
-
-An estimate for $\theta$ must concern the stability of the convective moment of inertia. 
+The advantage of writing this in the frame of the principal axes of $\mathbf{E}$ is that the entire equation
+may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
+As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that $\phi=0$
+(for nonzero $\phi$ the equations become more complex, but the basic physics is unchanged).
+The direction of $\Psi$ is given by colatitude $\alpha$ and longitude $\beta$, as shown in Figure~\ref{fig:reference_frames}.
+
+We may solve Equation~\eqref{eq:liouville_ode_new_frame} for $\dot{\theta}$ in terms of $\theta$, $\alpha$, and $\beta$:
+\begin{equation}
+\dot{\theta} = 
+-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\alpha \sin\beta
+\label{eq:theta_competition}
+\end{equation}
+Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
+The first term acts to reduce $\theta$ via TPW, the second term can increase (or decrease) it arbitrarily via
+relative motion between the geographic frame and the principal axes of the $\mathbf{E}$ frame.
+
+Note that comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
+then the rate of TPW $\vert \dot{\Theta} \vert$ is just $\vert \dot{\theta} \vert$.
+Conversely, when TPW is very slow $\theta$ is completely determined by $\mitbf{\Psi}$, and 
+$\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$ 
+(where we have neglected the orientation factor $\sin \alpha \sin \beta$ for the purposes of scaling).
+In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 
+at least occasionally, be larger than its decay.
+
+An estimate for $\vert \mitbf{\Psi} \vert$ must concern the stability of the convective moment of inertia. 
 Convection is continuously redistributing mass throughout the mantle, which will perturb the convective moment of inertia.  
-If the convective moment is stable to small perturbations, then $\theta$ should be small, but if it is not stable to perturbations, then $\theta$ could be up to $90^\circ$. 
+If the convective moment is stable to small perturbations, then $\vert \mitbf{\Psi} \vert$ should be small, 
+and $\theta$ will never be large.
+However, if $\mathbf{E}$ is not stable to perturbations, then $\theta$ could be up to $90^\circ$. 
 Given a random perturbation to that moment, we would like to give a bound on how big the perturbation to $\theta$ is. 
@@
 \begin{equation}
-\lVert \sin(2 \theta) \rVert \le \frac{ 2 \lVert \delta \rVert}{ \displaystyle \min_{i \neq j} \lVert \lambda_i - \lambda_j \rVert }
+\vert \sin(2 \theta) \vert \le \frac{ 2 \vert \delta \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }
 \label{eq:kahan}
@@
 
-We may arrive at a simple estimate of the growth rate of the angle $\theta$ by differentiating Equation~\eqref{eq:kahan}, tentatively holding $\lVert \lambda_i - \lambda_j \rVert$ fixed:
-
-\begin{equation}
-\lVert \dot{\theta} \rVert \le \frac{ \lVert \dot{\delta} \rVert}{ \displaystyle \min_{i \neq j} \lVert \lambda_i - \lambda_j \rVert }
+We may arrive at a simple estimate of the growth rate of the angle $\theta$ (and thus $\vert \mitbf{\Psi} \vert$) 
+by differentiating Equation~\eqref{eq:kahan}, tentatively holding $\vert \lambda_i - \lambda_j \vert$ fixed:
+
+\begin{equation}
+\vert \mitbf{\Psi} \vert \sim \vert \dot{\theta} \vert \le \frac{ \vert \dot{\delta} \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }
 \label{eq:grow_perturbation}
@@
 
-Furthermore, the quantity $(\lambda_i-\lambda_j)$ which appears in the denominator of these equations is precisely the same as the quantity which we estimated in the previous section to scale with $\sim \mathrm{Ra}^{-1}$.
+The quantity $(\lambda_i-\lambda_j)$ which appears in the denominator of these equations is precisely the same as the quantity which we estimated in the previous section to scale with $\sim \mathrm{Ra}^{-1}$.
 Therefore, as the Rayleigh number increases, the characteristic gap between the eigenvalues of the convective moment becomes smaller.  
@@
 \begin{equation}
-\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{m} \sin{2 \theta}
+\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{1}{m} \sin{2 \theta}
 \label{eq:simplest_milankovitch}
@@
 \begin{equation}
-\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}
+\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}
 \label{eq:simplest_milankovitch_defer_scaling}
@@
 \begin{equation}
-\max ( \dot{\theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 6^\circ / \mathrm{Myr}
+\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 6^\circ / \mathrm{Myr}
 \end{equation}
@@
 
-The most important parameters are the Rayleigh number and the $m$, which acts as a damper to TPW.
+The most important parameters are the Rayleigh number and $m$, which acts as a damper to TPW.
 The dependence on the Rayleigh number is more complicated, since it is a control on both the forcing 
@@
 \begin{acknowledgments}
-This work was supported by National Science Foundation grant EAR-1246670.
+This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for a thoughtful review.
 \end{acknowledgments}

### 2016-02-23 figures/reference_frames.py

@@ -196,9 +196,9 @@ text_options = {'horizontalalignment': 'center',
                 'fontsize': 16}
 
 # add labels for axes
-ax.text(1.1*x0[0],1.1*x0[1],1.1*x0[2],r'$e_1$', **text_options)
-ax.text(1.1*y0[0],1.1*y0[1],1.1*y0[2],r'$e_2$', **text_options)
-ax.text(1.1*z0[0],1.1*z0[1],1.1*z0[2],r'$e_3$', **text_options)
+ax.text(1.1*x0[0],1.1*x0[1],1.1*x0[2],r'$\mathbf{e}_1$', **text_options)
+ax.text(1.1*y0[0],1.1*y0[1],1.1*y0[2],r'$\mathbf{e}_2$', **text_options)
+ax.text(1.1*z0[0],1.1*z0[1],1.1*z0[2],r'$\mathbf{e}_3$', **text_options)
 ax.text(1.1*omega[0],1.1*omega[1],1.1*omega[2],r'$\omega$', **text_options)
 ax.text(1.1*psi[0],1.1*psi[1],1.1*psi[2],r'$\Psi$', **text_options)
 ax.text(omega[0] + omega_dot[0]*1.15, omega[1] + omega_dot[1]*1.15, omega[2]+omega_dot[2]*1.15, r'$\dot{\omega}$', **text_options)

### 2016-02-23 tpw_rate.tex
Text on second invariant

@@
 This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.
-Two special cases of this equation are of note. First, if $\theta=\phi=45^\circ$, then it can be expressed in terms of
+Two special cases of this equation are of note. First, if it is evaluated on the octahedral plane
+($\phi=45^\circ, \theta\approx 55^\circ$), then it can be expressed in terms of
 the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator ($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
@@
 \dot{\Theta}^2
-&= \frac{1}{4(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
-&= \frac{E_{\mathrm{II}}}{4(C-A)^2 T_1^2} \\
-\end{aligned}
-\end{equation}
-Second, if the rotation vector $\mitbf{\omega}$ lies in the plane of the $\lambda_1$ axis ($\phi = 0$), then 
-Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
+&= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
+&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2} \\
+\end{aligned}
+\end{equation}
+The second invariant of the stress deviator is commonly used in the theories of elasticity and plasticity
+as a convenient scalar approximation of the shear stress in a system.
+The second invariant of the moment of inertia deviator can be similarly useful as a scalar estimate of the ``rotational stress.''
+
+The second special case is if the rotation vector $\mitbf{\omega}$ lies in the plane of the $\lambda_1$ axis ($\phi = 0$).
+Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
 \begin{equation}
@@
 \end{equation}
-
-Written this way, it is clear that the important quantities for estimating $\dot{\Theta}$ are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.
+The maximum polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).
+
+From Equation~\eqref{eq:simple_milankovitch} it is clear that the important quantities for estimating $\dot{\Theta}$ are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.
 They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principal axis of the convective moment.

### 2016-03-02 tpw_rate.tex
Make equation referencing consistent

@@
 
-The term on the right-hand side of equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
+The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
 This, as well, may be parameterized using a viscoelastic Love number approach:
@@
 We are interested in the rate of true polar wander, which is the angular change of the pole
-per unit time. We can transform equation~\ref{eq:liouville_ode} into a scalar equation by dotting it
+per unit time. We can transform Equation~\eqref{eq:liouville_ode} into a scalar equation by dotting it
 with $\dot{\mitbf{\omega}}$, since we are interested in the magnitude of the rate, rather than
@@
 (a fourth one, defined by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations).
-Convenient choices for these numbers are listed in Table \ref{tab:nondim}, along with approximate Earth-like values for them.
+Convenient choices for these numbers are listed in Table~\ref{tab:nondim}, along with approximate Earth-like values for them.
 
@@
 
-By definition the hydrostatic reference state is a solution to~\eqref{eq:navier_stokes} where there is no flow:
+By definition the hydrostatic reference state is a solution to Equation~\eqref{eq:navier_stokes} where there is no flow:
 
@@
 \end{equation}
-Nondimensionalizing with the parameters in Table \ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
+Nondimensionalizing with the parameters in Table~\ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
 \begin{equation}
@@
 
-Splitting the volume into $V_H$ and $\Delta V$ and substituting~\eqref{eq:eos} into~\eqref{eq:quasistatic_angular_momentum}, we find
+Splitting the volume into $V_H$ and $\Delta V$ and substituting Equation~\eqref{eq:eos} into Equation~\eqref{eq:quasistatic_angular_momentum}, we find
 \begin{equation}
@@
 \end{equation}
-This equation may be identified with~\eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
-Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table \ref{tab:nondim}.
+This equation may be identified with Equation~\eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
+Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table~\ref{tab:nondim}.
 
@@
 We count the number of fundamental units for the problem, in this case mass (kg), length (m), time (s), and temperature (K), as well as the number of parameters in use for this problem.  
-These parameters are listed in Table \ref{tab:parameters}, and subtracting the ...
+These parameters are listed in Table~\ref{tab:parameters}, and subtracting the ...
 
@@
 where $V_S$ is the reference spherical volume of the mantle.
-As discussed in Section \ref{sec:convective_moment}, these internal density loads may be convolved with their surface responses to get the convective moment of inertia $E_{ij}$.
+As discussed in Section~\ref{sec:convective_moment}, these internal density loads may be convolved with their surface responses to get the convective moment of inertia $E_{ij}$.
 The surface response is a function of the viscosity structure of the planet and the wavelength and depth of the interal load.
@@
 
-For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix \ref{appendix:moments}).
+For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix~\ref{appendix:moments}).
 Therefore, an estimate of one allows an estimate of the other.
@@
 \end{equation}
-Inserting the temperature expansion~\eqref{eq:T_series} into~\eqref{eq:temperature_fluctuations} results in 
+Inserting the temperature expansion~\eqref{eq:T_series} into Equation~\eqref{eq:temperature_fluctuations} results in 
 a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
@@
 If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principal axes.
-This is illustrated schematically in Figure \ref{fig:perturb}
+This is illustrated schematically in Figure~\ref{fig:perturb}
 
@@
 Equation~\eqref{eq:grow_perturbation} and its relaxation through Equation~\eqref{eq:simple_milankovitch}.
-We may plug in the estimate for $\Lambda_{ij}$ into equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
+We may plug in the estimate for $\Lambda_{ij}$ into Equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
 \begin{equation}
@@
 where $\mathbf{1}$ is the identity matrix, and $\lambda_1$, $\lambda_2$, and $\lambda_3$ are the principal moments.
-From equation~\eqref{eq:milankovitch} we see that the important quantities are the differences between the 
+From Equation~\eqref{eq:milankovitch} we see that the important quantities are the differences between the 
 principal moments, $(\lambda_3-\lambda_1)$, $(\lambda_3-\lambda_2)$ and $(\lambda_2-\lambda_1)$.
@@
 picking out the degree-two part of the density field laterally, and low-order polynomials radially.
-When density is a function of temperature, we can insert the equation of state~\eqref{eq:eos} into 
+When density is a function of temperature, we can insert the equation of state, Equation~\eqref{eq:eos}, into 
 Equation~\eqref{eq:degree_two_moments} and integrate over a reference spherical volume $V_S$.

### 2016-03-02 tpw_rate.tex
Address some comments

@@
 This fluid-limit representation of the deformation does not allow for any disequilibrium between $J_{ij}$ and $E_{ij}$,
-and so cannot by itself capture the physics of TPW.
+so it does not permit TPW.
 
@@
 \end{equation}
-where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature
-(he remaining parameters are defined in Table~\ref{tab:parameters}).
+where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
 The gravity vector $\mathbf{g}$ is given by $\mathbf{g} = -\nabla V$, where the 
@@
 \end{equation}
+The remaining parameters are defined in Table~\ref{tab:parameters}.
 Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure
@@
 We consider relatively slowly rotating bodies here ($m \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
-In this regime, therefore, we neglect the dependence on $m$, and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra})$ and $\theta(\mathrm{Ra})$.
+In this regime, therefore, we neglect the dependence on $m$, and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra}, \Gamma)$ and $\theta(\mathrm{Ra}, \Gamma)$.
 
@@
 \end{equation}
-Of course, convective mixing can produce smaller scales, but the power in these scales
+Strictly speaking, convective mixing can produce smaller scales, but the power in these scales
 is greatly reduced by diffusion.
@@
 
-The mismatch between the current axis and the principal axis of the convective moment is perhaps the most important parameter in Equation~\eqref{eq:scaled_rotation}.  
+The mismatch between the current rotation axis and the principal axis of the convective moment is perhaps the most important parameter in Equation~\eqref{eq:scaled_rotation}.  
 Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.

### 2016-03-02 tpw_rate.tex
More edits

@@
 
+\subsection{A note on reference frames}
+\label{sec:reference_frames}
+
 \subsection{Rotational deformation}
@@
 An alternative to the Love number formulism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
-A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2015free}
+A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
 permits more general treatments of mantle rheology.
@@
 as a convenient scalar approximation of the shear stress in a system.
-The second invariant of the moment of inertia deviator can be similarly useful as a scalar estimate of the ``rotational stress.''
+The second invariant of the moment of inertia deviator can be similarly viewed as a scalar estimate of the ``rotational stress.''
 
@@
 Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
-We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$: 
+We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$ (see Appendix~\ref{appendix:moments} for more detail): 
 \begin{equation}
@@
 n_{\text{max}}, l_{\text{max}}, m_{\text{max}} \sim \frac{R}{d}
+\label{eq:max_wavenumber}
 \end{equation}
@@
 Any of a number of scaling laws can provide an estimate for the characteristic length scale of a convecting system which may depending on rheology, geometry, or density structure.
-The simplest, based on boundary layer theory \citep{turcotte1967finite}, finds $d/R \sim \mathrm{Ra}^{-1/3}$, thus furnishing us with an estimate of the power in the degree-two part of the field as a function of Rayleigh number:
+The simplest, based on boundary layer theory \citep{turcotte1967finite}, finds $d/R \sim \mathrm{Ra}^{-1/3}$.
+This scaling is roughly a measure of the diffusive lengthscale for the timescale of a convective overturn, consistent
+with the cutoff in Equation~\eqref{eq:max_wavenumber}.
+It thus furnishes us with an estimate of the power in 
+the degree-two part of the field as a function of Rayleigh number:
 
@@
 \Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }
+\label{eq:lambda_estimate}
 \end{equation}
@@
 of the principal axes of the convective moment $\mathbf{E}$.
-This reference frame rotates with respect to the geographic frame, which we can
-describe by introducing a rotation vector $\mathbf{\Psi}$.
+The $\mathbf{E}$-frame slowly rotates with respect to the geographic frame, which we can
+describe by introducing a rotation vector $\mathbf{\Psi}$ 
+(Here $\mitbf{\Psi}$ describes the rotation of the $\mathbf{E}$-frame in the body-fixed frame).
 We can then write the rate of change of $\mitbf{\omega}$ in the principal axis frame (denoted by primes)
@@
 \end{equation}
-The advantage of writing this in the frame of the principal axes of $\mathbf{E}$ is that the entire equation
+The advantage of expressing TPW in the frame of the principal axes of $\mathbf{E}$ is that the entire equation
 may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
@@
 Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
-The first term acts to reduce $\theta$ via TPW, the second term can increase (or decrease) it arbitrarily via
+The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it arbitrarily via
 relative motion between the geographic frame and the principal axes of the $\mathbf{E}$ frame.
-
-Note that comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
-then the rate of TPW $\vert \dot{\Theta} \vert$ is just $\vert \dot{\theta} \vert$.
-Conversely, when TPW is very slow $\theta$ is completely determined by $\mitbf{\Psi}$, and 
-$\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$ 
+Comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
+then the rate of TPW $\vert \dot{\Theta} \vert$ is identical to $\vert \dot{\theta} \vert$.
+Conversely, when the rate of TPW is very slow (i.e. $\vert \dot{\Theta} \vert \sim 0$), 
+then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$ 
+and $\theta$ is completely determined by $\mitbf{\Psi}$
 (where we have neglected the orientation factor $\sin \alpha \sin \beta$ for the purposes of scaling).
@@
 Equation~\eqref{eq:grow_perturbation} and its relaxation through Equation~\eqref{eq:simple_milankovitch}.
-We may plug in the estimate for $\Lambda_{ij}$ into Equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
+We may plug in the estimate for $\Lambda_{ij}$ (Equation~\eqref{eq:lambda_estimate}) into Equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
 \begin{equation}
@@
 Typical values for the time constant $T_1$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
-Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies)
-are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
+Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
+are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
 The largest question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
@@
 where $\mathbf{1}$ is the identity matrix, and $\lambda_1$, $\lambda_2$, and $\lambda_3$ are the principal moments.
-From Equation~\eqref{eq:milankovitch} we see that the important quantities are the differences between the 
+From Equation~\eqref{eq:polar_wander_rate} we see that the important quantities are the differences between the 
 principal moments, $(\lambda_3-\lambda_1)$, $(\lambda_3-\lambda_2)$ and $(\lambda_2-\lambda_1)$.
@@
 \fi
-Dividing by $I_0$ and nodimensionalizing the integrals results in a factor of $\Gamma = \alpha \Delta T$ and a normalized set of 
+Dividing Equation~\eqref{eq:degree_two_moments_of_temperature} by $I_0$ and nodimensionalizing 
+the integrals results in a factor of $\Gamma = \alpha \Delta T$ and a normalized set of 
 degree-two coefficients for the temperature field, which we abbreviate as $T_{\text{degree-two}}$:

### 2016-03-02 tpw_rate.tex
Punctuate equations
Fix label, add some sentences on linear momentum

@@
 \begin{equation}
-\mitbf{\Omega } (t)\times ( \mathbf{I}(t) \cdot \mitbf{\Omega }(t) ) = 0
+\mitbf{\Omega } (t)\times ( \mathbf{I}(t) \cdot \mitbf{\Omega }(t) ) = 0.
 \label{eq:quasistatic_liouville}
@@
 \begin{equation}
-\mathbf{\Omega} \times (\mathbf{J} \cdot \mitbf{\Omega}) = -\mitbf{\Omega} \times ( \mathbf{E} \cdot \mitbf{\Omega} )
+\mathbf{\Omega} \times (\mathbf{J} \cdot \mitbf{\Omega}) = -\mitbf{\Omega} \times ( \mathbf{E} \cdot \mitbf{\Omega} ).
 \label{eq:disequilibrium}
@@
 \begin{equation}
-k_f = \frac{3 G (C-A)}{\Omega^2 a^5}
+k_f = \frac{3 G (C-A)}{\Omega^2 a^5}.
 \label{eq:fluid_love}
@@
 J_{ij} = \frac{k_f a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right) -
- \frac{k_f a^5 T_1}{3G} \left(\dot{\Omega}_i \Omega_j + \Omega_i \dot{\Omega}_j - \frac{2}{3} \Omega_q \dot{\Omega}_q \delta_{ij} \right)
+ \frac{k_f a^5 T_1}{3G} \left(\dot{\Omega}_i \Omega_j + \Omega_i \dot{\Omega}_j - \frac{2}{3} \Omega_q \dot{\Omega}_q \delta_{ij} \right).
 \label{eq:rotational_deformation}
@@
 \begin{equation}
-E_{ij} = (1+k^L_f) C_{ij}
+E_{ij} = (1+k^L_f) C_{ij}.
 \end{equation}
@@
 \begin{equation}
-\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right)
+\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right).
 \label{eq:liouville}
@@
 \begin{equation}
- \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right]
+ \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right].
 \label{eq:liouville_ode}
@@
 \begin{equation}
-\dot{\Theta}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right]
+\dot{\Theta}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
 \label{eq:polar_wander_rate_tensor}
@@
 with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of 
-$\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$.
-Plugging these in, and after some tedious algebra, we find
+$\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ (see Figure~\ref{fig:reference_frames}).
+Plugging this description of $\mitbf{\omega}$ into Equation~\eqref{eq:polar_wander_rate_tensor},
+and after some tedious algebra, we find
 \ifdetail
@@
 A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2.
 \right]\\
@@
 \cos{\theta} \\
-\end{bmatrix}
+\end{bmatrix}.
 \end{equation}
@@
 &\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
-&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \\
+&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
 \label{eq:polar_wander_rate}
@@
 Two special cases of this equation are of note. First, if it is evaluated on the octahedral plane
-($\phi=45^\circ, \theta\approx 55^\circ$), then it can be expressed in terms of
-the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator ($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
+(the plane with direction cosines all $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
+then it can be expressed in terms of the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator 
+($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
 \begin{equation}
@@
 &= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
-&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2} \\
+&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2}. \\
 \end{aligned}
@@
 \vert \dot{\Theta} \vert = 
-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) 
+\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1).
 \label{eq:simple_milankovitch}
@@
 \begin{equation}
-\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right)
+\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right).
 \label{eq:eos}
@@
 \begin{aligned}
-- \nabla P_0 =  \rho_0 \mathbf{g} -  \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}
+- \nabla P_0 =  \rho_0 \mathbf{g} -  \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}.
 \label{eq:hydrostatic}
@@
 \begin{aligned}
- - \nabla P^* + \; \nabla^2 \mathbf{u} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra} \; m \; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0
+ - \nabla P^* + \; \nabla^2 \mathbf{u} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra} \; m \; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0.
 \end{aligned}
@@
 -\int_V \mathbf{r} \times \nabla P \,dV + \int_V \eta \mathbf{r} \times \nabla^2 \mathbf{u} \,dV - \int_V \rho \mathbf{r} \times \mathbf{g} \,dV 
-   + \int_V \rho \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}  \,dV &= 0 \\
+   + \int_V \rho \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}  \,dV &= 0. \\
 \end{aligned}
@@
 \begin{aligned}
- \mitbf{\Omega} \times \int_V \rho \mathbf{r} \times (\mitbf{\Omega}\times \mathbf{r}) \,dV &= 0 \\
+ \mitbf{\Omega} \times \int_V \rho \mathbf{r} \times (\mitbf{\Omega}\times \mathbf{r}) \,dV &= 0. \\
 \end{aligned}
@@
 
-Splitting the volume into $V_H$ and $\Delta V$ and substituting Equation~\eqref{eq:eos} into Equation~\eqref{eq:quasistatic_angular_momentum}, we find
+We now seek to evaluate Equation~\eqref{eq:quasistatic_angular_momentum} in the perturbed, convecting state.
+Hydrostatic balance (Equation~\eqref{eq:hydrostatic}) ensures that the integral over the reference shape $V_H$ vanishes when $\rho = \rho_0$.
+Nonzero contributions arise from perturbations in the density field or from perturbations in the shape.
+To make this dependence explicit, we split the shape into the reference volume $V_H$ and perturbations from it $\Delta V$,
+and use Equation~\eqref{eq:eos} to define density perturbations.
+Substituting this decomposition into Equation~\eqref{eq:quasistatic_angular_momentum} brings the integral into the form
 \begin{equation}
@@
 &\int_{\Delta V} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV + 
-\int_{\Delta V} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 0  \\
+\int_{\Delta V} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 0.  \\
 \end{aligned}
@@
 \int_{\Delta V} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 
--\int_{V_H} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV
+-\int_{V_H} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV.
 \end{aligned}
@@
 \begin{equation}
-\rho = 1 - \alpha \Delta T \;\;  T
+\rho = 1 - \alpha \Delta T \;\;  T.
 \end{equation}
@@
 \begin{equation}
- \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}= - \nabla P_0 + \rho_0 \mathbf{g}
+ \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}= - \nabla P_0 + \rho_0 \mathbf{g}.
 \end{equation}
@@
 \begin{equation}
-(C-A) \sim I_0 m
+(C-A) \sim I_0 m.
 \end{equation}
@@
 \begin{equation}
-T_1 \sim \frac{ \eta }{ \rho_0 g_0 R} = \frac{R^2}{\kappa} \frac{\Gamma}{\mathrm{Ra} }
+T_1 \sim \frac{ \eta }{ \rho_0 g_0 R} = \frac{R^2}{\kappa} \frac{\Gamma}{\mathrm{Ra} }.
 \end{equation}
@@
 \begin{equation}
-\frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma m} \Lambda_{31} \sin{2 \theta}
+\frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma m} \Lambda_{31} \sin{2 \theta}.
 \label{eq:scaled_rotation}
@@
 \includegraphics[width=0.9\textwidth]{figures/eigengap.pdf}
-\caption{ Left: Time series of the normalized difference between moments $\Lambda_{12}$ for convection in a 2D annulus at several different Rayleigh numbers.
+\caption{ Left: Time series of the normalized difference between moments $\Lambda_{21} = (\lambda_2 - \lambda_1)/I_0$ for convection in a 2D annulus at several different Rayleigh numbers.
 As the Rayleigh number increases, the average value of the relative moment decreases due to less low-degree coherence in the temperature structure.
-Right:  average value of $\Lambda_{12}$ for the different Rayleigh numbers.
+Right:  average value of $\Lambda_{21}$ for the different Rayleigh numbers.
 Also shown is a line with slope $\mathrm{Ra}^{-2/3}$, which is predicted from the scaling analysis (the exponent is $-2/3$ instead of $-1$ due to the reduced dimensionality of the simulations).}
@@
 \begin{equation} 
-T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi)
+T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi).
 \label{eq:T_series}
@@
 \begin{equation}
-\Lambda_{31} \sim \Gamma T_{\text{degree-two}}
+\Lambda_{31} \sim \Gamma T_{\text{degree-two}}.
 \end{equation}
@@
 \begin{equation}
-\max \left( \displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \right) = 1
+\max \left( \displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \right) = 1.
 \end{equation}
@@
 \begin{equation}
-n_{\text{max}}, l_{\text{max}}, m_{\text{max}} \sim \frac{R}{d}
+n_{\text{max}}, l_{\text{max}}, m_{\text{max}} \sim \frac{R}{d}.
 \label{eq:max_wavenumber}
@@
 \begin{equation}
-N_{\text{modes}} = n_{\text{max}} \times l_{\text{max}} \times m_{\text{max}} \sim \left( \frac{R}{d} \right)^{3}
+N_{\text{modes}} = n_{\text{max}} \times l_{\text{max}} \times m_{\text{max}} \sim \left( \frac{R}{d} \right)^{3}.
 \end{equation}
@@
 \begin{equation}
-T_{\text{degree-two}}(t) \sim \frac{1}{N_{\text{modes}}} \sim \left( \frac{d}{R}\right)^3
+T_{\text{degree-two}}(t) \sim \frac{1}{N_{\text{modes}}} \sim \left( \frac{d}{R}\right)^3.
 \end{equation}
@@
 \begin{equation}
-T_{\text{degree-two}}(t) \sim \mathrm{Ra}^{-1}
+T_{\text{degree-two}}(t) \sim \mathrm{Ra}^{-1}.
 \label{eq:degree_two_of_ra}
@@
 \begin{equation}
-\Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }
+\Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }.
 \label{eq:lambda_estimate}
@@
 \begin{equation}
-\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime
+\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
 \end{equation}
@@
 \begin{equation}
- \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime
+ \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
 \label{eq:liouville_ode_new_frame}
@@
 \dot{\theta} = 
--\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\alpha \sin\beta
+-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\alpha \sin\beta.
 \label{eq:theta_competition}
@@
 \begin{equation}
-\vert \sin(2 \theta) \vert \le \frac{ 2 \vert \delta \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }
+\vert \sin(2 \theta) \vert \le \frac{ 2 \vert \delta \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
 \label{eq:kahan}
@@
 \begin{equation}
-\vert \mitbf{\Psi} \vert \sim \vert \dot{\theta} \vert \le \frac{ \vert \dot{\delta} \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }
+\vert \mitbf{\Psi} \vert \sim \vert \dot{\theta} \vert \le \frac{ \vert \dot{\delta} \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
 \label{eq:grow_perturbation}
@@
 \begin{equation}
-\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{1}{m} \sin{2 \theta}
+\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{1}{m} \sin{2 \theta}.
 \label{eq:simplest_milankovitch}
@@
 \begin{equation}
-\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}
+\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
 \label{eq:simplest_milankovitch_defer_scaling}
@@
 Y_{20} &= \frac{1}{4} \sqrt{\frac{ 5}{\pi}} \frac{ 2 z^2 - x^2 - y^2}{r^2} \\ 
-Y_{22} &= \frac{1}{4} \sqrt{\frac{15}{\pi}} \frac{ x^2 - y^2}{r^2}
+Y_{22} &= \frac{1}{4} \sqrt{\frac{15}{\pi}} \frac{ x^2 - y^2}{r^2}.
 \end{aligned}
@@
 Y_{20}^* &= 2 z^2 - x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{5} } r^2 Y_{20} \\ 
-Y_{22}^* &= x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{15} } r^2 Y_{22}
+Y_{22}^* &= x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{15} } r^2 Y_{22}.
 \end{aligned}
@@
 (\lambda_3 - \lambda_1) &= \frac{1}{2} \int_V  \rho \left( Y_{22}^* - Y_{20}^* \right) dV \\
-(\lambda_3 - \lambda_2) &= -\frac{1}{2} \int_V \rho \left( Y_{22}^* + Y_{20}^*  \right) dV \\
+(\lambda_3 - \lambda_2) &= -\frac{1}{2} \int_V \rho \left( Y_{22}^* + Y_{20}^*  \right) dV. \\
 \end{aligned}
@@
 (\lambda_3 - \lambda_1) &= 2 \sqrt{ \frac{\pi}{15} } \int_V  \rho r^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV \\
-(\lambda_3 - \lambda_2) &= -2 \sqrt{ \frac{\pi}{15} } \int_V  \rho r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV \\
+(\lambda_3 - \lambda_2) &= -2 \sqrt{ \frac{\pi}{15} } \int_V  \rho r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV. \\
 \label{eq:degree_two_moments}
@@
 (\lambda_3 - \lambda_1) &= -2 \sqrt{ \frac{\pi}{15} } \alpha \rho_0 \int_{V_S} T r^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV \\
-(\lambda_3 - \lambda_2) &= 2 \sqrt{ \frac{\pi}{15} } \alpha \rho_0 \int_{V_S}  T r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV \\
+(\lambda_3 - \lambda_2) &= 2 \sqrt{ \frac{\pi}{15} } \alpha \rho_0 \int_{V_S}  T r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV. \\
 \label{eq:degree_two_moments_of_temperature}
@@
 \begin{equation}
-I_0 = \frac{2}{3} \int_{V_S} \rho_0 r^2 \,dV
+I_0 = \frac{2}{3} \int_{V_S} \rho_0 r^2 \,dV.
 \end{equation}
@@
 \begin{equation}
-\Lambda_{ij} \sim \Gamma T_{\text{degree-two}}
+\Lambda_{ij} \sim \Gamma T_{\text{degree-two}}.
 \end{equation}

### 2016-03-02 figures/eigengap.py

@@ -110,7 +110,7 @@ for i,f in enumerate(output_files):
 ax.set_xlim(0,4)
 ax.set_ylim(0, 8.0e-5)
 ax.set_xlabel(r'Time (Gyr)')
-ax.set_ylabel(r'$(\lambda_2-\lambda_1)/I_0$')
+ax.set_ylabel(r'$\Lambda_{21} = (\lambda_2-\lambda_1)/I_0$')
 ax.legend(loc="upper right")
 print np.polyfit(np.log(Rayleighs[3:]), np.log(amplitudes[3:]),1)
 

### 2016-03-03 figures/reference_frames.py
Fix latitude/colatitude error

@@ -168,7 +168,7 @@ guideline = np.array([np.cos(beta)*np.sin(alpha)*(span), np.sin(beta)*np.sin(alp
 ax.plot( guideline[0,:], guideline[1,:], guideline[2,:], 'k--', alpha=alpha_val)
 
 tmpvec = np.array([ np.cos(beta), np.sin(beta), 0.])
-guideline = np.array([tmpvec * s + psi * (1.-s) for s in span] )
+guideline = np.array([z0 * s + psi * (1.-s) for s in span] )
 for i,p in enumerate(guideline):
   guideline[i] = 0.45*p/norm(p)
 ax.plot( guideline[:,0], guideline[:,1], guideline[:,2], 'k')
@@ -207,7 +207,7 @@ ax.text(omega[0] + omega_dot_prime[0]*1.2, omega[1] + omega_dot_prime[1]*1.2, om
 # add text for angles
 p = 0.45*(omega + z0)/2.0
 ax.text( p[0], p[1], p[2], r'$\theta$', **text_options)
-p = 0.4* (tmpvec + psi) / 2.0
+p = 0.4* (z0 + psi) / 2.0
 ax.text( p[0], p[1], p[2], r'$\alpha$', **text_options)
 p = 0.6* (tmpvec + tmpvec2) / 2.0
 ax.text( p[0], p[1], p[2], r'$\beta$', **text_options)

### 2016-03-03 tpw_rate.tex
Work on reference frame text

@@
 \end{equation}
-where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity.
+where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity vector.
 On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
@@
 \label{sec:reference_frames}
+True polar wander can be described in different reference frames, and this choice is fundamentally an arbitrary one.
+However, certain aspects of the physics can be made much simpler by an appropriate choice of the reference frame.
+In our treatment of TPW, we will refer to three different reference frames:
+
+First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
+Second, there is the body-fixed (or geographic) frame, which corresponds roughly to a no-net-rotation or hotspot reference frame. 
+Treatments of gravitational or rotational deformation of a planet are most naturally expressed in the body-fixed frame,
+as described in Section~\ref{sec:rotational_deformation}.
+The relative rotation between the inertial and geographic reference frame is given by the planetary rotation vector $\mitbf{\Omega}$.
+Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
+which we will refer to as the ``$\mathbf{E}$-frame.'' 
+Redistribution of mantle mass anomalies due to convection changes the principal axes of $\mathbf{E}$, 
+rotating the $\mathbf{E}$-frame with respect to the geographic frame, which we will describe
+by a rotation vector $\mitbf{\Psi}$ (see Section~\ref{sec:theta}).
+
+These reference frames and vectors are illustrated in Figure~\ref{fig:reference_frames}.
 
@@
 
-The part of the moment of inertia due to the elastic rotational deformation is traditionally related to the degree-two part of the gravity field via MacCullagh's formula \citep{munk1960rotation}:
+The part of the moment of inertia due to the elastic rotational deformation in the body-fixed frame 
+is traditionally related to the degree-two part of the gravity field via MacCullagh's formula \citep{munk1960rotation}:
 \begin{equation}
@@
 Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
-The rate of polar wander is the angular rate of change of the spin axis $\left| \dot{\mitbf{\omega}} \right| = \dot{\Theta}$ in a body-fixed geographic reference frame.
+The rate of polar wander is the angular rate of change of the spin axis $\left| \dot{\mitbf{\omega}} \right| = \dot{\Theta}$ in the geographic reference frame.
 This can be calculated by dotting Equation~\eqref{eq:liouville_ode} with itself:
@@
 The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively.
-The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $z$-axis.
-For simplicity, we assume the longitude $\psi$ of the rotation axis is zero.
-True polar wander in the geographic reference frame moves the rotation axis towards the $z$-axis, shown by $\dot{\mitbf{\omega}}$.
-However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$.
-This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$, with colatitude $\alpha$ and longitude $\beta$.
-The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$ frame, shown as $\dot{\mitbf{\omega}}^\prime$.
-Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Section~\ref{sec:theta}}
+Since the choice of geographic axes is arbitrary, we assume that at this instant the $\mathbf{E}$-frame and the geographic frames are colocated (though at future times they will not be).
+The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $\mathbf{e}_3$-axis.
+For illustration, we assume the longitude $\psi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane.
+True polar wander in the geographic reference frame moves the rotation axis towards the $\mathbf{e}_3$-axis, shown by $\dot{\mitbf{\omega}}$.
+However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$-frame.
+This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$ in the geographic frame, which is defined by colatitude $\alpha$ and longitude $\beta$.
+The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$-frame.
+The motion of $\mitbf{\omega}$ in the $\mathbf{E}$-frame is shown by $\dot{\mitbf{\omega}}^\prime$.
+Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Sections~\ref{sec:reference_frames} and \ref{sec:theta}}
 \label{fig:reference_frames}
@@
 Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
-It would be straightforward to include them, and they do not affect the overall treatment of this scaling.
+It would be straightforward to include them, but they do not affect the overall treatment of this scaling.
 
@@
 its decay via relaxation of the pole towards the maximum moment of inertia.
-We can interrogate these by constructing Equation~\eqref{eq:liouville_ode} in the reference frame
-of the principal axes of the convective moment $\mathbf{E}$.
-The $\mathbf{E}$-frame slowly rotates with respect to the geographic frame, which we can
-describe by introducing a rotation vector $\mathbf{\Psi}$ 
-(Here $\mitbf{\Psi}$ describes the rotation of the $\mathbf{E}$-frame in the body-fixed frame).
-We can then write the rate of change of $\mitbf{\omega}$ in the principal axis frame (denoted by primes)
+We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
+from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
+the rotation of the body-fixed frame relative to inertial space, as shown in Equation~\eqref{eq:euler})
+The $\mathbf{E}$-frame slowly rotates with respect to the geographic frame, described by the rotation vector $\mathbf{\Psi}$ 
+(here $\mitbf{\Psi}$ defines the rotation of the $\mathbf{E}$-frame in the geographic frame, see Figure~\ref{fig:reference_frames}).
+We can then write the rate of change of $\mitbf{\omega}$ in the $\mathbf{E}$-frame (denoted by primes)
 \begin{equation}
@@
 \end{equation}
-The advantage of expressing TPW in the frame of the principal axes of $\mathbf{E}$ is that the entire equation
+The advantage of expressing TPW in the $\mathbf{E}$-frame is that the entire equation
 may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
@@
 The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it arbitrarily via
-relative motion between the geographic frame and the principal axes of the $\mathbf{E}$ frame.
+relative motion between the geographic frame and the principal axes of the $\mathbf{E}$-frame.
 Comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,

### 2016-03-08 tpw_rate.tex
Small rephrasing
Address comments

@@
 First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
-Second, there is the body-fixed (or geographic) frame, which corresponds roughly to a no-net-rotation or hotspot reference frame. 
+Second, there is the body-fixed (or geographic) frame.
+By definition, the rotaiton of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
+A terrestrial no-net-rotation or hotspot reference frame are common choices for the body-fixed frame.
 Treatments of gravitational or rotational deformation of a planet are most naturally expressed in the body-fixed frame,
 as described in Section~\ref{sec:rotational_deformation}.
-The relative rotation between the inertial and geographic reference frame is given by the planetary rotation vector $\mitbf{\Omega}$.
 Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
@@
 \end{equation}
-An alternative to the Love number formulism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
+An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
 A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
@@
 Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
-The rate of polar wander is the angular rate of change of the spin axis $\left| \dot{\mitbf{\omega}} \right| = \dot{\Theta}$ in the geographic reference frame.
-This can be calculated by dotting Equation~\eqref{eq:liouville_ode} with itself:
+This correspondence permits useful insights (see below).
+Let $\dot{\Theta} = \vert \dot{\mitbf{\omega}} \vert$ denote the rate of polar wander.
+Evaluating the scalar product $\dot{\Theta}^2 = \dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}}$ gives
 \begin{equation}
@@
 This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.
-Two special cases of this equation are of note. First, if it is evaluated on the octahedral plane
+Two special cases of this equation are of note. First, if $\dot{\Theta}^2$ is evaluated on the octahedral plane
 (the plane with direction cosines all $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
@@
 
-The second special case is if the rotation vector $\mitbf{\omega}$ lies in the plane of the $\lambda_1$ axis ($\phi = 0$).
+The second special case is if the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_2$ plane ($\phi = 0$).
 Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
@@
 and in the geodesy community has commonly been termed $m$ \citep[e.g.][]{nakiboglu1982hydrostatic, chambat2010flattening}, which we adopt here.
-Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
+Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers that characterize those effects (e.g., the Prandtl and dissipation numbers).
 It would be straightforward to include them, but they do not affect the overall treatment of this scaling.
@@
 \end{equation}
-The first term of this equation is zero due to the hydrostatic equation.  
+As previously noted, the first term of this equation is zero due to the hydrostatic equation.  
 The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma\equiv \alpha \Delta T$.
@@
 Having drawn the connection between the angular momentum equation (Section~\ref{sec:rotational_dynamics}) and the linear momentum 
-equation (Section~\ref{sec:internal}), we would like to apply insights from fluid dynamics and scaling of the 
+equation (Section~\ref{sec:internal}), we would like to apply the scaling of the 
 linear momentum equation to the problem of TPW. We would therefore like to rewrite Equation~\eqref{eq:simple_milankovitch}
@@
 This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
-The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces($m$ \citep{munk1960rotation}:
+The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the ratio of rotational to gravitational forces(\citep{munk1960rotation}:
 \begin{equation}
@@
 \begin{equation}
-\frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma m} \Lambda_{31} \sin{2 \theta}.
+\frac{\kappa}{R^2} \; \dot{\Theta} \sim - \frac{\mathrm{Ra}}{\Gamma m} \Lambda_{31} \sin{2 \theta}.
 \label{eq:scaled_rotation}
@@
 a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
-Orthogonality of the basis functions for the expansion results in the integral picking out degree-two 
-spherical harmonics in the lateral dimensions, and only having power in the lowest few radial functions $R_n(r)$.
+Orthogonality of the basis functions for the expansion means that the integral for $C_{ij}$ picks out degree-two 
+spherical harmonics in the lateral dimensions, and only the lowest few radial functions $R_n(r)$.
 Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
@@
 from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
-the rotation of the body-fixed frame relative to inertial space, as shown in Equation~\eqref{eq:euler})
+the rotation of the body-fixed frame relative to inertial space, as shown in Equation~\eqref{eq:euler}).
 The $\mathbf{E}$-frame slowly rotates with respect to the geographic frame, described by the rotation vector $\mathbf{\Psi}$ 
@@
 
-We may solve Equation~\eqref{eq:liouville_ode_new_frame} for $\dot{\theta}$ in terms of $\theta$, $\alpha$, and $\beta$:
+We may solve Equation~\eqref{eq:liouville_ode_new_frame} for $\dot{\theta}$ 
+(i.e., the colatitudinal velocity in the $\mathbf{E}$ frame, not the TPW rate $\dot{\Theta}$)
+in terms of $\theta$, $\alpha$, and $\beta$:
 \begin{equation}
@@
 However, if $\mathbf{E}$ is not stable to perturbations, then $\theta$ could be up to $90^\circ$. 
-Given a random perturbation to that moment, we would like to give a bound on how big the perturbation to $\theta$ is. 
+Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of the perturbation to $\theta$. 
 This may be done by application of a theorem due to \citet{davis1970rotation}.
@@
 still large TPW events associated with them.
+For example, at $\sim$0.3 Gyr the eigengap dips, with an associated large TPW event, even though
+there is technically no interchange of the axes.
 
 We suggest, then, that $90^\circ$ IITPW events are simply a special case of a broad class of events which can occur
-when the principal moments are close to each other.  Furthermore, a more vigorously convecting planet is much more likely to experience rapid TPW events
+when the principal moments are close to each other. 
+Furthermore, a more vigorously convecting planet is much more likely to experience rapid TPW events
 as it has a lower characteristic gap between principal moments and the gap is much more likely to go to zero.
@@
 are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
-The largest question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
+A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
 \citet{richards1997explanation} have argued that the convective planform of Earth has been stable for the last
@@
 
-We may explicitly draw a connection between the moment of inertia of a rotating object and it's degree-two density structure.  The moment of inertia tensor may be written in index notation
+There is a connection between the moment of inertia of a rotating object and the degree-two density structure.
+The moment of inertia tensor may be written in index notation
 \begin{equation}

### 2016-03-08 tpw_rate.tex
Change angles so as not to reuse alpha

@@
 However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$-frame.
-This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$ in the geographic frame, which is defined by colatitude $\alpha$ and longitude $\beta$.
+This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$ in the geographic frame, which is defined by colatitude $\beta$ and longitude $\gamma$.
 The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$-frame.
@@
 (for nonzero $\phi$ the equations become more complex, but the basic physics is unchanged).
-The direction of $\Psi$ is given by colatitude $\alpha$ and longitude $\beta$, as shown in Figure~\ref{fig:reference_frames}.
+The direction of $\Psi$ is given by colatitude $\beta$ and longitude $\gamma$, as shown in Figure~\ref{fig:reference_frames}.
 
@@
 (i.e., the colatitudinal velocity in the $\mathbf{E}$ frame, not the TPW rate $\dot{\Theta}$)
-in terms of $\theta$, $\alpha$, and $\beta$:
+in terms of $\theta$, $\beta$, and $\gamma$:
 \begin{equation}
 \dot{\theta} = 
--\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\alpha \sin\beta.
+-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\beta \sin\gamma.
 \label{eq:theta_competition}
@@
 and $\theta$ is completely determined by $\mitbf{\Psi}$
-(where we have neglected the orientation factor $\sin \alpha \sin \beta$ for the purposes of scaling).
+(where we have neglected the orientation factor $\sin \beta \sin \gamma$ for the purposes of scaling).
 In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 

### 2016-03-08 figures/reference_frames.py

@@ -208,9 +208,9 @@ ax.text(omega[0] + omega_dot_prime[0]*1.2, omega[1] + omega_dot_prime[1]*1.2, om
 p = 0.45*(omega + z0)/2.0
 ax.text( p[0], p[1], p[2], r'$\theta$', **text_options)
 p = 0.4* (z0 + psi) / 2.0
-ax.text( p[0], p[1], p[2], r'$\alpha$', **text_options)
-p = 0.6* (tmpvec + tmpvec2) / 2.0
 ax.text( p[0], p[1], p[2], r'$\beta$', **text_options)
+p = 0.6* (tmpvec + tmpvec2) / 2.0
+ax.text( p[0], p[1], p[2], r'$\gamma$', **text_options)
 
 # show figure
 ax.view_init(elev=25, azim=49)

### 2016-03-15 response_to_review.tex
Add supporting material for resubmission.

\documentclass[a4paper,12pt]{article}
\usepackage[utf8]{inputenc}

%opening
\title{Response to review}
\author{Ian Rose and Bruce Buffett}
\date{}

\begin{document}

\maketitle

We thank Yanick Ricard for his thoughtful review.
In addition to several smaller suggestions and corrections,
his review centers on our treatment of the reference frames relevant to the scaling of true polar pander (TPW).
We broadly agree with his criticism, and have substantially reworked our discussion of reference frames to address it.
Several of the key equations have been corrected, and we have changed some of the notation for clarity,
but the overall conclusions of the manuscript are unchanged.

In particular, we have done the following to clarify and correct the treatment of reference frames:
\\
\\
1. We have added a section (Section 2.2) which makes the differences between reference frames explict.
We identify three frames which are important to the problem: 
(1) the inertial, nonrotating frame, 
(2) A body-fixed geographic frame, rotating with respect to the inertial frame with relative rotation vector $\mathbf{\Omega}$, and
(3) The frame described by the principal axes of the convective moment of inertia $\mathbf{E}$,
rotating with respect to the body-fixed frame with relative rotation vector $\mathbf{\Psi}$.
\\
\\
2. We have added a new figure (Figure 1) which illustrates the different reference frames.
\\
\\
3. We have changed the discussion in Section 2.5 to clarify that the rate of TPW is measured in the body-fixed frame.
The rate of TPW we denote by $\dot{\Theta}$. The physics describing the magnitude of $\dot{\Theta}$ are 
more naturally expressed in the $\mathbf{E}$-frame, with colatitude $\theta$ and longitude $\phi$. 
For much of the discussion we make the simplifying assumption that $\phi=0$.
If the $\mathbf{E}$-frame is not rotating with respect to the body fixed frame ($\mathbf{\Psi}=0$) then $\dot{\Theta} = \dot{\theta}$.
\\
\\
4. We have updated the discussion of the time evolution of the mismatch angle $\theta$ (Section 4.2)
in light of the changes to the reference frames. The characteristic size of $\theta$ is set by 
the competition of its decay via TPW and its growth via $\mathbf{\Psi}$.
This is reflected in Equations (43) and (44).
Our scaling furnishes estimates of the sizes of these two processes.
\\
\\
Besides the discussion of reference frames, we have addressed these other issues raised in the review:

\begin{itemize}
\item We have added the clarifying intermediate step in Equation (7).
\item We have added Equation (21) on the source of the gravitational field.
\item We have changed the notation of the ratio of centrifugal to gravitational forces from
the fluid-dynamics-inspired ``Froude number'' to the symobl $m$ more commonly used in geodesy.
\item We have clarified discussion of transforming the linear momentum equation to the
angular momentum equation (Equations (25)-(29)). The review criticized this transformation as not very useful.
While we agree that these equations are not particularly easy to evaluate in practice,
they do serve two useful purposes for our scaling: First, they establish the overall consistency
between the approximations made for the linear momentum equation (Section 3) and those made for
the angular momentum equation (Section 2). Second, Equations (23)-(29) demonstrate the
origin of the nondimensionalization that we use for our scaling. In particular, Equations (28)-(29)
shows the origin of $\alpha \Delta T$ as an independent nondimensional number.
\item We have added additional note on the effect of dynamic compensation on the value of $(1+k_f^L)$
\item The review suggested that stirring and mixing can populate lengthscales smaller than $d$
in the convecting system, where $d$ is the injection lengthscale, roughly set by the thickness of
the boundary layers. While mixing can do this, we argue that those lengthscales are
quickly homogenized by thermal diffusion, whereas thermal anomalies at larger lengthscales 
persist for a much longer time.

\end{itemize}

\end{document}

### 2016-03-15 response_to_referee.tex
Updates to response.

--- 2016-03-15 response_to_review.tex
@@
 %opening
-\title{Response to review}
+\title{Response to referee}
 \author{Ian Rose and Bruce Buffett}
@@
 In addition to several smaller suggestions and corrections,
-his review centers on our treatment of the reference frames relevant to the scaling of true polar pander (TPW).
+his review centers on our treatment of the reference frames relevant to the scaling of true polar wander (TPW).
 We broadly agree with his criticism, and have substantially reworked our discussion of reference frames to address it.
@@
 (2) A body-fixed geographic frame, rotating with respect to the inertial frame with relative rotation vector $\mathbf{\Omega}$, and
-(3) The frame described by the principal axes of the convective moment of inertia $\mathbf{E}$,
-rotating with respect to the body-fixed frame with relative rotation vector $\mathbf{\Psi}$.
+(3) The frame described by the principal axes of the convective moment of inertia $\mathbf{E}$, which we call the $\mathbf{E}$-frame.
+This frame rotates slowly with respect to the body-fixed frame with rotation vector $\mathbf{\Psi}$.
 \\
@@
 \\
-Besides the discussion of reference frames, we have addressed these other issues raised in the review:
+Besides the discussion of reference frames, we have addressed the following other issues raised in the review:
 

### 2016-06-06 tpw_rate.tex
Add line numbers.
Edits from Bruce.

@@
 \usepackage{color,soul}
+\usepackage{lineno}
 
@@
 
+\linenumbers
+
 \section{Introduction}
@@
 Second, there is the body-fixed (or geographic) frame.
-By definition, the rotaiton of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
+By definition, the rotation of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
 A terrestrial no-net-rotation or hotspot reference frame are common choices for the body-fixed frame.
@@
 Redistribution of mantle mass anomalies due to convection changes the principal axes of $\mathbf{E}$, 
-rotating the $\mathbf{E}$-frame with respect to the geographic frame, which we will describe
+causing the $\mathbf{E}$-frame to slowly rotate with respect to the geographic frame. We denote this drift 
 by a rotation vector $\mitbf{\Psi}$ (see Section~\ref{sec:theta}).
@@
 \end{equation}
-where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, and $C_{ij}$ is the moment of inertia without that response.
+where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, 
+and $C_{ij}$ is the moment of inertia due solely to the internal load.
 Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:
@@
 For illustration, we assume the longitude $\psi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane.
-True polar wander in the geographic reference frame moves the rotation axis towards the $\mathbf{e}_3$-axis, shown by $\dot{\mitbf{\omega}}$.
+True polar wander moves the rotation axis towards the $\mathbf{e}_3$-axis.
+We denote this motion by $\dot{\mitbf{\omega}}$ in the geographic frame.
 However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$-frame.
-This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$ in the geographic frame, which is defined by colatitude $\beta$ and longitude $\gamma$.
+This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$, which is defined by colatitude $\beta$ and longitude $\gamma$.
 The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$-frame.
-The motion of $\mitbf{\omega}$ in the $\mathbf{E}$-frame is shown by $\dot{\mitbf{\omega}}^\prime$.
 Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Sections~\ref{sec:reference_frames} and \ref{sec:theta}}
@@
 
-The polar wander rate $\dot{\Theta}$ is measured in the body-fixed geographic frame, 
-but the physics of the right-hand-side of Equation~\eqref{eq:polar_wander_rate_tensor} 
-is more naturally expressed in the reference frame of the principal axes of $\mathbf{E}$,
-which is, in general, rotating with respect to the geographic frame.
-However, the right-hand-side is scalar-valued, and as such is invariant to rotations.
+Our goal is to quantify the polar wander rate $\dot{\Theta}$ due to a convective perturbation $\mathbf{E}$ in the moment of inertia. 
+The rate defined by Equation~\eqref{eq:polar_wander_rate_tensor} is expressed in the body-fixed geographic frame, 
+which is also the reference frame used to measure TPW. 
+However, the physics of the right-hand-side of Equation~\eqref{eq:polar_wander_rate_tensor} 
+is more naturally expressed in the reference frame of the principal axes of $\mathbf{E}$. 
+In general the $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, so the time derivative of $\omega$ is different in these two frames.
+However, the rate of true polar wander in Equation~\eqref{eq:polar_wander_rate_tensor} is a scalar quantity, which means it is invariant to rotations.
 Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
 with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of 
-$\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ (see Figure~\ref{fig:reference_frames}).
+$\mitbf{\omega}$ in the $\mathbf{E}$-frame with colatitude $\theta$ and longitude $\phi$ (see Figure~\ref{fig:reference_frames}).
 Plugging this description of $\mitbf{\omega}$ into Equation~\eqref{eq:polar_wander_rate_tensor},
@@
 From Equation~\eqref{eq:simple_milankovitch} it is clear that the important quantities for estimating $\dot{\Theta}$ are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.
-They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principal axis of the convective moment.
-Both $\theta$ and $\lambda_i$ depend upon the dynamics of the convecting planetary mantle.
-In order to derive estimates for them, we must consider those dynamics.
+They represent, respectively, the size of convective anomalies in the moment of inertia tensor 
+and the angular mismatch between the rotation axis and the principal axis of the convective moment. 
+The dynamics of mantle convection affects the rate of TPW in two ways; 
+it controls $\lambda_i$ and it represents the mechanism that drives the angular mismatch. 
+The value of $\theta$ at any time also depends on the relaxation of the rotation axis back toward the principal axis of $\mathbf{E}$. 
+We focus on the dynamics of the convective contribution in the next section.
 
@@
 where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
-The gravity vector $\mathbf{g}$ is given by $\mathbf{g} = -\nabla V$, where the 
-gravitational potential $V$ comes from solving Poisson's equation
+The vector $\mathbf{g}$ is the gravitational acceleration, 
+which defined in terms of a gravitational potential $V$ by $\mathbf{g} = -\nabla V$. 
+The gravitational potential obeys 
 \begin{equation}
@@
 \end{equation}
-For the purposes of scaling, we assume that gravity is the approximately constant
-vector $g_0 \mathbf{\hat{r}}$ (as is the case for Earth's mantle).
+where $G$ is the gravitational constant. 
+For the purposes of scaling, we assume that the magnitude of the gravitational acceleration, $g_0$, is the approximately constant (as is the case for Earth's mantle).
 In addition we use the simple equation of state
@@
 \end{equation}
+to define the density.
 The remaining parameters are defined in Table~\ref{tab:parameters}.
@@
 
-In particular, the two most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
-This second nondimensional number does not have a uniformly agreed-upon name: it has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence},
+Two dimensionless parameters have a prominent role in our scaling analysis. 
+The first is the Rayleigh number, which characterizes the  vigor of convection. 
+The second is the ratio of centrifugal to gravitational forces.
+This nondimensional number does not have a uniformly agreed-upon name: 
+it has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, 
 and in the geodesy community has commonly been termed $m$ \citep[e.g.][]{nakiboglu1982hydrostatic, chambat2010flattening}, which we adopt here.
@@
 
-In this case the dynamics can be characterized in terms of deviations from a reference hydrostatic state, which includes the dynamic pressure $P^* = P - P_0$ and density perturbations $\delta \rho = \rho- \rho_0 = - \rho_0 \alpha (T-T_0)$.
-In addition, we expect deviations in the figure of the planet from its hydrostatic shape, which we denote by $V = V_H + \Delta V$, where $V_H$ is the hydrostatic figure and $\Delta V$ is the deviation.
-The introduction of $\Delta V$ requires another nondimensional number to characterize it, and we find that the quantity $\Gamma \equiv \alpha \Delta T$ is convenient.
+In this case the dynamics can be characterized in terms of deviations from a reference hydrostatic state, 
+which includes the dynamic pressure $P^* = P - P_0$ and density perturbations $\delta \rho = \rho- \rho_0 = - \rho_0 \alpha (T-T_0)$.
+In addition, we expect deviations in the figure of the planet from its hydrostatic shape, 
+which we denote by $V = V_H + \Delta V$, where $V_H$ is the hydrostatic figure and $\Delta V$ is the deviation. 
+Our rationale for this decomposition is simply that the hydrostatic pressure is most naturally defined in the (oblate) hydrostatic configuration. 
+The introduction of $\Delta V$ requires another nondimensional number to characterize it, 
+and we find that the quantity $\Gamma \equiv \alpha \Delta T$ is convenient.
 Finally, we define $\mitbf{\Omega} = \Omega_0 \mitbf{\omega}$, where $\mitbf{\omega}$ is a unit vector in the direction of $\mitbf{\Omega}$.
@@
 \end{equation}
-This equation may be identified with Equation~\eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
+This equation may be identified with Equation~\eqref{eq:disequilibrium}, 
+where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of inertia with the spin axis (right side).
 Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table~\ref{tab:nondim}.
@@
 This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
-The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the ratio of rotational to gravitational forces(\citep{munk1960rotation}:
+The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is 
+proportional to the ratio of rotational to gravitational forces (\citep{munk1960rotation}:
 \begin{equation}
@@
 
-The quantites $\Lambda_{ij}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $m$, and $\Gamma$.  
+The quantity $\Lambda_{ij}$ should be a function of our nondimensional parameters $\mathrm{Ra}$, $m$, and $\Gamma$. 
+Similarly, the convective forcing that drives changes in $\theta$ must also be a function these nondimensional parameters. 
+To be explicit in our discussion, we denote this driving force by ${\Psi}$.
 We consider relatively slowly rotating bodies here ($m \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
-In this regime, therefore, we neglect the dependence on $m$, and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra}, \Gamma)$ and $\theta(\mathrm{Ra}, \Gamma)$.
-
-\subsection{An estimate of $\Lambda_{ij}$}
+In this regime, therefore, we neglect the dependence on $m$, 
+and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra}, \Gamma)$ and $\Psi(\mathrm{Ra}, \Gamma)$.
+
+\subsection{An estimate for $\Lambda_{ij}$}
 \label{sec:lambda}
@@
 For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix~\ref{appendix:moments}).
-Therefore, an estimate of one allows an estimate of the other.
+Therefore, an estimate for one specifies the other.
 We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
@@
 This is a strong constraint, but it gives very little information about the distribution of power across the $T_{lmn}$.  
-We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow (relatively low $\mathrm{Ra}$) and that of chaotic flow (relatively high $\mathrm{Ra}$).
+We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow (relatively low $\mathrm{Ra}$) 
+and that of chaotic flow (relatively high $\mathrm{Ra}$).
 The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
@@
 We performed a series of numerical simulations of mantle convection at different Rayleigh numbers to test this scaling.
-We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII82}.
+We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII82},
 which allows for flexible implementation of different rheologies, geometries, and postprocessors.
@@
 
-\subsection{Estimating the mismatch $\theta$}
+\subsection{An estimate for the angular mismatch $\theta$}
 \label{sec:theta}
 
-The mismatch between the current rotation axis and the principal axis of the convective moment is perhaps the most important parameter in Equation~\eqref{eq:scaled_rotation}.  
+Convection drives a mismatch $\theta$ between the current rotation axis and the principal axis of the convective moment. 
+This parameter is crucial for the rate of TPW, as described in is in Equation~\eqref{eq:scaled_rotation}.  
 Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
 
-There are two processes at work which control the evolution of $\theta$, its growth through perturbations in the convective moment and 
-its decay via relaxation of the pole towards the maximum moment of inertia.
+Two processes control the evolution of $\theta$. Its growth occurs through perturbations in the convective moment and 
+its decay occurs by relaxation of the pole towards the maximum moment of inertia.
 We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
 from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
-the rotation of the body-fixed frame relative to inertial space, as shown in Equation~\eqref{eq:euler}).
-The $\mathbf{E}$-frame slowly rotates with respect to the geographic frame, described by the rotation vector $\mathbf{\Psi}$ 
-(here $\mitbf{\Psi}$ defines the rotation of the $\mathbf{E}$-frame in the geographic frame, see Figure~\ref{fig:reference_frames}).
-We can then write the rate of change of $\mitbf{\omega}$ in the $\mathbf{E}$-frame (denoted by primes)
+the rotation of the body-fixed frame relative to inertial space in Equation~\eqref{eq:euler}).
+The $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, 
+which we describe by the rotation vector $\mathbf{\Psi}$. 
+(To be explicit in our discussion we let $\mitbf{\Psi}$ define the rotation of the $\mathbf{E}$-frame in the geographic frame, 
+although this choice is arbitrary as long as we are consistent). The time derivatives of $\mitbf{\omega}$ in the two frames are related by 
 \begin{equation}
@@
 \end{equation}
-Equation~\eqref{eq:liouville_ode} then becomes
+where primes are used to define quantities in the $\mathbf{E}$-frame. Rearranging gives 
+\begin{equation}
+\dot{\mitbf{\omega}}^{\prime} = \dot{\mitbf{\omega}} - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
+\end{equation}
+We now substitute for $\dot{\mitbf{\omega}}$ from Equation~\eqref{eq:liouville_ode}, 
+making the additional assumption that the geographic and $\mathbf{E}$-frames are momentarily aligned. 
+This assumption is not a serious restriction because our goal is simply to relate the time derivatives of $\mitbf{\omega}$ in the two frames, 
+rather than track the evolution of $\mitbf{\omega}$ over time. The expression for $\dot{\mitbf{\omega}}$ becomes
 \begin{equation}
@@
 \end{equation}
-The advantage of expressing TPW in the $\mathbf{E}$-frame is that the entire equation
+The advantage of expressing the rate of TPW in the $\mathbf{E}$-frame is that the entire equation
 may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
-As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that $\phi=0$
-(for nonzero $\phi$ the equations become more complex, but the basic physics is unchanged).
-The direction of $\Psi$ is given by colatitude $\beta$ and longitude $\gamma$, as shown in Figure~\ref{fig:reference_frames}.
-
-We may solve Equation~\eqref{eq:liouville_ode_new_frame} for $\dot{\theta}$ 
-(i.e., the colatitudinal velocity in the $\mathbf{E}$ frame, not the TPW rate $\dot{\Theta}$)
-in terms of $\theta$, $\beta$, and $\gamma$:
+As in Equation~\eqref{eq:simple_milankovitch}, we can choose a convenient coordinate system by aligning the axes so that 
+$\mitbf{\omega}^{\prime}$ lies in $\mathbf{e}_1^{\prime}-\mathbf{e}_3^{\prime} $ plane (as shown in Figure~\ref{fig:reference_frames}). 
+In this case the orientation of $\mitbf{\omega}$ is defined by colatitude $\theta$ and longitude $\phi=0$. 
+The colatitude $\theta$ is precisely the angular misalignment that specifies the rate of TPW in Equation~\eqref{eq:simple_milankovitch}.
+The orientation of $\Psi$ is not restricted in this frame, 
+so we specify the direction using colatitude $\beta$ and longitude $\gamma$ (see Figure~\ref{fig:reference_frames}).
+
+The time derivative of the unit vector $\mitbf{\omega}^{\prime}$ defines the changing orientation of the rotation axis in the $\mathbf{E}$-frame. 
+We can express this changing orientation in terms of changes in angles $\theta$ and $\phi$. The change in colatitude is
 \begin{equation}
@@
 \end{equation}
+where $\mitbf{\Psi}$ is expressed in terms of its amplitude $|\mitbf{\Psi}|$ and its orientation angles $\beta$ and $\gamma$.
 Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
@@
 Conversely, when the rate of TPW is very slow (i.e. $\vert \dot{\Theta} \vert \sim 0$), 
-then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$ 
-and $\theta$ is completely determined by $\mitbf{\Psi}$
-(where we have neglected the orientation factor $\sin \beta \sin \gamma$ for the purposes of scaling).
+then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$, so the evolution of $\theta$ is driven mostly by $\mitbf{\Psi}$. 
+For the purposes of scaling we will omit the orientation factor $\sin \beta \sin \gamma$ and focus on $|\mitbf{\Psi}|$.
 In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 
@@
 Convection is continuously redistributing mass throughout the mantle, which will perturb the convective moment of inertia.  
-If the convective moment is stable to small perturbations, then $\vert \mitbf{\Psi} \vert$ should be small, 
-and $\theta$ will never be large.
-However, if $\mathbf{E}$ is not stable to perturbations, then $\theta$ could be up to $90^\circ$. 
-Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of the perturbation to $\theta$. 
+If the convective moment is stable to small perturbations, then $\vert \mitbf{\Psi} \vert$ should be small, so
+ $\theta$ will never have the opportunity to grow.
+However, if $\mathbf{E}$ is not stable to perturbations, then $|\mitbf{\Psi}|$ can be large and $\theta$ can grow up to $90^\circ$. 
+Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of changes in the orientation of the principle axes. 
 This may be done by application of a theorem due to \citet{davis1970rotation}.
 
-Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.  
-Then the rotation of the principal axes of the tensor is bounded by
-
-\begin{equation}
-\vert \sin(2 \theta) \vert \le \frac{ 2 \vert \delta \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
+Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor after some time interval $\Delta t$, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.
+The corresponding rotation of the principal axes is defined by the angle $\xi$.
+A bound on $\xi$ is given by
+\begin{equation}
+\vert \sin(2 \xi) \vert \le \frac{ 2 \vert \delta \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
 \label{eq:kahan}
@@
 
-We may arrive at a simple estimate of the growth rate of the angle $\theta$ (and thus $\vert \mitbf{\Psi} \vert$) 
+We may arrive at a simple estimate for $|\mitbf{\Psi}|$  
 by differentiating Equation~\eqref{eq:kahan}, tentatively holding $\vert \lambda_i - \lambda_j \vert$ fixed:
-
-\begin{equation}
-\vert \mitbf{\Psi} \vert \sim \vert \dot{\theta} \vert \le \frac{ \vert \dot{\delta} \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
+\begin{equation}
+\vert \mitbf{\Psi} \vert \sim \vert \dot{\xi} \vert \le \frac{ \vert \dot{\delta} \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
 \label{eq:grow_perturbation}
 \end{equation} 
+As noted previously, $|\mitbf{\Psi}|\sim |\dot\theta |$ when the rate of TPW is small and the angular misalignment $\theta$ is small. 
+Consequently, we can interpret Equation~\eqref{eq:grow_perturbation} as a bound on how fast $\theta$ can grow.
 
@@
 
-We then suggest that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta(\mathrm{Ra})$.
-Whereas the rest of the expression has a weak dependence on $\mathrm{Ra}$, $\theta$ is expected to be strongly dependent on it.
+We then suggest that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta$.
+Whereas the prefactor $\mathrm{Ra}^{1-3\beta}$ in Equation~\eqref{eq:simplest_milankovitch} has a weak dependence on $\mathrm{Ra}$, 
+the misalignment angle $\theta$ is expected to be strongly dependent on it.
 
@@
 This is the regime considered in \citet{steinberger1997changes}, \citet{roberts2007cause}, and \citet{zhong2007supercontinent}, and can be considered the ``slow TPW'' regime.
-When convection is more chaotic, however, there may be large excursions in $\theta$, which can drastically increase the wander rate.  
+When convection is more chaotic, however, there may be large excursions in $\theta$, 
+which are driven by large values for $|\mitbf{\Psi}|$ when $(\lambda_i - \lambda_j)$ is small, according to Equation~\eqref{eq:grow_perturbation}. 
+In this case a dramatic increase in the rate of TPW is possible.
 If $\theta=90^\circ$, this corresponds to IITPW \citep{kirschvink1997evidence}.  
@@
 \begin{acknowledgments}
-This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for a thoughtful review.
+This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for thoughtful reviews and constructive comments that greatly improved the paper.
 \end{acknowledgments}

### 2016-06-06 tpw_rate.tex
Fix statement on choice of coordinate system.

@@
 
-The second special case is if the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_2$ plane ($\phi = 0$).
+The second special case is if the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_3$ plane ($\phi = 0$).
 Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
@@
 may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
-As in Equation~\eqref{eq:simple_milankovitch}, we can choose a convenient coordinate system by aligning the axes so that 
+As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that 
 $\mitbf{\omega}^{\prime}$ lies in $\mathbf{e}_1^{\prime}-\mathbf{e}_3^{\prime} $ plane (as shown in Figure~\ref{fig:reference_frames}). 
@@
 The colatitude $\theta$ is precisely the angular misalignment that specifies the rate of TPW in Equation~\eqref{eq:simple_milankovitch}.
-The orientation of $\Psi$ is not restricted in this frame, 
-so we specify the direction using colatitude $\beta$ and longitude $\gamma$ (see Figure~\ref{fig:reference_frames}).
+We furthermore specify the orientation of $\Psi$
+using colatitude $\beta$ and longitude $\gamma$ (see Figure~\ref{fig:reference_frames}).
 

### 2016-06-06 figures/reference_frames.py

@@ -115,11 +115,8 @@ beta = 110. * np.pi/180.
 psi = np.array([ np.sin(alpha)*np.cos(beta), np.sin(alpha)*np.sin(beta), np.cos(alpha)])
 # tpw vector
 omega_dot = np.dot(E, omega.T) - np.dot(omega, np.dot(E, omega.T))*omega.T
-omega_dot_prime = omega_dot + np.cross(psi, omega)
 #omega_dot = 0.3*omega_dot/norm(omega_dot)
-#omega_dot_prime = 0.3*omega_dot_prime/norm(omega_dot_prime)
 omega_dot *= 0.35
-omega_dot_prime *= 0.35
 
 # produce figure
 alpha_val = 0.4
@@ -184,11 +181,6 @@ omega_dot_arrow = Arrow3D([omega[0], omega[0]+omega_dot[0]], \
                           [omega[2], omega[2]+omega_dot[2]], \
                           **arrow_prop_dict)
 ax.add_artist(omega_dot_arrow)
-omega_dot_prime_arrow = Arrow3D([omega[0], omega[0]+omega_dot_prime[0]], \
-                                [omega[1], omega[1]+omega_dot_prime[1]], \
-                                [omega[2], omega[2]+omega_dot_prime[2]], \
-                                **arrow_prop_dict)
-ax.add_artist(omega_dot_prime_arrow)
 
 
 text_options = {'horizontalalignment': 'center',
@@ -202,7 +194,6 @@ ax.text(1.1*z0[0],1.1*z0[1],1.1*z0[2],r'$\mathbf{e}_3$', **text_options)
 ax.text(1.1*omega[0],1.1*omega[1],1.1*omega[2],r'$\omega$', **text_options)
 ax.text(1.1*psi[0],1.1*psi[1],1.1*psi[2],r'$\Psi$', **text_options)
 ax.text(omega[0] + omega_dot[0]*1.15, omega[1] + omega_dot[1]*1.15, omega[2]+omega_dot[2]*1.15, r'$\dot{\omega}$', **text_options)
-ax.text(omega[0] + omega_dot_prime[0]*1.2, omega[1] + omega_dot_prime[1]*1.2, omega[2]+omega_dot_prime[2]*1.2, r'$\dot{\omega}^\prime$', **text_options)
 
 # add text for angles
 p = 0.45*(omega + z0)/2.0

### 2016-06-07 figures/perturb.sh
Update symbol in figure.

@@ -14,18 +14,18 @@ gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
 EOF
 
 #lambda 1
-gmt pstext -R -J -K  -O >>perturb.ps <<EOF
+gmt pstext -R -J -K  -O -F+f9 >>perturb.ps <<EOF
 .04 -0.32 @~l@-1@-@~
 EOF
 
 #lambda 2
-gmt pstext -R -J -K -O >>perturb.ps <<EOF
+gmt pstext -R -J -K -O -F+f9 >>perturb.ps <<EOF
 .32 0.26 @~l@-2@-@~
 EOF
 
 #theta
-gmt pstext -R -J -K -O >>perturb.ps <<EOF
-.46 0.10 @~q@~
+gmt pstext -R -J -K -O -F+f9 >>perturb.ps <<EOF
+.46 0.10 @~x@~
 EOF
 
 #Rotation axis
@@ -58,18 +58,18 @@ gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
 EOF
 
 #lambda 1
-gmt pstext -R -J -K  -O >>perturb.ps <<EOF
+gmt pstext -R -J -K  -O -F+f9 >>perturb.ps <<EOF
 .36 -0.15 @~l@-1@-@~
 EOF
 
 #lambda 2
-gmt pstext -R -J -K -O >>perturb.ps <<EOF
+gmt pstext -R -J -K -O -F+f9 >>perturb.ps <<EOF
 -.02 0.46 @~l@-2@-@~
 EOF
 
 #theta
-gmt pstext -R -J -K -O >>perturb.ps <<EOF
-.1 0.10 @~q@~
+gmt pstext -R -J -K -O -F+f9 >>perturb.ps <<EOF
+.1 0.10 @~x@~
 EOF
 
 #rotation axis

### 2016-06-07 figures/perturb.py
Rewrite GMT script in Matplotlib

@@ -1,0 +1,85 @@
+from matplotlib import pyplot as plt
+import matplotlib.patches as patches
+import numpy as np
+
+plt.rc('text', usetex=True)
+
+d2r = np.pi/180.
+
+fig = plt.figure(figsize=(8,4), frameon=False)
+
+
+ax = fig.add_subplot(111)
+ax.axis('off')
+ax.axes.get_yaxis().set_major_formatter(plt.NullFormatter())
+ax.axes.get_xaxis().set_major_formatter(plt.NullFormatter())
+ax.tick_params( top='off', bottom='off', left='off', right='off')
+ax.set_xlim(-2,2)
+ax.set_ylim(-1,1)
+
+ellipse_args = { 'edgecolor': 'k', 'facecolor':'w', 'lw':2 } 
+arrow_args = {'arrowstyle':'-|>', 'lw':2, 'mutation_scale':20, 'facecolor':'k'}
+text_args = {'horizontalalignment': 'center', 'verticalalignment': 'center', 'fontsize': 20}
+
+##FLATTENED ELLIPSE
+
+#ellipse
+flat_ellipse = patches.Ellipse( xy = (-1,0), width=1.6, height=1.2, angle=20., **ellipse_args)
+ax.add_patch(flat_ellipse)
+#perturbation
+perturbation = patches.Ellipse( xy = (-1.3, 0.4), width=0.15, height=0.15, facecolor='darkgray', edgecolor='w')
+ax.add_patch(perturbation)
+
+#spin axis
+omega = patches.FancyArrowPatch( (-1,0), (-0., 0.1), **arrow_args)
+ax.add_patch(omega)
+ax.text( 0.04, 0.1, r'$\Omega$', **text_args)
+
+#principal axes
+lambda_1 = patches.FancyArrowPatch( (-1,0), (-1. + 0.8*np.cos(30.*d2r), 0.8*np.sin(30.*d2r)), **arrow_args)
+lambda_2 = patches.FancyArrowPatch( (-1,0), (-1. + 0.6*np.cos(-60.*d2r), 0.6*np.sin(-60.*d2r)), **arrow_args)
+ax.add_patch(lambda_1)
+ax.add_patch(lambda_2)
+ax.text( 1.5, -0.25, r'$\lambda_1$', **text_args)
+ax.text( 0.97, 0.5, r'$\lambda_2$', **text_args)
+
+#arc
[…]

### 2016-06-07 figures/perturb.sh

@@ -1,94 +1,0 @@
-
-#draw ellipse
-gmt psxy -R-2/2/-2/2 -Jx3 -Se -Wthicker,black -K > perturb.ps <<EOF
-0 0 20 3.6 2.8
-EOF
-
-# red axis
-gmt psxy -R -J -Sv0.3+e -Wthicker,red -K -O >> perturb.ps <<EOF
-0 0 20 1.8 1.8
-EOF
-#blue axis
-gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
-0 0 290 1.4 1.4
-EOF
-
-#lambda 1
-gmt pstext -R -J -K  -O -F+f9 >>perturb.ps <<EOF
-.04 -0.32 @~l@-1@-@~
-EOF
-
-#lambda 2
-gmt pstext -R -J -K -O -F+f9 >>perturb.ps <<EOF
-.32 0.26 @~l@-2@-@~
-EOF
-
-#theta
-gmt pstext -R -J -K -O -F+f9 >>perturb.ps <<EOF
-.46 0.10 @~x@~
-EOF
-
-#Rotation axis
-gmt psxy -R -J -Sv0.3+e -Wthicker,black -K -O >> perturb.ps <<EOF
-0 0 4 2.4 1 
-EOF
-
-#Omega
-gmt pstext -R -J -K -O >>perturb.ps <<EOF
-0.9 .05 @~W@~
-EOF
-
-#perturbation
-gmt psxy -R -J -Sc.1i -Gdarkgray -K -O >>perturb.ps <<EOF
--0.2 0.3
-EOF
-
-#draw ellipse
-gmt psxy -R-2/2/-2/2 -Jx3 -X2i -Se -Wthicker,black -K -O >> perturb.ps <<EOF
-0 0 70 3.2 3.2
-EOF
-
-#red axis
-gmt psxy -R -J -Sv0.3+e -Wthicker,red -K -O >> perturb.ps <<EOF
-0 0 80 1.6 1.6
-EOF
-#blue axis
-gmt psxy -R -J -Sv0.3+e -Wthicker,blue -K -O >> perturb.ps <<EOF
-0 0 350 1.6 1.6
-EOF
-
-#lambda 1
-gmt pstext -R -J -K  -O -F+f9 >>perturb.ps <<EOF
-.36 -0.15 @~l@-1@-@~
-EOF
-
-#lambda 2
[…]

### 2016-06-07 tpw_rate.tex
Small edits, and fix line numbering.

@@
 %\detailtrue
+
+\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+  \renewenvironment{#1}%
+    {\linenomath\csname old#1\endcsname}%
+    {\csname oldend#1\endcsname\endlinenomath}}% 
+  \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+    \patchAmsMathEnvironmentForLineno{#1}%
+    \patchAmsMathEnvironmentForLineno{#1*}}%
+  \AtBeginDocument{%
+  \patchBothAmsMathEnvironmentsForLineno{equation}%
+  \patchBothAmsMathEnvironmentsForLineno{align}%
+  \patchBothAmsMathEnvironmentsForLineno{flalign}%
+  \patchBothAmsMathEnvironmentsForLineno{alignat}%
+  \patchBothAmsMathEnvironmentsForLineno{gather}%
+  \patchBothAmsMathEnvironmentsForLineno{multline}%
+  }
 
@@
 The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $\mathbf{e}_3$-axis.
-For illustration, we assume the longitude $\psi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane.
+For illustration, we assume the longitude $\phi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane.
 True polar wander moves the rotation axis towards the $\mathbf{e}_3$-axis.
@@
 The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is 
-proportional to the ratio of rotational to gravitational forces (\citep{munk1960rotation}:
+proportional to the ratio of rotational to gravitational forces \citep{munk1960rotation}:
 \begin{equation}
@@
 Convection drives a mismatch $\theta$ between the current rotation axis and the principal axis of the convective moment. 
-This parameter is crucial for the rate of TPW, as described in is in Equation~\eqref{eq:scaled_rotation}.  
+This parameter is crucial for the rate of TPW, as previously noted in Equation~\eqref{eq:scaled_rotation}.  
 Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
@@
 $\mitbf{\omega}^{\prime}$ lies in $\mathbf{e}_1^{\prime}-\mathbf{e}_3^{\prime} $ plane (as shown in Figure~\ref{fig:reference_frames}). 
-In this case the orientation of $\mitbf{\omega}$ is defined by colatitude $\theta$ and longitude $\phi=0$. 
+In this case the orientation of $\mitbf{\omega}^\prime$ is defined by colatitude $\theta$ and longitude $\phi=0$. 
 The colatitude $\theta$ is precisely the angular misalignment that specifies the rate of TPW in Equation~\eqref{eq:simple_milankovitch}.
@@
 Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
-The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it arbitrarily via
+The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it via
 relative motion between the geographic frame and the principal axes of the $\mathbf{E}$-frame.
@@
 \includegraphics[width=0.9\textwidth]{figures/perturb.pdf}
-\caption{Graphical demonstration of the $\sin{2 \theta}$ theorem of \citet{davis1970rotation}.
+\caption{Graphical demonstration of the $\sin{2 \xi}$ theorem of \citet{davis1970rotation}.
 Two spheroidal bodies with eigenvalues $\lambda_2 > \lambda_1$ start out with the rotation axis $\Omega$ aligned with the $\lambda_2$ axis.
@@
 There is a constant competition between growth of the mismatch angle $\theta$ through 
-Equation~\eqref{eq:grow_perturbation} and its relaxation through Equation~\eqref{eq:simple_milankovitch}.
+Equation~\eqref{eq:grow_perturbation} and its relaxation through TPW.
 We may plug in the estimate for $\Lambda_{ij}$ (Equation~\eqref{eq:lambda_estimate}) into Equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
@@
 We then suggest that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta$.
-Whereas the prefactor $\mathrm{Ra}^{1-3\beta}$ in Equation~\eqref{eq:simplest_milankovitch} has a weak dependence on $\mathrm{Ra}$, 
+Whereas the prefactor $\mathrm{Ra}^{1-3\beta}$ in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} has a weak dependence on $\mathrm{Ra}$, 
 the misalignment angle $\theta$ is expected to be strongly dependent on it.

### 2016-06-07 tpw_rate.tex
Four nondimensional numbers, really.

@@
 (in the latter case the boundary conditions on $P$ must be modified).
-Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires three nondimensional numbers to characterize it
-(a fourth one, defined by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations).
+Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires four nondimensional numbers to characterize it
+(a fifth one, defined by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations).
 Convenient choices for these numbers are listed in Table~\ref{tab:nondim}, along with approximate Earth-like values for them.

### 2016-06-08 figures/perturb.py
Crop excess whitespace.

@@ -82,4 +82,4 @@ ax.text( 1.15,  0.13, r'$\xi$', **text_args)
 ax.add_patch( patches.Ellipse(xy=(1,0), width=0.02, height=0.02, facecolor='k', edgecolor='k') )
 
 #plt.show()
-plt.savefig('perturb.pdf')
+plt.savefig('perturb.pdf', bbox_inches='tight')

### 2016-06-08 tpw_rate.tex
Start to address some of Isamu's comments.

@@
 In general the $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, so the time derivative of $\omega$ is different in these two frames.
-However, the rate of true polar wander in Equation~\eqref{eq:polar_wander_rate_tensor} is a scalar quantity, which means it is invariant to rotations.
+However, Equation~\eqref{eq:polar_wander_rate_tensor} is a scalar equation, which means it is invariant to rotations.
 Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
@@
 (in the latter case the boundary conditions on $P$ must be modified).
+The centrifugal term, although the largest of the terms neglected in typical mantle convection models,
+is generally small compared to gravitational forces (at least for Earthlike parameters),
+and likely does not have a strong influence on the style of convection.
+However, it is critical for determining the the size of the rotational bulge which must be moved around via TPW, 
+and so we must include it to establish the connection between the linear and angular momentum equations.
+
 Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires four nondimensional numbers to characterize it
@@
 \begin{equation}
-C_{ij} = \int_{V_S} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) \,dV
+C_{ij} = \int_{V_H} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) \,dV
 \label{eq:temperature_fluctuations}
 \end{equation}
-where $V_S$ is the reference spherical volume of the mantle.
 As discussed in Section~\ref{sec:convective_moment}, these internal density loads may be convolved with their surface responses to get the convective moment of inertia $E_{ij}$.
@@
 Equation~\eqref{eq:grow_perturbation} and its relaxation through TPW.
+\citet{goldreich1969some} envisioned an analogy of beetles crawling on the surface of the globe, with
+the spin axis trying to keep up with the instantaneous figure axis set by the beetles.
+Our analysis begins the answer the questions ``how big are the beetles?'' and ``how fast are they crawling?''
+
 We may plug in the estimate for $\Lambda_{ij}$ (Equation~\eqref{eq:lambda_estimate}) into Equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression

### 2016-07-11 tpw_rate.tex
Fix typo
Use elsarticle template.

@@
-% gjilguid2e.tex
-
-%\documentclass[extra,onecolumn]{gji}
-\documentclass[extra,mreferee]{gji}
-\usepackage{timet}
+%% 
+%% Copyright 2007, 2008, 2009 Elsevier Ltd
+%% 
+%% This file is part of the 'Elsarticle Bundle'.
+%% ---------------------------------------------
+%% 
+%% It may be distributed under the conditions of the LaTeX Project Public
+%% License, either version 1.2 of this license or (at your option) any
+%% later version.  The latest version of this license is in
+%%    http://www.latex-project.org/lppl.txt
+%% and version 1.2 or later is part of all distributions of LaTeX
+%% version 1999/12/01 or later.
+%% 
+%% The list of all files belonging to the 'Elsarticle Bundle' is
+%% given in the file `manifest.txt'.
+%%
+
+%% Template article for Elsevier's document class `elsarticle'
+%% with numbered style bibliographic references
+%% SP 2008/03/01
+
+\documentclass[preprint,12pt,authoryear]{elsarticle}
+
+%% Use the option review to obtain double line spacing
+%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
+
+%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
+%% for a journal layout:
+%% \documentclass[final,1p,times]{elsarticle}
+%% \documentclass[final,1p,times,twocolumn]{elsarticle}
+%% \documentclass[final,3p,times]{elsarticle}
+%% \documentclass[final,3p,times,twocolumn]{elsarticle}
+%% \documentclass[final,5p,times]{elsarticle}
+%% \documentclass[final,5p,times,twocolumn]{elsarticle}
+
+%% For including figures, graphicx.sty has been loaded in
+%% elsarticle.cls. If you prefer to use the old commands
+%% please give \usepackage{epsfig}
+
+%% The amssymb package provides various useful mathematical symbols
+\usepackage{amssymb}
+\usepackage{amsmath}
 \usepackage{url}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{color,soul}
+%% The amsthm package provides extended theorem environments
+%% \usepackage{amsthm}
+
+%% The lineno packages adds line numbers. Start line numbering with
+%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
+%% for the whole article with \linenumbers.
 \usepackage{lineno}
 
-\title[TPW Rates]
-  {Scaling rates of true polar wander in convecting planets and moons}
-\author[I. Rose and B. Buffett]
-  {Ian Rose$^1$ and Bruce A. Buffett$^1$ \\
-  $^1$ Department of Earth \& Planetary Science, University of California, Berkeley, CA 94720, USA.  E-mail: ian.rose@berkeley.edu
+%% Block of code for fixing corresponding author bug 
+%% in elsarticle template... Don't really understand it
+\makeatletter
+\def\@author#1{\g@addto@macro\elsauthors{\normalsize%
+    \def\baselinestretch{1}%
+    \upshape\authorsep#1\unskip\textsuperscript{%
+      \ifx\@fnmark\@empty\else\unskip\sep\@fnmark\let\sep=,\fi
+      \ifx\@corref\@empty\else\unskip\sep\@corref\let\sep=,\fi
+      }%
+    \def\authorsep{\unskip,\space}%
+    \global\let\@fnmark\@empty
+    \global\let\@corref\@empty  %% Added
+    \global\let\sep\@empty}%
+    \@eadauthor={#1}
 }
-\date{}
-\pagerange{\pageref{firstpage}--\pageref{lastpage}}
-\volume{}
-\pubyear{}
-
-\let\leqslant=\leq
+\makeatother
+%% End block of weird code
+
+%% Command for bold Greek symbols
+\newcommand{\mitbf}[1]{\hbox{\mathversion{bold}$#1$}}
 
@@
 
-\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
-  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
-  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
-  \renewenvironment{#1}%
-    {\linenomath\csname old#1\endcsname}%
-    {\csname oldend#1\endcsname\endlinenomath}}% 
-  \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
-    \patchAmsMathEnvironmentForLineno{#1}%
-    \patchAmsMathEnvironmentForLineno{#1*}}%
-  \AtBeginDocument{%
-  \patchBothAmsMathEnvironmentsForLineno{equation}%
-  \patchBothAmsMathEnvironmentsForLineno{align}%
-  \patchBothAmsMathEnvironmentsForLineno{flalign}%
-  \patchBothAmsMathEnvironmentsForLineno{alignat}%
-  \patchBothAmsMathEnvironmentsForLineno{gather}%
-  \patchBothAmsMathEnvironmentsForLineno{multline}%
-  }
+\journal{Physics of Earth and Planetary Interiors}
 
@@
 
-\label{firstpage}
-
-\maketitle
-
-\begin{summary}
+\begin{frontmatter}
+
+%% Title, authors and addresses
+
+%% use the tnoteref command within \title for footnotes;
+%% use the tnotetext command for theassociated footnote;
+%% use the fnref command within \author or \address for footnotes;
+%% use the fntext command for theassociated footnote;
+%% use the corref command within \author for corresponding author footnotes;
+%% use the cortext command for theassociated footnote;
+%% use the ead command for the email address,
+%% and the form \ead[url] for the home page:
+%% \title{Title\tnoteref{label1}}
+%% \tnotetext[label1]{}
+%% \author{Name\corref{cor1}\fnref{label2}}
+%% \ead{email address}
+%% \ead[url]{home page}
+%% \fntext[label2]{}
+%% \cortext[cor1]{}
+%% \address{Address\fnref{label3}}
+%% \fntext[label3]{}
+
+\title{Scaling rates of true polar wander in convecting planets and moons}
+
+%% use optional labels to link authors explicitly to addresses:
+%% \author[label1,label2]{}
+%% \address[label1]{}
+%% \address[label2]{}
+
+\author{Ian Rose\corref{cor1}\fnref{ref1}}
+\author{Bruce Buffett\fnref{ref1}}
+
+\fntext[ref1]{University of California, Berkeley}
+\cortext[cor1]{Corresponding author, \url{ian.rose@berkeley.edu}}
+
+\address{}
+
+\begin{abstract}
 Mass redistribution in the convecting mantle of a planet can cause perturbations in its moment of inertia tensor. 
@@
 Finally, we discuss implications for large-scale TPW in Earth's past, which has been suggested to be important for life and climate history.
-\end{summary}
-
-\begin{keywords}
-Earth rotation variations; Mantle processes; Dynamics: convection currents and mantle plumes; Planetary interiors; Numerical solutions; Paleomagnetism applied to tectonics.
-\end{keywords}
+\end{abstract}
+
+\begin{keyword}
+Rotational variations; polar wobble \sep
+Paleomagnetism \sep
+Models of interior structure \sep
+Planetary interiors \sep
+Convection currents and mantle plumes \sep
+\PACS 91.10.Nj \sep 91.25.Ng \sep 91.35.Cb \sep 91.45.Bg \sep 91.45.Fj
+%% keywords here, in the form: keyword \sep keyword
+
+%% PACS codes here, in the form: \PACS code \sep code
+
+%% MSC codes here, in the form: \MSC code \sep code
+%% or \MSC[2008] code \sep code (2000 is the default)
+
+\end{keyword}
+
+\end{frontmatter}
 
@@
 the spin axis trying to keep up with the instantaneous figure axis set by the beetles.
-Our analysis begins the answer the questions ``how big are the beetles?'' and ``how fast are they crawling?''
+Our analysis begins to answer the questions ``how big are the beetles?'' and ``how fast are they crawling?''
 
@@
 
-\begin{acknowledgments}
-This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for thoughtful reviews and constructive comments that greatly improved the paper.
-\end{acknowledgments}
-
-\bibliographystyle{gji}
-\bibliography{tpw_rate.bib}
+\section*{Acknowledgments}
+This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for careful and constructive comments that greatly improved the paper.
+
+\section*{Bibliography}
+\bibliographystyle{elsarticle-harv} 
+\bibliography{tpw_rate}
 
@@
 
-\label{lastpage}
-
 \end{document}
+\endinput

### 2016-07-11 highlights.tex
Add highlights document.

\documentclass{article}
\begin{document}

\section*{Highlights}

\begin{itemize}
\item We develop a framework for analyzing true polar wander (TPW) from a perspective of fluid dynamics.
\item We identify a small set of nondimensional numbers which characterize the rates of true polar wander for a convecting planet.
\item We develop scalings for both the driving forces of TPW and the planetary response in terms of the nondimensional numbers.
\item We present numerical simulations supporting our scaling analyses.
\end{itemize}

\end{document}

### 2016-07-18 tpw_rate.tex
Small fixes.
Light edits to help with parsing.

@@
 %% for the whole article with \linenumbers.
-\usepackage{lineno}
+%%\usepackage{lineno}
 
@@
 
-\linenumbers
+%%\linenumbers
+
+%%BEGINCLIP
 
@@
 As such, they do not vary independently, and first-order questions about the propensity for planets to experience TPW remain: how are rates of TPW expected to vary with the vigor of convection? 
-Are other planetary bodies more or less likely than Earth to experience TPW?  And are these rates expected to vary in Earth history?
+Are other planetary bodies more or less likely than Earth to experience TPW?  And are these rates expected to vary through Earth history?
 
@@
 On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
-In this case Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
-For a slowly convecting fluid such as a planetary mantle the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
+In this case, Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
+For a slowly convecting fluid, such as a planetary mantle, the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
 \begin{equation}
@@
 In our treatment of TPW, we will refer to three different reference frames:
-
-First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
-Second, there is the body-fixed (or geographic) frame.
+\begin{itemize}
+\item First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
+\item Second, there is the body-fixed (or geographic) frame.
 By definition, the rotation of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
@@
 as described in Section~\ref{sec:rotational_deformation}.
-Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
+\item Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
 which we will refer to as the ``$\mathbf{E}$-frame.'' 
@@
 by a rotation vector $\mitbf{\Psi}$ (see Section~\ref{sec:theta}).
+\end{itemize}
 
@@
 The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
-This, as well, may be parameterized using a viscoelastic Love number approach:
+This part of the moment of inertia may also be parameterized using a viscoelastic Love number approach:
 \begin{equation} 
@@
 
-As a warmup, consider the case where $\phi = 0$, then  
+As a warm-up, consider the case where $\phi = 0$, then  
 \begin{equation}
@@
 
-Now consider the more general case whhere $\phi \ne 0$.
+Now consider the more general case where $\phi \ne 0$.
 Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
@@
 \end{equation}
-The maximum polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).
-
-From Equation~\eqref{eq:simple_milankovitch} it is clear that the important quantities for estimating $\dot{\Theta}$ are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.
-They represent, respectively, the size of convective anomalies in the moment of inertia tensor 
-and the angular mismatch between the rotation axis and the principal axis of the convective moment. 
+The maximum true polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).
+
+From Equation~\eqref{eq:simple_milankovitch} it is clear that the important quantities for estimating the rate of true polar wander ($\dot{\Theta}$)
+are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.  
+They represent, respectively, the angular mismatch between the rotation axis and the principal axis of the convective moment 
+and the size of convective anomalies in the moment of inertia tensor. 
 The dynamics of mantle convection affects the rate of TPW in two ways; 
@@
 Furthermore, since our goal is to establish a scaling for TPW rate, we must identify a minimal set of nondimensional numbers which describe its physics.
-For simplicty we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.
+For simplicity we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.
 The equations for mass, momentum, and energy then read
@@
 The centrifugal term, although the largest of the terms neglected in typical mantle convection models,
-is generally small compared to gravitational forces (at least for Earthlike parameters),
+is generally small compared to gravitational forces (at least for Earth-like parameters),
 and likely does not have a strong influence on the style of convection.
-However, it is critical for determining the the size of the rotational bulge which must be moved around via TPW, 
+However, this term is critical for determining the size of the rotational bulge which must move relative to the mantle,
 and so we must include it to establish the connection between the linear and angular momentum equations.
@@
 
-In this case the dynamics can be characterized in terms of deviations from a reference hydrostatic state, 
+In this case, the dynamics can be characterized in terms of deviations from a reference hydrostatic state, 
 which includes the dynamic pressure $P^* = P - P_0$ and density perturbations $\delta \rho = \rho- \rho_0 = - \rho_0 \alpha (T-T_0)$.
@@
 \end{equation}
-This can be directly identified with Equation~\eqref{eq:quasistatic_liouville},
+This equation can be directly identified with Equation~\eqref{eq:quasistatic_liouville},
 and is a statement that a quasistatic body will rotate around the principal axis of its total moment of inertia.
@@
 
-It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
+It is useful to define a nondimensional eigenvalue difference, which may also be described as a normalized ``eigengap'': $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
 This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
@@
 Similarly, the convective forcing that drives changes in $\theta$ must also be a function these nondimensional parameters. 
-To be explicit in our discussion, we denote this driving force by ${\Psi}$.
+To be explicit in our discussion, we denote the angular velocity induced by forcing by ${\Psi}$ (see Figure~\ref{fig:reference_frames}).
 We consider relatively slowly rotating bodies here ($m \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
@@
 
-Fluctuations in the nonhydrostatic moment of inertia are due to temperature fluctuations in the mantle and their spatial structure:
+Fluctuations in the nonhydrostatic moment of inertia are caused by density variations 
+due to the temperature fluctuations in the mantle and their spatial structure:
 
@@
 As discussed in Section~\ref{sec:convective_moment}, these internal density loads may be convolved with their surface responses to get the convective moment of inertia $E_{ij}$.
-The surface response is a function of the viscosity structure of the planet and the wavelength and depth of the interal load.
+The surface response is a function of the viscosity structure of the planet and the wavelength and depth of the internal load.
 The factor $(1+k^L_f)$ is generally an order-one parameter 
 (strictly speaking, dynamic compensation usually makes it less than one, \citep[e.g.][]{richards1984geoid}). 
-For the purposes of scaling it is reasonable to neglect this multiplicative factor.
+For the purposes of scaling, it is reasonable to neglect this multiplicative factor.
 
@@
 where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
-$T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$:
+$T_{nlm}$ are the coefficients for the expansion which have been normalized by $\Delta T$:
 \begin{equation} 
@@
 This is a strong constraint, but it gives very little information about the distribution of power across the $T_{lmn}$.  
-We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow (relatively low $\mathrm{Ra}$) 
+We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow, (relatively low $\mathrm{Ra}$) 
 and that of chaotic flow (relatively high $\mathrm{Ra}$).
@@
 \end{equation}
-Other power spectra for the temperature field are possible. Isoviscous models tend to be ``bluer,'' and models with viscosity stratification tend to be ``redder'' \citep{richards1999polar}.
+Other power spectra for the temperature field are possible. Isoviscous models tend to be ``bluer,'' (dominated by small wavelengths) 
+and models with viscosity stratification tend to be ``redder'' (dominated by long wavelengths) \citep{richards1999polar}.
 Present day Earth seems to have a fairly ``red'' spectrum, with large low-degree seismic anomalies due to 
@@
 
-\subsection{An estimate for the angular mismatch $\theta$}
+\subsection{An estimate for the angular mismatch angle ($\theta$)}
 \label{sec:theta}
 
-Convection drives a mismatch $\theta$ between the current rotation axis and the principal axis of the convective moment. 
+Convection can drive both growth and decay in the mismatch angle ($\theta$) between the current rotation axis and the principal axis of the convective moment. 
 This parameter is crucial for the rate of TPW, as previously noted in Equation~\eqref{eq:scaled_rotation}.  
@@
 If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principal axes.
-This is illustrated schematically in Figure~\ref{fig:perturb}
+This is illustrated schematically in Figure~\ref{fig:perturb}.
+\citet{evans1998true} described a situation where the convective moment of inertia of Earth is prolate, with the
+maximum and intermediate moments of inertia nearly equal.
+He argued that perturbations in this case would result in TPW paths that are approximately
+on the great circle path between the intermediate and maximum axes.
+This behavior is a consequence of the instability described in Equation~\eqref{eq:kahan}.
 
@@
 
-In the limit that the eigengap becomes zero, the rotation of the pinciple axes can be arbitrary. 
+In the limit that the eigengap becomes zero, the rotation of the principle axes can be arbitrary. 
 This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}, where 
@@
 When the eigengap gets small, the misfit angle becomes much larger, and the rate of polar wander becomes much faster.
-At $\sim$1 Gyr the eigengap goes to zero, and the misfit angle goes to approximately $90^\circ$, 
+At $\sim$1 Gyr into the model run the eigengap goes to zero, and the misfit angle goes to approximately $90^\circ$, 
 an IITPW event. But there are several other events where the eigengap becomes small, and there are 
@@
 as it has a lower characteristic gap between principal moments and the gap is much more likely to go to zero.
+
+\citet{tsai2007theoretical} suggested that interchanges in the principal moments are unlikely to produce large
+TPW events, since (1) during an inertial interchange the total moment of inertia anomaly is small, and
+(2) the TPW event can only asymptotically approach 90$^\circ$.
+This is true for the case of a single interchange. However, we argue that this is overly restrictive
+for our generalization of IITPW.
+Our numerical experiments show that in a vigorously convecting system the principal moments can
+interchange, or almost interchange, quite rapidly, and the perturbation sizes can be large soon after an interchange.
+Furthermore, the location of the principal axes need not stay put during a TPW event,
+which can lead to extended polar wandering of distances greater than 90$^\circ$.
 
@@
 The preceding results clarify the complex relationship between the Rayleigh number, $m$, and the rates of TPW for a convecting planet.
-As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principal axes of the convective moment of inertia. 
+As mantle convection redistributes mass in the planet's interior, the spin axis moves around to stay aligned with the principal axes of the convective moment of inertia. 
 There is a constant competition between growth of the mismatch angle $\theta$ through 
@@
 \begin{equation}
-\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 6^\circ / \mathrm{Myr}
-\end{equation}
-which is similar to the timescales discussed by \citet{cambiotti2011new}, and within the range 
-suggested by paleomagnetic observations.
+\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 6^\circ / \mathrm{Myr},
+\end{equation}
+which corresponds to about 66 cm/yr 90$^\circ$ from the TPW axis.
+This is similar to the rates discussed by \citet{cambiotti2011new}, (though our rate is larger since we allow for the
+possibility of a larger mismatch angle), and is within the range suggested by some interpretations of paleomagnetic data.
 The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
@@
 though the response to internal variations in the moment of inertia becomes more limited 
-(and potentially richer, as in the oscillitory motions suggested by \citet{creveling2012mechanisms}).
+(and potentially richer, as in the oscillatory motions suggested by \citet{creveling2012mechanisms}).
 
@@
 
-\section*{Acknowledgments}
-This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for careful and constructive comments that greatly improved the paper.
-
-\section*{Bibliography}
-\bibliographystyle{elsarticle-harv} 
-\bibliography{tpw_rate}
+%\section*{Acknowledgments}
+%This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for careful and constructive comments that greatly improved the paper.
+%We thank the Computational Infrastructure for Geodynamics (geodynamics.org) which is funded by the National Science Foundation under award NSF-0949446.
+
+%\section*{Bibliography}
+%\bibliographystyle{elsarticle-harv} 
+%\bibliography{tpw_rate}
 
@@
 principal moments, $(\lambda_3-\lambda_1)$, $(\lambda_3-\lambda_2)$ and $(\lambda_2-\lambda_1)$.
-These quantities may be rewritten in terms of degree-two real sperical harmonics \citep[e.g.][]{dahlen1999theoretical}.
+These quantities may be rewritten in terms of degree-two real spherical harmonics \citep[e.g.][]{dahlen1999theoretical}.
 The relevant (fully normalized) harmonics are, in Cartesian coordinates:
@@
 
-It is useful to define unnormalized verions of the spherical harmonics as intermediate products:
+It is useful to define unnormalized versions of the spherical harmonics as intermediate products:
 \begin{equation}
@@
 \end{equation}
-Up to the normalization constants, these expressions are idential to multipole expansions, 
+Up to the normalization constants, these expressions are identical to multipole expansions, 
 picking out the degree-two part of the density field laterally, and low-order polynomials radially.
@@
 \fi
-Dividing Equation~\eqref{eq:degree_two_moments_of_temperature} by $I_0$ and nodimensionalizing 
+Dividing Equation~\eqref{eq:degree_two_moments_of_temperature} by $I_0$ and nondimensionalizing 
 the integrals results in a factor of $\Gamma = \alpha \Delta T$ and a normalized set of 
@@
 
+%%ENDCLIP
+
 \end{document}

### 2016-07-18 frontmatter.tex
Add frontmatter.

--- 2016-06-08 tpw_rate.tex
@@
-% gjilguid2e.tex
+\chapter{Rates of true polar wander on convecting planets}
+\label{chap:intro}
 
-%\documentclass[extra,onecolumn]{gji}
-\documentclass[extra,mreferee]{gji}
-\usepackage{timet}
-\usepackage{url}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{color,soul}
-\usepackage{lineno}
-
-\title[TPW Rates]
-  {Scaling rates of true polar wander in convecting planets and moons}
-\author[I. Rose and B. Buffett]
-  {Ian Rose$^1$ and Bruce A. Buffett$^1$ \\
-  $^1$ Department of Earth \& Planetary Science, University of California, Berkeley, CA 94720, USA.  E-mail: ian.rose@berkeley.edu
-  }
-\date{}
-\pagerange{\pageref{firstpage}--\pageref{lastpage}}
-\volume{}
-\pubyear{}
-
-\let\leqslant=\leq
-
-%in case I want to add detail to any derivation
-\newif\ifdetail
-\detailfalse
-%\detailtrue
-
-\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
-  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
-  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
-  \renewenvironment{#1}%
-    {\linenomath\csname old#1\endcsname}%
-    {\csname oldend#1\endcsname\endlinenomath}}% 
-  \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
-    \patchAmsMathEnvironmentForLineno{#1}%
-    \patchAmsMathEnvironmentForLineno{#1*}}%
-  \AtBeginDocument{%
-  \patchBothAmsMathEnvironmentsForLineno{equation}%
-  \patchBothAmsMathEnvironmentsForLineno{align}%
-  \patchBothAmsMathEnvironmentsForLineno{flalign}%
-  \patchBothAmsMathEnvironmentsForLineno{alignat}%
-  \patchBothAmsMathEnvironmentsForLineno{gather}%
-  \patchBothAmsMathEnvironmentsForLineno{multline}%
-  }
-
-\begin{document}
-
-\label{firstpage}
-
-\maketitle
-
-\begin{summary}
 Mass redistribution in the convecting mantle of a planet can cause perturbations in its moment of inertia tensor. 
@@
 Finally, we discuss implications for large-scale TPW in Earth's past, which has been suggested to be important for life and climate history.
-\end{summary}
-
-\begin{keywords}
-Earth rotation variations; Mantle processes; Dynamics: convection currents and mantle plumes; Planetary interiors; Numerical solutions; Paleomagnetism applied to tectonics.
-\end{keywords}
-
-\linenumbers
-
-\section{Introduction}
-\label{sec:intro}
-
-A rotating, quasistatic body like a planetary mantle will tend to spin about the axis of its maximum moment of inertia.
-Convection in a planetary mantle continuously redistributes mass, which can change the moment of inertia tensor, necessitating a change in the spin axis of the planet to conserve angular momentum, a process known as true polar wander (TPW).
-
-TPW was first considered in detail by \citet{darwin1887influence}, and the theory has been subsequently developed by many \citep[e.g.][]{munk1960rotation, goldreich1969some, ricard1993polar}. 
-Despite this, the ability of internal mass anomalies to drive large-scale TPW remains controversial. 
-Paleomagnetic data have been interpreted to require up to $3^\circ-12^\circ$/Myr rates of TPW \citep{mitchell2011sutton}, but the ability of the mantle to respond at such rates has been questioned \citep{tsai2007theoretical}.
-
-The primary uncertainties in assigning a maximum TPW rate to a convecting planet are the size of convective anomalies, which drive the rotational adjustment, and the viscosity structure of the mantle, which retards it.
-These two uncertainties are not unrelated: they are both expected to be functions of the geometric and material properties of the mantle.
-As such, they do not vary independently, and first-order questions about the propensity for planets to experience TPW remain: how are rates of TPW expected to vary with the vigor of convection? 
-Are other planetary bodies more or less likely than Earth to experience TPW?  And are these rates expected to vary in Earth history?
-
-These questions suggest that an approach rooted in dimensional analysis and fluid dynamics can clarify the rates and magnitudes of TPW.
-Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment of inertia variations \citep[e.g.][]{tsai2007theoretical, creveling2012mechanisms}.
-\citet{richards1999polar} coupled thermal convection models to a polar wander model, but did not address in detail the scaling relationships between the two.
-
-Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle,  which we support with numerical simulations.
-
-\section{Rotational dynamics}
-\label{sec:rotational_dynamics}
-
-\subsection{The Liouville equation}
-\label{sec:liouville}
-Conservation of angular momentum for a torque-free system in a rotating reference frame requires
-
-\begin{equation}
-\frac{d \mathbf{H}}{dt} + \mitbf{\Omega} \times \mathbf{H} = 0
-\label{eq:euler}
-\end{equation}
-where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity vector.
-On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
-In this case Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
-For a slowly convecting fluid such as a planetary mantle the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
-\begin{equation}
-\mitbf{\Omega } (t)\times ( \mathbf{I}(t) \cdot \mitbf{\Omega }(t) ) = 0.
-\label{eq:quasistatic_liouville}
-\end{equation}
-Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for $\mathbf{I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
-In practice, this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
-The moment of inertia tensor in Equation~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
-of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
-Here we are interested in contributions from mantle convection, so we restrict our attention to these processes.
-
-For mantle convection problems the moment of inertia tensor is commonly separated into three parts \citep{sabadini1981pleistocene, spada1992excitation}:
-\begin{equation}
-I_{ij}(t) = I_0 \delta_{ij} + J_{ij}(t) + E_{ij}(t)
-\label{eq:separation}
-\end{equation}
-where $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
-If we plug this decomposition into Equation~\eqref{eq:quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
-\begin{equation}
-\mathbf{\Omega} \times (\mathbf{J} \cdot \mitbf{\Omega}) = -\mitbf{\Omega} \times ( \mathbf{E} \cdot \mitbf{\Omega} ).
-\label{eq:disequilibrium}
-\end{equation}
-This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches
-of the convective part of the moment of inertia ($\mathbf{E}$) and the rotational deformation part of the moment of inertia ($\mathbf{J}$).
-Our goal is to characterize this balance from a perspective of scaling and fluid dynamics.
-
-\subsection{A note on reference frames}
-\label{sec:reference_frames}
-True polar wander can be described in different reference frames, and this choice is fundamentally an arbitrary one.
-However, certain aspects of the physics can be made much simpler by an appropriate choice of the reference frame.
-In our treatment of TPW, we will refer to three different reference frames:
-
-First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
-Second, there is the body-fixed (or geographic) frame.
-By definition, the rotation of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
-A terrestrial no-net-rotation or hotspot reference frame are common choices for the body-fixed frame.
-Treatments of gravitational or rotational deformation of a planet are most naturally expressed in the body-fixed frame,
-as described in Section~\ref{sec:rotational_deformation}.
-Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
-which we will refer to as the ``$\mathbf{E}$-frame.'' 
-Redistribution of mantle mass anomalies due to convection changes the principal axes of $\mathbf{E}$, 
-causing the $\mathbf{E}$-frame to slowly rotate with respect to the geographic frame. We denote this drift 
-by a rotation vector $\mitbf{\Psi}$ (see Section~\ref{sec:theta}).
-
-These reference frames and vectors are illustrated in Figure~\ref{fig:reference_frames}.
-
-\subsection{Rotational deformation}
-\label{sec:rotational_deformation}
-
-The part of the moment of inertia due to the elastic rotational deformation in the body-fixed frame 
-is traditionally related to the degree-two part of the gravity field via MacCullagh's formula \citep{munk1960rotation}:
-\begin{equation}
-J_{ij} = \frac{k a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right)
-\label{eq:elastic_deformation}
-\end{equation}
-where $k$ is an elastic Love number, $a$ is the semimajor axis of the planet, and $G$ is the gravitational constant.
-This result may be extended to a viscoelastic rheology via the viscoelastic correspondence principle \citep[e.g.][]{peltier1974impulse}:
-\begin{equation}
-J_{ij} = \frac{k(t) a^5}{3 G} * \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right)
-\label{eq:viscoelastic_deformation}
-\end{equation}
-where $k$ is now a time-dependent viscoelastic Love number which is convolved with the time-dependent rotation vector.
-
-The infinite-time limit of Equation~\eqref{eq:viscoelastic_deformation} for a 
-constant rotation vector around the $z$-axis implies
-\begin{equation}
-\begin{aligned}
-J_{zz} = C = &\frac{2}{3} \frac{k_f a^5 \Omega^2}{3 G} \\
-J_{xx} = J_{yy} = A = -&\frac{1}{3} \frac{k_f a^5 \Omega^2}{3 G} \\
-\end{aligned}
-\end{equation}
-where $C$ and $A$ are the polar and equatorial moments of inertia, respectively, 
-and $k_f$ is the fluid limit of $k$. 
-We can solve for $k_f$ in terms of $C-A$:
-\begin{equation}
-k_f = \frac{3 G (C-A)}{\Omega^2 a^5}.
-\label{eq:fluid_love}
-\end{equation}
-This fluid-limit representation of the deformation does not allow for any disequilibrium between $J_{ij}$ and $E_{ij}$,
-so it does not permit TPW.
-
-\citet{ricard1993polar} obtain an approximation to Equation~\eqref{eq:viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.
-This introduces a new parameter, termed $T_1$, which can be seen as a weighted relaxation time for the system.  
-This simple approximation in the Laplace domain allows for an analytical transformation back into the time domain, and neglecting second order terms in $\dot{\Omega}$ we find:
-\begin{equation}
-J_{ij} = \frac{k_f a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right) -
- \frac{k_f a^5 T_1}{3G} \left(\dot{\Omega}_i \Omega_j + \Omega_i \dot{\Omega}_j - \frac{2}{3} \Omega_q \dot{\Omega}_q \delta_{ij} \right).
-\label{eq:rotational_deformation}
-\end{equation}
-The two terms of this equation have simple interpretations.  
-The first term corresponds to the fluid limit of rotational deformation (in the absence of any long-term elastic strength).  
-The second term represents the lag in the moment of inertia due to the viscous adjustment of the rotational bulge, where $T_1$ is the characteristic time constant for this adjustment.
-Since the first term represents the fluid limit of rotational deformation, it automatically satisfies Equation~\eqref{eq:quasistatic_liouville}, and hence does not contribute to the polar wander problem.
-
-\subsection{The convective moment of inertia}
-\label{sec:convective_moment}
-
-The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
-This, as well, may be parameterized using a viscoelastic Love number approach:
-\begin{equation} 
-\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
-\end{equation}
-where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, 
-and $C_{ij}$ is the moment of inertia due solely to the internal load.
-Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:
-\begin{equation}
-E_{ij} = (1+k^L_f) C_{ij}.
-\end{equation}
-An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
-A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
-permits more general treatments of mantle rheology.
-
-\subsection{Rate of true polar wander}
-\label{sec:tpw_rate}
-
-We are in a position to address the rates of true polar wander for a given convective moment $\mathbf{E}$.
-A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
-Substituting Equation~\eqref{eq:rotational_deformation} into Equation~\eqref{eq:disequilibrium} we find
-\begin{equation}
-\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right).
-\label{eq:liouville}
-\end{equation}
-Introducing a unit vector $\mitbf{\omega} = \mitbf{\Omega}/\|\mitbf{\Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for $\dot{\mitbf{\omega}}$:
-\begin{equation}
- \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right].
-\label{eq:liouville_ode}
-\end{equation}
-Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
-This correspondence permits useful insights (see below).
-Let $\dot{\Theta} = \vert \dot{\mitbf{\omega}} \vert$ denote the rate of polar wander.
-Evaluating the scalar product $\dot{\Theta}^2 = \dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}}$ gives
-\begin{equation}
-\dot{\Theta}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
-\label{eq:polar_wander_rate_tensor}
-\end{equation}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{figures/reference_frames.pdf}
-\caption{Relevant vectors and angles for the TPW analysis.
-The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively.
-Since the choice of geographic axes is arbitrary, we assume that at this instant the $\mathbf{E}$-frame and the geographic frames are colocated (though at future times they will not be).
-The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $\mathbf{e}_3$-axis.
-For illustration, we assume the longitude $\phi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane.
-True polar wander moves the rotation axis towards the $\mathbf{e}_3$-axis.
-We denote this motion by $\dot{\mitbf{\omega}}$ in the geographic frame.
-However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$-frame.
-This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$, which is defined by colatitude $\beta$ and longitude $\gamma$.
-The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$-frame.
-Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Sections~\ref{sec:reference_frames} and \ref{sec:theta}}
-\label{fig:reference_frames}
-\end{figure*}
-
-Our goal is to quantify the polar wander rate $\dot{\Theta}$ due to a convective perturbation $\mathbf{E}$ in the moment of inertia. 
-The rate defined by Equation~\eqref{eq:polar_wander_rate_tensor} is expressed in the body-fixed geographic frame, 
-which is also the reference frame used to measure TPW. 
-However, the physics of the right-hand-side of Equation~\eqref{eq:polar_wander_rate_tensor} 
-is more naturally expressed in the reference frame of the principal axes of $\mathbf{E}$. 
-In general the $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, so the time derivative of $\omega$ is different in these two frames.
-However, Equation~\eqref{eq:polar_wander_rate_tensor} is a scalar equation, which means it is invariant to rotations.
-Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
-with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of 
-$\mitbf{\omega}$ in the $\mathbf{E}$-frame with colatitude $\theta$ and longitude $\phi$ (see Figure~\ref{fig:reference_frames}).
-Plugging this description of $\mitbf{\omega}$ into Equation~\eqref{eq:polar_wander_rate_tensor},
-and after some tedious algebra, we find
-\ifdetail
-
-And here is the tedious algebra.  Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$.
-We are interested in the rate of true polar wander, which is the angular change of the pole
-per unit time. We can transform Equation~\eqref{eq:liouville_ode} into a scalar equation by dotting it
-with $\dot{\mitbf{\omega}}$, since we are interested in the magnitude of the rate, rather than
-the direction at the moment.
-
-\begin{equation}
-\begin{aligned}
-\dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}} &= 
-A^2 \left[ \mathbf{E}\cdot \mitbf{\omega} - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)\mitbf{\omega}\right]
-\cdot \left[ \mathbf{E}\cdot \mitbf{\omega} - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)\mitbf{\omega}\right] \\
-\dot{\mitbf{\omega}}^2  &= 
-A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)\cdot \left(\mathbf{E}\cdot \mitbf{\omega}\right)
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right) \left( \mitbf{\omega}\cdot \mathbf{E} \cdot \mitbf{\omega} \right) 
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right) \left( \mitbf{\omega}\cdot \mathbf{E} \cdot \mitbf{\omega} \right) 
- + \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 \mitbf{\omega}\cdot \mitbf{\omega} 
-\right]\\
-\dot{\mitbf{\omega}}^2  &= 
-A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2.
-\right]\\
-\end{aligned}
-\end{equation}
-Since this is a scalar equation, we can arbitrarily choose a coordinate system in which to evaluate it.
-It is most convenient to choose the principal axes of $\mathbf{E}$, where
-\begin{equation}
-\mathbf{E} = 
-\begin{bmatrix}
-\lambda_1 & 0 & 0 \\
-0 & \lambda_2 & 0 \\
-0 & 0 & \lambda_3 \\
-\end{bmatrix}
-\end{equation}
-\begin{equation}
-\mitbf{\omega} = 
-\begin{bmatrix}
-\sin{\theta} \cos{\phi} \\
-\sin{\theta} \sin{\phi} \\
-\cos{\theta} \\
-\end{bmatrix}.
-\end{equation}
-
-As a warmup, consider the case where $\phi = 0$, then  
-\begin{equation}
-\mitbf{\omega} = 
-\begin{bmatrix}
-\sin{\theta}  \\
-0 \\
-\cos{\theta} \\
-\end{bmatrix}
-\end{equation}
-We can evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
-\begin{equation}
-\mathbf{E}\cdot \mitbf{\omega} = 
-\begin{bmatrix}
-\lambda_1 \sin{\theta} \\
-0 \\
-\lambda_3 \cos{\theta}
-\end{bmatrix}
-\end{equation}
-
-\begin{equation}
-\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
-\end{equation}
-
-\begin{equation}
-\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
-\lambda_1 \sin^2{\theta} + \lambda_3 \cos^2{\theta}
-\end{equation}
-
-\begin{equation}
-\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
-\lambda_1^2 \sin^4{\theta} + \lambda_3^2 \cos^4{\theta} + 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta}
-\end{equation}
-Therefore, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
-\begin{equation}
-\begin{aligned}
-\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
- \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
- &= \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
- - \lambda_1^2 \sin^4{\theta} - \lambda_3^2 \cos^4{\theta} - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
- &= \lambda_1^2 \sin^2{\theta} (1 - \sin^2{\theta}) + \lambda_3^2 \cos^2{\theta} (1 - \cos^2{\theta})
-  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
- &= \lambda_1^2 \sin^2{\theta} \cos^2{\theta} + \lambda_3^2 \cos^2{\theta} \sin^2{\theta}
-  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
- &= (\lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \sin^2{\theta} \cos^2{\theta} \\
- &= \frac{1}{2} (\lambda_3 - \lambda_1)^2 \sin{2 \theta} \\
-\end{aligned}
-\end{equation}
-
-Now consider the more general case whhere $\phi \ne 0$.
-Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
-\begin{equation}
-\mathbf{E}\cdot \mitbf{\omega} = 
-\begin{bmatrix}
-\lambda_1 \sin{\theta} \cos{\phi}\\
-\lambda_2 \sin{\theta} \sin{\phi} \\
-\lambda_3 \cos{\theta}
-\end{bmatrix}
-\end{equation}
-
-\begin{equation}
-\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta}
-\end{equation}
-
-\begin{equation}
-\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
-\lambda_1 \sin^2{\theta} \cos^2{\phi} + \lambda_2 \sin^2{\theta} \sin^2{\phi} + \lambda_3 \cos^2{\theta}
-\end{equation}
-
-\begin{equation}
-\begin{aligned}
-\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
-&\lambda_1^2 \sin^4{\theta} \cos^4{\phi} + 
-\lambda_2^2 \sin^4{\theta} \sin^4{\phi} +
-\lambda_3^2 \cos^4{\theta} \\
-&+ 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
-&+ 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
-&+ 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
-\end{aligned}
-\end{equation}
-Again, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
-\begin{equation}
-\begin{aligned}
-\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
- \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
-&=
-\lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta} \\
-&-\lambda_1^2 \sin^4{\theta} \cos^4{\phi} - 
-\lambda_2^2 \sin^4{\theta} \sin^4{\phi} -
-\lambda_3^2 \cos^4{\theta} \\
-&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
-&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
-&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
-&=
-\sin^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} \right] \\
-&- \sin^2{\theta} \left[ \lambda_1^2 \sin^2{\theta} \cos^2{\phi}(1-\sin^2{\phi}) + \lambda_2^2 \sin^2{\theta} \sin^2{\phi}(1-\cos^2{\phi}) \right] \\
-&+ \lambda_3^2 \cos^2{\theta}(1-\cos^2{\theta}) \\
-&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
-&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
-&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
-&=
-\sin^2{\theta} \left[\lambda_1^2 \cos^2{\phi} (1 - \sin^2{\theta}) + \lambda_2^2 \sin^2{\phi} (1-\sin^2{\theta}) \right] \\
-&+ \sin^2{\theta} \left[ \lambda_1^2 \sin^2{\theta} \cos^2{\phi}\sin^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi}\cos^2{\phi} \right] \\
-&+ \lambda_3^2 \cos^2{\theta}\sin^2{\theta} \\
-&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
-&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
-&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
-&=
-\sin^2{\theta}\cos^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} \right] \\
-&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left[ \lambda_1^2 + \lambda_2^2 -2 \lambda_1\lambda_2 \right] \\
-&+ \lambda_3^2 \cos^2{\theta}\sin^2{\theta} \\
-&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
-&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
-&=
-\sin^2{\theta}\cos^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} + \lambda_3^2 - 2 \lambda_1 \lambda_3 \cos^2{\phi} - 2 \lambda_2 \lambda_3 \sin^2{\phi} \right] \\
-&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
-&=
-\sin^2{\theta}\cos^2{\theta} \left[( \lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \cos^2{\phi} + (\lambda_2^2 + \lambda_3^2 - 2 \lambda_2 \lambda_3 ) \sin^2{\phi} \right] \\
-&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
-&=
-\sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] \\
-&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
-&=
-\sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right]
-+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
-\end{aligned}
-\end{equation}
-
-At long last, we get
-\begin{equation}
-\begin{aligned}
-\frac{\dot{\mitbf{\omega}}^2}{A^2}  = 
-&\frac{1}{4}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
-&\frac{1}{4}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \\
-\end{aligned}
-\end{equation}
-
-\else
-\fi
-
-\begin{equation}
-\begin{aligned}
-\dot{\Theta}^2 = 
-&\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
-&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
-\label{eq:polar_wander_rate}
-\end{aligned}
-\end{equation}
-This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.
-Two special cases of this equation are of note. First, if $\dot{\Theta}^2$ is evaluated on the octahedral plane
-(the plane with direction cosines all $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
-then it can be expressed in terms of the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator 
-($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
-\begin{equation}
-\begin{aligned}
-\dot{\Theta}^2
-&= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
-&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2}. \\
-\end{aligned}
-\end{equation}
-The second invariant of the stress deviator is commonly used in the theories of elasticity and plasticity
-as a convenient scalar approximation of the shear stress in a system.
-The second invariant of the moment of inertia deviator can be similarly viewed as a scalar estimate of the ``rotational stress.''
-
-The second special case is if the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_3$ plane ($\phi = 0$).
-Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
-\begin{equation}
-\vert \dot{\Theta} \vert = 
-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1).
-\label{eq:simple_milankovitch}
-\end{equation}
-The maximum polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).
-
-From Equation~\eqref{eq:simple_milankovitch} it is clear that the important quantities for estimating $\dot{\Theta}$ are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.
-They represent, respectively, the size of convective anomalies in the moment of inertia tensor 
-and the angular mismatch between the rotation axis and the principal axis of the convective moment. 
-The dynamics of mantle convection affects the rate of TPW in two ways; 
-it controls $\lambda_i$ and it represents the mechanism that drives the angular mismatch. 
-The value of $\theta$ at any time also depends on the relaxation of the rotation axis back toward the principal axis of $\mathbf{E}$. 
-We focus on the dynamics of the convective contribution in the next section.
-
-\section{Internal dynamics}
-\label{sec:internal}
-
-Mantle convection and rotational dynamics of planetary bodies are usually considered separately, yet the processes are based on a common set of governing equations. 
-As such, some extra care must be taken to ensure that the equations we consider are self-consistent. 
-Furthermore, since our goal is to establish a scaling for TPW rate, we must identify a minimal set of nondimensional numbers which describe its physics.
-For simplicty we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.
-The equations for mass, momentum, and energy then read
-
-\begin{equation}
-\nabla \cdot \mathbf{u} = 0
-\label{eq:conserve_mass}
-\end{equation}
-
-\begin{equation}
-\begin{aligned}
-- \nabla P + \eta \nabla^2 \mathbf{u} =  \rho \mathbf{g} -  \rho \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}
-\label{eq:navier_stokes}
-\end{aligned}
-\end{equation}
-
-\begin{equation}
-\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \kappa \nabla^2 T
-\label{eq:energy}
-\end{equation}
-where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
-The vector $\mathbf{g}$ is the gravitational acceleration, 
-which defined in terms of a gravitational potential $V$ by $\mathbf{g} = -\nabla V$. 
-The gravitational potential obeys 
-\begin{equation}
-\nabla^2 V = 4 \pi G \rho.
-\end{equation}
-where $G$ is the gravitational constant. 
-For the purposes of scaling, we assume that the magnitude of the gravitational acceleration, $g_0$, is the approximately constant (as is the case for Earth's mantle).
-In addition we use the simple equation of state
-\begin{equation}
-\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right).
-\label{eq:eos}
-\end{equation}
-to define the density.
-The remaining parameters are defined in Table~\ref{tab:parameters}.
-Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure
-(in the latter case the boundary conditions on $P$ must be modified).
-The centrifugal term, although the largest of the terms neglected in typical mantle convection models,
-is generally small compared to gravitational forces (at least for Earthlike parameters),
-and likely does not have a strong influence on the style of convection.
-However, it is critical for determining the the size of the rotational bulge which must be moved around via TPW, 
-and so we must include it to establish the connection between the linear and angular momentum equations.
-
-Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires four nondimensional numbers to characterize it
-(a fifth one, defined by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations).
-Convenient choices for these numbers are listed in Table~\ref{tab:nondim}, along with approximate Earth-like values for them.
-
-Two dimensionless parameters have a prominent role in our scaling analysis. 
-The first is the Rayleigh number, which characterizes the  vigor of convection. 
-The second is the ratio of centrifugal to gravitational forces.
-This nondimensional number does not have a uniformly agreed-upon name: 
-it has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, 
-and in the geodesy community has commonly been termed $m$ \citep[e.g.][]{nakiboglu1982hydrostatic, chambat2010flattening}, which we adopt here.
-Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers that characterize those effects (e.g., the Prandtl and dissipation numbers).
-It would be straightforward to include them, but they do not affect the overall treatment of this scaling.
-
-In this case the dynamics can be characterized in terms of deviations from a reference hydrostatic state, 
-which includes the dynamic pressure $P^* = P - P_0$ and density perturbations $\delta \rho = \rho- \rho_0 = - \rho_0 \alpha (T-T_0)$.
-In addition, we expect deviations in the figure of the planet from its hydrostatic shape, 
-which we denote by $V = V_H + \Delta V$, where $V_H$ is the hydrostatic figure and $\Delta V$ is the deviation. 
-Our rationale for this decomposition is simply that the hydrostatic pressure is most naturally defined in the (oblate) hydrostatic configuration. 
-The introduction of $\Delta V$ requires another nondimensional number to characterize it, 
-and we find that the quantity $\Gamma \equiv \alpha \Delta T$ is convenient.
-Finally, we define $\mitbf{\Omega} = \Omega_0 \mitbf{\omega}$, where $\mitbf{\omega}$ is a unit vector in the direction of $\mitbf{\Omega}$.
-
-By definition the hydrostatic reference state is a solution to Equation~\eqref{eq:navier_stokes} where there is no flow:
-
-\begin{equation}
-\begin{aligned}
-- \nabla P_0 =  \rho_0 \mathbf{g} -  \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}.
-\label{eq:hydrostatic}
-\end{aligned}
-\end{equation}
-Nondimensionalizing with the parameters in Table~\ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
-\begin{equation}
-\begin{aligned}
- - \nabla P^* + \; \nabla^2 \mathbf{u} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra} \; m \; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0.
-\end{aligned}
-\end{equation}
-
-We can explicitly draw the connection between the angular and linear momentum equations by returning to the dimensional
-Equation~\eqref{eq:navier_stokes}, crossing it with $\mathbf{r}$ and integrating over the volume of the mantle:
-\begin{equation}
-\begin{aligned}
--\int_V \mathbf{r} \times \nabla P \,dV + \int_V \eta \mathbf{r} \times \nabla^2 \mathbf{u} \,dV - \int_V \rho \mathbf{r} \times \mathbf{g} \,dV 
-   + \int_V \rho \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}  \,dV &= 0. \\
-\end{aligned}
-\end{equation}
-
-The first three terms represent pressure, viscous torques, and gravitational torques on the mantle.  
-Convection in the outer core, atmospheres, and oceans is not strong enough to provide significant pressure and viscous torques over geologic timescales, and a self-gravitating body cannot self-torque \citep{braginsky1995equations}.
-Therefore we can neglect those terms, and we are left with
-
-\begin{equation}
-\begin{aligned}
-\int_V \rho \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV &= 0 \\
-\label{eq:quasistatic_angular_momentum}
-\end{aligned}
-\end{equation}
-which may be rewritten via the Jacobi identity to find
-\ifdetail
-\begin{equation}
-\begin{aligned}
-- \int_V \rho (\mitbf{\Omega}\times \mathbf{r}) \times ( \mathbf{r} \times \mitbf{\Omega} ) \,dV + 
-- \int_V \rho \mitbf{\Omega}\times(\mitbf{\Omega}\times \mathbf{r}) \times \mathbf{r} \,dV &= 0 \\
-\end{aligned}
-\end{equation}
-\fi
-\begin{equation}
-\begin{aligned}
- \mitbf{\Omega} \times \int_V \rho \mathbf{r} \times (\mitbf{\Omega}\times \mathbf{r}) \,dV &= 0. \\
-\end{aligned}
-\end{equation}
-This can be directly identified with Equation~\eqref{eq:quasistatic_liouville},
-and is a statement that a quasistatic body will rotate around the principal axis of its total moment of inertia.
-
-We now seek to evaluate Equation~\eqref{eq:quasistatic_angular_momentum} in the perturbed, convecting state.
-Hydrostatic balance (Equation~\eqref{eq:hydrostatic}) ensures that the integral over the reference shape $V_H$ vanishes when $\rho = \rho_0$.
-Nonzero contributions arise from perturbations in the density field or from perturbations in the shape.
-To make this dependence explicit, we split the shape into the reference volume $V_H$ and perturbations from it $\Delta V$,
-and use Equation~\eqref{eq:eos} to define density perturbations.
-Substituting this decomposition into Equation~\eqref{eq:quasistatic_angular_momentum} brings the integral into the form
-\begin{equation}
-\begin{aligned}
-&\int_{V_H} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV + 
-\int_{V_H} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV +  \\
-&\int_{\Delta V} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV + 
-\int_{\Delta V} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 0.  \\
-\end{aligned}
-\end{equation}
-As previously noted, the first term of this equation is zero due to the hydrostatic equation.  
-The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma\equiv \alpha \Delta T$.
-Removing these, we find
-\begin{equation}
-\begin{aligned}
-\int_{\Delta V} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 
--\int_{V_H} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV.
-\end{aligned}
-\end{equation}
-This equation may be identified with Equation~\eqref{eq:disequilibrium}, 
-where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of inertia with the spin axis (right side).
-Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table~\ref{tab:nondim}.
-
-\ifdetail
-\subsection{Buckingham Pi theorem and nondimensionalization}
-\label{sec:buckingham}
-The Buckingham Pi theorem (essentially an application of the rank-nullity theorem) allows us to determine the number of nondimensional numbers for this problem.
-We count the number of fundamental units for the problem, in this case mass (kg), length (m), time (s), and temperature (K), as well as the number of parameters in use for this problem.  
-These parameters are listed in Table~\ref{tab:parameters}, and subtracting the ...
-
-\begin{table}
-\centering
-\caption{Nondimensionalizations for variables, primed variables denote dimensional variables}
-\label{tab:nondim_convert}
-\begin{tabular}{@{}lll}
-$x^\prime$ &=& $R \;\; x$ \\
-$t^\prime$ &=& $R^2/\kappa \;\; t$ \\
-$T^\prime$ &=& $\Delta T \;\; T$ \\
-$\rho^\prime$ &=& $\rho_0 \;\; \rho$\\
-$u^\prime$ &=& $\kappa/R \;\; u$ \\
-$P^\prime$ &=& $\eta \kappa/R^2 \;\; P$ \\
-$\Omega^\prime$ &=& $\Omega_0 \;\; \Omega$ \\
-$g^\prime$ &=& $g_0 \;\; g$
-\end{tabular}
-\end{table}
-
-If we plug these nondimensionalizations into the governing equations we find that the incompressibility constraint remains unchanged.  
-The temperature equation and equation of state become
-\begin{equation}
-\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \nabla^2 T
-\end{equation}
-
-\begin{equation}
-\rho = 1 - \alpha \Delta T \;\;  T.
-\end{equation}
-We may furthermore define a hydrostatic reference state where the density is the reference density everywhere and the velocity is zero, i.e.
-
-\begin{equation}
- \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}= - \nabla P_0 + \rho_0 \mathbf{g}.
-\end{equation}
-Subtracting this from the momentum equation and defining a dynamic pressure $P^* = P - P_0$, we find
-
-\begin{equation}
-\begin{aligned}
- - & \Omega_0^2  \alpha  \Delta T T R \; \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} = - \eta \kappa / R^2 \; \nabla P^* \\ 
-&+ \eta \kappa / R^2 \; \nabla \cdot \left( \eta \epsilon (\mathbf{u}) \right) - \alpha \Delta T g_0 \; \mathbf{g}
-\end{aligned}
-\end{equation}
-
-\begin{equation}
-\begin{aligned}
- - & \mathrm{Ra} \; m \; T \; \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} = - \nabla P^* \\ 
-&+ \; \nabla \cdot \left( \eta \epsilon (\mathbf{u}) \right) - \mathrm{Ra} \; T \; \mathbf{g}
-\end{aligned}
-\end{equation}
-
-\fi
-
-\begin{table}
-\centering
-\caption{Parameters for rotating mantle convection}
-\label{tab:parameters}
-\begin{tabular}{@{}lcc}
-Symbol & Definition\\
-\hline
-$R_i$ & inner radius \\
-$R$ & outer radius \\
-$G$ & gravitational constant \\
-$V$ & gravitational potential \\
-$M$ & mass of the planet \\
-$\Omega_0$ & reference rotation rate \\
-$\eta$ & viscosity \\
-$\kappa$ & thermal diffusivity \\
-$\alpha$ &  thermal expansivity \\ 
-$g_0$ & reference gravity \\
-$I_0$ & reference moment of inertia \\
-$T_0$ & reference temperature \\
-$\rho_0$ & reference density \\ 
-$\Delta T$ & temperature drop across mantle
-\end{tabular}
-\end{table}
-
-\begin{table}
-\centering
-\caption{Nondimensional numbers with approximate Earth-like values}
-\label{tab:nondim}
-\begin{tabular}{@{}lcccc}
-Symbol &  Name & Definition & Approximate value \\
-\hline
-Ra & Rayleigh &  $\rho_0 g_0 \alpha \Delta T R^3/\eta \kappa$ & $10^7$\\
-$m$ & Froude & $\Omega_0^2 R/g_0 = \Omega_0^2 R^3/GM$ & $10^{-3}$ \\
-A & aspect ratio & $R_i/R$ & $0.54$ \\
-$\Gamma$ & density deficit &$ \alpha \Delta T$ & $10^{-2}$ \\
-\end{tabular}
-\end{table}
-
-\section{Scaling}
-\label{sec:scaling}
-
-Having drawn the connection between the angular momentum equation (Section~\ref{sec:rotational_dynamics}) and the linear momentum 
-equation (Section~\ref{sec:internal}), we would like to apply the scaling of the 
-linear momentum equation to the problem of TPW. We would therefore like to rewrite Equation~\eqref{eq:simple_milankovitch}
-in terms of the nondimensional numbers identified in Section~\ref{sec:internal}.
-Specifically, we will need estimates for $(C-A)$, $T_1$, $(\lambda_3-\lambda_1)$, and $\theta$.
-
-It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
-This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
-The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is 
-proportional to the ratio of rotational to gravitational forces \citep{munk1960rotation}:
-\begin{equation}
-(C-A) \sim I_0 m.
-\end{equation}
-
-The time constant $T_1$ is a viscous relaxation time of the planetary mantle.
-It is frequently represented as a weighted average of the different relaxation modes \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
-\begin{equation}
-T_1 \sim \frac{ \eta }{ \rho_0 g_0 R} = \frac{R^2}{\kappa} \frac{\Gamma}{\mathrm{Ra} }.
-\end{equation}
-Plugging these scalings into Equation~\eqref{eq:simple_milankovitch} we find
-\begin{equation}
-\frac{\kappa}{R^2} \; \dot{\Theta} \sim - \frac{\mathrm{Ra}}{\Gamma m} \Lambda_{31} \sin{2 \theta}.
-\label{eq:scaled_rotation}
-\end{equation}
-
-At this point we do not have estimates for the characteristic magnitudes of $\Lambda_{ij}$ or $\theta$, both of which are crucial for predicting characteristic rates of TPW.
-They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principal axis of the convective moment.
-
-The quantity $\Lambda_{ij}$ should be a function of our nondimensional parameters $\mathrm{Ra}$, $m$, and $\Gamma$. 
-Similarly, the convective forcing that drives changes in $\theta$ must also be a function these nondimensional parameters. 
-To be explicit in our discussion, we denote this driving force by ${\Psi}$.
-We consider relatively slowly rotating bodies here ($m \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
-In this regime, therefore, we neglect the dependence on $m$, 
-and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra}, \Gamma)$ and $\Psi(\mathrm{Ra}, \Gamma)$.
-
-\subsection{An estimate for $\Lambda_{ij}$}
-\label{sec:lambda}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{figures/eigengap.pdf}
-\caption{ Left: Time series of the normalized difference between moments $\Lambda_{21} = (\lambda_2 - \lambda_1)/I_0$ for convection in a 2D annulus at several different Rayleigh numbers.
-As the Rayleigh number increases, the average value of the relative moment decreases due to less low-degree coherence in the temperature structure.
-Right:  average value of $\Lambda_{21}$ for the different Rayleigh numbers.
-Also shown is a line with slope $\mathrm{Ra}^{-2/3}$, which is predicted from the scaling analysis (the exponent is $-2/3$ instead of $-1$ due to the reduced dimensionality of the simulations).}
-\label{fig:eigengap}
-\end{figure*}
-
-Fluctuations in the nonhydrostatic moment of inertia are due to temperature fluctuations in the mantle and their spatial structure:
-
-\begin{equation}
-C_{ij} = \int_{V_H} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) \,dV
-\label{eq:temperature_fluctuations}
-\end{equation}
-As discussed in Section~\ref{sec:convective_moment}, these internal density loads may be convolved with their surface responses to get the convective moment of inertia $E_{ij}$.
-The surface response is a function of the viscosity structure of the planet and the wavelength and depth of the interal load.
-The factor $(1+k^L_f)$ is generally an order-one parameter 
-(strictly speaking, dynamic compensation usually makes it less than one, \citep[e.g.][]{richards1984geoid}). 
-For the purposes of scaling it is reasonable to neglect this multiplicative factor.
-
-For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix~\ref{appendix:moments}).
-Therefore, an estimate for one specifies the other.
-We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
-where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
-$T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$:
-\begin{equation} 
-T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi).
-\label{eq:T_series}
-\end{equation}
-Inserting the temperature expansion~\eqref{eq:T_series} into Equation~\eqref{eq:temperature_fluctuations} results in 
-a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
-Orthogonality of the basis functions for the expansion means that the integral for $C_{ij}$ picks out degree-two 
-spherical harmonics in the lateral dimensions, and only the lowest few radial functions $R_n(r)$.
-Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
-We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$ (see Appendix~\ref{appendix:moments} for more detail): 
-\begin{equation}
-\Lambda_{31} \sim \Gamma T_{\text{degree-two}}.
-\end{equation}
-The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by
-
-\begin{equation}
-\max \left( \displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \right) = 1.
-\end{equation}
-This is a strong constraint, but it gives very little information about the distribution of power across the $T_{lmn}$.  
-We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow (relatively low $\mathrm{Ra}$) 
-and that of chaotic flow (relatively high $\mathrm{Ra}$).
-The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
-the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is a broadening of the spatial 
-and temporal spectra \citep{mclaughlin1982transition}.
-
-At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes which are largely influenced by the aspect ratio.
-This spectrum may or may not have a lot of power in the degree-two modes, and does not depend strongly on time.
-
-At high Rayleigh number we expect the shortest lengthscales to be limited by the effects of thermal diffusion,
-which tends to wipe out thermal heterogeneity at small scales.  Consequently, there will be little power in modes
-with shorter lengthscales than that allowed by diffusion. 
-Therefore we expect, to a good approximation, that the infinite sum in Equation~\eqref{eq:T_series} can be truncated at some maximum wavenumber, set by the smallest lengthscale $d$:
-\begin{equation}
-n_{\text{max}}, l_{\text{max}}, m_{\text{max}} \sim \frac{R}{d}.
-\label{eq:max_wavenumber}
-\end{equation}
-Strictly speaking, convective mixing can produce smaller scales, but the power in these scales
-is greatly reduced by diffusion.
-Thus total number of modes that are accessible to the system are
-
-\begin{equation}
-N_{\text{modes}} = n_{\text{max}} \times l_{\text{max}} \times m_{\text{max}} \sim \left( \frac{R}{d} \right)^{3}.
-\end{equation}
-The value of each $T_{lmn}(t)$ will in general be some complex function of time, but for a given style of convection we expect there to be some average value.
-For chaotic flow the power should be spread out amongst the modes accessible to it.
-We may make the hypothesis that each of the modes are roughly as likely as any of the others, which implies
-
-\begin{equation}
-T_{\text{degree-two}}(t) \sim \frac{1}{N_{\text{modes}}} \sim \left( \frac{d}{R}\right)^3.
-\end{equation}
-
-Any of a number of scaling laws can provide an estimate for the characteristic length scale of a convecting system which may depending on rheology, geometry, or density structure.
-The simplest, based on boundary layer theory \citep{turcotte1967finite}, finds $d/R \sim \mathrm{Ra}^{-1/3}$.
-This scaling is roughly a measure of the diffusive lengthscale for the timescale of a convective overturn, consistent
-with the cutoff in Equation~\eqref{eq:max_wavenumber}.
-It thus furnishes us with an estimate of the power in 
-the degree-two part of the field as a function of Rayleigh number:
-
-\begin{equation}
-T_{\text{degree-two}}(t) \sim \mathrm{Ra}^{-1}.
-\label{eq:degree_two_of_ra}
-\end{equation}
-
-We performed a series of numerical simulations of mantle convection at different Rayleigh numbers to test this scaling.
-We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII82},
-which allows for flexible implementation of different rheologies, geometries, and postprocessors.
-In order to test a wide range of Rayleigh numbers, we ran the simulations in a 2D annulus, tracking the eigenvalues of the 
-moment of inertia tensor and integrating Equation~\eqref{eq:liouville_ode} in time.
-For the 2D simulations there is a reduced dimensionality when calculating the number of modes,
-so $N_\text{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\text{degree-two}} \sim \mathrm{Ra}^{-2/3}$, 
-which is shown as a dashed line in Figure~\ref{fig:eigengap}.
-This result has a simple interpretation.
-As the Rayleigh number of the system increases, the smallest lengthscale of convective features gets smaller.
-The total power in the temperature field is spread across a larger spectrum, leaving less total power for the degree-two part, which is what drives TPW.
-
-With an estimate for the power in the degree-two part of the temperature field, we may finally estimate $\Lambda_{ij}$:
-\begin{equation}
-\Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }.
-\label{eq:lambda_estimate}
-\end{equation}
-Other power spectra for the temperature field are possible. Isoviscous models tend to be ``bluer,'' and models with viscosity stratification tend to be ``redder'' \citep{richards1999polar}.
-Present day Earth seems to have a fairly ``red'' spectrum, with large low-degree seismic anomalies due to 
-Cenozoic subduction history and the lower mantle LLSVPs \citep{dziewonski2010mantle}.
-Nevertheless, at high Rayleigh number the expectation is that the power will be distributed across many length scales.
-
-\subsection{An estimate for the angular mismatch $\theta$}
-\label{sec:theta}
-
-Convection drives a mismatch $\theta$ between the current rotation axis and the principal axis of the convective moment. 
-This parameter is crucial for the rate of TPW, as previously noted in Equation~\eqref{eq:scaled_rotation}.  
-Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
-
-Two processes control the evolution of $\theta$. Its growth occurs through perturbations in the convective moment and 
-its decay occurs by relaxation of the pole towards the maximum moment of inertia.
-We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
-from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
-the rotation of the body-fixed frame relative to inertial space in Equation~\eqref{eq:euler}).
-The $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, 
-which we describe by the rotation vector $\mathbf{\Psi}$. 
-(To be explicit in our discussion we let $\mitbf{\Psi}$ define the rotation of the $\mathbf{E}$-frame in the geographic frame, 
-although this choice is arbitrary as long as we are consistent). The time derivatives of $\mitbf{\omega}$ in the two frames are related by 
-\begin{equation}
-\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
-\end{equation}
-where primes are used to define quantities in the $\mathbf{E}$-frame. Rearranging gives 
-\begin{equation}
-\dot{\mitbf{\omega}}^{\prime} = \dot{\mitbf{\omega}} - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
-\end{equation}
-We now substitute for $\dot{\mitbf{\omega}}$ from Equation~\eqref{eq:liouville_ode}, 
-making the additional assumption that the geographic and $\mathbf{E}$-frames are momentarily aligned. 
-This assumption is not a serious restriction because our goal is simply to relate the time derivatives of $\mitbf{\omega}$ in the two frames, 
-rather than track the evolution of $\mitbf{\omega}$ over time. The expression for $\dot{\mitbf{\omega}}$ becomes
-\begin{equation}
- \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
-\label{eq:liouville_ode_new_frame}
-\end{equation}
-The advantage of expressing the rate of TPW in the $\mathbf{E}$-frame is that the entire equation
-may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
-As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that 
-$\mitbf{\omega}^{\prime}$ lies in $\mathbf{e}_1^{\prime}-\mathbf{e}_3^{\prime} $ plane (as shown in Figure~\ref{fig:reference_frames}). 
-In this case the orientation of $\mitbf{\omega}^\prime$ is defined by colatitude $\theta$ and longitude $\phi=0$. 
-The colatitude $\theta$ is precisely the angular misalignment that specifies the rate of TPW in Equation~\eqref{eq:simple_milankovitch}.
-We furthermore specify the orientation of $\Psi$
-using colatitude $\beta$ and longitude $\gamma$ (see Figure~\ref{fig:reference_frames}).
-
-The time derivative of the unit vector $\mitbf{\omega}^{\prime}$ defines the changing orientation of the rotation axis in the $\mathbf{E}$-frame. 
-We can express this changing orientation in terms of changes in angles $\theta$ and $\phi$. The change in colatitude is
-\begin{equation}
-\dot{\theta} = 
--\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\beta \sin\gamma.
-\label{eq:theta_competition}
-\end{equation}
-where $\mitbf{\Psi}$ is expressed in terms of its amplitude $|\mitbf{\Psi}|$ and its orientation angles $\beta$ and $\gamma$.
-Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
-The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it via
-relative motion between the geographic frame and the principal axes of the $\mathbf{E}$-frame.
-Comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
-then the rate of TPW $\vert \dot{\Theta} \vert$ is identical to $\vert \dot{\theta} \vert$.
-Conversely, when the rate of TPW is very slow (i.e. $\vert \dot{\Theta} \vert \sim 0$), 
-then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$, so the evolution of $\theta$ is driven mostly by $\mitbf{\Psi}$. 
-For the purposes of scaling we will omit the orientation factor $\sin \beta \sin \gamma$ and focus on $|\mitbf{\Psi}|$.
-In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 
-at least occasionally, be larger than its decay.
-
-An estimate for $\vert \mitbf{\Psi} \vert$ must concern the stability of the convective moment of inertia. 
-Convection is continuously redistributing mass throughout the mantle, which will perturb the convective moment of inertia.  
-If the convective moment is stable to small perturbations, then $\vert \mitbf{\Psi} \vert$ should be small, so
- $\theta$ will never have the opportunity to grow.
-However, if $\mathbf{E}$ is not stable to perturbations, then $|\mitbf{\Psi}|$ can be large and $\theta$ can grow up to $90^\circ$. 
-Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of changes in the orientation of the principle axes. 
-This may be done by application of a theorem due to \citet{davis1970rotation}.
-
-Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor after some time interval $\Delta t$, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.
-The corresponding rotation of the principal axes is defined by the angle $\xi$.
-A bound on $\xi$ is given by
-\begin{equation}
-\vert \sin(2 \xi) \vert \le \frac{ 2 \vert \delta \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
-\label{eq:kahan}
-\end{equation} 
-That is to say, if there is a large difference between the eigenvalues, this stabilizes axes to perturbations.  
-If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principal axes.
-This is illustrated schematically in Figure~\ref{fig:perturb}
-
-We may arrive at a simple estimate for $|\mitbf{\Psi}|$  
-by differentiating Equation~\eqref{eq:kahan}, tentatively holding $\vert \lambda_i - \lambda_j \vert$ fixed:
-\begin{equation}
-\vert \mitbf{\Psi} \vert \sim \vert \dot{\xi} \vert \le \frac{ \vert \dot{\delta} \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
-\label{eq:grow_perturbation}
-\end{equation} 
-As noted previously, $|\mitbf{\Psi}|\sim |\dot\theta |$ when the rate of TPW is small and the angular misalignment $\theta$ is small. 
-Consequently, we can interpret Equation~\eqref{eq:grow_perturbation} as a bound on how fast $\theta$ can grow.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{figures/perturb.pdf}
-\caption{Graphical demonstration of the $\sin{2 \xi}$ theorem of \citet{davis1970rotation}.
-Two spheroidal bodies with eigenvalues $\lambda_2 > \lambda_1$ start out with the rotation axis $\Omega$ aligned with the $\lambda_2$ axis.
-However, on the left the eigengap $\lVert \lambda_2 - \lambda_1 \rVert$ is large, while on the right it is small.
-A negative mass perturbation is instantaneously added to both bodies, which effects a small rotation of the principal axes on the left, but a large one on the right.}
-\label{fig:perturb}
-\end{figure*}
-
-The quantity $(\lambda_i-\lambda_j)$ which appears in the denominator of these equations is precisely the same as the quantity which we estimated in the previous section to scale with $\sim \mathrm{Ra}^{-1}$.
-Therefore, as the Rayleigh number increases, the characteristic gap between the eigenvalues of the convective moment becomes smaller.  
-Additionally, the timescale of fluctuations in these values goes down. 
-Overall, this makes the principal axes of high Rayleigh number systems much less stable.
-This is consistent with the result of \citet{richards1999polar}.
-
-In the limit that the eigengap becomes zero, the rotation of the pinciple axes can be arbitrary. 
-This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}, where 
-the mismatch angle is $90^\circ$. However, the eigengap does not need to be zero for there to be large displacements polar wander, 
-and if it does go to zero, the wander does not need to be $90^\circ$.
-
-Figure~\ref{fig:misfit} shows a representative timeseries for annular convection, where 
-we track the eigenvalues of the moment of inertia and integrate Equation~\eqref{eq:liouville_ode} in time.
-Since it is a 2D model, the spin axis may be represented by a single angle.
-When the eigengap gets small, the misfit angle becomes much larger, and the rate of polar wander becomes much faster.
-At $\sim$1 Gyr the eigengap goes to zero, and the misfit angle goes to approximately $90^\circ$, 
-an IITPW event. But there are several other events where the eigengap becomes small, and there are 
-still large TPW events associated with them.
-For example, at $\sim$0.3 Gyr the eigengap dips, with an associated large TPW event, even though
-there is technically no interchange of the axes.
-
-We suggest, then, that $90^\circ$ IITPW events are simply a special case of a broad class of events which can occur
-when the principal moments are close to each other. 
-Furthermore, a more vigorously convecting planet is much more likely to experience rapid TPW events
-as it has a lower characteristic gap between principal moments and the gap is much more likely to go to zero.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
-\caption{Top: Time series of principal moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.
-Bottom: Time series of spin axis and mismatch angle $\theta$.
-When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger.
-At $\sim$1 Gyr the gap goes to zero and there is a nearly $90^\circ$ TPW event, with $\sim80^\circ$ degrees of polar wander in $\sim$30 Myr.
-However, there are several other large TPW events which happen when the eigengap is small.}
-\label{fig:misfit}
-\end{figure*}
-
-\section{Discussion}
-\label{sec:discussion}
-
-The preceding results clarify the complex relationship between the Rayleigh number, $m$, and the rates of TPW for a convecting planet.
-As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principal axes of the convective moment of inertia. 
-There is a constant competition between growth of the mismatch angle $\theta$ through 
-Equation~\eqref{eq:grow_perturbation} and its relaxation through TPW.
-\citet{goldreich1969some} envisioned an analogy of beetles crawling on the surface of the globe, with
-the spin axis trying to keep up with the instantaneous figure axis set by the beetles.
-Our analysis begins the answer the questions ``how big are the beetles?'' and ``how fast are they crawling?''
-
-We may plug in the estimate for $\Lambda_{ij}$ (Equation~\eqref{eq:lambda_estimate}) into Equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
-\begin{equation}
-\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{1}{m} \sin{2 \theta}.
-\label{eq:simplest_milankovitch}
-\end{equation}
-Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
-Response timescales for relaxation of the mismatch angle go down at high Rayleigh numbers.
-At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
-That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
-Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
-However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
-With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
-\begin{equation}
-\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
-\label{eq:simplest_milankovitch_defer_scaling}
-\end{equation}
-In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
-$d(\mathrm{Ra})$ will still result in a weak dependence of the prefactor in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} 
-on the Rayleigh number.
-
-We then suggest that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta$.
-Whereas the prefactor $\mathrm{Ra}^{1-3\beta}$ in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} has a weak dependence on $\mathrm{Ra}$, 
-the misalignment angle $\theta$ is expected to be strongly dependent on it.
-
-Indeed, we can identify two endmember behaviors of Equation~\eqref{eq:simplest_milankovitch_defer_scaling}.
-When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.
-This is the regime considered in \citet{steinberger1997changes}, \citet{roberts2007cause}, and \citet{zhong2007supercontinent}, and can be considered the ``slow TPW'' regime.
-When convection is more chaotic, however, there may be large excursions in $\theta$, 
-which are driven by large values for $|\mitbf{\Psi}|$ when $(\lambda_i - \lambda_j)$ is small, according to Equation~\eqref{eq:grow_perturbation}. 
-In this case a dramatic increase in the rate of TPW is possible.
-If $\theta=90^\circ$, this corresponds to IITPW \citep{kirschvink1997evidence}.  
-This, however, is a special case in the large $\theta$, ``fast TPW'' regime.
-
-As an example, we may consider the early Earth, when the mantle was presumably hotter and less viscous, leading to a 
-higher Rayleigh number. We would then predict that convection was more vigorous, leading to 
-a less stable $\theta(t)$, and thus more TPW and more frequent ``fast TPW'' events.
-
-For Cenozoic Earth we can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch}.
-Typical values for the time constant $T_1$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
-Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
-are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
-A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
-\citet{richards1997explanation} have argued that the convective planform of Earth has been stable for the last
-few hundred million years. On the other hand, we know that there have been large reorganizations of that planform 
-during Earth history, so this recent geologic stability may not hold in general \citep{evans2003true}.
-Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
-Allowing for such a large mismatch angle ($\theta = 45^\circ$) we may estimate the maximum polar wander rate
-\begin{equation}
-\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 6^\circ / \mathrm{Myr}
-\end{equation}
-which is similar to the timescales discussed by \citet{cambiotti2011new}, and within the range 
-suggested by paleomagnetic observations.
-The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
-which results in a corresponding uncertainty for the relaxation time $T_1$ and the maximum polar wander rate.
-
-Thus far we have restricted our discussion to planets with lithospheres lacking long-term elastic strength.
-For this case the long-time limit of the planetary figure is coaxial with the convective 
-moment of inertia. This assumption is not necessarily true in all cases.
-Earth's lithosphere is pervasively fractured and hydrated, and may not have much strength when subjected to rotational changes on geologic timescales.
-However, a planet with a stagnant lid (such as Mars) may have considerable strength, preventing the figure of the planet 
-from reaching the fluid limit of Equation~\eqref{eq:fluid_love}.
-
-The theory of TPW response for the case of elastic lithospheres has been developed in, among other places, 
-\citet{matsuyama2006rotational}, \citet{creveling2012mechanisms}, and \citet{chan2014time}.
-The formalism developed in Section~\ref{sec:scaling} can still be applied to this case, 
-though the response to internal variations in the moment of inertia becomes more limited 
-(and potentially richer, as in the oscillitory motions suggested by \citet{creveling2012mechanisms}).
-
-\section{Conclusion}
-We have developed a framework for discussing the rates of true polar wander for a convecting planet 
-from a perspective of scaling and fluid dynamics.
-We identified a small number of dimensionless parameters which describe the system, and showed how they affect the overall dynamics of the system.
-
-The most important parameters are the Rayleigh number and $m$, which acts as a damper to TPW.
-The dependence on the Rayleigh number is more complicated, since it is a control on both the forcing 
-of TPW and the response, which act in opposite directions.
-Overall, however, we expect that more vigorously convecting planets should be less rotationally stable, and experience more TPW.
-This perspective allows us to consider not only the polar wandering of Phanerozoic Earth,
-but also allows us to hypothesize about polar wandering during the Archean and Proterozoic, or 
-on other planetary bodies.
-
-\begin{acknowledgments}
-This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for thoughtful reviews and constructive comments that greatly improved the paper.
-\end{acknowledgments}
-
-\bibliographystyle{gji}
-\bibliography{tpw_rate.bib}
-
-\appendix
-
-\section{Degree-two moments}
-\label{appendix:moments}
-
-There is a connection between the moment of inertia of a rotating object and the degree-two density structure.
-The moment of inertia tensor may be written in index notation
-\begin{equation}
-I_{ij} = \int_V \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) dV
-\label{eq:inertia}
-\end{equation}
-where $\mathbf{r}$ is the Eulerian coordinate, $\rho$ is the density, and $V$ is the volume of the material.  
-It is useful to enter the principal axes of the moment of inertia:
-\begin{equation}
-\mathbf{I} = \mathbf{1} \begin{bmatrix}
-\lambda_1 \\ \lambda_2 \\ \lambda_3
-\end{bmatrix} = 
-\mathbf{1} \begin{bmatrix}
-\int_V \rho (y^2+z^z) dV\\
-\int_V \rho (x^2+z^2) dV\\
-\int_V \rho (x^2+y^2) dV
-\end{bmatrix}
-\end{equation}
-where $\mathbf{1}$ is the identity matrix, and $\lambda_1$, $\lambda_2$, and $\lambda_3$ are the principal moments.
-From Equation~\eqref{eq:polar_wander_rate} we see that the important quantities are the differences between the 
-principal moments, $(\lambda_3-\lambda_1)$, $(\lambda_3-\lambda_2)$ and $(\lambda_2-\lambda_1)$.
-These quantities may be rewritten in terms of degree-two real sperical harmonics \citep[e.g.][]{dahlen1999theoretical}.
-The relevant (fully normalized) harmonics are, in Cartesian coordinates:
-\begin{equation}
-\begin{aligned}
-Y_{20} &= \frac{1}{4} \sqrt{\frac{ 5}{\pi}} \frac{ 2 z^2 - x^2 - y^2}{r^2} \\ 
-Y_{22} &= \frac{1}{4} \sqrt{\frac{15}{\pi}} \frac{ x^2 - y^2}{r^2}.
-\end{aligned}
-\end{equation}
-Solving for $( \lambda_i - \lambda_j )$ in terms of these harmonics, we find
-\ifdetail
-
-It is useful to define unnormalized verions of the spherical harmonics as intermediate products:
-\begin{equation}
-\begin{aligned}
-Y_{20}^* &= 2 z^2 - x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{5} } r^2 Y_{20} \\ 
-Y_{22}^* &= x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{15} } r^2 Y_{22}.
-\end{aligned}
-\end{equation}
-These make it easy to rewrite the differences between the $\lambda_n$:
-\begin{equation}
-\begin{aligned}
-(\lambda_2 - \lambda_1) &= \int_V \rho Y_{22}^* \, dV \\
-(\lambda_3 - \lambda_1) &= \frac{1}{2} \int_V  \rho \left( Y_{22}^* - Y_{20}^* \right) dV \\
-(\lambda_3 - \lambda_2) &= -\frac{1}{2} \int_V \rho \left( Y_{22}^* + Y_{20}^*  \right) dV. \\
-\end{aligned}
-\end{equation}
-Substituting back in for the normalized spherical harmonics we find
-\fi
-\begin{equation}
-\begin{aligned}
-(\lambda_2 - \lambda_1) &= 4 \sqrt{\frac{\pi}{15} } \int_V   \rho r^2  Y_{22} \,dV \\
-(\lambda_3 - \lambda_1) &= 2 \sqrt{ \frac{\pi}{15} } \int_V  \rho r^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV \\
-(\lambda_3 - \lambda_2) &= -2 \sqrt{ \frac{\pi}{15} } \int_V  \rho r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV. \\
-\label{eq:degree_two_moments}
-\end{aligned}
-\end{equation}
-Up to the normalization constants, these expressions are idential to multipole expansions, 
-picking out the degree-two part of the density field laterally, and low-order polynomials radially.
-When density is a function of temperature, we can insert the equation of state, Equation~\eqref{eq:eos}, into 
-Equation~\eqref{eq:degree_two_moments} and integrate over a reference spherical volume $V_S$.
-This allows us to drop the terms which integrate to zero due to the orthogonality of spherical harmonics, and we are left with:
-\begin{equation}
-\begin{aligned}
-(\lambda_2 - \lambda_1) &= -4 \sqrt{\frac{\pi}{15} } \alpha \rho_0 \int_{V_S} T r^2  Y_{22} \,dV \\
-(\lambda_3 - \lambda_1) &= -2 \sqrt{ \frac{\pi}{15} } \alpha \rho_0 \int_{V_S} T r^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV \\
-(\lambda_3 - \lambda_2) &= 2 \sqrt{ \frac{\pi}{15} } \alpha \rho_0 \int_{V_S}  T r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV. \\
-\label{eq:degree_two_moments_of_temperature}
-\end{aligned}
-\end{equation}
-We normalize the differences in eigenvalues by the reference moment $I_0$:
-\ifdetail
-\begin{equation}
-\begin{aligned}
-\int_{V_S} \rho_0 \frac{2 r^2 - Y_{20}^*}{3} &= \int_{V_S} \rho_0 (x^2 + y^2) dV = I_0 \\
-\end{aligned}
-\end{equation}
-\begin{equation}
-I_0 = \frac{2}{3} \int_{V_S} \rho_0 r^2 \left(1 - 2 \sqrt{\frac{\pi}{5}} Y_{20}\right) \,dV
-\end{equation}
-\fi
-\begin{equation}
-I_0 = \frac{2}{3} \int_{V_S} \rho_0 r^2 \,dV.
-\end{equation}
-\ifdetail
-\begin{equation}
-\begin{aligned}
-\Lambda_{21} &= \alpha \Delta T \frac{-6 \sqrt{\frac{\pi}{15} } \int_{V_S} T^\prime r^2  Y_{22} \,dV}{\int_{V_S} r^2 \, dV} \\
-\Lambda_{31} &= \alpha \Delta T \frac{-3 \sqrt{ \frac{\pi}{15} } \int_{V_S} T^\prime r^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV}{\int_{V_S} r^2 \, dV}\\
-\Lambda_{32} &= \alpha \Delta T \frac{3 \sqrt{ \frac{\pi}{15} }  \int_{V_S}  T^\prime r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV}{\int_{V_S} r^2 \, dV} \\
-\end{aligned}
-\end{equation}
-\begin{equation}
-\begin{aligned}
-\Lambda_{21} &= \Gamma \frac{-6 \sqrt{\frac{\pi}{15} } \int_{V_S^\prime} T^\prime {r^\prime}^2  Y_{22} \,dV^\prime}{\int_{V_S^\prime} {r^\prime}^2 \,dV^\prime} \\
-\Lambda_{31} &= \Gamma \frac{-3 \sqrt{ \frac{\pi}{15} } \int_{V_S^\prime} T^\prime {r^\prime}^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV^\prime}{\int_{V_S} {r^\prime}^2 \, dV^\prime}\\
-\Lambda_{32} &= \Gamma \frac{3 \sqrt{ \frac{\pi}{15} }  \int_{V_S^\prime}  T^\prime {r^\prime}^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV^\prime}{\int_{V_S} {r^\prime}^2 \, dV^\prime} \\
-\end{aligned}
-\end{equation}
-\fi
-Dividing Equation~\eqref{eq:degree_two_moments_of_temperature} by $I_0$ and nodimensionalizing 
-the integrals results in a factor of $\Gamma = \alpha \Delta T$ and a normalized set of 
-degree-two coefficients for the temperature field, which we abbreviate as $T_{\text{degree-two}}$:
-\begin{equation}
-\Lambda_{ij} \sim \Gamma T_{\text{degree-two}}.
-\end{equation}
-
-\label{lastpage}
-
-\end{document}

### 2016-07-18 tpw_rate.tex
Dont use appendix for now.
Add short captions.

@@
 \includegraphics[width=0.9\textwidth]{figures/reference_frames.pdf}
-\caption{Relevant vectors and angles for the TPW analysis.
+\caption[Relevant vectors and angles for the TPW analysis.]{Relevant vectors and angles for the TPW analysis.
 The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively.
@@
 \includegraphics[width=0.9\textwidth]{figures/eigengap.pdf}
-\caption{ Left: Time series of the normalized difference between moments $\Lambda_{21} = (\lambda_2 - \lambda_1)/I_0$ for convection in a 2D annulus at several different Rayleigh numbers.
+\caption[Normalized difference between convective moments of inertia in the mantle.]{ Left: Time series of the normalized difference between moments $\Lambda_{21} = (\lambda_2 - \lambda_1)/I_0$ for convection in a 2D annulus at several different Rayleigh numbers.
 As the Rayleigh number increases, the average value of the relative moment decreases due to less low-degree coherence in the temperature structure.
@@
 \includegraphics[width=0.9\textwidth]{figures/perturb.pdf}
-\caption{Graphical demonstration of the $\sin{2 \xi}$ theorem of \citet{davis1970rotation}.
+\caption[Graphical demonstration of the $\sin{2 \xi}$ theorem of \citet{davis1970rotation}.] {Graphical demonstration of the $\sin{2 \xi}$ theorem of \citet{davis1970rotation}.
 Two spheroidal bodies with eigenvalues $\lambda_2 > \lambda_1$ start out with the rotation axis $\Omega$ aligned with the $\lambda_2$ axis.
@@
 \includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
-\caption{Top: Time series of principal moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.
+\caption[Principal moments and mismatch angle in a convecting matle.]{Top: Time series of principal moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.
 Bottom: Time series of spin axis and mismatch angle $\theta$.
@@
 
-\appendix
+%\appendix
 

### 2016-07-31 tpw_rate.tex
Fix appendices.
Break derivation into appendix.

@@
 %% \usepackage{amsthm}
+
+\usepackage{appendix}
 
@@
 
-%\appendix
+\begin{subappendices}
 
@@
 
+\section{Derivation of the polar wander rate equation}
+\label{appendix:scalar_liouville}
+
+Here is the tedious algebra for deriving Equation~\eqref{eq:polar_wander_rate}.  
+We begin with the scalar equation Equation~\eqref{eq:polar_wander_rate_tensor}.
+Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$:
+
+\begin{equation}
+\begin{aligned}
+\dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}} = \dot{\Theta}^2  &= 
+A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2.
+\right]\\
+\end{aligned}
+\end{equation}
+Since this is a scalar equation, we can arbitrarily choose a coordinate system in which to evaluate it.
+It is most convenient to choose the principal axes of $\mathbf{E}$, where
+\begin{equation}
+\mathbf{E} = 
+\begin{bmatrix}
+\lambda_1 & 0 & 0 \\
+0 & \lambda_2 & 0 \\
+0 & 0 & \lambda_3 \\
+\end{bmatrix}
+\end{equation}
+\begin{equation}
+\mitbf{\omega} = 
+\begin{bmatrix}
+\sin{\theta} \cos{\phi} \\
+\sin{\theta} \sin{\phi} \\
+\cos{\theta} \\
+\end{bmatrix}.
+\end{equation}
+As a warmup, consider the case where $\phi = 0$, then  
+\begin{equation}
+\mitbf{\omega} = 
+\begin{bmatrix}
+\sin{\theta}  \\
+0 \\
+\cos{\theta} \\
+\end{bmatrix}
+\end{equation}
+We can evaluate some of the terms of the scalar equation for $\dot{\Theta}^2$:
+\begin{equation}
+\mathbf{E}\cdot \mitbf{\omega} = 
+\begin{bmatrix}
+\lambda_1 \sin{\theta} \\
+0 \\
+\lambda_3 \cos{\theta}
+\end{bmatrix}
+\end{equation}
+
+\begin{equation}
+\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
+\lambda_1 \sin^2{\theta} + \lambda_3 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
+\lambda_1^2 \sin^4{\theta} + \lambda_3^2 \cos^4{\theta} + 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta}
+\end{equation}
+Therefore, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
+\begin{equation}
+\begin{aligned}
+\frac{\dot{\Theta}}{A^2}  &= 
+ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
+ &= \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
+ - \lambda_1^2 \sin^4{\theta} - \lambda_3^2 \cos^4{\theta} - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
+ &= \lambda_1^2 \sin^2{\theta} (1 - \sin^2{\theta}) + \lambda_3^2 \cos^2{\theta} (1 - \cos^2{\theta})
+  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
+ &= \lambda_1^2 \sin^2{\theta} \cos^2{\theta} + \lambda_3^2 \cos^2{\theta} \sin^2{\theta}
+  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
+ &= (\lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \sin^2{\theta} \cos^2{\theta} \\
+ &= \frac{1}{2} (\lambda_3 - \lambda_1)^2 \sin{2 \theta} \\
+\end{aligned}
+\end{equation}
+which is equivalent to Equation~\eqref{eq:simple_milankovitch}.
+
+Now consider the more general case where $\phi \ne 0$.
+Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
+\begin{equation}
+\mathbf{E}\cdot \mitbf{\omega} = 
+\begin{bmatrix}
+\lambda_1 \sin{\theta} \cos{\phi}\\
+\lambda_2 \sin{\theta} \sin{\phi} \\
+\lambda_3 \cos{\theta}
+\end{bmatrix}
+\end{equation}
+
+\begin{equation}
+\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
+\lambda_1 \sin^2{\theta} \cos^2{\phi} + \lambda_2 \sin^2{\theta} \sin^2{\phi} + \lambda_3 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\begin{aligned}
+\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
+&\lambda_1^2 \sin^4{\theta} \cos^4{\phi} + 
+\lambda_2^2 \sin^4{\theta} \sin^4{\phi} +
+\lambda_3^2 \cos^4{\theta} \\
+&+ 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
+&+ 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
+&+ 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
+\end{aligned}
+\end{equation}
+Again, plugging these into the scalar equation for $\dot{\Theta}^2$, we find:
+\begin{equation}
+\begin{aligned}
+\frac{\dot{\Theta}^2}{A^2}  &= 
+ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
+&=
+\lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta} \\
+&-\lambda_1^2 \sin^4{\theta} \cos^4{\phi} - 
+\lambda_2^2 \sin^4{\theta} \sin^4{\phi} -
+\lambda_3^2 \cos^4{\theta} \\
+&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
+&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
+&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi}. 
+\end{aligned}
+\end{equation}
+After a lot of simplification, we get
+\begin{equation}
+\begin{aligned}
+\frac{\dot{\Theta}^2}{A^2}  = 
+&\frac{1}{4}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
+&\frac{1}{4}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \\
+\end{aligned}
+\end{equation}
+which is Equation~\eqref{eq:polar_wander_rate}.
+
+\end{subappendices}
+
 %%ENDCLIP

### 2016-08-01 tpw_rate.tex
Address some comments from Nick.

--- 2016-07-18 tpw_rate.tex
@@
 %% for the whole article with \linenumbers.
-%%\usepackage{lineno}
+\usepackage{lineno}
 
@@
 
-%%\linenumbers
-
-%%BEGINCLIP
+\linenumbers
 
@@
 
-As a warm-up, consider the case where $\phi = 0$, then  
+As a warmup, consider the case where $\phi = 0$, then  
 \begin{equation}
@@
 
-Now consider the more general case where $\phi \ne 0$.
+Now consider the more general case whhere $\phi \ne 0$.
 Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
@@
 Furthermore, since our goal is to establish a scaling for TPW rate, we must identify a minimal set of nondimensional numbers which describe its physics.
-For simplicity we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.
+For simplicty we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.
 The equations for mass, momentum, and energy then read
@@
 The centrifugal term, although the largest of the terms neglected in typical mantle convection models,
-is generally small compared to gravitational forces (at least for Earth-like parameters),
+is generally small compared to gravitational forces (at least for Earthlike parameters),
 and likely does not have a strong influence on the style of convection.
@@
 where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
-$T_{nlm}$ are the coefficients for the expansion which have been normalized by $\Delta T$:
+$T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$:
 \begin{equation} 
@@
 If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principal axes.
-This is illustrated schematically in Figure~\ref{fig:perturb}.
-\citet{evans1998true} described a situation where the convective moment of inertia of Earth is prolate, with the
-maximum and intermediate moments of inertia nearly equal.
-He argued that perturbations in this case would result in TPW paths that are approximately
-on the great circle path between the intermediate and maximum axes.
-This behavior is a consequence of the instability described in Equation~\eqref{eq:kahan}.
+This is illustrated schematically in Figure~\ref{fig:perturb}
 
@@
 
-In the limit that the eigengap becomes zero, the rotation of the principle axes can be arbitrary. 
+In the limit that the eigengap becomes zero, the rotation of the pinciple axes can be arbitrary. 
 This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}, where 
@@
 as it has a lower characteristic gap between principal moments and the gap is much more likely to go to zero.
-
-\citet{tsai2007theoretical} suggested that interchanges in the principal moments are unlikely to produce large
-TPW events, since (1) during an inertial interchange the total moment of inertia anomaly is small, and
-(2) the TPW event can only asymptotically approach 90$^\circ$.
-This is true for the case of a single interchange. However, we argue that this is overly restrictive
-for our generalization of IITPW.
-Our numerical experiments show that in a vigorously convecting system the principal moments can
-interchange, or almost interchange, quite rapidly, and the perturbation sizes can be large soon after an interchange.
-Furthermore, the location of the principal axes need not stay put during a TPW event,
-which can lead to extended polar wandering of distances greater than 90$^\circ$.
 
@@
 though the response to internal variations in the moment of inertia becomes more limited 
-(and potentially richer, as in the oscillatory motions suggested by \citet{creveling2012mechanisms}).
+(and potentially richer, as in the oscillitory motions suggested by \citet{creveling2012mechanisms}).
 
@@
 
-%\section*{Acknowledgments}
-%This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for careful and constructive comments that greatly improved the paper.
-%We thank the Computational Infrastructure for Geodynamics (geodynamics.org) which is funded by the National Science Foundation under award NSF-0949446.
-
-%\section*{Bibliography}
-%\bibliographystyle{elsarticle-harv} 
-%\bibliography{tpw_rate}
+\section*{Acknowledgments}
+This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for careful and constructive comments that greatly improved the paper.
+We thank the Computational Infrastructure for Geodynamics (geodynamics.org) which is funded by the National Science Foundation under award NSF-0949446.
+
+\section*{Bibliography}
+\bibliographystyle{elsarticle-harv} 
+\bibliography{tpw_rate}
 
@@
 principal moments, $(\lambda_3-\lambda_1)$, $(\lambda_3-\lambda_2)$ and $(\lambda_2-\lambda_1)$.
-These quantities may be rewritten in terms of degree-two real spherical harmonics \citep[e.g.][]{dahlen1999theoretical}.
+These quantities may be rewritten in terms of degree-two real sperical harmonics \citep[e.g.][]{dahlen1999theoretical}.
 The relevant (fully normalized) harmonics are, in Cartesian coordinates:
@@
 
-It is useful to define unnormalized versions of the spherical harmonics as intermediate products:
+It is useful to define unnormalized verions of the spherical harmonics as intermediate products:
 \begin{equation}
@@
 \end{equation}
-Up to the normalization constants, these expressions are identical to multipole expansions, 
+Up to the normalization constants, these expressions are idential to multipole expansions, 
 picking out the degree-two part of the density field laterally, and low-order polynomials radially.
@@
 \fi
-Dividing Equation~\eqref{eq:degree_two_moments_of_temperature} by $I_0$ and nondimensionalizing 
+Dividing Equation~\eqref{eq:degree_two_moments_of_temperature} by $I_0$ and nodimensionalizing 
 the integrals results in a factor of $\Gamma = \alpha \Delta T$ and a normalized set of 
@@
 
-%%ENDCLIP
-
 \end{document}

### 2016-08-01 frontmatter.tex
Remove abstract from frontmatter.

@@
 \chapter{Rates of true polar wander on convecting planets}
-\label{chap:intro}
-
-Mass redistribution in the convecting mantle of a planet can cause perturbations in its moment of inertia tensor. 
-Conservation of angular momentum dictates that these perturbations can change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). 
-Although the existence of TPW on Earth is well-verified, its rate and magnitude over geologic time scales remain controversial. 
-Here we present scaling analyses and numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors. 
-For simple rotating convection, the most important parameters are the Rayleigh number, the rotation rate, and the size of relative density fluctuations (i.e.
-thermal expansivity times the temperature variations).
-We identify timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander. 
-These timescales, as well as the relative sizes of convective anomalies, control the rate and magnitude of TPW.
-This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and when they are likely to occur.
-Finally, we discuss implications for large-scale TPW in Earth's past, which has been suggested to be important for life and climate history.
+\label{chap:tpw_rate}

### 2016-08-06 tpw_rate.tex
Add some more discussion.

@@
 
-\linenumbers
+%%\linenumbers
 
@@
 If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principal axes.
-This is illustrated schematically in Figure~\ref{fig:perturb}
+This is illustrated schematically in Figure~\ref{fig:perturb}.
+\citet{evans1998} described a situation where the convective moment of inertia of Earth is prolate, with the
+maximum and intermediate moments of inertia nearly equal.
+He argued that perturbations in this case would result in TPW paths that are approximately
+on the great circle path between the intermediate and maximum axes.
+This behavior is a consequence of the instability described in Equation~\eqref{eq:kahan}.
 

### 2016-08-06 tpw_rate.tex
Fix reference.

@@
 This is illustrated schematically in Figure~\ref{fig:perturb}.
-\citet{evans1998} described a situation where the convective moment of inertia of Earth is prolate, with the
+\citet{evans1998true} described a situation where the convective moment of inertia of Earth is prolate, with the
 maximum and intermediate moments of inertia nearly equal.

### 2016-08-08 tpw_rate.tex
Reference chapter.

@@
 %%\linenumbers
+
+%%BEGINCLIP
 
@@
 An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
-A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
+In Chapter~\ref{chap:free_surface} we describe the implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT}.
 permits more general treatments of mantle rheology.
@@
 The centrifugal term, although the largest of the terms neglected in typical mantle convection models,
-is generally small compared to gravitational forces (at least for Earthlike parameters),
+is generally small compared to gravitational forces (at least for Earth-like parameters),
 and likely does not have a strong influence on the style of convection.
@@
 
-\section*{Acknowledgments}
-This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for careful and constructive comments that greatly improved the paper.
-We thank the Computational Infrastructure for Geodynamics (geodynamics.org) which is funded by the National Science Foundation under award NSF-0949446.
-
-\section*{Bibliography}
-\bibliographystyle{elsarticle-harv} 
-\bibliography{tpw_rate}
-
-\appendix
+%\section*{Acknowledgments}
+%This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for careful and constructive comments that greatly improved the paper.
+%We thank the Computational Infrastructure for Geodynamics (geodynamics.org) which is funded by the National Science Foundation under award NSF-0949446.
+
+%\section*{Bibliography}
+%\bibliographystyle{elsarticle-harv} 
+%\bibliography{tpw_rate}
+
+\begin{subappendices}
 
@@
 
+\section{Derivation of the polar wander rate equation}
+\label{appendix:scalar_liouville}
+
+Here is the tedious algebra for deriving Equation~\eqref{eq:polar_wander_rate}.  
+We begin with the scalar equation Equation~\eqref{eq:polar_wander_rate_tensor}.
+Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$:
+
+\begin{equation}
+\begin{aligned}
+\dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}} = \dot{\Theta}^2  &= 
+A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2.
+\right]\\
+\end{aligned}
+\end{equation}
+Since this is a scalar equation, we can arbitrarily choose a coordinate system in which to evaluate it.
+It is most convenient to choose the principal axes of $\mathbf{E}$, where
+\begin{equation}
+\mathbf{E} = 
+\begin{bmatrix}
+\lambda_1 & 0 & 0 \\
+0 & \lambda_2 & 0 \\
+0 & 0 & \lambda_3 \\
+\end{bmatrix}
+\end{equation}
+\begin{equation}
+\mitbf{\omega} = 
+\begin{bmatrix}
+\sin{\theta} \cos{\phi} \\
+\sin{\theta} \sin{\phi} \\
+\cos{\theta} \\
+\end{bmatrix}.
+\end{equation}
+As a warmup, consider the case where $\phi = 0$, then  
+\begin{equation}
+\mitbf{\omega} = 
+\begin{bmatrix}
+\sin{\theta}  \\
+0 \\
+\cos{\theta} \\
+\end{bmatrix}
+\end{equation}
+We can evaluate some of the terms of the scalar equation for $\dot{\Theta}^2$:
+\begin{equation}
+\mathbf{E}\cdot \mitbf{\omega} = 
+\begin{bmatrix}
+\lambda_1 \sin{\theta} \\
+0 \\
+\lambda_3 \cos{\theta}
+\end{bmatrix}
+\end{equation}
+
+\begin{equation}
+\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
+\lambda_1 \sin^2{\theta} + \lambda_3 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
+\lambda_1^2 \sin^4{\theta} + \lambda_3^2 \cos^4{\theta} + 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta}
+\end{equation}
+Therefore, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
+\begin{equation}
+\begin{aligned}
+\frac{\dot{\Theta}}{A^2}  &= 
+ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
+ &= \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
+ - \lambda_1^2 \sin^4{\theta} - \lambda_3^2 \cos^4{\theta} - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
+ &= \lambda_1^2 \sin^2{\theta} (1 - \sin^2{\theta}) + \lambda_3^2 \cos^2{\theta} (1 - \cos^2{\theta})
+  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
+ &= \lambda_1^2 \sin^2{\theta} \cos^2{\theta} + \lambda_3^2 \cos^2{\theta} \sin^2{\theta}
+  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
+ &= (\lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \sin^2{\theta} \cos^2{\theta} \\
+ &= \frac{1}{2} (\lambda_3 - \lambda_1)^2 \sin{2 \theta} \\
+\end{aligned}
+\end{equation}
+which is equivalent to Equation~\eqref{eq:simple_milankovitch}.
+
+Now consider the more general case where $\phi \ne 0$.
+Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
+\begin{equation}
+\mathbf{E}\cdot \mitbf{\omega} = 
+\begin{bmatrix}
+\lambda_1 \sin{\theta} \cos{\phi}\\
+\lambda_2 \sin{\theta} \sin{\phi} \\
+\lambda_3 \cos{\theta}
+\end{bmatrix}
+\end{equation}
+
+\begin{equation}
+\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
+\lambda_1 \sin^2{\theta} \cos^2{\phi} + \lambda_2 \sin^2{\theta} \sin^2{\phi} + \lambda_3 \cos^2{\theta}
+\end{equation}
+
+\begin{equation}
+\begin{aligned}
+\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
+&\lambda_1^2 \sin^4{\theta} \cos^4{\phi} + 
+\lambda_2^2 \sin^4{\theta} \sin^4{\phi} +
+\lambda_3^2 \cos^4{\theta} \\
+&+ 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
+&+ 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
+&+ 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
+\end{aligned}
+\end{equation}
+Again, plugging these into the scalar equation for $\dot{\Theta}^2$, we find:
+\begin{equation}
+\begin{aligned}
+\frac{\dot{\Theta}^2}{A^2}  &= 
+ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
+ - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
+&=
+\lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta} \\
+&-\lambda_1^2 \sin^4{\theta} \cos^4{\phi} - 
+\lambda_2^2 \sin^4{\theta} \sin^4{\phi} -
+\lambda_3^2 \cos^4{\theta} \\
+&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
+&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
+&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi}. 
+\end{aligned}
+\end{equation}
+After a lot of simplification, we get
+\begin{equation}
+\begin{aligned}
+\frac{\dot{\Theta}^2}{A^2}  = 
+&\frac{1}{4}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
+&\frac{1}{4}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \\
+\end{aligned}
+\end{equation}
+which is Equation~\eqref{eq:polar_wander_rate}.
+
+\end{subappendices}
+
+%%ENDCLIP
+
 \end{document}

### 2017-04-17 tpw_rate.tex
Spelling.
Edits from Bruce.

@@
 \begin{abstract}
-Mass redistribution in the convecting mantle of a planet can cause perturbations in its moment of inertia tensor. 
-Conservation of angular momentum dictates that these perturbations can change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). 
-Although the existence of TPW on Earth is well-verified, its rate and magnitude over geologic time scales remain controversial. 
+Mass redistribution in the convecting mantle of a planet causes perturbations in its moment of inertia tensor. 
+Conservation of angular momentum dictates that these perturbations change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). 
+Although the existence of TPW on Earth is firmly established, its rate and magnitude over geologic time scales remain controversial. 
 Here we present scaling analyses and numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors. 
-For simple rotating convection, the most important parameters are the Rayleigh number, the rotation rate, and the size of relative density fluctuations (i.e.
-thermal expansivity times the temperature variations).
-We identify timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander. 
+For simple rotating convection, we identify a set of dimensionless parameters that characterize true polar wander.
+We use these parameters to define  timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander.
 These timescales, as well as the relative sizes of convective anomalies, control the rate and magnitude of TPW.
-This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and when they are likely to occur.
-Finally, we discuss implications for large-scale TPW in Earth's past, which has been suggested to be important for life and climate history.
+This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and relates them to a  broader class of events that enable large and often rapid TPW.
+We expect these events to have been more frequent in Earth’s past.
 \end{abstract}
@@
 
-Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle,  which we support with numerical simulations.
+Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle.
+We identify a set of dimensionless parameters that fully characterize TPW and use numerical simulations to quantify the expected dependence on these parameters.
+Finally, we use the scaling results to make predictions about TPW over Earth's history and for other planetary bodies.
 
@@
 \end{equation}
-Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for $\mathbf{I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
+Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for the principle values of $\mathbf{I}(t)$;  the eigenvectors define the principal axes and the eigenvalues correspond to the principal moments of inertia.
+The rotation axis is expected to align with the axis for the largest moment.
 In practice, this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
 The moment of inertia tensor in Equation~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
-of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
-Here we are interested in contributions from mantle convection, so we restrict our attention to these processes.
+of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and the associated surface deflections.
+Here we are interested in those contributions that arise from mantle convection.
 
@@
 However, certain aspects of the physics can be made much simpler by an appropriate choice of the reference frame.
-In our treatment of TPW, we will refer to three different reference frames:
-\begin{itemize}
-\item First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
-\item Second, there is the body-fixed (or geographic) frame.
+In our treatment of TPW, we will refer to three different reference frames.  First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
+ Second, there is the body-fixed (or geographic) frame.
 By definition, the rotation of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
@@
 as described in Section~\ref{sec:rotational_deformation}.
-\item Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
+ Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
 which we will refer to as the ``$\mathbf{E}$-frame.'' 
 Redistribution of mantle mass anomalies due to convection changes the principal axes of $\mathbf{E}$, 
-causing the $\mathbf{E}$-frame to slowly rotate with respect to the geographic frame. We denote this drift 
-by a rotation vector $\mitbf{\Psi}$ (see Section~\ref{sec:theta}).
-\end{itemize}
-
-These reference frames and vectors are illustrated in Figure~\ref{fig:reference_frames}.
+causing the $\mathbf{E}$-frame to slowly rotate with respect to the geographic frame.
 
@@
 
-The part of the moment of inertia due to the elastic rotational deformation in the body-fixed frame 
-is traditionally related to the degree-two part of the gravity field via MacCullagh's formula \citep{munk1960rotation}:
+Rotational deformation in an elastic body is traditionally related to the degree-two part of the 
+centrifugal potential using a Love-number formalism. When the perturbation in gravitational
+potential is related to the moments of inertia using MacCullagh's formula  \citep{munk1960rotation}, we find:
 \begin{equation}
@@
 \end{equation}
-where $k$ is an elastic Love number, $a$ is the semimajor axis of the planet, and $G$ is the gravitational constant.
-This result may be extended to a viscoelastic rheology via the viscoelastic correspondence principle \citep[e.g.][]{peltier1974impulse}:
+where $k$ is an elastic Love number, $a$ is the semimajor axis of the planet, and $G$ is the gravitational constant. Both $\Omega_i$ and
+$J_{ij}$ are defined in the body-fixed frame. An extension to  viscoelastic rheology is permitted by the viscoelastic correspondence principle \citep[e.g.][]{peltier1974impulse}:
 \begin{equation}
@@
 \end{equation}
-This fluid-limit representation of the deformation does not allow for any disequilibrium between $J_{ij}$ and $E_{ij}$,
-so it does not permit TPW.
+This infinite-time limit  ensures that the left-hand side of Equation (4) vanishes. It follows that the contribution from the convective perturbation $E_{ij}$ also vanishes, so TPW is not permitted.
 
 \citet{ricard1993polar} obtain an approximation to Equation~\eqref{eq:viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.
-This introduces a new parameter, termed $T_1$, which can be seen as a weighted relaxation time for the system.  
-This simple approximation in the Laplace domain allows for an analytical transformation back into the time domain, and neglecting second order terms in $\dot{\Omega}$ we find:
+This introduces a new parameter, $\tau_R$, which can be seen as a weighted relaxation time for the system.  
+This simple approximation in the Laplace domain allows for an analytical transformation back into the time domain. Neglecting second order terms in $\dot{\Omega}$ we find:
 \begin{equation}
 J_{ij} = \frac{k_f a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right) -
- \frac{k_f a^5 T_1}{3G} \left(\dot{\Omega}_i \Omega_j + \Omega_i \dot{\Omega}_j - \frac{2}{3} \Omega_q \dot{\Omega}_q \delta_{ij} \right).
+ \frac{k_f a^5 \tau_R}{3G} \left(\dot{\Omega}_i \Omega_j + \Omega_i \dot{\Omega}_j - \frac{2}{3} \Omega_q \dot{\Omega}_q \delta_{ij} \right).
 \label{eq:rotational_deformation}
@@
 The first term corresponds to the fluid limit of rotational deformation (in the absence of any long-term elastic strength).  
-The second term represents the lag in the moment of inertia due to the viscous adjustment of the rotational bulge, where $T_1$ is the characteristic time constant for this adjustment.
-Since the first term represents the fluid limit of rotational deformation, it automatically satisfies Equation~\eqref{eq:quasistatic_liouville}, and hence does not contribute to the polar wander problem.
+The second term represents the lag in the moment of inertia due to the viscous adjustment of the rotational bulge, where $\tau_R$ is the characteristic time constant for this adjustment.
+Since the first term represents the hydrostatic limit of rotational deformation, it automatically satisfies Equation~\eqref{eq:quasistatic_liouville}, and hence does not contribute to the polar wander problem.
 
@@
 The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
-This part of the moment of inertia may also be parameterized using a viscoelastic Love number approach:
+Both of these contributions may also be parameterized using a viscoelastic Love number approach:
 \begin{equation} 
@@
 and $C_{ij}$ is the moment of inertia due solely to the internal load.
-Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:
+When the timescale of the surface response is short compared to the true polar wander timescale, it is reasonable to  use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
 \begin{equation}
@@
 \end{equation}
-Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
-This correspondence permits useful insights (see below).
-Let $\dot{\Theta} = \vert \dot{\mitbf{\omega}} \vert$ denote the rate of polar wander.
-Evaluating the scalar product $\dot{\Theta}^2 = \dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}}$ gives
+In order to evaluate $\dot{\mitbf{\omega}}$ we require knowledge of both $\mathbf{E}$ and $\mitbf{\omega}$ in the geographic frame (i.e.
+the frame used to derive Equation (13)).
+The same frame is also used to measure TPW, but the expressions on the right-hand-side of Equation (13) are more naturally evaluated in the reference frame of the principal axes of $\mathbf{E}$.
+We can avoid the need to reconcile these two frames by confining our attention to the rate of polar wander (We denoted  the rate of TPW by 
+ $\dot{W} = \vert \dot{\mitbf{\omega}} \vert$) .
+Evaluating the scalar product $\dot{W}^2 = \dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}}$ gives
 \begin{equation}
@@
 \end{equation}
+The scalar terms on the right-hand-side of Equation (14) are now independent of the choice of coordinate system as long as we are consistent.
+Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
+with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$. In this case the orientation of 
+$\mitbf{\omega}$ is defined in the $\mathbf{E}$-frame using colatitude $\theta$ and longitude $\phi$ (see Figure~\ref{fig:reference_frames}).
+Plugging this description of $\mitbf{\omega}$ into Equation~\eqref{eq:polar_wander_rate_tensor},
+and after some tedious algebra, we find
+
+\begin{equation}
+\begin{aligned}
+\dot{W}^2 = 
+&\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
+&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
+\label{eq:polar_wander_rate}
+\end{aligned}
+\end{equation}
+This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.
+Once the orientation of $\mitbf{\omega}$ is specified in the E-frame by angles $\theta$ and $\phi$, we can evaluate the rate of TPW in terms of the principal moments of $\mathbf{E}$.
+Two special cases of this equation are worth  noting. First, if $\mitbf{\omega}$  coincides with  the normal to the octahedral plane
+(i.e. the plane defined by a normal with direction cosines all equal to $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
+then it can be expressed in terms of the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator 
+($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
+\begin{equation}
+\begin{aligned}
+\dot{W}^2
+&= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
+&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2}. \\
+\end{aligned}
+\end{equation}
+The second invariant of the stress deviator is commonly used in the theories of plasticity
+to characterize the state of shear stress.
+The second invariant of the moment of inertia deviator can be viewed as a ``rotational stress" that drives a TPW response.
 
@@
 The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively.
-Since the choice of geographic axes is arbitrary, we assume that at this instant the $\mathbf{E}$-frame and the geographic frames are colocated (though at future times they will not be).
 The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $\mathbf{e}_3$-axis.
-For illustration, we assume the longitude $\phi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane.
+For illustration, the longitude $\phi$ of the rotation axis is taken to be zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane.
+In general $\phi$ is not zero.
 True polar wander moves the rotation axis towards the $\mathbf{e}_3$-axis.
-We denote this motion by $\dot{\mitbf{\omega}}$ in the geographic frame.
-However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$-frame.
-This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$, which is defined by colatitude $\beta$ and longitude $\gamma$.
-The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$-frame.
-Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Sections~\ref{sec:reference_frames} and \ref{sec:theta}}
+Observations of $\dot{\mitbf{\omega}}$ are made in the geographic frame, which drifts relative to the E-frame.
+}
 \label{fig:reference_frames}
@@
 
-Our goal is to quantify the polar wander rate $\dot{\Theta}$ due to a convective perturbation $\mathbf{E}$ in the moment of inertia. 
-The rate defined by Equation~\eqref{eq:polar_wander_rate_tensor} is expressed in the body-fixed geographic frame, 
-which is also the reference frame used to measure TPW. 
-However, the physics of the right-hand-side of Equation~\eqref{eq:polar_wander_rate_tensor} 
-is more naturally expressed in the reference frame of the principal axes of $\mathbf{E}$. 
-In general the $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, so the time derivative of $\omega$ is different in these two frames.
-However, Equation~\eqref{eq:polar_wander_rate_tensor} is a scalar equation, which means it is invariant to rotations.
-Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
-with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of 
-$\mitbf{\omega}$ in the $\mathbf{E}$-frame with colatitude $\theta$ and longitude $\phi$ (see Figure~\ref{fig:reference_frames}).
-Plugging this description of $\mitbf{\omega}$ into Equation~\eqref{eq:polar_wander_rate_tensor},
-and after some tedious algebra, we find
-\ifdetail
-
-And here is the tedious algebra.  Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$.
-We are interested in the rate of true polar wander, which is the angular change of the pole
-per unit time. We can transform Equation~\eqref{eq:liouville_ode} into a scalar equation by dotting it
-with $\dot{\mitbf{\omega}}$, since we are interested in the magnitude of the rate, rather than
-the direction at the moment.
-
-\begin{equation}
-\begin{aligned}
-\dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}} &= 
-A^2 \left[ \mathbf{E}\cdot \mitbf{\omega} - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)\mitbf{\omega}\right]
-\cdot \left[ \mathbf{E}\cdot \mitbf{\omega} - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)\mitbf{\omega}\right] \\
-\dot{\mitbf{\omega}}^2  &= 
-A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)\cdot \left(\mathbf{E}\cdot \mitbf{\omega}\right)
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right) \left( \mitbf{\omega}\cdot \mathbf{E} \cdot \mitbf{\omega} \right) 
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right) \left( \mitbf{\omega}\cdot \mathbf{E} \cdot \mitbf{\omega} \right) 
- + \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 \mitbf{\omega}\cdot \mitbf{\omega} 
-\right]\\
-\dot{\mitbf{\omega}}^2  &= 
-A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2.
-\right]\\
-\end{aligned}
-\end{equation}
-Since this is a scalar equation, we can arbitrarily choose a coordinate system in which to evaluate it.
-It is most convenient to choose the principal axes of $\mathbf{E}$, where
-\begin{equation}
-\mathbf{E} = 
-\begin{bmatrix}
-\lambda_1 & 0 & 0 \\
-0 & \lambda_2 & 0 \\
-0 & 0 & \lambda_3 \\
-\end{bmatrix}
-\end{equation}
-\begin{equation}
-\mitbf{\omega} = 
-\begin{bmatrix}
-\sin{\theta} \cos{\phi} \\
-\sin{\theta} \sin{\phi} \\
-\cos{\theta} \\
-\end{bmatrix}.
-\end{equation}
-
-As a warm-up, consider the case where $\phi = 0$, then  
-\begin{equation}
-\mitbf{\omega} = 
-\begin{bmatrix}
-\sin{\theta}  \\
-0 \\
-\cos{\theta} \\
-\end{bmatrix}
-\end{equation}
-We can evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
-\begin{equation}
-\mathbf{E}\cdot \mitbf{\omega} = 
-\begin{bmatrix}
-\lambda_1 \sin{\theta} \\
-0 \\
-\lambda_3 \cos{\theta}
-\end{bmatrix}
-\end{equation}
-
-\begin{equation}
-\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
-\end{equation}
-
-\begin{equation}
-\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
-\lambda_1 \sin^2{\theta} + \lambda_3 \cos^2{\theta}
-\end{equation}
-
-\begin{equation}
-\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
-\lambda_1^2 \sin^4{\theta} + \lambda_3^2 \cos^4{\theta} + 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta}
-\end{equation}
-Therefore, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
-\begin{equation}
-\begin{aligned}
-\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
- \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
- &= \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
- - \lambda_1^2 \sin^4{\theta} - \lambda_3^2 \cos^4{\theta} - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
- &= \lambda_1^2 \sin^2{\theta} (1 - \sin^2{\theta}) + \lambda_3^2 \cos^2{\theta} (1 - \cos^2{\theta})
-  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
- &= \lambda_1^2 \sin^2{\theta} \cos^2{\theta} + \lambda_3^2 \cos^2{\theta} \sin^2{\theta}
-  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
- &= (\lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \sin^2{\theta} \cos^2{\theta} \\
- &= \frac{1}{2} (\lambda_3 - \lambda_1)^2 \sin{2 \theta} \\
-\end{aligned}
-\end{equation}
-
-Now consider the more general case where $\phi \ne 0$.
-Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
-\begin{equation}
-\mathbf{E}\cdot \mitbf{\omega} = 
-\begin{bmatrix}
-\lambda_1 \sin{\theta} \cos{\phi}\\
-\lambda_2 \sin{\theta} \sin{\phi} \\
-\lambda_3 \cos{\theta}
-\end{bmatrix}
-\end{equation}
-
-\begin{equation}
-\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta}
-\end{equation}
-
-\begin{equation}
-\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
-\lambda_1 \sin^2{\theta} \cos^2{\phi} + \lambda_2 \sin^2{\theta} \sin^2{\phi} + \lambda_3 \cos^2{\theta}
-\end{equation}
-
-\begin{equation}
-\begin{aligned}
-\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
-&\lambda_1^2 \sin^4{\theta} \cos^4{\phi} + 
-\lambda_2^2 \sin^4{\theta} \sin^4{\phi} +
-\lambda_3^2 \cos^4{\theta} \\
-&+ 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
-&+ 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
-&+ 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
-\end{aligned}
-\end{equation}
-Again, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
-\begin{equation}
-\begin{aligned}
-\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
- \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
- - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
-&=
-\lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta} \\
-&-\lambda_1^2 \sin^4{\theta} \cos^4{\phi} - 
-\lambda_2^2 \sin^4{\theta} \sin^4{\phi} -
-\lambda_3^2 \cos^4{\theta} \\
-&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
-&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
-&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
-&=
-\sin^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} \right] \\
-&- \sin^2{\theta} \left[ \lambda_1^2 \sin^2{\theta} \cos^2{\phi}(1-\sin^2{\phi}) + \lambda_2^2 \sin^2{\theta} \sin^2{\phi}(1-\cos^2{\phi}) \right] \\
-&+ \lambda_3^2 \cos^2{\theta}(1-\cos^2{\theta}) \\
-&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
-&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
-&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
-&=
-\sin^2{\theta} \left[\lambda_1^2 \cos^2{\phi} (1 - \sin^2{\theta}) + \lambda_2^2 \sin^2{\phi} (1-\sin^2{\theta}) \right] \\
-&+ \sin^2{\theta} \left[ \lambda_1^2 \sin^2{\theta} \cos^2{\phi}\sin^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi}\cos^2{\phi} \right] \\
-&+ \lambda_3^2 \cos^2{\theta}\sin^2{\theta} \\
-&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
-&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
-&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
-&=
-\sin^2{\theta}\cos^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} \right] \\
-&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left[ \lambda_1^2 + \lambda_2^2 -2 \lambda_1\lambda_2 \right] \\
-&+ \lambda_3^2 \cos^2{\theta}\sin^2{\theta} \\
-&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
-&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
-&=
-\sin^2{\theta}\cos^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} + \lambda_3^2 - 2 \lambda_1 \lambda_3 \cos^2{\phi} - 2 \lambda_2 \lambda_3 \sin^2{\phi} \right] \\
-&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
-&=
-\sin^2{\theta}\cos^2{\theta} \left[( \lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \cos^2{\phi} + (\lambda_2^2 + \lambda_3^2 - 2 \lambda_2 \lambda_3 ) \sin^2{\phi} \right] \\
-&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
-&=
-\sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] \\
-&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
-&=
-\sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right]
-+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
-\end{aligned}
-\end{equation}
-
-At long last, we get
-\begin{equation}
-\begin{aligned}
-\frac{\dot{\mitbf{\omega}}^2}{A^2}  = 
-&\frac{1}{4}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
-&\frac{1}{4}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \\
-\end{aligned}
-\end{equation}
-
-\else
-\fi
-
-\begin{equation}
-\begin{aligned}
-\dot{\Theta}^2 = 
-&\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
-&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
-\label{eq:polar_wander_rate}
-\end{aligned}
-\end{equation}
-This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.
-Two special cases of this equation are of note. First, if $\dot{\Theta}^2$ is evaluated on the octahedral plane
-(the plane with direction cosines all $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
-then it can be expressed in terms of the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator 
-($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
-\begin{equation}
-\begin{aligned}
-\dot{\Theta}^2
-&= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
-&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2}. \\
-\end{aligned}
-\end{equation}
-The second invariant of the stress deviator is commonly used in the theories of elasticity and plasticity
-as a convenient scalar approximation of the shear stress in a system.
-The second invariant of the moment of inertia deviator can be similarly viewed as a scalar estimate of the ``rotational stress.''
-
-The second special case is if the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_3$ plane ($\phi = 0$).
-Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
-\begin{equation}
-\vert \dot{\Theta} \vert = 
-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1).
+The second special case occurs when the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_3$ plane ($\phi = 0$).
+Equation~\eqref{eq:polar_wander_rate} reduces to
+\begin{equation}
+\vert \dot{W} \vert = 
+\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1)\, .
 \label{eq:simple_milankovitch}
 \end{equation}
-The maximum true polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).
-
-From Equation~\eqref{eq:simple_milankovitch} it is clear that the important quantities for estimating the rate of true polar wander ($\dot{\Theta}$)
-are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.  
-They represent, respectively, the angular mismatch between the rotation axis and the principal axis of the convective moment 
-and the size of convective anomalies in the moment of inertia tensor. 
-The dynamics of mantle convection affects the rate of TPW in two ways; 
-it controls $\lambda_i$ and it represents the mechanism that drives the angular mismatch. 
-The value of $\theta$ at any time also depends on the relaxation of the rotation axis back toward the principal axis of $\mathbf{E}$. 
-We focus on the dynamics of the convective contribution in the next section.
+An equivalent expression was given by \citet{tsai2007theoretical}. We adopt Equation (17) as the basis for our dimensional analysis. Note that
+the maximum  rate is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).
+
+From Equation~\eqref{eq:simple_milankovitch} it is clear that the important quantities for estimating the rate of true polar wander ($\dot{W}$)
+are $\theta$ and $\lambda_3 - \lambda_1$. Both of these quantities depend on the structure and dynamics of mantle convection.  The size of
+convective anomalies directly determines $\mathbf{E}$, and hence $\lambda_3-\lambda_1$.
+We also expect convection to alter the orientation of the principal axes of $\mathbf{E}$, which drives a change in $\theta$.
+Relaxation of the rotation axis back toward the principal axis of $\mathbf{E}$ acts to reduce $\theta$.
+In the next section we focus on the role of mantle convection in setting $\lambda_3-\lambda_1$ and in driving changes in $\theta$.
 
@@
 As such, some extra care must be taken to ensure that the equations we consider are self-consistent. 
-Furthermore, since our goal is to establish a scaling for TPW rate, we must identify a minimal set of nondimensional numbers which describe its physics.
+Furthermore, since our goal is to establish a scaling for TPW rate, we must identify a minimal set of nondimensional numbers that fully characterize the problem.
 For simplicity we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.
@@
 The remaining parameters are defined in Table~\ref{tab:parameters}.
-Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure
+Note that we retain the centrifugal term in the momentum equation, which is normally either neglected or absorbed into a modified pressure
 (in the latter case the boundary conditions on $P$ must be modified).
-The centrifugal term, although the largest of the terms neglected in typical mantle convection models,
-is generally small compared to gravitational forces (at least for Earth-like parameters),
-and likely does not have a strong influence on the style of convection.
-However, this term is critical for determining the size of the rotational bulge which must move relative to the mantle,
-and so we must include it to establish the connection between the linear and angular momentum equations.
-
-Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires four nondimensional numbers to characterize it
-(a fifth one, defined by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations).
+The centrifugal term
+is generally small compared to gravitational forces (at least for Earth-like parameters). Consequently, it does  not have a strong influence on the style of convection.
+However, this term is critical for determining the size of the rotational bulge. It is also essential for establishing a connection between the linear and angular momentum equations.
+
+Four dimensionless parameters are required to characterize the dynamics of true polar wander  (cf.
+\citet{barenblatt1996scaling}).
+A fifth parameter, defined  by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations.
 Convenient choices for these numbers are listed in Table~\ref{tab:nondim}, along with approximate Earth-like values for them.
-
-Two dimensionless parameters have a prominent role in our scaling analysis. 
+This particular choice relies on $R$ for the length scale, $R^2/\kappa$ for the time scale and $\Delta T$ for the temperature scale.
+
+Two of these four dimensionless parameters have a prominent role in our scaling analysis. 
 The first is the Rayleigh number, which characterizes the  vigor of convection. 
@@
 
-In this case, the dynamics can be characterized in terms of deviations from a reference hydrostatic state, 
-which includes the dynamic pressure $P^* = P - P_0$ and density perturbations $\delta \rho = \rho- \rho_0 = - \rho_0 \alpha (T-T_0)$.
+The dynamics can be characterized in terms of deviations from a reference hydrostatic state. We define a
+ dynamic pressure by $P^* = P - P_0$ and introduce density perturbations $\delta \rho = \rho- \rho_0 = - \rho_0 \alpha (T-T_0)$.
 In addition, we expect deviations in the figure of the planet from its hydrostatic shape, 
 which we denote by $V = V_H + \Delta V$, where $V_H$ is the hydrostatic figure and $\Delta V$ is the deviation. 
-Our rationale for this decomposition is simply that the hydrostatic pressure is most naturally defined in the (oblate) hydrostatic configuration. 
-The introduction of $\Delta V$ requires another nondimensional number to characterize it, 
-and we find that the quantity $\Gamma \equiv \alpha \Delta T$ is convenient.
+Our rationale for this decomposition is simply that the hydrostatic pressure is defined in the hydrostatic configuration $V_H$ rather than $V$. 
+Introducing $\Delta V$ requires another nondimensional number to characterize it, 
+and we find that the quantity $\Gamma \equiv \alpha \Delta T$ is convenient. (The motivation for this choice is explained below.)
 Finally, we define $\mitbf{\Omega} = \Omega_0 \mitbf{\omega}$, where $\mitbf{\omega}$ is a unit vector in the direction of $\mitbf{\Omega}$.
@@
 By definition the hydrostatic reference state is a solution to Equation~\eqref{eq:navier_stokes} where there is no flow:
-
 \begin{equation}
@@
 \end{equation}
-Nondimensionalizing with the parameters in Table~\ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
+Removing this reference state from the full momentum equation defines an equation for the perturbations. 
+When the perturbed momentum equation is written in nondimensional form using the parameters in Table~\ref{tab:nondim} we obtain:
 \begin{equation}
@@
 \end{equation}
-
-We can explicitly draw the connection between the angular and linear momentum equations by returning to the dimensional
-Equation~\eqref{eq:navier_stokes}, crossing it with $\mathbf{r}$ and integrating over the volume of the mantle:
+Only two dimensionless parameters appear in Equation (24); the buoyancy force is specified by $Ra$ and the centrifugal force depends on the product $Ra\; m$.
+No additional parameters appear in the energy (temperature) equation because $\Delta T$ is used to scale temperature and time is scaled by the thermal diffusion time $\tau_T = R^2/\kappa$.
+The aspect ratio $A$ arises in the problem through definition of the boundary conditions, whereas the dependence on the dimensionless density deficit $\Gamma$ has not yet emerged. This dependence is
+revealed by drawing a connection between the linear and angular momentum equations.
+
+We start with the dimensional form of 
+Equation~\eqref{eq:navier_stokes}.  Crossing it with $\mathbf{r}$ and integrating over the volume of the mantle gives:
 \begin{equation}
@@
 \end{equation}
-
 The first three terms represent pressure, viscous torques, and gravitational torques on the mantle.  
@@
 Therefore we can neglect those terms, and we are left with
-
 \begin{equation}
@@
 \end{equation}
-This equation can be directly identified with Equation~\eqref{eq:quasistatic_liouville},
-and is a statement that a quasistatic body will rotate around the principal axis of its total moment of inertia.
-
-We now seek to evaluate Equation~\eqref{eq:quasistatic_angular_momentum} in the perturbed, convecting state.
+This equation can be directly identified with Equation~\eqref{eq:quasistatic_liouville}; it 
+ is a statement that a quasistatic body will rotate around the principal axis of its total moment of inertia.
+
+We now seek to quantify the perturbation to Equation~\eqref{eq:quasistatic_angular_momentum} due to dynamics.
 Hydrostatic balance (Equation~\eqref{eq:hydrostatic}) ensures that the integral over the reference shape $V_H$ vanishes when $\rho = \rho_0$.
@@
 This equation may be identified with Equation~\eqref{eq:disequilibrium}, 
-where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of inertia with the spin axis (right side).
-Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table~\ref{tab:nondim}.
-
-\ifdetail
-\subsection{Buckingham Pi theorem and nondimensionalization}
-\label{sec:buckingham}
-The Buckingham Pi theorem (essentially an application of the rank-nullity theorem) allows us to determine the number of nondimensional numbers for this problem.
-We count the number of fundamental units for the problem, in this case mass (kg), length (m), time (s), and temperature (K), as well as the number of parameters in use for this problem.  
-These parameters are listed in Table~\ref{tab:parameters}, and subtracting the ...
-
-\begin{table}
-\centering
-\caption{Nondimensionalizations for variables, primed variables denote dimensional variables}
-\label{tab:nondim_convert}
-\begin{tabular}{@{}lll}
-$x^\prime$ &=& $R \;\; x$ \\
-$t^\prime$ &=& $R^2/\kappa \;\; t$ \\
-$T^\prime$ &=& $\Delta T \;\; T$ \\
-$\rho^\prime$ &=& $\rho_0 \;\; \rho$\\
-$u^\prime$ &=& $\kappa/R \;\; u$ \\
-$P^\prime$ &=& $\eta \kappa/R^2 \;\; P$ \\
-$\Omega^\prime$ &=& $\Omega_0 \;\; \Omega$ \\
-$g^\prime$ &=& $g_0 \;\; g$
-\end{tabular}
-\end{table}
-
-If we plug these nondimensionalizations into the governing equations we find that the incompressibility constraint remains unchanged.  
-The temperature equation and equation of state become
-\begin{equation}
-\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \nabla^2 T
-\end{equation}
-
-\begin{equation}
-\rho = 1 - \alpha \Delta T \;\;  T.
-\end{equation}
-We may furthermore define a hydrostatic reference state where the density is the reference density everywhere and the velocity is zero, i.e.
-
-\begin{equation}
- \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}= - \nabla P_0 + \rho_0 \mathbf{g}.
-\end{equation}
-Subtracting this from the momentum equation and defining a dynamic pressure $P^* = P - P_0$, we find
-
-\begin{equation}
-\begin{aligned}
- - & \Omega_0^2  \alpha  \Delta T T R \; \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} = - \eta \kappa / R^2 \; \nabla P^* \\ 
-&+ \eta \kappa / R^2 \; \nabla \cdot \left( \eta \epsilon (\mathbf{u}) \right) - \alpha \Delta T g_0 \; \mathbf{g}
-\end{aligned}
-\end{equation}
-
-\begin{equation}
-\begin{aligned}
- - & \mathrm{Ra} \; m \; T \; \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} = - \nabla P^* \\ 
-&+ \; \nabla \cdot \left( \eta \epsilon (\mathbf{u}) \right) - \mathrm{Ra} \; T \; \mathbf{g}
-\end{aligned}
-\end{equation}
-
-\fi
+where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of inertia with the spin axis (right side). Equation (29) also reveals
+the need for  the dimensionless parameter $\Gamma = \alpha \Delta T$.
+The perturbation  $\alpha(T-T_0)$ on the right-hand side must be related to the perturbation $\Delta V$ on the left-hand side because all other terms in Equation (29) are the same.
+A change in the scale of $\Delta V$ implies a change in the scale of $\alpha(T-T_0)$, and vice versa.
+
+We now proceed to characterize the rate of TPW in terms of the dimensionless number identified in Table~\ref{tab:nondim}.
+This approach ensures that we reduce the problem to the minimum set of parameters.
 
@@
 
-Having drawn the connection between the angular momentum equation (Section~\ref{sec:rotational_dynamics}) and the linear momentum 
-equation (Section~\ref{sec:internal}), we would like to apply the scaling of the 
-linear momentum equation to the problem of TPW. We would therefore like to rewrite Equation~\eqref{eq:simple_milankovitch}
-in terms of the nondimensional numbers identified in Section~\ref{sec:internal}.
-Specifically, we will need estimates for $(C-A)$, $T_1$, $(\lambda_3-\lambda_1)$, and $\theta$.
-
-It is useful to define a nondimensional eigenvalue difference, which may also be described as a normalized ``eigengap'': $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
-This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
-The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is 
-proportional to the ratio of rotational to gravitational forces \citep{munk1960rotation}:
+Equation (17) defines the rate of TPW in terms of four physical parameters: $(C-A)$, $T_1$, $(\lambda_3-\lambda_1)$ and $\theta$.
+This equation can be expressed in nondimensional form using the dimensionless numbers from Table 2.
+We begin by expressing $\dot{W}$ in terms of a nondimensional time $t^{\prime} = t /\tau_T$, where $\tau_T = R^2/\kappa$ is the thermal diffusion time.
+Similarly, the time constant $\tau_R$ can be expressed in terms of $\tau_T$ and other dimensionless numbers.
+In detail $\tau_R$ represents a weighted average of different relaxation modes, but it is sufficient for our purposes to let
+\begin{equation}
+\tau_R \sim \frac{ \eta }{ \rho_0 g_0 R} = \tau_T \, \frac{\Gamma}{\mathrm{Ra} }.
+\end{equation}
+Combining these results gives the dimensionless rate of TPW
+\begin{equation}
+\frac{dW}{dt'} \sim \frac{Ra}{\Gamma} \frac{(\lambda_3 - \lambda_1)}{(C-A)} \sin 2\theta\, .
+\end{equation}
+A more convenient form is obtained by noting that difference in the polar and equatorial moments $(C-A)$ is proportional to the ratio of rotational to gravitational forces
+\citep{munk1960rotation}:
 \begin{equation}
@@
 \end{equation}
-
-The time constant $T_1$ is a viscous relaxation time of the planetary mantle.
-It is frequently represented as a weighted average of the different relaxation modes \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
-\begin{equation}
-T_1 \sim \frac{ \eta }{ \rho_0 g_0 R} = \frac{R^2}{\kappa} \frac{\Gamma}{\mathrm{Ra} }.
-\end{equation}
-Plugging these scalings into Equation~\eqref{eq:simple_milankovitch} we find
-\begin{equation}
-\frac{\kappa}{R^2} \; \dot{\Theta} \sim - \frac{\mathrm{Ra}}{\Gamma m} \Lambda_{31} \sin{2 \theta}.
-\label{eq:scaled_rotation}
-\end{equation}
-
-At this point we do not have estimates for the characteristic magnitudes of $\Lambda_{ij}$ or $\theta$, both of which are crucial for predicting characteristic rates of TPW.
-They represent, respectively, the size of convective anomalies in the moment of inertia tensor and the angular mismatch between the rotation axis and the principal axis of the convective moment.
-
-The quantity $\Lambda_{ij}$ should be a function of our nondimensional parameters $\mathrm{Ra}$, $m$, and $\Gamma$. 
-Similarly, the convective forcing that drives changes in $\theta$ must also be a function these nondimensional parameters. 
-To be explicit in our discussion, we denote the angular velocity induced by forcing by ${\Psi}$ (see Figure~\ref{fig:reference_frames}).
-We consider relatively slowly rotating bodies here ($m \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
-In this regime, therefore, we neglect the dependence on $m$, 
-and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra}, \Gamma)$ and $\Psi(\mathrm{Ra}, \Gamma)$.
+We can also define a nondimensional eigenvalue differencing using $\Lambda_{ij} = (\lambda_i-\lambda_j)/I_0$. Introducing these results into Equation (31) gives
+\begin{equation}
+\frac{d W}{dt'} \sim \frac{Ra\, \Lambda_{31}}{\Gamma\, m} \sin 2\theta
+\end{equation}
+which defines the rate of TPW in terms of dimensionless parameters $Ra$, $m$ and $\Gamma$.
+At this point we do not have estimates for $\Lambda_{ij}$ or $\theta$.
+However, we do know that they must be functions of the dimensionless numbers in Table 2 and our dimensionless time $t^{\prime}$.
+
+For the case of slowly rotating bodies ($m \ll 1$), we do not expect rotation have a large influence on the style of convection, so it would be reasonable to  
+ look for scalings of the form $\Lambda_{ij}(\mathrm{Ra}, \Gamma, t^{\prime})$ and $\theta(\mathrm{Ra}, \Gamma, t^{\prime})$.
+We avoid the complications of dealing with the time dependence by considering time averages for $\Lambda_{ij}$ and characterizing an upper bound on $\theta$.
 
@@
 \label{sec:lambda}
+
+Fluctuations in the nonhydrostatic moment of inertia are caused by a redistribution of density anomalies 
+ in the mantle. The direct contribution from 
+internal temperature variations is given by
+\begin{equation}
+C_{ij} = \int_{V_H} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) \,dV
+\label{eq:temperature_fluctuations}
+\end{equation}
+An additional contribution arises from the deflection of the surface and internal interfaces.
+The surface response is a function of the viscosity structure of the planet and the wavelength and depth of the internal load.
+It is customary to represent the surface response using the fluid-limit geoid kernals,  where the factor
+ $1+k^L_f$ is close to 1. 
+(Strictly speaking, dynamic compensation usually makes $1 + k_f^L$ less than one, \citep[e.g.][]{richards1984geoid}). 
+For the purposes of scaling, we can omit this multiplicative factor and estimate $E_{ij}$ from the direct contribution $C_{ij}$.
+
+For thermal convection, the convective part of the moment of inertia is related to the degree-two part of the temperature field (see Appendix A).
+To extract this component of the temperature field we can expand $T-T_0$ in a sum  of orthonormal basis functions $R_n Y_{lm}$, 
+where $Y_{lm}$ are spherical harmonics and $R_n$ are some set of orthogonal radial polynomials. The coefficients
+$T_{nlm}$ of the series expansion are normalized by $\Delta T$, so we represent the temperature anomaly in the form
+\begin{equation} 
+T( r , \theta, \phi, t ) - T_0 = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi).
+\label{eq:T_series}
+\end{equation}
+Orthogonality of the basis functions means that the integral for $C_{ij}$ picks out degree-two 
+spherical harmonics in the lateral dimensions, and only the lowest few radial functions $R_n(r)$.
+Consequently, only a few terms in the expansion matter for TPW.
+We seek to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$ (see Appendix A for more detail).
+Once the contribution to $T_{\text{degree-two}}$ is established, the integral for $C_{ij}$ can be approximated by
+\begin{equation}
+C_{ij} \sim I_0 (\alpha \Delta T) T_{\text{degree-two}} \, .
+\end{equation}
+The nondimensional difference in the principal moments becomes
+\begin{equation}
+\Lambda_{31} \sim \Gamma T_{\text{degree-two}}.
+\end{equation}
+
+Normalizing the 
+ temperature field by $\Delta T$ means the summation over all basis functions in Equation (35) cannot exceed 1.
+This is a strong constraint, but it gives very little information about the distribution of power across the $T_{lmn}$.
+We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow, (relatively low $\mathrm{Ra}$) 
+and that of chaotic flow (relatively high $\mathrm{Ra}$).
+At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes which are largely influenced by the aspect ratio.
+At high Rayleigh number become chaotic and there is a broadening of the spatial and temporal spectra  \citep{mclaughlin1982transition}.
 
@@
 
-Fluctuations in the nonhydrostatic moment of inertia are caused by density variations 
-due to the temperature fluctuations in the mantle and their spatial structure:
-
-\begin{equation}
-C_{ij} = \int_{V_H} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) \,dV
-\label{eq:temperature_fluctuations}
-\end{equation}
-As discussed in Section~\ref{sec:convective_moment}, these internal density loads may be convolved with their surface responses to get the convective moment of inertia $E_{ij}$.
-The surface response is a function of the viscosity structure of the planet and the wavelength and depth of the internal load.
-The factor $(1+k^L_f)$ is generally an order-one parameter 
-(strictly speaking, dynamic compensation usually makes it less than one, \citep[e.g.][]{richards1984geoid}). 
-For the purposes of scaling, it is reasonable to neglect this multiplicative factor.
-
-For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix~\ref{appendix:moments}).
-Therefore, an estimate for one specifies the other.
-We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
-where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
-$T_{nlm}$ are the coefficients for the expansion which have been normalized by $\Delta T$:
-\begin{equation} 
-T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi).
-\label{eq:T_series}
-\end{equation}
-Inserting the temperature expansion~\eqref{eq:T_series} into Equation~\eqref{eq:temperature_fluctuations} results in 
-a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
-Orthogonality of the basis functions for the expansion means that the integral for $C_{ij}$ picks out degree-two 
-spherical harmonics in the lateral dimensions, and only the lowest few radial functions $R_n(r)$.
-Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
-We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$ (see Appendix~\ref{appendix:moments} for more detail): 
-\begin{equation}
-\Lambda_{31} \sim \Gamma T_{\text{degree-two}}.
-\end{equation}
-The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by
-
-\begin{equation}
-\max \left( \displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \right) = 1.
-\end{equation}
-This is a strong constraint, but it gives very little information about the distribution of power across the $T_{lmn}$.  
-We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow, (relatively low $\mathrm{Ra}$) 
-and that of chaotic flow (relatively high $\mathrm{Ra}$).
-The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
-the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is a broadening of the spatial 
-and temporal spectra \citep{mclaughlin1982transition}.
-
-At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes which are largely influenced by the aspect ratio.
-This spectrum may or may not have a lot of power in the degree-two modes, and does not depend strongly on time.
-
-At high Rayleigh number we expect the shortest lengthscales to be limited by the effects of thermal diffusion,
-which tends to wipe out thermal heterogeneity at small scales.  Consequently, there will be little power in modes
-with shorter lengthscales than that allowed by diffusion. 
-Therefore we expect, to a good approximation, that the infinite sum in Equation~\eqref{eq:T_series} can be truncated at some maximum wavenumber, set by the smallest lengthscale $d$:
+The lengthscale of temperature anomalies in the mantle is limited by the effects of thermal diffusion. 
+Structure at very short lengthscales is erased by thermal diffusion, so there will be little power in modes
+with lengthscales shorter than that allowed by diffusion. 
+Therefore the infinite sum in Equation~\eqref{eq:T_series} can be truncated at some maximum wavenumber, set by the smallest lengthscale $d$:
 \begin{equation}
@@
 \end{equation}
-The value of each $T_{lmn}(t)$ will in general be some complex function of time, but for a given style of convection we expect there to be some average value.
+The value of each $T_{lmn}(t)$ will in general be some complex function of time, but for a given style of convection we expect the time average to define a representative power spectrum.
 For chaotic flow the power should be spread out amongst the modes accessible to it.
@@
 
-Any of a number of scaling laws can provide an estimate for the characteristic length scale of a convecting system which may depending on rheology, geometry, or density structure.
-The simplest, based on boundary layer theory \citep{turcotte1967finite}, finds $d/R \sim \mathrm{Ra}^{-1/3}$.
+Any of a number of scaling laws can provide an estimate for the characteristic length scale of a convecting system. The specifics are likely to depend on rheology, geometry, and density structure.
+The simplest model is based on boundary layer theory \citep{turcotte1967finite}, which finds $d/R \sim \mathrm{Ra}^{-1/3}$.
 This scaling is roughly a measure of the diffusive lengthscale for the timescale of a convective overturn, consistent
@@
 
-We performed a series of numerical simulations of mantle convection at different Rayleigh numbers to test this scaling.
-We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII82},
-which allows for flexible implementation of different rheologies, geometries, and postprocessors.
-In order to test a wide range of Rayleigh numbers, we ran the simulations in a 2D annulus, tracking the eigenvalues of the 
-moment of inertia tensor and integrating Equation~\eqref{eq:liouville_ode} in time.
+We test this prediction using a series of numerical simulations of mantle convection at different Rayleigh numbers. 
+All calculations are based on the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, which is built on the finite element library \texttt{deal.II} \citep{dealII82}. This package
+ allows for flexible implementation of different rheologies, geometries, and postprocessors.
+In order to test a wide range of Rayleigh numbers, we ran the simulations in a 2D annulus.
+At each time step we evaluate moment of inertia tensor in the frame of the convection calculation and evolve an initial rotation vector in time using  Equation~\eqref{eq:liouville_ode}.
+Recall that Equation (13) describes the evolution of the rotation vector in a geographic frame.
+Here we adopt the coordinate frame of the convection calculations as the geographic frame, so off-diagonal elements of $\mathbf{E}$ will generally be non-zero.
+We also calculate the eigenvalues of $\mathbf{E}$ at each time step to evaluate $\Lambda_{21} = (\lambda_2 - \lambda_1)/I_0$.
+
 For the 2D simulations there is a reduced dimensionality when calculating the number of modes,
-so $N_\text{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\text{degree-two}} \sim \mathrm{Ra}^{-2/3}$, 
-which is shown as a dashed line in Figure~\ref{fig:eigengap}.
+so $N_\text{modes} \sim \left(R/d \right)^2$.
+This leads us to a scaling of $T_{\text{degree-two}} \sim \mathrm{Ra}^{-2/3}$ and $\Lambda_{21} \sim \Gamma Ra^{-2/3}$, according to Equation (37).
+A fit to the average value of $\Lambda_{21}$ as a function of $Ra$ is shown as a dashed line in Figure~\ref{fig:eigengap}.
 This result has a simple interpretation.
 As the Rayleigh number of the system increases, the smallest lengthscale of convective features gets smaller.
-The total power in the temperature field is spread across a larger spectrum, leaving less total power for the degree-two part, which is what drives TPW.
-
-With an estimate for the power in the degree-two part of the temperature field, we may finally estimate $\Lambda_{ij}$:
+The total power in the temperature field is spread across a larger spectrum, leaving less total power for the degree-two part to drive TPW.
+
+Returning to the three-dimensional problem we expect $\Lambda_{ij}$ to scale as
 \begin{equation}
@@
 \end{equation}
-Other power spectra for the temperature field are possible. Isoviscous models tend to be ``bluer,'' (dominated by small wavelengths) 
-and models with viscosity stratification tend to be ``redder'' (dominated by long wavelengths) \citep{richards1999polar}.
-Present day Earth seems to have a fairly ``red'' spectrum, with large low-degree seismic anomalies due to 
-Cenozoic subduction history and the lower mantle LLSVPs \citep{dziewonski2010mantle}.
-Nevertheless, at high Rayleigh number the expectation is that the power will be distributed across many length scales.
+Using this estimate in Equation (33) gives
+\begin{equation}
+\frac{d W}{dt^\prime} = \frac{1}{m} \sin 2 \theta
+\end{equation}
+Remarkably, $Ra$ and $\Gamma$ have completely dropped from the expression for the rate of TPW.
+The time constant $\tau_R$ goes down at high Rayleigh numbers.
+At the same time, the  coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
+That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
+Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
+However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
+With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
+\begin{equation}
+\frac{d\Theta}{dt^\prime} \sim \frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
+\label{eq:simplest_milankovitch_defer_scaling}
+\end{equation}
+In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
+$d(\mathrm{Ra})$ will still result in a weak dependence 
+on the Rayleigh number.
+
+ Much of the debate about the magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
+In the next section we seek to characterize the maximum value for $\theta$ in terms of the dimensionless parameters.
 
@@
 
-Convection can drive both growth and decay in the mismatch angle ($\theta$) between the current rotation axis and the principal axis of the convective moment. 
-This parameter is crucial for the rate of TPW, as previously noted in Equation~\eqref{eq:scaled_rotation}.  
-Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
-
-Two processes control the evolution of $\theta$. Its growth occurs through perturbations in the convective moment and 
-its decay occurs by relaxation of the pole towards the maximum moment of inertia.
-We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
-from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
-the rotation of the body-fixed frame relative to inertial space in Equation~\eqref{eq:euler}).
-The $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, 
-which we describe by the rotation vector $\mathbf{\Psi}$. 
-(To be explicit in our discussion we let $\mitbf{\Psi}$ define the rotation of the $\mathbf{E}$-frame in the geographic frame, 
-although this choice is arbitrary as long as we are consistent). The time derivatives of $\mitbf{\omega}$ in the two frames are related by 
-\begin{equation}
-\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
-\end{equation}
-where primes are used to define quantities in the $\mathbf{E}$-frame. Rearranging gives 
-\begin{equation}
-\dot{\mitbf{\omega}}^{\prime} = \dot{\mitbf{\omega}} - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
-\end{equation}
-We now substitute for $\dot{\mitbf{\omega}}$ from Equation~\eqref{eq:liouville_ode}, 
-making the additional assumption that the geographic and $\mathbf{E}$-frames are momentarily aligned. 
-This assumption is not a serious restriction because our goal is simply to relate the time derivatives of $\mitbf{\omega}$ in the two frames, 
-rather than track the evolution of $\mitbf{\omega}$ over time. The expression for $\dot{\mitbf{\omega}}$ becomes
-\begin{equation}
- \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
-\label{eq:liouville_ode_new_frame}
-\end{equation}
-The advantage of expressing the rate of TPW in the $\mathbf{E}$-frame is that the entire equation
-may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
-As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that 
-$\mitbf{\omega}^{\prime}$ lies in $\mathbf{e}_1^{\prime}-\mathbf{e}_3^{\prime} $ plane (as shown in Figure~\ref{fig:reference_frames}). 
-In this case the orientation of $\mitbf{\omega}^\prime$ is defined by colatitude $\theta$ and longitude $\phi=0$. 
-The colatitude $\theta$ is precisely the angular misalignment that specifies the rate of TPW in Equation~\eqref{eq:simple_milankovitch}.
-We furthermore specify the orientation of $\Psi$
-using colatitude $\beta$ and longitude $\gamma$ (see Figure~\ref{fig:reference_frames}).
-
-The time derivative of the unit vector $\mitbf{\omega}^{\prime}$ defines the changing orientation of the rotation axis in the $\mathbf{E}$-frame. 
-We can express this changing orientation in terms of changes in angles $\theta$ and $\phi$. The change in colatitude is
-\begin{equation}
-\dot{\theta} = 
--\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\beta \sin\gamma.
-\label{eq:theta_competition}
-\end{equation}
-where $\mitbf{\Psi}$ is expressed in terms of its amplitude $|\mitbf{\Psi}|$ and its orientation angles $\beta$ and $\gamma$.
-Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
-The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it via
-relative motion between the geographic frame and the principal axes of the $\mathbf{E}$-frame.
-Comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
-then the rate of TPW $\vert \dot{\Theta} \vert$ is identical to $\vert \dot{\theta} \vert$.
-Conversely, when the rate of TPW is very slow (i.e. $\vert \dot{\Theta} \vert \sim 0$), 
-then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$, so the evolution of $\theta$ is driven mostly by $\mitbf{\Psi}$. 
-For the purposes of scaling we will omit the orientation factor $\sin \beta \sin \gamma$ and focus on $|\mitbf{\Psi}|$.
-In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 
-at least occasionally, be larger than its decay.
-
-An estimate for $\vert \mitbf{\Psi} \vert$ must concern the stability of the convective moment of inertia. 
-Convection is continuously redistributing mass throughout the mantle, which will perturb the convective moment of inertia.  
-If the convective moment is stable to small perturbations, then $\vert \mitbf{\Psi} \vert$ should be small, so
- $\theta$ will never have the opportunity to grow.
-However, if $\mathbf{E}$ is not stable to perturbations, then $|\mitbf{\Psi}|$ can be large and $\theta$ can grow up to $90^\circ$. 
+Two processes control the evolution of $\theta$.
+It grows through perturbations in the convective moment and  decays by relaxation of the rotation axis towards the principal axis of $\mathbf{E}$.
+Large values of $\theta$ are achieved when
+convective perturbations deflect the principal axis away from the rotation axis faster than relaxation can reduce $\theta$. Here we focus on the nature of
+perturbations that drive large increases in $\theta$ with the understanding that a complete description of $\theta(t)$ as a function of time must include both the growth and decay of the misalignment.
+
+An instantaneous change in the convective moment of inertia shifts the orientation of the principal axis, causing a change in the misalignment angle.
+A random change could make the angle larger or smaller, so there is no guarantee that a given perturbation will cause $\theta$ to grow.
+On the other hand,  we know that $\theta$ only grows through perturbation in the convective moment.
+Of course no change in the moment of inertia happens instantaneously.
+Instead,  changes in $\mathbf{E}$ accumulate over  finite durations and these gradual changes must compete with the relaxation process to define the evolution of $\theta$.
+Still, the concept of an instantaneous perturbation is useful because the conditions that facilitate large changes in $\theta$ also promote rapid growth rates.
+
 Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of changes in the orientation of the principle axes.
-This may be done by application of a theorem due to \citet{davis1970rotation}.
-
 Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor after some time interval $\Delta t$, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.
 The corresponding rotation of the principal axes is defined by the angle $\xi$.
-A bound on $\xi$ is given by
+A bound on $\xi$ is given by \citep{davis1970rotation}
 \begin{equation}
@@
 \end{equation} 
-That is to say, if there is a large difference between the eigenvalues, this stabilizes axes to perturbations.  
-If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principal axes.
-This is illustrated schematically in Figure~\ref{fig:perturb}.
-\citet{evans1998true} described a situation where the convective moment of inertia of Earth is prolate, with the
-maximum and intermediate moments of inertia nearly equal.
-He argued that perturbations in this case would result in TPW paths that are approximately
-on the great circle path between the intermediate and maximum axes.
-This behavior is a consequence of the instability described in Equation~\eqref{eq:kahan}.
-
-We may arrive at a simple estimate for $|\mitbf{\Psi}|$  
-by differentiating Equation~\eqref{eq:kahan}, tentatively holding $\vert \lambda_i - \lambda_j \vert$ fixed:
-\begin{equation}
-\vert \mitbf{\Psi} \vert \sim \vert \dot{\xi} \vert \le \frac{ \vert \dot{\delta} \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
-\label{eq:grow_perturbation}
-\end{equation} 
-As noted previously, $|\mitbf{\Psi}|\sim |\dot\theta |$ when the rate of TPW is small and the angular misalignment $\theta$ is small. 
-Consequently, we can interpret Equation~\eqref{eq:grow_perturbation} as a bound on how fast $\theta$ can grow.
+Small changes in orientation (i.e.
+small $\xi$) occur when the difference in eigenvalues is large.
+Conversely, large changes in orientation are possible when differences between the eigenvalues are small.
+This dependence is illustrated schematically in Figure~\ref{fig:perturb}.
+Unstable conditions prevail when the maximum and intermediate moments of inertia are nearly equal, leading to large TPW events \citep{evans1998true}.
+(\citet{tsai2007theoretical} argue that large TPW events take a long time because the rate of TPW is small when $\Lambda_{31}$ is small or when $\theta = 0$ or $\pi/2$ (see Equation (33).
+If the maximum and intermediate axes are interchanged when the misalignment angle is small $\theta \approx 0$, then the new definition of the maximum axis after the interchange ensures that $\theta \approx \pi/2$.
+In this case the rate of TPW may be small before and after the interchange).
 
@@
 
-The quantity $(\lambda_i-\lambda_j)$ which appears in the denominator of these equations is precisely the same as the quantity which we estimated in the previous section to scale with $\sim \mathrm{Ra}^{-1}$.
-Therefore, as the Rayleigh number increases, the characteristic gap between the eigenvalues of the convective moment becomes smaller.  
-Additionally, the timescale of fluctuations in these values goes down. 
-Overall, this makes the principal axes of high Rayleigh number systems much less stable.
-This is consistent with the result of \citet{richards1999polar}.
-
-In the limit that the eigengap becomes zero, the rotation of the principle axes can be arbitrary. 
-This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}, where 
-the mismatch angle is $90^\circ$. However, the eigengap does not need to be zero for there to be large displacements polar wander, 
-and if it does go to zero, the wander does not need to be $90^\circ$.
-
-Figure~\ref{fig:misfit} shows a representative timeseries for annular convection, where 
-we track the eigenvalues of the moment of inertia and integrate Equation~\eqref{eq:liouville_ode} in time.
-Since it is a 2D model, the spin axis may be represented by a single angle.
-When the eigengap gets small, the misfit angle becomes much larger, and the rate of polar wander becomes much faster.
-At $\sim$1 Gyr into the model run the eigengap goes to zero, and the misfit angle goes to approximately $90^\circ$, 
-an IITPW event. But there are several other events where the eigengap becomes small, and there are 
-still large TPW events associated with them.
-For example, at $\sim$0.3 Gyr the eigengap dips, with an associated large TPW event, even though
-there is technically no interchange of the axes.
-
-We suggest, then, that $90^\circ$ IITPW events are simply a special case of a broad class of events which can occur
+The angle $\xi$ would correspond to a change in $\theta$ if relaxation was not allowed to operate.
+While this approximation is not realistic, it does suggest that larger perturbations in $\theta$ are possible when  $(\lambda_i-\lambda_j)$ small.
+Scaling results and numerical simulations from the previous section show that the difference in eigenvalues scales with $\sim \mathrm{Ra}^{-1}$.
+Therefore, the characteristic gap between the eigenvalues of the convective moment becomes smaller as the Rayleigh number increases.
+Additionally, the timescale of fluctuations in these values goes down because the density anomalies are moving faster.
+Overall, this makes the principal axes of high Rayleigh number systems much less stable, consistent with the result of \citet{richards1999polar}.
+
+In the limit that the eigengap becomes zero, a small perturbation $\delta$ can produce an arbitrary rotation of the principle axes.
+In effect, the location of the principal axes are set by the location of a small density anomaly.
+There is a close correspondence with  hypothesized ``inertial interchange true polar wander'' (IITPW) \citep{kirschvink1997evidence}.
+As noted above, the interchange of the maximum and intermediate axis can produce a large change in $\theta$ because the misalignment angle is suddenly measured from a new axis.
+An IITPW event can produce a large change in the orientation of the rotation axis (potentially $90^{\circ}$ if there are not further perturbations to the $\mathbf{E}$ tensor).
+However, the eigengap does not need to be zero for there to be large polar wander events.
+In addition, wander does not need to be $90^\circ$ after an interchange of maximum and intermediate axes due to the subsequent evolution of density anomalies during the event.
+We clarify these points using numerical simulations of convection in a 2D spherical annulus.
+
+Figure~\ref{fig:misfit} shows a representative time series from the convection calculations.
+As before, we compute the convective moment of inertia in the geographic frame and evaluate the eigenvalues of $\mathbf{E}$ to determine the eigengap.
+We also evolve the orientation of the rotation axis by integrating  Equation~\eqref{eq:liouville_ode} in time.
+Since it is a 2D model, the spin axis is described by a single angle. The misalignment angle $\theta$ also lies in the plane of the 2D model.
+When the eigengap gets small,  the corresponding misalignment angle increases and the rate of polar wander becomes much faster.
+This behavior is quantitatively consistent with the predictions from Equation (33).
+We draw special attention to two events in the time series.
+At $\sim$0.3 Gyr the eigengap dips and the misalignment angle increases to $\theta \sim 20^{\circ}$.
+There is a large change in the orientation of the spin axis, even though
+there is technically no interchange of the maximum and intermediate axes.
+A conventional IITPW event occurs at $\sim 1$ Gyr;  the eigengap goes to zero, interchanging the maximum and intermediate axes.
+The misfit angle goes to approximately $90^\circ$.
+During this time the orientation of the spin axis changes by roughly 90$^{\circ}$.
+
+Large TPW events are associated with inertial interchanges, but a broader class of events are evident
 when the principal moments are close to each other.
-Furthermore, a more vigorously convecting planet is much more likely to experience rapid TPW events
-as it has a lower characteristic gap between principal moments and the gap is much more likely to go to zero.
+We argue that large TPW events do not (necessarily) imply an inertial interchange.
+Conversely, we do not expect all large TPW events to be associated with a 90$^{\circ}$ change in the spin.
+Our scaling results suggest that large TPW events are more likely to occur in the early Earth when the planet is convecting more vigorously and the characteristic gap between principal moments is small.
 
 \citet{tsai2007theoretical} suggested that interchanges in the principal moments are unlikely to produce large
-TPW events, since (1) during an inertial interchange the total moment of inertia anomaly is small, and
-(2) the TPW event can only asymptotically approach 90$^\circ$.
-This is true for the case of a single interchange. However, we argue that this is overly restrictive
-for our generalization of IITPW.
-Our numerical experiments show that in a vigorously convecting system the principal moments can
-interchange, or almost interchange, quite rapidly, and the perturbation sizes can be large soon after an interchange.
-Furthermore, the location of the principal axes need not stay put during a TPW event,
-which can lead to extended polar wandering of distances greater than 90$^\circ$.
+TPW events because the rate of TPW is low. This implies that a large 90$^{\circ}$ change in the orientation of the rotation axis occurs over a long time.  While
+this is true, the numerical convection calculation reveals a more complex behavior. 
+First, the principal moments can interchange, or almost interchange, quite rapidly. In these cases  the perturbations in the convective moments can become large soon after an interchange. Second, the
+ the location of the principal axes do not stay put during a IITPW event.
+A persistent drift of the principal axes can can lead to extended polar wandering, yielding angular changes greater than 90$^\circ$.
 
@@
 
-The preceding results clarify the complex relationship between the Rayleigh number, $m$, and the rates of TPW for a convecting planet.
+The preceding results clarify the complex relationships that determine the rates of TPW for a convecting planet.
 As mantle convection redistributes mass in the planet's interior, the spin axis moves around to stay aligned with the principal axes of the convective moment of inertia. 
-There is a constant competition between growth of the mismatch angle $\theta$ through 
-Equation~\eqref{eq:grow_perturbation} and its relaxation through TPW.
+There is a constant competition between growth of the mismatch angle $\theta$ through convective perturbations
+ and its relaxation through TPW.
 \citet{goldreich1969some} envisioned an analogy of beetles crawling on the surface of the globe, with
 the spin axis trying to keep up with the instantaneous figure axis set by the beetles.
-Our analysis begins to answer the questions ``how big are the beetles?'' and ``how fast are they crawling?''
-
-We may plug in the estimate for $\Lambda_{ij}$ (Equation~\eqref{eq:lambda_estimate}) into Equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
-\begin{equation}
-\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{1}{m} \sin{2 \theta}.
-\label{eq:simplest_milankovitch}
-\end{equation}
-Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
-Response timescales for relaxation of the mismatch angle go down at high Rayleigh numbers.
-At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
-That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
-Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
-However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
-With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
-\begin{equation}
-\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
-\label{eq:simplest_milankovitch_defer_scaling}
-\end{equation}
-In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
-$d(\mathrm{Ra})$ will still result in a weak dependence of the prefactor in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} 
-on the Rayleigh number.
-
-We then suggest that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta$.
-Whereas the prefactor $\mathrm{Ra}^{1-3\beta}$ in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} has a weak dependence on $\mathrm{Ra}$, 
-the misalignment angle $\theta$ is expected to be strongly dependent on it.
+Our analysis begins to answer the questions ``how big are the beetles?'' and ``how fast are they crawling?’’.
+We find that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta$.
+Surprisingly, the direct dependence of $\dot{W}$ on $Ra$ is weak (see Equation (44)), but the indirect influence of $Ra$ on the misalignment angle $\theta$ is quite strong.
 
@@
 When convection is more chaotic, however, there may be large excursions in $\theta$, 
-which are driven by large values for $|\mitbf{\Psi}|$ when $(\lambda_i - \lambda_j)$ is small, according to Equation~\eqref{eq:grow_perturbation}. 
+which are driven by large changes in the orientation of the principal axes when $(\lambda_i - \lambda_j)$ is small.  
 In this case a dramatic increase in the rate of TPW is possible.
-If $\theta=90^\circ$, this corresponds to IITPW \citep{kirschvink1997evidence}.  
-This, however, is a special case in the large $\theta$, ``fast TPW'' regime.
-
-As an example, we may consider the early Earth, when the mantle was presumably hotter and less viscous, leading to a 
-higher Rayleigh number. We would then predict that convection was more vigorous, leading to 
-a less stable $\theta(t)$, and thus more TPW and more frequent ``fast TPW'' events.
+We refer to this as the `` fast TPW’’.
+Inertial interchanges occur when the maximum and intermediate axis are swapped and this can lead to a large change in
+ $\theta$ \citep{kirschvink1997evidence}.
+Increasing $\theta$ to $\pi/2$ does not necessarily require fast TPW, but our numerical simulations show that this is often the case.
+Moreover, the inertial interchange is only  a special case of the large $\theta$, ``fast TPW'' regime.
+We expect large $\theta$ and ``fast TPW’’ to be more prevalent in the early Earth, when the mantle was presumably hotter and less viscous (i.e.
+higher $Ra$).
+More vigorous convection increases the rate of change of density anomalies  and  makes
+$\theta(t)$ less stable. Large (and potentially fast) TPW events should have been more frequent.
 
 For Cenozoic Earth we can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch}.
-Typical values for the time constant $T_1$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
+Typical values for the time constant $\tau_R$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
 Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
@@
 which corresponds to about 66 cm/yr 90$^\circ$ from the TPW axis.
-This is similar to the rates discussed by \citet{cambiotti2011new}, (though our rate is larger since we allow for the
-possibility of a larger mismatch angle), and is within the range suggested by some interpretations of paleomagnetic data.
+This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the
+possibility of a larger mismatch angle. Such large angles may be necessary to explain the range of rates suggested by some interpretations of paleomagnetic data \citep{mitchell2011sutton}.
 The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
@@
 
-The most important parameters are the Rayleigh number and $m$, which acts as a damper to TPW.
-The dependence on the Rayleigh number is more complicated, since it is a control on both the forcing 
-of TPW and the response, which act in opposite directions.
-Overall, however, we expect that more vigorously convecting planets should be less rotationally stable, and experience more TPW.
+The most important parameters are the Rayleigh number and $m$, which defines the ratio of centrifugal to gravitational forces.
+Rates of TPW are proportional to  $m^{1}$ because the size of the equatorial bulge acts as a damper to TPW.
+The dependence on the Rayleigh number is more complicated.
+Large $Ra$ shorten the response time of the mantle, but also reduce the power in the degree-two part of the temperature field, which drives TPW.
+These two effects nearly cancel, although there is also an indirect influence of $Ra$ on the misalignment angle $\theta$.
+Overall, we expect a more vigorously convecting planet to be less rotationally stable, and experience more frequent large TPW events.
 This perspective allows us to consider not only the polar wandering of Phanerozoic Earth,
@@
 \section*{Acknowledgments}
-This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for careful and constructive comments that greatly improved the paper.
+This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard and Victor Tsai for careful and constructive comments that greatly improved the paper.
 We thank the Computational Infrastructure for Geodynamics (geodynamics.org) which is funded by the National Science Foundation under award NSF-0949446.
@@
 \bibliographystyle{elsarticle-harv} 
-\bibliography{tpw_rate}
+\bibliography{tpw_rate.bib}
 

### 2017-04-18 tpw_rate.tex
principle -> principal

@@
 \end{equation}
-Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for the principle values of $\mathbf{I}(t)$;  the eigenvectors define the principal axes and the eigenvalues correspond to the principal moments of inertia.
+Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for the principal values of $\mathbf{I}(t)$;  the eigenvectors define the principal axes and the eigenvalues correspond to the principal moments of inertia.
 The rotation axis is expected to align with the axis for the largest moment.
@@
 
-Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of changes in the orientation of the principle axes.
+Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of changes in the orientation of the principal axes.
 Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor after some time interval $\Delta t$, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.
@@
 
-In the limit that the eigengap becomes zero, a small perturbation $\delta$ can produce an arbitrary rotation of the principle axes.
+In the limit that the eigengap becomes zero, a small perturbation $\delta$ can produce an arbitrary rotation of the principal axes.
 In effect, the location of the principal axes are set by the location of a small density anomaly.

### 2017-05-01 tpw_rate.tex
More edits from Bruce.

--- 2017-04-17 tpw_rate.tex
@@
 Here we present scaling analyses and numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors. 
-For simple rotating convection, we identify a set of dimensionless parameters that characterize true polar wander.
+For simple rotating convection, we identify a set of dimensionless parameters that fully characterize true polar wander.
 We use these parameters to define  timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander.
@@
 The same frame is also used to measure TPW, but the expressions on the right-hand-side of Equation (13) are more naturally evaluated in the reference frame of the principal axes of $\mathbf{E}$.
-We can avoid the need to reconcile these two frames by confining our attention to the rate of polar wander (We denoted  the rate of TPW by 
+We can avoid the need to reconcile these two frames by confining our attention to the rate of TPW because scalar quantities are invariant under coordinate rotations.
+(We denoted  the rate of wander by
  $\dot{W} = \vert \dot{\mitbf{\omega}} \vert$) .
@@
 \begin{equation}
-\dot{\Theta}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
+\dot{W}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
 \label{eq:polar_wander_rate_tensor}
 \end{equation}
-The scalar terms on the right-hand-side of Equation (14) are now independent of the choice of coordinate system as long as we are consistent.
+The scalar terms on the right-hand-side of Equation (14) can now be evaluated in either the geographic or the $E$ frames.
 Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
@@
 Once the orientation of $\mitbf{\omega}$ is specified in the E-frame by angles $\theta$ and $\phi$, we can evaluate the rate of TPW in terms of the principal moments of $\mathbf{E}$.
-Two special cases of this equation are worth  noting. First, if $\mitbf{\omega}$  coincides with  the normal to the octahedral plane
+Two special cases of this equation are identified by noting a similarity in the form of Equation (13) to the shear stress on a plane in classical elastostatics.
+First, if $\mitbf{\omega}$  coincides with  the normal to the octahedral plane
 (i.e. the plane defined by a normal with direction cosines all equal to $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
@@
 \begin{equation}
-\frac{d\Theta}{dt^\prime} \sim \frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
+\frac{d W}{dt^\prime} \sim \frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
 \label{eq:simplest_milankovitch_defer_scaling}
@@
 Unstable conditions prevail when the maximum and intermediate moments of inertia are nearly equal, leading to large TPW events \citep{evans1998true}.
-(\citet{tsai2007theoretical} argue that large TPW events take a long time because the rate of TPW is small when $\Lambda_{31}$ is small or when $\theta = 0$ or $\pi/2$ (see Equation (33).
+On the hand hand, \citet{tsai2007theoretical} argue that large TPW events take a long time because the rate of TPW is small when $\Lambda_{31}$ is small or when $\theta = 0$ or $\pi/2$ (see Equation (33)).
 If the maximum and intermediate axes are interchanged when the misalignment angle is small $\theta \approx 0$, then the new definition of the maximum axis after the interchange ensures that $\theta \approx \pi/2$.
-In this case the rate of TPW may be small before and after the interchange).
+In this case the rate of TPW may be small before and after the interchange.
 
@@
 Additionally, the timescale of fluctuations in these values goes down because the density anomalies are moving faster.
-Overall, this makes the principal axes of high Rayleigh number systems much less stable, consistent with the result of \citet{richards1999polar}.
+Overall, this makes the principal axes of high Rayleigh number systems much less stable, consistent with the result of \citet{richards1999polar}, based on mantle convection calculations.
 
@@
 When the eigengap gets small,  the corresponding misalignment angle increases and the rate of polar wander becomes much faster.
-This behavior is quantitatively consistent with the predictions from Equation (33).
+This behavior is consistent with the predictions from Equation (33).
 We draw special attention to two events in the time series.
@@
 We argue that large TPW events do not (necessarily) imply an inertial interchange.
-Conversely, we do not expect all large TPW events to be associated with a 90$^{\circ}$ change in the spin.
+Conversely, we do not expect all large TPW events to be associated with a 90$^{\circ}$ change in the spin axis.
 Our scaling results suggest that large TPW events are more likely to occur in the early Earth when the planet is convecting more vigorously and the characteristic gap between principal moments is small.
 
-\citet{tsai2007theoretical} suggested that interchanges in the principal moments are unlikely to produce large
-TPW events because the rate of TPW is low. This implies that a large 90$^{\circ}$ change in the orientation of the rotation axis occurs over a long time.  While
-this is true, the numerical convection calculation reveals a more complex behavior. 
+\citet{tsai2007theoretical} suggested that large TPW events do not always accompany an interchange of the principal moments because the rate of TPW is low.
+A large 90$^{\circ}$ change in the orientation of the rotation axis can only occur over a long time.
+While
+this is true for idealized cases, the numerical convection calculation reveals a more complex behavior. 
 First, the principal moments can interchange, or almost interchange, quite rapidly. In these cases  the perturbations in the convective moments can become large soon after an interchange. Second, the
@@
  $\theta$ \citep{kirschvink1997evidence}.
-Increasing $\theta$ to $\pi/2$ does not necessarily require fast TPW, but our numerical simulations show that this is often the case.
+Increasing $\theta$ to $\pi/2$ does not necessarily require fast TPW, but our numerical simulations show that rapid motion often occurs because $\theta$ does not remain close to $\pi/2$.
 Moreover, the inertial interchange is only  a special case of the large $\theta$, ``fast TPW'' regime.
@@
 The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
-which results in a corresponding uncertainty for the relaxation time $T_1$ and the maximum polar wander rate.
+which results in a corresponding uncertainty for the relaxation time $\tau_R$ and the maximum polar wander rate.
 

### 2017-05-01 figures/reference_frames.py
Remove relative rotation vector from reference frames figure.

@@ -109,10 +109,6 @@ E = np.array([ [L1, 0, 0], [0, L2, 0], [0, 0, L3] ])
 theta = 30. * np.pi/180.
 phi = 0. * np.pi/180.
 omega = np.array([ np.sin(theta)*np.cos(phi), np.sin(theta)*np.sin(phi), np.cos(theta)])
-#relative motion of the two frames
-alpha = 50. * np.pi/180.
-beta = 110. * np.pi/180.
-psi = np.array([ np.sin(alpha)*np.cos(beta), np.sin(alpha)*np.sin(beta), np.cos(alpha)])
 # tpw vector
 omega_dot = np.dot(E, omega.T) - np.dot(omega, np.dot(E, omega.T))*omega.T
 #omega_dot = 0.3*omega_dot/norm(omega_dot)
@@ -153,28 +149,6 @@ span = np.linspace(0, 1., 100)
 guideline = np.array([np.cos(phi)*np.sin(theta)*np.ones_like(span), np.sin(phi)*np.sin(theta)*np.ones_like(span), np.cos(theta)*span])
 ax.plot( guideline[0,:], guideline[1,:], guideline[2,:], 'k--', alpha=alpha_val)
 
-# plot relative rotation
-psi_arrow = Arrow3D([o[0], psi[0]], [o[1], psi[1]], [o[2], psi[2]], **arrow_prop_dict)
-ax.add_artist(psi_arrow)
-plot_rotation_arrow(ax, psi_arrow, ccw=False)
-# add helper lines for psi
-span = np.linspace(0, 1., 100)
-guideline = np.array([np.cos(beta)*np.sin(alpha)*np.ones_like(span), np.sin(beta)*np.sin(alpha)*np.ones_like(span), np.cos(alpha)*span])
-ax.plot( guideline[0,:], guideline[1,:], guideline[2,:], 'k--', alpha=alpha_val)
-guideline = np.array([np.cos(beta)*np.sin(alpha)*(span), np.sin(beta)*np.sin(alpha)*(span), np.zeros_like(span)])
-ax.plot( guideline[0,:], guideline[1,:], guideline[2,:], 'k--', alpha=alpha_val)
-
-tmpvec = np.array([ np.cos(beta), np.sin(beta), 0.])
-guideline = np.array([z0 * s + psi * (1.-s) for s in span] )
[…]

### 2017-05-01 reply.tex
Add reply doc.

\documentclass[a4paper,12pt]{article}
\usepackage[utf8]{inputenc}

%opening
\title{Response to reviewer 1 (Victor Tsai)}
\author{}
\date{}

\begin{document}

\maketitle

We thank Victor for a comprehensive review of our paper. We agree that there are new and intriguing results in this paper.
We also acknowledge that the original manuscript lacked clarity on several key concepts.
We also value and appreciate the suggestion of making greater points of contact with the current literature.
The revised manuscript has been substantially modified to address all of the reviewer's comments.
We have elaborated on key points, expanded our discussion of prior work, and altered the organization of the paper such that each new section builds directly on the earlier results.
All of these changes have improved the presentation.
We think these changes will also erase any confusion between $\Theta$ and $\theta$.

The confusion between $\theta$ and $\Theta$ is partly connected to the definition of Inertial Interchange True Polar Wander (IITPW).
Victor's view is clearly expressed in his review:
\begin{quote}
Only if Theta eventually ends up close to 90 degrees would it be IITPW.
\end{quote}
We have a slightly different view, but this difference is largely a question of semantics.
To us an inertial interchange occurs when the maximum and intermediate moments of inertia are interchanged.
Victor describes IITPW as the long-time consequence of the interchange.
Both perspectives are reasonable, and actually describe slightly different things, so the key to resolving this question in the paper is to emphasize the physics of the process.
On this point, we do not think there is any disagreement between us and the reviewer.
We have added text throughout the paper to address all of the points in the second paragraph of the review (lines 478, 483, 530) and earlier in the paper (e.g. line 104).
We also follow the advice of renaming one of the thetas.
We hesitate to annotate these changes in the text using a different colored font because these changes are fairly pervasive and (in places) extensive.

The third paragraph of the review deals with the difficulty for the reader to ``understand what the main results of the manuscript were''.
We readily acknowledge this point.
Several key concepts were developed in isolated, self-standing sections.
We impose a heavy burden on the readers by asking them to put the pieces together for themselves.
A reorganization of the paper permits a more systematic development of the ideas; each new results builds on the previous results, so the connections are much clearer.
We also place signposts at key points to ensure that the readers does not lose sight of the end goal.
The reviewer specifically mentions the appears of additional equations in the Discussion section.
All of these concerns are eliminated by the reorganization.

Victor goes on to make valuable suggestions in the fourth paragraph of the review about linking our work to existing literature.
This is a fair criticism.
This is not only reasonable, but it also helps to emphasize the main contributions. We implements all of the specific suggestions connected with Tsai \& Stevenson (2007), and we make similar additions connected with the studies
of Ricard et al. (1993); Richards et al. (1999) and Cambiotti et al. (2011).

Additional Comments/suggestions:

\begin{itemize}

\item Paragraphs added to introductory clauses.

\item Caption to Fig. 1 is reworded to clarify.

\item Use of $T$ to denote temperature and $T_1$ denote relaxation time is avoiding by using $\tau$ for all timescales.

\item Line 396 is reworded, as suggested.

\item The reference to a theorem in D\&K1970 is move to the place where the theorem is stated.

\item The text around line 456 is substantially modified. It was possible to streamline the text and improve clarity.

\item Lines 468/469 were confusing. The text is revised and the typo is corrected.

\item How representative is Fig. 4? The answer is ``very''. There was no need to be selective about the examples we showed. The events in question are always present. We completely agree that the rate of TPW is small when theta = 90. This fact is acknowledged in the revised manuscript and supported by pointed to this expectation in our equation for the rate of TPW. However, embedding this equation in a mantle convection simulation reveals some interesting complications because the mass anomalies that cause TPW are always in motion.

\end{itemize}

We appreciate Victor's thoughtful suggestions.

\end{document}

### 2017-05-01 tpw_rate.tex
T_1 => \tau_R

@@
 
-%in case I want to add detail to any derivation
-\newif\ifdetail
-\detailfalse
-%\detailtrue
-
 \journal{Physics of Earth and Planetary Interiors}
@@
 \begin{equation}
-\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right).
+\frac{k_f a^5 \tau_R \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right).
 \label{eq:liouville}
@@
 \begin{equation}
- \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right].
+ \dot{\mitbf{\omega}}  = \frac{1}{(C-A)\tau_R} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right].
 \label{eq:liouville_ode}
@@
 \begin{equation}
-\dot{W}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
+\dot{W}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 \tau_R^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
 \label{eq:polar_wander_rate_tensor}
@@
 \dot{W}^2 = 
-&\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
-&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
+&\frac{1}{4(C-A)^2 \tau_R^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
+&\frac{1}{4(C-A)^2 \tau_R^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
 \label{eq:polar_wander_rate}
@@
 \dot{W}^2
-&= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
-&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2}. \\
+  &= \frac{1}{9(C-A)^2 ]\tau_R^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
+  &= \frac{E_{\mathrm{II}}}{9(C-A)^2 \tau_R^2}. \\
 \end{aligned}
@@
 \vert \dot{W} \vert = 
-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1)\, .
+\frac{1}{2(C-A) \tau_R}\sin{2\theta} ( \lambda_3 - \lambda_1)\, .
 \label{eq:simple_milankovitch}
@@
 which may be rewritten via the Jacobi identity to find
-\ifdetail
-\begin{equation}
-\begin{aligned}
-- \int_V \rho (\mitbf{\Omega}\times \mathbf{r}) \times ( \mathbf{r} \times \mitbf{\Omega} ) \,dV + 
-- \int_V \rho \mitbf{\Omega}\times(\mitbf{\Omega}\times \mathbf{r}) \times \mathbf{r} \,dV &= 0 \\
-\end{aligned}
-\end{equation}
-\fi
 \begin{equation}
@@
 
-Equation (17) defines the rate of TPW in terms of four physical parameters: $(C-A)$, $T_1$, $(\lambda_3-\lambda_1)$ and $\theta$.
+Equation (17) defines the rate of TPW in terms of four physical parameters: $(C-A)$, $\tau_R$, $(\lambda_3-\lambda_1)$ and $\theta$.
 This equation can be expressed in nondimensional form using the dimensionless numbers from Table 2.
@@
 \begin{equation}
-\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 6^\circ / \mathrm{Myr},
+\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{\tau_R} \sim 6^\circ / \mathrm{Myr},
 \end{equation}
@@
 Solving for $( \lambda_i - \lambda_j )$ in terms of these harmonics, we find
-\ifdetail
-
-It is useful to define unnormalized versions of the spherical harmonics as intermediate products:
-\begin{equation}
-\begin{aligned}
-Y_{20}^* &= 2 z^2 - x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{5} } r^2 Y_{20} \\ 
-Y_{22}^* &= x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{15} } r^2 Y_{22}.
-\end{aligned}
-\end{equation}
-These make it easy to rewrite the differences between the $\lambda_n$:
-\begin{equation}
-\begin{aligned}
-(\lambda_2 - \lambda_1) &= \int_V \rho Y_{22}^* \, dV \\
-(\lambda_3 - \lambda_1) &= \frac{1}{2} \int_V  \rho \left( Y_{22}^* - Y_{20}^* \right) dV \\
-(\lambda_3 - \lambda_2) &= -\frac{1}{2} \int_V \rho \left( Y_{22}^* + Y_{20}^*  \right) dV. \\
-\end{aligned}
-\end{equation}
-Substituting back in for the normalized spherical harmonics we find
-\fi
 \begin{equation}
@@
 We normalize the differences in eigenvalues by the reference moment $I_0$:
-\ifdetail
-\begin{equation}
-\begin{aligned}
-\int_{V_S} \rho_0 \frac{2 r^2 - Y_{20}^*}{3} &= \int_{V_S} \rho_0 (x^2 + y^2) dV = I_0 \\
-\end{aligned}
-\end{equation}
-\begin{equation}
-I_0 = \frac{2}{3} \int_{V_S} \rho_0 r^2 \left(1 - 2 \sqrt{\frac{\pi}{5}} Y_{20}\right) \,dV
-\end{equation}
-\fi
 \begin{equation}
@@
 \end{equation}
-\ifdetail
-\begin{equation}
-\begin{aligned}
-\Lambda_{21} &= \alpha \Delta T \frac{-6 \sqrt{\frac{\pi}{15} } \int_{V_S} T^\prime r^2  Y_{22} \,dV}{\int_{V_S} r^2 \, dV} \\
-\Lambda_{31} &= \alpha \Delta T \frac{-3 \sqrt{ \frac{\pi}{15} } \int_{V_S} T^\prime r^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV}{\int_{V_S} r^2 \, dV}\\
-\Lambda_{32} &= \alpha \Delta T \frac{3 \sqrt{ \frac{\pi}{15} }  \int_{V_S}  T^\prime r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV}{\int_{V_S} r^2 \, dV} \\
-\end{aligned}
-\end{equation}
-\begin{equation}
-\begin{aligned}
-\Lambda_{21} &= \Gamma \frac{-6 \sqrt{\frac{\pi}{15} } \int_{V_S^\prime} T^\prime {r^\prime}^2  Y_{22} \,dV^\prime}{\int_{V_S^\prime} {r^\prime}^2 \,dV^\prime} \\
-\Lambda_{31} &= \Gamma \frac{-3 \sqrt{ \frac{\pi}{15} } \int_{V_S^\prime} T^\prime {r^\prime}^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV^\prime}{\int_{V_S} {r^\prime}^2 \, dV^\prime}\\
-\Lambda_{32} &= \Gamma \frac{3 \sqrt{ \frac{\pi}{15} }  \int_{V_S^\prime}  T^\prime {r^\prime}^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV^\prime}{\int_{V_S} {r^\prime}^2 \, dV^\prime} \\
-\end{aligned}
-\end{equation}
-\fi
 Dividing Equation~\eqref{eq:degree_two_moments_of_temperature} by $I_0$ and nondimensionalizing 

### 2017-05-01 tpw_rate.tex
Turn around a couple of sentences.

@@
 where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity vector.
-On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
+On a dynamic planet $\mathbf{I}$ may be a function of time, so $\mitbf{\Omega}$ must be also vary with time to conserve angular momentum.
 In this case, Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
@@
 
-For mantle convection problems the moment of inertia tensor is commonly separated into three parts \citep{sabadini1981pleistocene, spada1992excitation}:
+The moment of inertia tensor is commonly separated into three parts for mantle convection problems \citep{sabadini1981pleistocene, spada1992excitation}:
 \begin{equation}

### 2017-05-01 reply.tex
Edits to reply.

@@
 
-We thank Victor for a comprehensive review of our paper. We agree that there are new and intriguing results in this paper.
+We thank Victor for his comprehensive review of our paper.
+We agree that there are new and intriguing results in this paper.
 We also acknowledge that the original manuscript lacked clarity on several key concepts.
@@
 
-The confusion between $\theta$ and $\Theta$ is partly connected to the definition of Inertial Interchange True Polar Wander (IITPW).
+The confusion between $\theta$ and $\Theta$ is partly connected to the definition of inertial interchange true polar wander (IITPW).
 Victor's view is clearly expressed in his review:
@@
 We also place signposts at key points to ensure that the readers does not lose sight of the end goal.
-The reviewer specifically mentions the appears of additional equations in the Discussion section.
-All of these concerns are eliminated by the reorganization.
+The reviewer specifically mentions the appears of additional equations in the ``Discussion'' section.
+We hope that these concerns are eliminated by the reorganization.
 
@@
 This is not only reasonable, but it also helps to emphasize the main contributions.
-We implements all of the specific suggestions connected with Tsai \& Stevenson (2007), and we make similar additions connected with the studies
-of Ricard et al. (1993); Richards et al. (1999) and Cambiotti et al. (2011).
+We implement all of the specific suggestions connected with Tsai \& Stevenson (2007), and we make similar additions connected with the studies of 
+Ricard et al. (1993); Richards et al. (1999) and Cambiotti et al. (2011).
 
-Additional Comments/suggestions:
+Responses to the additional comments/suggestions:
 
-\begin{itemize}
+\begin{enumerate}
 
@@
 There was no need to be selective about the examples we showed.
-The events in question are always present.
+The events in question are always present when convection is sufficiently vigorous.
 We completely agree that the rate of TPW is small when theta = 90.
@@
 
-\end{itemize}
+\end{enumerate}
 

### 2017-05-01 reply.tex
Reword statement in response.
Edits to reply.

@@
 
-\item Paragraphs added to introductory clauses.
+\item We add commas to or invert the relevant sentences.
 
@@
 
-\item Line 396 is reworded, as suggested.
+\item Line 396 does not assume $\theta$ to be small. We have elaborated on the cited literature to clarify this point.
 
@@
 We completely agree that the rate of TPW is small when theta = 90.
-This fact is acknowledged in the revised manuscript and supported by pointed to this expectation in our equation for the rate of TPW.
+This fact is acknowledged in the revised manuscript and confirmed in our equation for the rate of TPW.
 However, embedding this equation in a mantle convection simulation reveals some interesting complications because the mass anomalies that cause TPW are always in motion.

### 2017-05-01 tpw_rate.tex
Edits to reply.

@@
 
- Much of the debate about the magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.
+ Much of the debate about the magnitude of TPW on Earth comes down to the question of how big $\theta$ can be.
+\citet{kirschvink1997evidence} hypothesized about an event where $\theta$ was approximately 90$^\circ$, while many geodynamic simulations of recent TPW assume that $\theta$ is small \citep{steinberger1997changes}.
 In the next section we seek to characterize the maximum value for $\theta$ in terms of the dimensionless parameters.

### 2017-05-01 tpw_rate.tex
Formatting.

@@
 the spin axis trying to keep up with the instantaneous figure axis set by the beetles.
-Our analysis begins to answer the questions ``how big are the beetles?'' and ``how fast are they crawling?’’.
+Our analysis begins to answer the questions ``how big are the beetles?'' and ``how fast are they crawling?''.
 We find that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta$.
@@
 In this case a dramatic increase in the rate of TPW is possible.
-We refer to this as the `` fast TPW’’.
+We refer to this as the ``fast TPW''.
 Inertial interchanges occur when the maximum and intermediate axis are swapped and this can lead to a large change in
@@
 Moreover, the inertial interchange is only  a special case of the large $\theta$, ``fast TPW'' regime.
-We expect large $\theta$ and ``fast TPW’’ to be more prevalent in the early Earth, when the mantle was presumably hotter and less viscous (i.e.
+We expect large $\theta$ and ``fast TPW'' to be more prevalent in the early Earth, when the mantle was presumably hotter and less viscous (i.e.
 higher $Ra$).

### 2017-05-01 tpw_rate.tex
Add a sentence on Cambiotti.

@@
 
-For Cenozoic Earth we can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch}.
+We can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch} to obtain an estimate of maximum TPW rates for Earth.
 Typical values for the time constant $\tau_R$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
@@
 A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
-\citet{richards1997explanation} have argued that the convective planform of Earth has been stable for the last
-few hundred million years. On the other hand, we know that there have been large reorganizations of that planform 
-during Earth history, so this recent geologic stability may not hold in general \citep{evans2003true}.
+\citet{richards1997explanation} performed TPW simulations based on Cenozoic plate motion reconstructions, and argued that the convective planform of Earth has been stable for the last few hundred million years.
+\citet{cambiotti2011new} took a similar approach, but allowed for the position of the spin axis to lag the principal axes (i.e.
+nonzero $\theta$).
+In that study they found values for $\theta$ of up to 3-7$^\circ$.
+On the other hand, we know that there have been large reorganizations of that planform during Earth history,
+so this relative stability of the spin-axis may not hold in general \citep{evans2003true}.
 Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.

### 2017-09-11 figures/eigengap.py
Add environment info.

@@ -1,3 +1,5 @@
+from __future__ import print_function
+
 import matplotlib
 import numpy as np
 import scipy as sp
@@ -13,9 +15,9 @@ import glob
 try:
   plt.style.use('./tpw_rate.mplstyle')
 except AttributeError:
-  print "old matplotlib version?"
+  print("old matplotlib version?")
 except ValueError:
-  print "Cannot find the requested stylesheet"
+  print("Cannot find the requested stylesheet")
 
 #c = ['#0A5F02', '#1c567a', '#814292', '#d7383b', '#fdae61', '#c0f8b8']
 #c = ['#e41a1c', '#377eb8', '#4daf4a', '#984ea3', '#ff7f00', '#ffff33'] 
@@ -38,7 +40,7 @@ frac = 50
 
 def get_spin_spectrum(text_file):
 
-  print text_file
+  print(text_file)
   #load the data
   data = np.loadtxt(text_file,skiprows=50, usecols=(1,17,18,19,20))
   #get rid of duplicate times (why might these exist?)
@@ -72,7 +74,7 @@ def get_spin_spectrum(text_file):
 output_files = [f for f in glob.glob("statistics_*")]
 sort_key = lambda s: float(s.split('_')[1])
 output_files.sort(key=sort_key)
-print output_files
+print(output_files)
 
 amplitudes = []
 Rayleighs = []
@@ -112,7 +114,7 @@ ax.set_ylim(0, 8.0e-5)
 ax.set_xlabel(r'Time (Gyr)')
 ax.set_ylabel(r'$\Lambda_{21} = (\lambda_2-\lambda_1)/I_0$')
 ax.legend(loc="upper right")
-print np.polyfit(np.log(Rayleighs[3:]), np.log(amplitudes[3:]),1)
+print(np.polyfit(np.log(Rayleighs[3:]), np.log(amplitudes[3:]),1))
 
 ax = plt.subplot(122)
 ax.grid(False)

### 2017-09-11 figures/misfit.py

@@ -9,9 +9,9 @@ import scipy.interpolate as interp
 try:
   plt.style.use('./tpw_rate.mplstyle')
 except AttributeError:
-  print "old matplotlib version?"
+  print("old matplotlib version?")
 except ValueError:
-  print "Cannot find the requested stylesheet"
+  print("Cannot find the requested stylesheet")
 
 data = np.loadtxt('statistics_500', skiprows=50, usecols=(1,17,18,19,20))
 

### 2017-09-11 tpw_rate.tex
Updates for submitting revisions.
Make simple fixes.

@@
 This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and relates them to a  broader class of events that enable large and often rapid TPW.
-We expect these events to have been more frequent in Earth’s past.
+We expect these events to have been more frequent in Earth's past.
 \end{abstract}
@@
 
-%%\linenumbers
+\linenumbers
 
@@
 \end{equation}
-An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
+An alternative to the Love number formalism is to calculate surface deflections directly using mantle convection simulations with a true free surface boundary condition.
 A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
@@
 this is true for idealized cases, the numerical convection calculation reveals a more complex behavior. 
-First, the principal moments can interchange, or almost interchange, quite rapidly. In these cases  the perturbations in the convective moments can become large soon after an interchange. Second, the
- the location of the principal axes do not stay put during a IITPW event.
+First, the principal moments can interchange, or almost interchange, quite rapidly. In these cases  the perturbations in the convective moments can become large soon after an interchange.
+Second, the location of the principal axes do not stay put during a IITPW event.
 A persistent drift of the principal axes can can lead to extended polar wandering, yielding angular changes greater than 90$^\circ$.

### 2017-09-11 figures/misfit.py
Add tpw rate to figure.

@@ -41,33 +41,46 @@ eigmean = np.mean(eig1 + eig2)/2.
 eigdiff = (eig1-eig2)/eigmean
 
 spin = data[:,1]
+spin_unwrap = np.unwrap(spin*np.pi*2./180.)*180./np.pi/2.
+spin_rate = interp.UnivariateSpline(time, spin_unwrap).derivative()(time)
 
 pos = np.where(np.abs(np.diff(spin)) >= 10.0)[0]
 time[pos] = np.nan
 spin[pos] = np.nan
 
 fig = plt.figure()
-ax = fig.add_subplot(211)
+ax = fig.add_subplot(311)
 
 ax.plot(time, eigdiff, label=r'$\Lambda_{21} = (\lambda_2-\lambda_1)/I_0$')
-plt.xlim(tmin, tmax)
-plt.ylim(0., 2.e-5)
+ax.set_xlim(tmin, tmax)
+ax.set_ylim(0., 2.e-5)
 ax.legend(loc='upper right')
 ax.xaxis.set_major_formatter(matplotlib.ticker.NullFormatter())
 ax.yaxis.set_major_locator(matplotlib.ticker.MaxNLocator(nbins=4, prune=None))
-plt.ylabel(r'Eigengap')
+ax.set_ylabel(r'Eigengap')
 
-ax = fig.add_subplot(212)
+ax = fig.add_subplot(312)
 
 ax.plot(np.array([]), np.array([]))
 ax.plot(time, spin, label=r'Spin axis')
 ax.plot(np.array([]), np.array([]))
 ax.plot(time, theta(time), label=r'$\theta$')
-plt.xlim(tmin, tmax)
-plt.ylim(0., 180.)
+ax.set_xlim(tmin, tmax)
+ax.set_ylim(0., 180.)
 ax.legend(loc='upper right')
+ax.xaxis.set_major_formatter(matplotlib.ticker.NullFormatter())
 ax.yaxis.set_major_locator(matplotlib.ticker.MaxNLocator(nbins=3, prune=None))
-plt.xlabel(r"Time (Gyr)")
-plt.ylabel(r"Angle ($^\circ$)")
+ax.set_ylabel(r"Angle ($^\circ$)")
+
+ax = fig.add_subplot(313)
+
+ax.plot(np.array([]), np.array([]))
+ax.plot(np.array([]), np.array([]))
[…]

### 2017-09-11 tpw_rate.tex
Move text block for better flow.
Minor edits.

@@
 
- Much of the debate about the magnitude of TPW on Earth comes down to the question of how big $\theta$ can be.
-\citet{kirschvink1997evidence} hypothesized about an event where $\theta$ was approximately 90$^\circ$, while many geodynamic simulations of recent TPW assume that $\theta$ is small \citep{steinberger1997changes}.
-In the next section we seek to characterize the maximum value for $\theta$ in terms of the dimensionless parameters.
-
 \subsection{An estimate for the angular mismatch angle ($\theta$)}
 \label{sec:theta}
+
+Much of the debate about the magnitude of TPW on Earth comes down to the question of how big $\theta$ can be.
+\citet{kirschvink1997evidence} hypothesized about an event where $\theta$ was approximately 90$^\circ$,
+while many geodynamic simulations of recent TPW assume that $\theta$ is small \citep{steinberger1997changes}.
+In this section we seek to characterize the maximum value for $\theta$ in terms of the dimensionless parameters.
 
@@
 \caption{Top: Time series of principal moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.
-Bottom: Time series of spin axis and mismatch angle $\theta$.
+Center: Time series of spin axis and mismatch angle $\theta$.
+Bottom: Time series of TPW rate.
 When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger.
@@
 nonzero $\theta$).
-In that study they found values for $\theta$ of up to 3-7$^\circ$.
+In that study they found values for $\theta$ of up to 3$^\circ$-7$^\circ$.
 On the other hand, we know that there have been large reorganizations of that planform during Earth history,
@@
 \end{equation}
-which corresponds to about 66 cm/yr 90$^\circ$ from the TPW axis.
+which corresponds to about 66 cm/yr at a point 90$^\circ$ from the TPW axis.
 This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the

### 2017-09-11 figures/misfit.py

@@ -76,11 +76,12 @@ ax = fig.add_subplot(313)
 
 ax.plot(np.array([]), np.array([]))
 ax.plot(np.array([]), np.array([]))
-ax.plot(time, spin_rate/1000., label=r'Spin axis')
+ax.plot(time, np.abs(spin_rate)/1000., label=r'Spin axis')
 ax.set_xlim(tmin, tmax)
+ax.set_ylim(0., 6.)
 ax.set_xlabel(r"Time (Gyr)")
 ax.set_ylabel(r'TPW rate ($^\circ$/Myr)')
 
 
 plt.savefig("misfit.pdf", bbox_inches='tight')
-plt.show()
+#plt.show()

### 2017-09-11 tpw_rate.tex
Wordcrafting.
Stronger statement.

@@
 where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity vector.
-On a dynamic planet $\mathbf{I}$ may be a function of time, so $\mitbf{\Omega}$ must be also vary with time to conserve angular momentum.
+On a dynamic planet $\mathbf{I}$ is a function of time, so $\mitbf{\Omega}$ must also vary with time to conserve angular momentum.
 In this case, Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
@@
 \end{equation}
-This infinite-time limit  ensures that the left-hand side of Equation (4) vanishes. It follows that the contribution from the convective perturbation $E_{ij}$ also vanishes, so TPW is not permitted.
+This infinite-time limit  ensures that the left-hand side of Equation (4) vanishes.
+It follows that the contribution from the convective perturbation $E_{ij}$ also vanishes, so no further TPW is permitted.
 

### 2017-09-11 tpw_rate.tex
Add clarifying clause.

@@
 Since it is a 2D model, the spin axis is described by a single angle. The misalignment angle $\theta$ also lies in the plane of the 2D model.
-When the eigengap gets small,  the corresponding misalignment angle increases and the rate of polar wander becomes much faster.
+When the eigengap gets small,  the corresponding misalignment angle increases.
+Even though at that moment the driving force may be small,
+it can recover quickly due to convection, and the rate of polar wander becomes much faster.
 This behavior is consistent with the predictions from Equation (33).

### 2017-09-13 tpw_rate.tex
Edits after conversation with Bruce.

@@
 where $G$ is the gravitational constant. 
-For the purposes of scaling, we assume that the magnitude of the gravitational acceleration, $g_0$, is the approximately constant (as is the case for Earth's mantle).
+For the purposes of scaling, we assume that the magnitude of the gravitational acceleration, $g_0$, is approximately constant (as is the case for Earth's mantle).
 In addition we use the simple equation of state
@@
 T_{\text{degree-two}}(t) \sim \frac{1}{N_{\text{modes}}} \sim \left( \frac{d}{R}\right)^3.
+\label{eq:degree_two_of_lengthscale}
 \end{equation}
@@
 At the same time, the  coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
-That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
+\citet{tsai2007theoretical} arrive at a similar result by a different path.
+Their estimate for the ratio of the convective timescale to the timescale for TPW depends only on $m$ (see their Equation~(4)).
+That the two results are similar is surprising, as their estimate of $T_\text{degree-two}$ varies linearly with $d/R$,
+rather than with $(d/R)^3$ (Equation~\eqref{eq:degree_two_of_lengthscale}).
+
+The cancellation of explicit $\mathrm{Ra}$ dependence is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
 Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
@@
 Much of the debate about the magnitude of TPW on Earth comes down to the question of how big $\theta$ can be.
-\citet{kirschvink1997evidence} hypothesized about an event where $\theta$ was approximately 90$^\circ$,
-while many geodynamic simulations of recent TPW assume that $\theta$ is small \citep{steinberger1997changes}.
+\citet{kirschvink1997evidence} hypothesized about an inertial interchange event, where $\theta$ alters by approximately 90$^\circ$,
+and can lead to large ($\sim90^\circ$) TPW over time (a process which they termed ``inertial interchange TPW, or IITPW).
+Conversely, many geodynamic simulations of recent TPW assume that $\theta$ is small \citep{steinberger1997changes}.
 In this section we seek to characterize the maximum value for $\theta$ in terms of the dimensionless parameters.
@@
 Unstable conditions prevail when the maximum and intermediate moments of inertia are nearly equal, leading to large TPW events \citep{evans1998true}.
-On the hand hand, \citet{tsai2007theoretical} argue that large TPW events take a long time because the rate of TPW is small when $\Lambda_{31}$ is small or when $\theta = 0$ or $\pi/2$ (see Equation (33)).
+On the other hand, \citet{tsai2007theoretical} point out that large TPW events take a long time because the rate of TPW is small when $\Lambda_{31}$ is small or when $\theta = 0$ or $\pi/2$ (see Equation (33)).
 If the maximum and intermediate axes are interchanged when the misalignment angle is small $\theta \approx 0$, then the new definition of the maximum axis after the interchange ensures that $\theta \approx \pi/2$.
@@
 In effect, the location of the principal axes are set by the location of a small density anomaly.
-There is a close correspondence with  hypothesized ``inertial interchange true polar wander'' (IITPW) \citep{kirschvink1997evidence}.
-As noted above, the interchange of the maximum and intermediate axis can produce a large change in $\theta$ because the misalignment angle is suddenly measured from a new axis.
-An IITPW event can produce a large change in the orientation of the rotation axis (potentially $90^{\circ}$ if there are not further perturbations to the $\mathbf{E}$ tensor).
+This has a close correspondence with hypothesized IITPW \citep{kirschvink1997evidence}.
+As noted above, the interchange of the maximum and intermediate axex can produce a large change in $\theta$ because the misalignment angle is suddenly measured from a new axis.
+The resulting TPW event can produce a large change in the orientation of the rotation axis (potentially $90^{\circ}$ if there are not further perturbations to the $\mathbf{E}$ tensor).
 However, the eigengap does not need to be zero for there to be large polar wander events.
@@
 At $\sim$1 Gyr the gap goes to zero and there is a nearly $90^\circ$ TPW event, with $\sim80^\circ$ degrees of polar wander in $\sim$30 Myr.
-However, there are several other large TPW events which happen when the eigengap is small.}
+However, there are several other large TPW events which happen when the eigengap is small.
+These results are fairly characteristic of high $\mathrm{Ra}$ convection.}
 \label{fig:misfit}
@@
 \end{equation}
-which corresponds to about 66 cm/yr at a point 90$^\circ$ from the TPW axis.
+which corresponds to about 66 cm/yr at a point 90$^\circ$ from the TPW axis (and is comparable to the maximum rates seen in Figure~\ref{fig:misfit}.
 This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the

### 2017-09-13 tpw_rate.tex
Close quotes

@@
 \citet{kirschvink1997evidence} hypothesized about an inertial interchange event, where $\theta$ alters by approximately 90$^\circ$,
-and can lead to large ($\sim90^\circ$) TPW over time (a process which they termed ``inertial interchange TPW, or IITPW).
+and can lead to large ($\sim90^\circ$) TPW over time (a process which they termed ``inertial interchange'' TPW, or IITPW).
 Conversely, many geodynamic simulations of recent TPW assume that $\theta$ is small \citep{steinberger1997changes}.

### 2017-09-13 tpw_rate.tex
Fix typo.

@@
 This has a close correspondence with hypothesized IITPW \citep{kirschvink1997evidence}.
-As noted above, the interchange of the maximum and intermediate axex can produce a large change in $\theta$ because the misalignment angle is suddenly measured from a new axis.
+As noted above, the interchange of the maximum and intermediate axes can produce a large change in $\theta$ because the misalignment angle is suddenly measured from a new axis.
 The resulting TPW event can produce a large change in the orientation of the rotation axis (potentially $90^{\circ}$ if there are not further perturbations to the $\mathbf{E}$ tensor).
@@
 \end{equation}
-which corresponds to about 66 cm/yr at a point 90$^\circ$ from the TPW axis (and is comparable to the maximum rates seen in Figure~\ref{fig:misfit}.
+which corresponds to about 66 cm/yr at a point 90$^\circ$ from the TPW axis (and is comparable to the maximum rates seen in Figure~\ref{fig:misfit}).
 This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the

### 2017-09-14 reply_0917.tex
Add yet another reply.

\documentclass[a4paper,12pt]{article}
\usepackage[utf8]{inputenc}

%opening
\title{Response to reviewer 1 (Victor Tsai)}
\author{}
\date{}

\begin{document}

\maketitle

We thank Victor for his thorough review of our paper.
We reproduce his comments here, with inline replies.

\begin{quote}
In abstract, ``Earths'' should be ``Earth's''

Line 44: ``must be also'' should be ``must also''
\end{quote}
These have been fixed.

\begin{quote}
Line 106-107: It is not clear what is meant by this statement. Do the authors mean ``no \_further\_ TPW is permitted'' after the fluid limit is achieved? Clearly there is finite TPW for the infinite time point relative to a given earlier time point.
\end{quote}
This clarification is a nice suggestion, and we have included it.

\begin{quote}
Line 132: ``calculate to'' should be ``calculate''
\end{quote}
This has been fixed.

\begin{quote}
Line 395-400: A result similar to this is discussed in Section 2, paragraph 15 of Tsai \& Stevenson (2007) and their Equation 4. In particular, they also find that many factors cancel out when comparing how TPW scales with Rayleigh number. The result in Equation 43 appears to be different than the T\&S result, but it would be useful to compare the present result to that earlier result, and note any similarities and differences.
\end{quote}
We have included a few sentences discussing the similar result in T\&S.
In particular, there is a different scaling for the dependence of the moment-of-inertia
anomalies upon the Rayleigh number.

\begin{quote}
Line 409-410: There appears to be a jump in topic, without a transition. A transition would be welcome.
\end{quote}
We have reorganized the sentences here, moving them to the start of the appropriate section rather than the end of the preceding section. We hope that this clarifies the narrative.

\begin{quote}
Line 410-415: Apparently there still appears to be differences in semantics between the authors’ usage and what I am familiar with as common usage. As far as I am aware, TPW is always defined as the relative reorientation of the rotation axis and the geographical axes. IITPW is thus a specific type of (hypothesized) TPW in which this relative reorientation reaches angles close to 90 degrees. It is true that the mechanism proposed behind “inertial interchange” relates to theta being close to 90 degrees, but I think the authors should still be clearer about this. Perhaps it is a difference between ``II'' and ``IITPW'': ``II'' is the mechanism (related to theta); ``IITPW'' is a resulting TPW (not related to theta directly, but indirectly through its suspected mechanism). Although this is indeed a semantic issue, it is an important semantic issue that I am sure many readers might be confused by. Partly, this paragraph is also confusing because many authors (like Tsai \& Stevenson) clearly recognize that theta can be large (close to 90 degrees), but still have a limit on TPW rates, which makes the earlier statement misleading.
\end{quote}
This semantic issue has been a bit sticky, but we take Victor's point.
We have clarified some of the language around use of IITPW, using ``inertial interchange''
when referring to the forcing for this particular mode of polar wander, and IITPW
for the 90 degree response to that forcing. In particular, we have changed some of the language
around lines 410-423 and 475-480.

\begin{quote}
Line 485-486: It is not clear why this statement is true. In particular, Equation 33 has dW/dt=0 if Lambda\_31 = 0. So, it is not clear why TPW is fast. Presumably, it is because Lambda\_31 grows fast enough that it quickly becomes large enough to drive significant TPW, but this is not obviously true. In any case, some further discussion would be warranted. It would also be useful if TPW rates were shown on Fig4, since it is difficult to tell how fast the TPW is (due to the compressed x axis).
\end{quote}
We have added an additional sentence clarifying our intent. We do indeed mean the $\Lambda_{31}$
grows fast enough that it quickly is able to drive TPW.
Furthermore, we have added the suggested panel to Figure 4, showing TPW rates of a few $^\circ$/Myr.
These rates are consistent with the rates derived in our scaling analyses.

\begin{quote}
Line 502-504: I believe the Tsai \& Stevenson (2007) suggestion is not so different from what is proposed. T\&S certainly emphasized that immediately after interchange, the forcing is zero, but that it can grow as long as there is sufficient time for that growth to occur. As mentioned in the previous review, the question appears to be how representative Fig4 is. In their response, the authors respond that it is ``very'' representative at high Rayleigh number. I do not doubt that, but the text is not very clear about this, and many readers might misinterpret the authors’ statements as saying Fig4 is representative of all TPW (i.e., even lower Rayleigh number TPW, despite the labeling of Rayleigh number in the figure caption). I would suggest that it behooves the authors to be crystal clear(er) about such statements.
\end{quote}
We have added an additional panel to Figure 4, and accompanying text in that figure caption, as well as at line 577 regarding the representativeness of this result.

\begin{quote}
Line 509: “the the” should be “the”
\end{quote}
This has been fixed.

\begin{quote}
Line 565: Equation 46 appears to be the same formula as in Tsai \& Stevenson (2007), but the numerical result appears to be 2-3 times larger (c.f. 2.4 degrees/Myr). Is this because of different choices of viscosities or moment of inertia magnitudes? Given the similarities/differences of the results, the Tsai \& Stevenson (2007) result should be referenced here and it would be useful to discuss the differences/similarities.
\end{quote}
This is indeed a similar formula to that of Tsai \& Stevenson (2007).
We were unable to reproduce the $2.4^\circ$/Myr figure, and so are unable to comment on specific differences, except to note that T\&S states
"The maximum TPW speed is given by equation (18) and is $2.4^\circ \textrm{Myr}^{-1}$ multiplied by a factor of order 1."
\\
\\
We appreciate Victor's thoughtful suggestions.

\end{document}
